### 2021-02-17 Main/Table_Simulation_ReadCounts/Table_Simulation_ReadCounts.tex
Adding latex table

\begin{table}

\caption{\label{tab:}Read count simulation. True values and average relative bias of parameter estimates (and $2.5^{th}$, $97.5^{th}$ percentiles) across a hundred simulated data sets are shown for H3K27me3 data with $10^6$ genomic windows and ENCODE-estimated SNR.\label{Table:table_simulation}}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{llrrrrrrr}
\toprule
\multicolumn{3}{c}{ } & \multicolumn{2}{c}{One Replicate} & \multicolumn{2}{c}{Two Replicates} & \multicolumn{2}{c}{Four Replicates} \\
\cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
Conditions & Parameter & True Value & Relative Bias & ($P_{2.5}$, $P_{97.5}$) & Relative Bias & ($P_{2.5}$, $P_{97.5}$) & Relative Bias & ($P_{2.5}$, $P_{97.5}$)\\
\midrule
 & $\beta_{1}$ & 1.116 & 0.000 & (-0.001, 0.001) & 0.000 & (-0.001, 0.001) & 0.000 & (-0.001, 0.001)\\

 & $\beta_{3}$ & 1.165 & 0.000 & (-0.002, 0.001) & 0.000 & (-0.001, 0.001) & 0.000 & (-0.001, 0.001)\\

 & $\lambda_{1}$ & 1.281 & 0.000 & (-0.004, 0.004) & 0.000 & (-0.003, 0.003) & 0.000 & (-0.002, 0.002)\\

\multirow{-4}{*}{\raggedright\arraybackslash Two} & $\lambda_{3}$ & 0.124 & 0.004 & (-0.057, 0.066) & -0.004 & (-0.047, 0.038) & -0.003 & (-0.035, 0.034)\\
\cmidrule{1-9}
 & $\beta_{1}$ & 1.116 & -0.003 & (-0.005, -0.002) & 0.000 & (-0.001, 0.000) & 0.000 & (-0.001, 0.001)\\

 & $\beta_{3}$ & 1.165 & -0.007 & (-0.010, -0.005) & 0.000 & (-0.001, 0.001) & 0.000 & (-0.001, 0.001)\\

 & $\lambda_{1}$ & 1.281 & -0.001 & (-0.006, 0.003) & 0.001 & (-0.002, 0.004) & 0.000 & (-0.002, 0.002)\\

\multirow{-4}{*}{\raggedright\arraybackslash Three} & $\lambda_{3}$ & 0.124 & -0.317 & (-0.404, -0.223) & -0.023 & (-0.057, 0.012) & -0.003 & (-0.025, 0.024)\\
\cmidrule{1-9}
 & $\beta_{1}$ & 1.116 & -0.014 & (-0.015, -0.012) & -0.007 & (-0.008, -0.006) & 0.000 & (-0.001, 0.000)\\

 & $\beta_{3}$ & 1.165 & -0.111 & (-0.114, -0.109) & -0.003 & (-0.004, -0.002) & 0.000 & (-0.001, 0.001)\\

 & $\lambda_{1}$ & 1.281 & -0.012 & (-0.016, -0.007) & 0.007 & (0.004, 0.010) & 0.000 & (-0.001, 0.002)\\

\multirow{-4}{*}{\raggedright\arraybackslash Four} & $\lambda_{3}$ & 0.124 & -3.520 & (-3.589, -3.439) & -0.360 & (-0.446, -0.282) & -0.005 & (-0.025, 0.017)\\
\bottomrule
\end{tabular}}
\end{table}

### 2021-02-17 Main/Table_Simulation_ReadCounts/Table_Simulation_ReadCounts.R

@@ -1,0 +1,277 @@
+# Table: simulation results
+
+library(data.table)
+library(ggplot2)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(RColorBrewer)
+library(kableExtra)
+
+### Simulation scnearios
+# Marker:   H3K27me3    H3K36me3
+# Groups:   2, 3, 4
+# Replicates: 1, 2, 4
+# Nwindow: 1e5, 5e5, 1e6
+# Nsim: 100
+
+scenario = expand.grid(
+  Marker = c('H3K27me3', 'H3K36me3'),
+  Groups = c(2, 3, 4),
+  Replicates = c(1, 2, 4),
+  Nwindow = c(1e5, 5e5, 1e6),
+  Nsim = 100,
+  Pct = round(100 * c(0.7, 0.8, 0.9, 1))
+)
+scenario$Label = with(scenario,
+                      paste(
+                        Marker,
+                        paste0(Groups, 'G'),
+                        paste0(Replicates, 'R'),
+                        paste0(Nwindow, 'W'),
+                        paste0(Pct, 'SNR'),
+                        sep = "_"
+                      ))
+scenario <-
+  scenario[order(scenario$Groups,
+                 scenario$Replicates,
+                 scenario$Nwindow,
+                 -scenario$Pct), ]
+# print(scenario)
+
+### True parameter functions
+
+genpar = function(Marker, Groups, pct) {
+  ## Initial probability (always begin with background) and Transition Matrix
+  pi = c(1, 0)
+  ## Number of states
+  N.states = 2 ^ Groups
+  if (Marker == 'H3K36me3') {
+    ## See 01_H3K36me3_Parameters.R
+    ## Transition Probability Matrix
+    d1 = 1 - 1 / (131.92967 + 1)
+    d2 = 1 - 1 / (43.75663 + 1)
+    gamma = matrix(0, N.states, N.states)
+    diag(gamma) = c(d1, rep(d2, N.states - 1))
+    gamma[1, 2:N.states] = (1 - d1) / (N.states - 1)
+    gamma[gamma == 0] = (1 - d2) / (N.states - 1)
+    
+    ## ChIP parameters
+    par.background = c(0.7560612, 0, 3.230474)
+    par.enrichment = c(2.4169519, 0, 2.891908)
+    
+    ## Offsets
[…]

### 2021-02-17 Latex/paper/paper.tex
Adding paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[useAMS,usenatbib,referee]{biom}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand*\red{\color{red}}

\usepackage[figuresright]{rotating}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{booktabs}
%\usepackage{lipsum}
\usepackage{array}
\usepackage{float}
\usepackage{multirow}
\usepackage{appendix}
\usepackage{makecell}
\usepackage{color}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{footnote}{2}

\title{Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions}

\author{Pedro L. Baldoni$^*$\email{baldoni@email.unc.edu},
Naim U. Rashid$^{**}$\email{naim@unc.edu}, and
Joseph G. Ibrahim$^{***}$\email{ibrahim@bios.unc.edu} \\
Department of Biostatistics, University of North Carolina at Chapel Hill, 
Chapel Hill, North Carolina, USA}

\begin{document}

\date{{\it Received May} 2020. {\it Revised May} 2020. {\it Accepted May} 2020.}

\pagerange{\pageref{firstpage}--\pageref{lastpage}}\pubyear{2020}

\volume{1}
\artmonth{May}
\doi{00.0000/0.000.0.0.0}

\label{firstpage}

\begin{abstract}
Epigenomics, the study of the human genome and its interactions with proteins and other cellular elements, has become of significant interest in recent years. Such interactions have been shown to regulate essential cellular functions and are associated with multiple complex diseases. Therefore, understanding how these interactions may change across conditions is central in biomedical research. Chromatin immunoprecipitation followed by massively-parallel sequencing (ChIP-seq) is one of several techniques to detect local changes in epigenomic activity (peaks). However, existing methods for differential peak calling are not optimized for the diversity in ChIP-seq signal profiles, are limited to the analysis of two conditions, or cannot classify specific patterns of differential change when multiple patterns exist. To address these limitations, we present a flexible and efficient method for the detection of differential epigenomic activity across multiple conditions. We utilize data from the ENCODE Consortium and show that the presented method, epigraHMM, exhibits superior performance to current tools and it is among the fastest algorithms available, while allowing the classification of combinatorial patterns of differential epigenomic activity and the characterization of chromatin regulatory states.
\end{abstract}

\begin{keywords}
%  Please place your key words in alphabetical order, separated
%  by semicolons, with the first letter of the first word capitalized,
%  and a period at the end of the list.
ChIP-seq; Differential Peak Call; Epigenomics; Hidden Markov Model; Mixture Model.
\end{keywords}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction}

% Paragraph 1: Why does epigenetics matter? Nail down the problem 
Epigenomics, the study of the human genome and its interactions with proteins and other cellular elements, has become of significant interest in recent years. Such interactions have been shown to regulate essential cellular functions such as gene expression and DNA packaging \citep{kim2018polycomb}, resulting in downstream phenotypic impact. Therefore, the interrogation of how these interactions may change across conditions, such as cell types or treatments, is of marked interest in biomedical research. Several landmark articles have identified specific genomic regions of changing (differential) epigenomic activity between conditions as drivers of cell differentiation \citep{creyghton2010histone}, cancer progression \citep{varambally2002polycomb}, and a number of human diseases \citep{portela2010epigenetic}. Within differential regions, the delineation of specific patterns of change across conditions is also of interest, for example classifying the gain-of- or loss-of-activity in genomic loci due to treatment \citep{clouaire2014cfp1}. The identification of specific combinations of processes acting locally may also be informative, such as for segmenting the genome into regulatory states \citep{kundaje2015integrative}.

% Paragraph 3: Introduce ChIP-seq
To quantify local epigenomic activity on a genome-wide scale, a common high-throughput assay is chromatin immunoprecipitation followed by massively parallel sequencing (ChIP-seq). ChIP-seq experiments begin with cross-linking DNA and proteins within chromatin structures, which are then fragmented by sonication in a particular sample. DNA fragments bound to the protein of interest are isolated by chromatin immunoprecipitation, which are then sequenced via massively parallel high-throughput sequencing to generate short sequencing reads pertaining to the original fragments. Sequences are then mapped onto a reference genome through sequence alignment to determine their likely locations of origin. Genomic coordinates containing a high density of mapped reads, often referred to as enrichment regions (peaks), indicate likely locations of protein-DNA interaction sites, and all other regions are referred to as background regions. This local read density is often summarized by counting the number of reads mapped onto non-overlapping windows of fixed length tiling the genome (window read counts), forming the basis for downstream analyses. Across multiple conditions, regions exhibiting enrichment in at least one condition, but not across all conditions, indicate the presence of differential activity pertaining to the protein-DNA interaction of interest.

% Paragraph 4: Issues with the data/methods
To date, many differential peak callers (DPCs) have been proposed \citep{song2011identifying,stark2011diffbind,shen2013diffreps,chen2015novel,lun2015csaw,allhoff2016differential}. However, several challenges affect their ability to accurately detect regions of differential activity from the wide range of ChIP-seq experiments (Section \ref{sec:background}). First, differential regions may be both short or broad in length, causing difficulty for methods optimized for a particular type of signal profile \citep{stark2011diffbind,chen2015novel}. Second, methods that pool experimental replicates together \citep{song2011identifying} often exhibit more false positive calls compared to methods that jointly model replicates from each condition \citep{steinhauser2016comprehensive}. Third, the analysis of ChIP-seq data is often subject to complex biases that may vary across the genome, as differences in local read enrichment may depend on quantities such as the total read abundance in a given region or GC content \citep{teng2017accounting}. DPCs that disregard such complex effects or that solely rely on sample-specific global scaling factors or control subtraction methods \citep{stark2011diffbind,shen2013diffreps,chen2015novel,allhoff2016differential} may be prone to detecting spurious differences due to the lack of non-linear normalization methods \citep{lun2015csaw}. Reflecting these limitations, a recent comparison of DPCs demonstrated that current methods tend to detect either a large number of short peaks (low sensitivity) or exhibit a high number of false positive calls (low specificity) in ChIP-seq experiments with broad regions of enrichment \citep{steinhauser2016comprehensive}. Moreover, few methods are able to simultaneously test for differential activity across three or more conditions \citep{chen2015novel,lun2015csaw}, or can classify specific differential combinatorial patterns. Altogether, these limitations can impact the drawing of accurate insights from modern epigenomic studies.

% Paragraph 4: What we propose? Broad peaks, mixture model, and multiple conditions
Here, we propose an efficient and flexible statistical method to identify differential regions of enrichment from epigenomic experiments with diverse signal profiles and collected under common multi-replicate, multi-condition settings. Our method overcomes the limitations of current DPCs with three major features. First, it uses a hidden Markov model (HMM) to account for the diversity in differential enrichment profiles that may result from short and broad epigenomic ChIP-seq data sets. Second, it captures specific differential combinatorial patterns through a novel finite mixture model emission distribution within the HMM's differential state. Each mixture component pertains to a particular differential combinatorial pattern that is formed by the presence or absence of local enrichment across conditions, where a generalized linear model (GLM) is used to model the specific differential combinatorial pattern while accounting for sample- and window-specific normalization factors via offsets. Third, it enables the simultaneous detection and classification of epigenomic changes under three or more conditions, a novelty not yet available in any other DPC algorithm. The presented method offers additional benefits over current HMM-based DPC algorithms \citep{song2011identifying,allhoff2016differential} that include a GLM-based framework with an embedded mixture model, which allows the modeling of covariates of interest as well as the inclusion of model offsets for non-linear normalization, and a fast and accurate parameter estimation scheme via rejection-controlled EM algorithm (RCEM).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data}
\label{sec:background}

% What are histones and why are they important
Histones are proteins that interact and condense  DNA in eukaryotic cells into structural units called nucleosomes. Multiple types of enzymatic modifications may be applied to histones, resulting in changes in local DNA packaging and chromatin accessibility mediated by nucleosomes \citep{bannister2011regulation}. In turn, cellular processes such as gene transcription, gene silencing, DNA repair, replication, and recombination are also affected. Proteins that interact with DNA and alter its functional properties are often referred to as epigenomic marks. For example, the trimethylation of histone H3 at lysines 36 and 27 (H3K36me3 and H3K27me3) are two types of histone modifications that tend to occur in genomic loci containing actively transcribed and repressed genes \citep{liu2016distinct}, respectively, and exhibit broad enrichment profiles. These marks have been investigated in cancer studies, where their absence is often observed in multiple cancer types \citep{wei2008loss}. As a result, H3K36me3 and H3K27me3 are considered to be key prognostic indicators in patients with breast, ovarian, and pancreatic cancer. The enhancer of zeste homolog 2 (EZH2), a major component of the polycomb complex PRC2 that catalyzes the methylation of H3K27me3 \citep{margueron2011polycomb}, is another example of a protein with experimental signal characterized by broad enrichment domains and co-occurs with the activity of H3K27me3.

Using ChIP-seq data pertaining to histone modifications H3K27me3 and H3K36me3, and to EZH2 from the ENCODE Consortium, we find that current DPCs have difficulty in accurately detecting broad regions of differential enrichment between several common cell lines (Figure~\ref{fig:fig_motivation}). In line with previous findings \citep{steinhauser2016comprehensive}, we observe that even current DPCs designed for broad data \citep{song2011identifying,allhoff2016differential} tend to detect either overly fragmented differential peaks or call regions exhibiting no difference in experimental signal between conditions as differential (Figure~\ref{fig:fig_motivation}A). The low specificity and sensitivity of such methods may impair the biological interpretation of the resulting peak calls in downstream analyses. Methods that rely on candidate peaks may also exhibit a compromised performance due to the limitations of single-sample peak callers in broad data \citep{stark2011diffbind,chen2015novel}. In addition, most current DPCs restrict their application to the analysis of two experimental conditions. For methods that are tailored for the analysis of three or more conditions, the classification of specific differential combinatorial patterns across conditions (or across various epigenomic processes) is still an open problem. The classification of such patterns would allow researchers to, for example, quantify treatment responses on the epigenomic level \citep{clouaire2014cfp1}, or identify sets of processes working together to regulate local chromatin state. We find that the performance of such methods exhibit low sensitivity and specificity in calling differential regions in broad marks (Figure~\ref{fig:fig_motivation}B).
\begin{figure}
\begin{center}
\includegraphics[scale = 0.75]{../../Main/Figure_epigraHMM/Figure_epigraHMM.pdf}
\end{center}
\caption{At the top, differential peak calls from current methods (under false discovery rate control of 0.05) for the histone modification H3K27me3 among cell lines Helas3, Hepg2, and Huvec. Shaded regions indicate  observed differential enrichment, and each vertical line type bordering each region represents a different combinatorial pattern of enrichment across cell lines. Optimal methods would call broad peaks inside shaded regions and no peaks outside them. At the bottom, general overview of the presented method for simultaneous detection and classification of broad and short differential regions of enrichment. This figure appears in color in the electronic version of this article.
\label{fig:fig_motivation}}
\end{figure}

We assessed the performance of our model on ChIP-seq experiments characterized by broad peaks (H3K36me3, H3K27me3, and EZH2) and short peaks (H3K27ac, H3K4me3, and the transcription factor CTCF). In simulations (Section \ref{sec:simulation}) and in data sets from the ENCODE Consortium \citep{landt2012chip}, we show that our model addresses the issues of the current peak callers in broad data (Section \ref{sec:broad}), while being flexible for short peaks (Section \ref{sec:sharp}) and comparable to the fastest DPCs regarding the computation time. We show that our method can also be utilized for genomic regulatory state segmentation when studying multiple types of epigenomic processes from a single condition or cell line (Section \ref{sec:predpostprob}). The Web Appendices A1 and A2 present the data accession codes and the data pre-processing steps, respectively. Code implementing the method and to replicate the presented results are available in the Web Appendices A3 and A4, respectively.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
\label{sec:methods}

\subsection{Statistical Model}
\label{subsec:hmmmixt}
Let $Y_{hij}$ denote the random variable pertaining to the ChIP read count for genomic window $j$ from sample $i$ of condition $h$, where $j=1,\ldots,M$, $i=1,\ldots,n_{h}$, $h=1,\ldots,G$, and let $y_{hij}$ be the observed count. Here, $n_h$ is the number of samples in condition $h$ and $N = \sum_{h = 1}^G n_h$ is the total number of samples across the $G$ conditions.  At the $j^{th}$ window, let $\mb{y}_{..j}=(y_{11j},\ldots,y_{Gn_{G}j})^\prime$ denote the $N \times 1$ vector of ChIP window read counts across all samples and conditions, and let $\mb{y}=(\mb{y}_{..1},\ldots,\mb{y}_{..M})$ denote the corresponding $N \times M$ matrix of window read counts spanning all windows, samples, and conditions. We assume that each window belongs to one of three possible hidden states: consensus background (state 1), differential (state 2), and consensus enrichment (state 3). Windows exhibiting low (high) enrichment across all conditions will be modeled by an emission distribution pertaining to the consensus background (enrichment) state. Windows exhibiting enrichment under at least one condition, but not all conditions, will be modeled by an emission distribution pertaining to the differential state. If $G$ conditions are of interest, there are $L = 2^G-2$ possible differential combinatorial patterns of enrichment and background across conditions at a given window. The emission distribution pertaining to the differential state models all $L$ possible differential combinatorial patterns via a mixture model with mixture proportions $\bs{\delta} = (\delta_{1},\ldots,\delta_{L})^\prime$, such that $\sum_{l=1}^{L}\delta_{l}=1$ (Web Figure 1).

To model transitions between states, we assume a single latent discrete time stationary Markov chain $\mb{Z}=\{Z_{j}\}_{j=1}^{M}$, $Z_{j}\in\{1,2,3\}$, with state-to-state transition probabilities $\bs{\gamma}=(\gamma_{11},\gamma_{12},\ldots,\gamma_{33})^\prime$ and initial probabilities $\bs{\pi}=(\pi_{1},\pi_{2},\pi_{3})^\prime$, such that $\sum_{s=1}^{3}\gamma_{rs}=1$ and $\sum_{s=1}^{3}\pi_{s}=1$ for $r\in\{1,2,3\}$.  To facilitate the notation, let $f_{r}(\mb{y}_{..j}| \bs{\psi}_{r})$ denote the emission distribution corresponding  to the $r^{th}$ hidden state, where $\bs{\Psi} = (\bs{\pi}^\prime,\bs{\gamma}^\prime,\bs{\delta}^\prime,\bs{\psi}^\prime)^\prime$ denotes the vector of all model parameters, $\bs{\psi}=(\bs{\psi}_{1}^\prime,\bs{\psi}_{2}^\prime,\bs{\psi}_{3}^\prime)^\prime$ denotes each state's set of emission distribution-specific parameters, and $\mathcal{Z}$ denotes the set of $3^M$ possible state paths of length $M$. Then, the likelihood function pertaining to the proposed HMM may be written as
\begin{align}\label{eq:difflik}
f(\mathbf{y}|\mathbf{x};\bs{\Psi}) &= \sum_{\mb{Z}\in\mathcal{Z}}\left\{\prod_{r=1}^{3}\pi_{r}^{I(Z_{1}=r)}\times\left(\prod_{j=2}^{M}\prod_{r=1}^{3}\prod_{s=1}^{3}\gamma_{rs}^{I(Z_{j-1}=r,Z_{j}=s)}\right)\right.\times\\
&\times\left.\left(\prod_{j=1}^{M}f_{1}\left(\mb{y}_{..j}|\bs{\psi}_{1}\right)^{I(Z_{j}=1)}f_{2}\left(\mb{y}_{..j}|\mb{x};\bs{\delta},\bs{\psi}_{2}\right)^{I(Z_{j}=2)}f_{3}\left(\mb{y}_{..j}|\bs{\psi}_{3}\right)^{I(Z_{j}=3)}\right)\right\}.\nonumber
\end{align}
Here, $\mb{x}$ is a fixed $G \times L$ design matrix enumerating each of the $L$ possible differential combinatorial patterns in terms of the presence or absence of enrichment across each of the $G$ conditions, only in the emission distribution of the differential state.

We assume that read counts pertaining to genomic windows from the consensus background ($r=1$) and consensus enrichment  ($r=3$) states follow a Negative Binomial (NB) distribution with state-specific parameters $\bs{\psi}_{r}=(\mu_{(r,hij)},\phi_{r})^\prime$, with mean $\mu_{(r,hij)}$ and variance $\mu_{(r,hij)}(1+\mu_{(r,hij)}/\phi_{r})$. Assuming independence of read counts across experiments and samples, conditional upon the HMM state, the emission distribution of the consensus background and consensus enrichment states, respectively, can be written as
\begin{align}\label{eq:nondiffdensity}
f_{r}(\mb{y}_{..j}|\bs{\psi}_{r}) &= \prod^{G}_{h=1}\prod^{n_{h}}_{i=1}\frac{\Gamma(y_{hij}+\phi_{r})}{y_{hij}!\Gamma(\phi_{r})}\left(\frac{\phi_{r}}{\mu_{(r,hij)}+\phi_{r}}\right)^{\phi_{r}}\left(\frac{\mu_{(r,hij)}}{\mu_{(r,hij)}+\phi_{r}}\right)^{y_{hij}},\quad r\in\{1,3\},
\end{align}
with $y_{hij}\in\{0,1,2,\ldots\}$, such that $\log(\mu_{(1,hij)})=\beta_{1} + u_{hij}$, $\log(\phi_{1})=\lambda_{1}$, $\log(\mu_{(3,hij)})=\beta_{1} + \beta_{3} + u_{hij}$, and $\log(\phi_{3})=\lambda_{1} + \lambda_{3}$. The offset $u_{hij}$ adjusts for technical artifacts and allows the non-linear normalization of the signal profile across genomic windows, conditions, and samples (Web Appendix B2). When $u_{hij}=0$, $\beta_{1}$ and $\lambda_{1}$ represent the log-mean and log-dispersion, respectively, of read counts pertaining to consensus background state windows, whereas $\beta_{3}$ and $\lambda_{3}$ represent the difference in log-mean and log-dispersion of read counts from consensus enrichment state windows relative to consensus background state windows.

For windows belonging to the differential state ($r = 2$), we assume that the corresponding read counts are modeled by a $L$-component finite mixture model with mixture components that follow a Negative Binomial distribution, where each component corresponds to a particular differential combinatorial pattern. To define these patterns, let us consider the sets $S_{1},\ldots,S_{L}$ that delineate the subset of the $G$ conditions that are enriched in each of the $L$ differential combinatorial patterns. For instance, if $G=3$, the sets $S_{1} =\{1\}$, $S_{2} =\{2\}$, $S_{3} =\{3\}$, $S_{4} =\{1,2\}$, $S_{5} =\{1,3\}$, and $S_{6} =\{2,3\}$ define the six possible differential combinatorial patterns of enrichment and background across three conditions. That is, the set  $S_{1}$ denotes enrichment in only the first condition and background in all others, whereas the set $S_{6}$ denotes enrichment in conditions 2 and 3 and background in condition 1. The presence or absence of enrichment in each of the $L$ sets is encoded into each column of $\mb{x}=(\mb{x}_{1},\ldots,\mb{x}_{L})$, such that $\mb{x}_{l} = (x_{1l},\ldots,x_{Gl})'$, and $x_{hl} = I(h\in S_{l})$ for $l=1,\ldots,L$ and $h = 1,\ldots,G$. That is, $\mb{x}_{l}$ is the $G \times 1$ vector of binary indicator variables denoting which subset of conditions are enriched in pattern (mixture component) $l$. A graphical illustration of our proposed model is provided in Web Figure 1.

Let $\bs{\psi}_{2}$ denote the state-specific parameter vector pertaining to the differential state and let $\bs{\psi}_{(2,l)}$ denote the set of parameters pertaining to the $l^{th}$ mixture component.  Assuming independence of read counts across conditions and samples, conditional upon the differential HMM state, the finite mixture model emission distribution can be written as
\begin{align}\label{eq:diffdensity}
f_{2}(\mb{y}_{..j}|\mb{x};\bs{\delta},\bs{\psi}_{2}) &= \sum_{l=1}^{L}\delta_{l}\left[\prod_{h=1}^{G}\prod_{i=1}^{n_{h}}\frac{\Gamma\left(y_{hij}+\phi_{(2,l,h)}\right)}{y_{hij}!\Gamma\left(\phi_{(2,l,h)}\right)}\left(\frac{\phi_{(2,l,h)}}{\mu_{(2,l,hij)}+\phi_{(2,l,h)}}\right)^{\phi_{(2,l,h)}}\times\right.\nonumber\\
&\left.\times\left(\frac{\mu_{(2,l,hij)}}{\mu_{(2,l,hij)}+\phi_{(2,l,h)}}\right)^{y_{hij}}\right],\quad y_{hij}\in\{0,1,2,\ldots\}, % &= \sum_{l=1}^{L}\delta_{l}f_{(2,l)}\left(\mb{y}_{..j}|\mb{x}_l;\bs{\psi}_{(2,l)}\right),\nonumber\\
\end{align}
where $\mu_{(2,l,hij)}$ and $\phi_{(2,l,h)}$ are the mean and dispersion, respectively, pertaining to read counts originating from window $j$ and sample $i$ in condition $h$ from the mixture component $l$. We assume that $\log(\mu_{(2,l,hij)}) = \beta_{1} + \beta_{3}x_{hl} + u_{hij}$ and $\log(\phi_{(2,l,h)}) = \lambda_{1} + \lambda_{3}x_{hl}$. That is, in the mixture component $l$, we utilize the same consensus background (consensus enriched) log-mean and log-dispersion from (\ref{eq:nondiffdensity}) in all conditions that are specified by $\mb{x}_{l}$ to be background (enriched) in the $l^{th}$ differential combinatorial pattern. There are several advantages to such a parametrization for the differential emission distribution. For example, it ensures that windows exhibiting differential enrichment across conditions share means and dispersions that are common between the consensus background and consensus enrichment states, a reasonable assumption that significantly increases computational efficiency.  Utilizing a mixture model as the differential state emission distribution avoids the computational burden that would come from assuming separate hidden states for each of the $L$ differential combinatorial patterns, particularly as $G$ increases.  We evaluate the strength of these assumptions through multiple simulations and a real data benchmarking analysis in Sections \ref{sec:simulation} and \ref{sec:realdata}.

Two novel features result from our proposed approach that are relevant to the context of differential enrichment detection from ChIP-seq experiments. By using a modified version of the Expectation-Maximization (EM) algorithm to estimate the model parameters, we are able not only to detect differential enrichment regions across multiple conditions,  but we can also classify various differential combinatorial patterns of enrichment within broad and short differential enrichment domains. With state-specific parameters, the current implementation of the method allows the direct modeling of continuous covariates (e.g. input controls; Web Appendix A3), for which a state-level testing of their effects on the read count distribution could be performed. In a simulation study and in real data analyses, however, we did not observe a significant improvement in performance in differential peak detection after accounting for the effect of input controls (Web Appendix B2), a fact that has also been observed by others \citep{lun2015csaw}.

\subsection{Estimation}
\label{subsec:hmmest}
To simplify the parameter estimation in (\ref{eq:obslik}), we introduce another set of latent variables $\mb{W}=(\mb{W}_{1}^\prime,\ldots,\mb{W}_{M}^\prime)^\prime$, such that $\mb{W}_{j}=(W_{j1},\ldots,W_{jL})^\prime$ for $j=1,\ldots,M$. We assume that $\mb{W}$ is a sequence of independent random vectors such that $\mb{W}_{j}|(Z_{j}=2)\sim\text{Multinomial}(1,\bs{\delta})$ and $\mb{W}_{j}|(Z_{j}=r)=0$ with probability 1 if $r=\{1,3\}$. Under this setup, one may define the data generating mechanism when $Z_{j}=2$ (differential state) and $W_{jl}=1$ ($l^{th}$ differential combinatorial pattern) such that read counts pertaining to genomic window $j$ are sampled from $f_{(2,l)}$ given $\bs{\psi}_{(2,l)}$ and $\mb{x}_{l}$. Let $\mathcal{W}$ denote the set of $L^M$ possible combinations of latent vectors $\mb{W}$. Hence, the likelihood function of the observed data (\ref{eq:difflik}) can be rewritten as
\begin{align}\label{eq:obslik}
f(\mathbf{y}|\mathbf{x};\bs{\Psi}) &= \sum_{\mb{Z}\in\mathcal{Z}}\sum_{\mb{W}\in\mathcal{W}}\left\{\left[\prod_{r=1}^{3}\pi_{r}^{I(Z_{1}=r)}\prod_{j=2}^{M}\prod_{r=1}^{3}\prod_{s=1}^{3}\gamma_{rs}^{I(Z_{j-1}=r,Z_{j}=s)}\right.\right]\times\left[\prod_{j=1}^{M}\left(\prod_{l=1}^{L}\delta_{l}^{W_{jl}}\right)^{I(Z_{j}=2)}\right]\times\nonumber\\
&\times \left.\left[\prod_{j=1}^{M}f_{1}\left(\mb{y}_{..j}|\bs{\psi}_{1}\right)^{I(Z_{j}=1)}\left(\prod_{l=1}^{L}f_{(2,l)}(\mb{y}_{..j}|\mb{x}_l;\bs{\psi}_{(2,l)})^{W_{jl}}\right)^{I(Z_{j}=2)}f_{3}\left(\mb{y}_{..j}|\bs{\psi}_{3}\right)^{I(Z_{j}=3)}\right]\right\}, %&= \sum_{\mb{Z}\in\mathcal{Z}}\left\{\left[\sum_{\mb{W}\in\mathcal{W}}f(\mb{y}|\mb{w},\mb{z},\mb{x};\bs{\psi})f(\mb{w}|\mb{z},\mb{x};\bs{\delta})\right]f(\mb{z}|\mb{x};\bs{\pi},\bs{\gamma})\right\},\\
\end{align}
where $f_{(2,l)}(\mb{y}_{..j}|\mb{x}_l;\bs{\psi}_{(2,l)})$ is defined as in (\ref{eq:diffdensity}). In the $t^{th}$ iteration of the EM algorithm, the $Q$ function of the complete data log-likelihood can be written as
\begin{align}\label{eq:diffQfunc}
Q\left(\bs{\Psi}|\bs{\Psi}^{(t)}\right) &= E_{\mb{Z}}\left(E_{\mb{W}|\mb{Z}}\left(\log\left(f(\mb{y},\mb{W},\mb{Z}|\mb{x};\bs{\Psi})\right)|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right),\nonumber\\
&= Q_{0}(\bs{\pi},\bs{\gamma}|\bs{\Psi}^{(t)})+Q_{1}(\bs{\psi}_{1}|\bs{\Psi}^{(t)})+Q_{2}(\bs{\delta},\bs{\psi}_{2}|\bs{\Psi}^{(t)})+Q_{3}(\bs{\psi}_{3}|\bs{\Psi}^{(t)}),
\end{align}
where $Q_{0}(\bs{\pi},\bs{\gamma}|\bs{\Psi}^{(t)})$, $Q_{1}(\bs{\psi}_{1}|\bs{\Psi}^{(t)})$, $Q_{2}(\bs{\delta},\bs{\psi}_{2}|\bs{\Psi}^{(t)})$, and $Q_{3}(\bs{\psi}_{3}|\bs{\Psi}^{(t)})$ are defined in Web Appendix B5. In the E-step of the EM algorithm, we compute the posterior probabilities from (\ref{eq:diffQfunc}). The quantities $Pr\left(Z_{j}=r|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)$ and $Pr\left(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)$, defined in Web Appendix B5, can be calculated through the Forward-Backward algorithm and $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})=f_{(2,l)}(\mb{y}_{..j}|\mb{x}_l;\bs{\psi}_{(2,l)}^{(t)})\delta_{l}^{(t)}/\sum_{k=1}^{L}f_{(2,k)}(\mb{y}_{..j}|\mb{x}_k;\bs{\psi}_{(2,k)}^{(t)})\delta_{k}^{(t)}$ for $l=1,\ldots,L$.

The $Q$ function is maximized with respect to the parameters $\bs{\Psi}=(\bs{\pi}^\prime,\bs{\gamma}^\prime,\bs{\delta}^\prime,\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$ during the M-step of the algorithm. Estimates for the initial and transition probabilities can be directly calculated as $\hat{\pi}^{(t+1)}_{r}=Pr\left(Z_{1}=r|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)$ and $\hat{\gamma}^{(t+1)}_{rs}=\sum_{j=2}^{M}Pr(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t)})/\sum_{j=2}^{M}Pr(Z_{j-1}=r|\mb{y},\mb{x};\bs{\Psi}^{(t)})$, respectively, restricted to $\sum_{r=1}^{3}\hat{\pi}^{(t+1)}_{r}=1$ and $\sum_{s=1}^{3}\hat{\gamma}^{(t+1)}_{rs} = 1$, for $r\in\{1,2,3\}$. We perform conditional maximizations to compute estimates of the remaining model parameters $(\bs{\delta}^\prime,\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$. First, mixture proportions can be estimated as $\hat{\delta}_{l}^{(t+1)}=\sum_{j=1}^{M}Pr(Z_{j}=2|\mb{y},\mb{x};\bs{\Psi}^{(t)})Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})/\sum_{j=1}^{M}Pr(Z_{j}=2|\mb{y},\mb{x};\bs{\Psi}^{(t)})$. Estimating $(\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$ from (\ref{eq:diffQfunc}) can be seen as obtaining parameter estimates from a series of weighted NB regression models with shared mean and dispersion parameters (Web Appendix B5). We jointly estimate these quantities via the algorithm BFGS \citep{fletcher2013practical}.

The estimation scheme is robust to situations where certain differential combinatorial patterns of enrichment are rare (Figure \ref{fig:fig_multimarks}). This unique characteristic results from the fact that ChIP-seq experiments often provide enough data (usually $M>10^7$ non-overlapping windows of 250 bp fixed size for the human reference genome) to estimate the parameters $(\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$, which are shared across all $L$ mixture components and HMM states. If pruning differential combinatorial patterns of the differential mixture component is of interest, the optimal number of mixture components $L^{*}$, $L^{*}<L$, can be selected via the Bayesian Information Criterion (BIC) for HMMs. We observed that selecting the optimal number of mixture components based on BIC agrees with the pruning of rare differential combinatorial patterns that we would not biologically expect to observe in real data (Web Appendix B3).

To obtain the parameter estimates $\hat{\bs{\Psi}}$, the EM algorithm iterates until the maximum absolute relative change in the parameter estimates three iterations apart is less than $10^{-3}$ for three consecutive iterations. To reduce the computation time, we make use of a RCEM algorithm with threshold 0.05. Briefly, the RCEM algorithm substantially reduces the dimensionality of the data during the M-step by randomly assigning a zero posterior probability to genomic windows unlikely to belong to each of the HMM states. The current estimation set up allows genomic windows exhibiting equal distribution of read counts to have their posterior probability aggregated during the M-step of the algorithm. Often, the distribution of read counts along the genome is highly concentrated on a particular set of values, such as 0, 1, and 2 for instance. Genomic windows exhibiting a particular pattern of counts across samples and conditions can have their posterior probability aggregated during the M-step, which further reduces the dimensionality of the objective function during the numerical optimization and leads to a fast gradient-based optimization. %Genomic windows with assigned zero posterior probability due to the RCEM approach do not contribute to the final set of aggregated posterior probabilities.

Once the algorithm reaches convergence, the final set of HMM posterior probabilities can be used to segment the genome into consensus background, differential, or consensus enrichment windows. Approaches that control the total false discovery rate (FDR) via posterior probabilities \citep{efron2001empirical} or that estimate the most likely sequence of hidden states \citep{viterbi1967error} can be used for such purposes. Let $\hat{\rho}_{j2} = Pr\left(Z_{j}=2|\mb{y},\mb{x};\hat{\bs{\Psi}}\right)$ denote the estimated posterior probability that the $j^{th}$ genomic window belongs to the differential HMM state, $j=1,\ldots,M$. For a cutoff of posterior probability $\alpha$, the nominal FDR is $\sum_{j=1}^{M}(1-\hat{\rho}_{j2})I(\hat{\rho}_{j2}\geq 1-\alpha)/\sum_{j=1}^{M}I(\hat{\rho}_{j2}\geq 1-\alpha)$, where $I(\cdot)$ is an indicator function. The posterior probability cutoff is then chosen by controlling the total FDR. Differential regions of enrichment are formed by merging adjacent windows that either meet a nominal FDR threshold level for the differential HMM state or belong to the same Viterbi's predicted state. A discussion about these two approaches under the proposed model is presented in the Web Appendix B4. Additional details of proposed EM algorithm and the implemented code are available in the Web Appendices B5 and A3, respectively.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Simulation Studies}
\label{sec:simulation}
We evaluate the presented model in two independent simulation studies of broad epigenomic marks. In the first study (Section~\ref{sec:sim1}), we simulated read count-based data to assess the precision of the parameter estimation scheme, the performance of differential peak detection, and the accuracy of the classification of specific differential combinatorial patterns of enrichment within differential peaks. In the second simulation study (Section~\ref{sec:sim2}), we utilize the simulation pipeline presented in \cite{lun2015csaw} to generate synthetic ChIP-seq reads from \textit{in silico} experiments with broad differential peaks. The aim of the second simulation study was to compare our model with other DPCs in a more realistic scenario with broad peaks, while also avoiding the assumption of a parametric model for the data.

\subsection{Read Count Simulation}
\label{sec:sim1}
Read counts were simulated under different scenarios that varied regarding the type of histone modification mark (H3K36me3 and H3K27me3), number of genomic windows ($M$, $10^5$, $5\times 10^5$, and $10^6$ windows), number of conditions ($G$, 2, 3, and 4), and number of replicates per condition ($n$, 1, 2, and 4). We further assessed our model under different Signal-to-Noise Ratio (SNR) levels. We define the SNR as the ratio between the means of consensus enrichment and consensus background emission distributions. Mean and dispersion parameters used in this simulation study were estimated from ENCODE data and are presented in the Web Appendix C1 for all the scenarios. Different SNR levels were defined by decreasing the ratio of the means in decrements of 10\% while maintaining the mean-variance relationship. Read counts were assumed to follow a NB distribution and were simulated using a first-order Markov chain with $2^G$ states, representing every combination of background and enrichment across $G$ conditions. We aimed to assess whether our model was able to assign all $2^G-2$ simulated differential states to the differential HMM state, while maintaining a precise parameter estimation scheme and accurate classification of differential combinatorial patterns.

\subsubsection{Simulation Results}
Table~\ref{Table:table_simulation} shows the true values and the average relative bias of parameter estimates (and the $2.5^{th}$, $97.5^{th}$ percentiles) from a hundred simulated data sets relative to the scenario of H3K27me3 with $10^5$ genomic windows (see Web Appendix C1 for additional details). Results are shown for different levels of SNR, number of conditions, and number of replicates per condition. Overall, no significant differences regarding the relative bias of parameter estimates were observed across simulations under different number of genomic windows. Depending on the number of conditions, the observed relative bias and the range of the reported percentiles tended to decrease as more replicates were included in the analyses. This effect was particularly significant in scenarios with four conditions with respect to parameters $\beta_{3}$ and $\lambda_{3}$. In general, scenarios with higher SNR showed lower relative bias and variability of the parameter estimates in comparison to scenarios with lower SNR, regardless of the number of conditions or replicates per condition. In scenarios with lower SNR levels or higher number of conditions, these results also highlight the importance of experimental replicates to achieve precise parameter estimates. The proposed estimation approach via EM algorithm led to precise parameter estimates and was robust to a data generating mechanism that was different than the one assumed by the proposed model.
\begin{center}
\input{../../Main/Table_Simulation_ReadCounts/Table_Simulation_ReadCounts.tex}
\end{center}

\subsection{Sequencing Read Simulation}
\label{sec:sim2}
We performed a second simulation study aiming to compare the proposed model with the current DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR. We used the simulation pipeline presented by \cite{lun2015csaw} where data were generated in a more general scheme without a particular read count model assumption. Here, sequencing reads from broad ChIP-seq experiments were generated for two conditions and two replicates per condition. For the differential peaks callers ChIPComp and DiffBind that require sets of candidate regions, we followed the analyses presented by \cite{lun2015csaw} and called peaks in advance using HOMER. Peaks were then used as input in the respective software for differential call. A hundred simulated data sets were generated and peaks were called by all the methods under multiple nominal FDR thresholds. For our method and RSEG, window-based posterior probabilities were used to control the total FDR as described in Section \ref{subsec:hmmest}.% We refer to the method presented in this article as epigraHMM. We attempted to utilize MACS in this simulation study but the peak caller failed to output peaks for the simulated data.

\subsubsection{Simulation Results}
Figure~\ref{fig:fig_simulation_csaw} shows the main results of our second simulation study. Out of 100 simulated data sets, RSEG either failed to analyze the data due to internal errors or called the entire genome as differential in 26 and in 3 instances, respectively. Problematic cases from RSEG were not considered in this simulation study. Similar issues have been previously reported in other studies \citep{starmer2016detecting}. We observed that our method showed the highest observed sensitivity among all DPCs, regardless of the nominal FDR thresholding level, while maintaining a moderate observed FDR (Figure~\ref{fig:fig_simulation_csaw}A). Here, the observed FDR is the proportion of genomic windows incorrectly called as differential out of the total number of windows called as differential. Methods such as diffReps, RSEG, and THOR showed higher observed FDR levels than the nominal threshold due to the excessive number of differential peaks called outside true differential regions (shaded area in Figure~\ref{fig:fig_simulation_csaw}D). While diffReps and THOR called an excessive number of short and discontiguous peaks, RSEG called regions that were usually wider than the observed differential enrichment regions. These results are further illustrated in Figure~\ref{fig:fig_simulation_csaw}B, where we present the average ratio of the number of called and simulated peaks (y-axis) and the average number of called peaks intersecting true differential regions (x-axis). Regarding the computation time, the HMM-based algorithms RSEG and THOR appeared to be the most computationally intensive and required longer amounts of time to analyze the data. In Figure~\ref{fig:fig_simulation_csaw}C, we present the box plots of computing time (in minutes) across a hundred simulated data sets for all benchmarked methods. While still being an HMM-based algorithm, our method was among the fastest tools for differential peak detection due to the implemented strategies to improve the computation time of the EM algorithm (Section \ref{subsec:hmmest}). Figure~\ref{fig:fig_simulation_csaw}D shows an example of a genomic region with simulated data and called peaks from various methods using nominal FDR threshold 0.05. As shown, our method was able to consistently cover most of true differential regions with broad peaks while exhibiting a limited number of false discoveries (see Web Appendix C2 for additional results).
\begin{figure}
\begin{center}
\includegraphics[scale = 0.65]{../../Main/Figure_Simulation_SequencingReads/Figure_Simulation_SequencingReads.pdf}
\end{center}
\caption{Sequencing read-based simulation from the csaw pipeline. (A): average observed sensitivity and FDR for various methods. (B): scatter plot of average ratio of called and simulated peaks (y-axis) and number of called peaks intersecting true differential regions (x-axis). (C): box plot of computing time (in minutes) for various algorithms. (D): an example of differential peak calls under a nominal FDR control of 0.05. Shaded areas indicate true differential peaks. This figure appears in color in the electronic version of this article.
\label{fig:fig_simulation_csaw}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Application to ENCODE Data}
\label{sec:realdata}
We applied our method to ChIP-seq data from the ENCODE Consortium (Section \ref{sec:background}) to detect differential regions of enrichment of several epigenomic marks across distinct cell lines. First, we analyzed broad data from the histone modifications H3K36me3 and H3K27me3 as well as data from EZH2 (Section \ref{sec:broad}). Secondly, we assessed the performance of the presented model on ChIP-seq experiments from the transcription factor CTCF and the histone modifications H3K27ac and H3K4me3 characterized by short peaks (Section \ref{sec:sharp}). For H3K27ac and H3K4me3, enrichment peaks are usually deposited on the promoter regions of actively transcribed genes and several studies have associated their role with gene transcription \citep{creyghton2010histone,lauberth2013h3k4me3}. The transcription factor CTCF is a protein that binds to short DNA motifs and is responsible for several cellular processes that include the regulation of the chromatin 3D structure and mRNA splicing \citep{shukla2011ctcf}. Two isogenic replicates for each epigenomic mark were used in the analysis. Using RNA-seq data, we assessed the practical significance of our results by associating the detection and classification of differential combinatorial patterns from called peaks of H3K36me3, H3K27me3, and EZH2 with the direction of gene expression (Section \ref{sec:predpostprob}; see Web Appendix D1 for the RNA-seq data processing).

We compared the genome-wide performance of the presented model with the current DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR. For the methods that require a set of candidate regions to be specified \textit{a priori}, ChIPComp and DiffBind, peaks were called in advance using MACS2 (\cite{zhang2008model}; available at \url{https://github.com/macs3-project/MACS}) and used as input to the software for differential call. We benchmarked methods regarding the coverage of differentially transcribed gene bodies, the number and average size of differential peak calls, $\log_{2}$ fold change (LFC) of read counts, Spearman correlation of $\log_{2}$-transformed read counts between cell lines, and computation time. Metrics for sensitivity and specificity were defined on the window level and based on the coverage of differentially transcribed gene bodies by called peaks (Web Appendix D1). For broad marks, read counts were computed using non-overlapping windows of 500bp. For the remaining short marks, we computed read counts using non-overlapping windows of 250bp. Results presented in this section pertain to the analysis of two cell lines, namely Helas3 and Hepg2. A discussion about the choice of the window size is presented in the Web Appendix D2. Results from the analysis of more than two cell lines are presented in the Web Appendix D3. Data accessing code, data pre-processing steps, method-specific parameters, and code to replicate the presented results are detailed in the Web Appendix A.

\subsection{Analysis of ChIP-seq Data From Broad Marks}
\label{sec:broad}
Methods were benchmarked regarding the coverage of differentially transcribed gene bodies. The histone modification H3K36me3 is known to be associated with gene transcription and enriched regions of this mark are usually deposited on transcribed gene bodies. Hence, the location of differential peaks of H3K36me3 is expected to agree with the location of differentially expressed genes. Following the analysis presented by \cite{steinhauser2016comprehensive}, we defined a set of protein coding genes exhibiting $|\text{LFC}|>2$ of ChIP-seq read counts between the two analyzed cell lines as true differentially transcribed genes. Results using different threshold levels are presented in the Web Appendix D1 and agree with those presented here. Protein-coding genes with total read count across cell lines under the $25^{th}$ percentile were excluded from the analysis. Normalization by the median log-ratios of each replicate over the geometric mean was performed to avoid spurious differences due to sequencing depth.%The analysis of H3K27me3 is challenging as its enrichment is often associated with both intergenic and silenced genes. Methods that utilize a gold-standard for benchmarking purposes for this mark rely on real-time PCR results on specific loci \citep{zang2009clustering}. Here, we compare methods for this mark with respect to the characteristics of called peaks.

In Figure~\ref{fig:fig_broad}A, we show receiver operating characteristic (ROC) curves for various methods and different nominal levels of FDR threshold for differential peaks of H3K36me3. Similar to the analysis of broad histone modification marks presented by \citet{xing2012genome}, we computed the observed true positive rates (false positive rates) on the window-level as the proportion of windows called as differential out of the total number of windows associated (not associated) with differentially transcribed genes (Web Appendix D1). Our method had the best overall performance among all DPCs as its differential peaks were able to cover most of differentially transcribed gene bodies while still maintaining a low number of false positives. Methods that tended to call short peaks, such as ChIPComp and DiffBind, were the ones with the lowest sensitivity among all methods. ChIPComp and DiffBind have been previously shown to be dependent on the set of candidate peaks and to perform best in scenarios with short peaks (Figure~\ref{fig:fig_broad}B; \citet{steinhauser2016comprehensive,lun2015csaw}). In Figure~\ref{fig:fig_broad}C, we show the observed sensitivity (y-axis) and the average differential peak size (kbp; x-axis) for various methods under different nominal FDR levels (the observed FDR is annotated next to each data point). Our model and RSEG, two HMM-based methods, tended to call broader differential peaks and exhibited better sensitivity than other methods. Yet, differential peaks called by RSEG often did not correspond to differential regions of enrichment (Figure~\ref{fig:fig_broad}D), a behavior that has been noted by others \citep{starmer2016detecting} and also seen in simulated data (Figure~\ref{fig:fig_simulation_csaw}). Our HMM-based method with a non-linear normalization scheme via model offsets allowed us to maintain a low observed FDR and a higher sensitivity than other DPCs. In Figure~\ref{fig:fig_broad}F, we show examples of differential peak calls for EZH2. Our method was among the fastest algorithms due to our computational scheme, taking approximately 1 hour to analyze genome-wide data (Figure~\ref{fig:fig_broad}E).
\begin{figure}
\begin{center}
\includegraphics[scale = 0.65]{../../Main/Figure_Broad/Figure_Broad.pdf}
\end{center}
\caption{Analysis of broad ENCODE data. (A): ROC curves of H3K36me3 differential peak calls. (C): sensitivity (y-axis) and average H3K36me3 differential peak size (kbp; x-axis) of various methods under different nominal FDR thresholds (observed FDR annotated next to data points). (B), (D), and (F): example of peak calls from H3K36me3, H3K27me3, and EZH2, respectively, under a nominal FDR control of 0.05. Posterior probabilities of the HMM differential state are shown at the bottom of each panel. (E): computing time of genome-wide analysis from various methods. This figure appears in color in the electronic version of this article.
\label{fig:fig_broad}}
\end{figure}

\subsection{Analysis of ChIP-seq Data From Short Marks}
\label{sec:sharp}
We further evaluated the performance of the proposed method on data sets characterized by short peaks, namely the histone modifications H3K4me3 and H3K27ac and the transcription factor CTCF. The goal of our analysis was to assess whether our method was robust to different types of data and still able to call short differential regions of enrichment. In these scenarios, differential peaks are usually observed in isolated genomic regions and exhibit a high SNR. It has been shown that certain HMM-based approaches, including RSEG, have low accuracy under short histone modification marks and TF data \citep{hocking2016optimizing}. As we show, the model proposed in this article performs comparably to the evaluated DPCs known to perform best in short data (ChIPComp, DiffBind; \cite{steinhauser2016comprehensive}) and appeared to be more efficient regarding the computation time in certain scenarios.

We calculated the LFC and the Spearman correlation between cell lines Helas3 and Hepg2 based on ChIP-seq read counts mapped onto differential peaks called by each method. Read counts were previously normalized by the median log-ratios of each replicate over the geometric mean to avoid spurious differences due to sequencing depth. As these marks are characterized by short peaks, ideal methods would show high absolute LFC and negative correlation between read counts mapped on differential peaks. Figure~\ref{fig:fig_sharp} shows the main results from our analysis using short data sets. In Figures~\ref{fig:fig_sharp} A and C we show the median LFC and the Spearman correlation of ChIP-seq counts for differential CTCF and H3K4me3 peak calls (sorted by the absolute LFC), respectively, under a nominal FDR control of 0.05. We present separate curves regarding the signal of observed enrichment to better characterize the direction of change. The results show that the HMM-based methods RSEG and THOR were among those with the lowest absolute LFC among all methods, which confirms their sub optimal performance in the scenario of short peaks \citep{hocking2016optimizing}. In addition, we observed that ChIPComp had the best performance overall as it was able to call differential peaks with the highest absolute LFC and the lowest correlation between read counts of the two analyzed cell lines. Our model was able to properly call truly short differential peaks (Figures~\ref{fig:fig_sharp} B, D, and F) and was comparable to the non-HMM based methods regarding the computation time (Figure~\ref{fig:fig_sharp}E), jointly calling differential peaks in less than 1.5 hour.
\begin{figure}
\begin{center}
\includegraphics[scale = 0.65]{../../Main/Figure_Narrow/Figure_Narrow.pdf}
\end{center}
\caption{Analysis of short ENCODE data. (A) and (C): median LFC and correlation between cell lines of ChIP-seq counts from differential peaks for CTCF and H3K4me3, respectively. (B), (D), and (F): example of peak calls from CTCF, H3K4me3, and H3K27ac, respectively. Posterior probabilities of the HMM differential state are shown at the bottom of each panel. (E): computing time of genome-wide analysis from various methods. Results are shown under a nominal FDR control of 0.05. This figure appears in color in the electronic version of this article.
\label{fig:fig_sharp}}
\end{figure}

\subsection{Genomic Segmentation and Classification of Chromatin States}
\label{sec:predpostprob}
Lastly, we analyzed data from the cell line Helas3 to segment its genome regarding the activity of marks H3K36me3, H3K27me3, and EZH2. We considered each mark as a separate experimental condition ($G=3$) and sought to jointly classify local chromatin states in Helas3 based upon the presence or absence of enrichment from each mark. It is known that EZH2 catalyzes the methylation of H3K27me3, a repressive mark, and H3K36me3 is associated with transcribed genes (Section \ref{sec:background}). Hence, we expected regions of enrichment in consensus for these marks to be rare and differential regions to be mostly represented by either transcribed chromatin states (enrichment for H3K36me3 alone) or repressed chromatin states (enrichment co-occurrence for H3K27me3 and EZH2). The analyses presented in this section highlight the applicability of our method in the context of genomic segmentation \citep{ernst2012chromhmm}, a distinct problem not tackled by current DPCs.

First, we segmented the genome using the Viterbi sequence of most likely HMM states to understand the distribution of genomic regions associated with consensus background, differential, and consensus enrichment states (Figure~\ref{fig:fig_multimarks}). While the majority of the genomic regions exhibited no enrichment for any of the analyzed marks, regions of consensus enrichment were rare and represented only 2\% of the analyzed genome (Figure~\ref{fig:fig_multimarks}A), as expected. Consensus background and differential regions mostly covered intergenic (66\%) and protein-coding genic regions (61\%), respectively. Differential genomic windows were mostly representing either transcribed chromatin states or repressed chromatin states (Figure~\ref{fig:fig_multimarks}B). All differential combinatorial patterns expected to be rare had associated mixture proportion estimates less than 0.02 (see Web Appendix B3 for a discussion on pruning rare states). %. The Viterbi algorithm allows the simultaneous segmentation of the genome regarding the three HMM states without the necessity of choosing state-specific FDR control thresholds
\begin{figure}
\begin{center}
\includegraphics[scale = 0.7]{../../Main/Figure_MultiMarks/Figure_MultiMarks.pdf}
\end{center}
\caption{Genomic chromatin state segmentation and classification. (A): distribution of base pairs (y-axis) and the Viterbi sequence of states (x-axis). (B): estimated mixture probabilities and the associated differential combinatorial patterns. (C): density estimate from expression of genes intersecting differential peaks associated with the enrichment of H3K36me3 alone or the enrichment of H3K27me3 and EZH2 in consensus. (D): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively. This figure appears in color in the electronic version of this article.
\label{fig:fig_multimarks}}
\end{figure}

These results suggest that protein-coding genes overlapping differential regions should be either silenced (e.g. genes associated with repressed chromatin states) or highly expressed (e.g. genes associated with transcribed chromatin states). To assign combinatorial patterns to differential peaks, we chose the combination pertaining to the most frequent mixture component across windows by using the maximum estimated mixture model posterior probability, $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})$, $j=1,\ldots,M$. For genes overlapping differential regions associated with either transcribed or repressed chromatin states, we computed the distribution of transcripts per million (TPM) from matching RNA-seq experiments. Genes associated with transcribed chromatin states had a significantly higher distribution of TPM counts than genes associated with repressed chromatin states (Figure~\ref{fig:fig_multimarks}C). We detected broad differential regions of enrichment and the classification of differential combinatorial patterns agreed with their biological roles as well as the expression levels of associated gene bodies. Figure~\ref{fig:fig_multimarks}D shows an example of a genomic region with differential enrichment for the analyzed marks. We compare our results to ChromHMM with 3 states, a method developed for chromatin segmentation. Our method offers the benefit of simultaneously detecting differential peaks and classifying the combinatorial pattern of enrichment through mixture model posterior probabilities even in the context of genomic segmentation with highly diverse epigenomic marks (see Web Appendix D4). By using the BIC for model selection (Web Appendix B3), one can choose the number of biologically relevant mixture components to be included in the model, a task that may not be as straightforward in methods such as ChromHMM (see Supplementary Figure 4 in \citet{ernst2012chromhmm}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{sec:discussion}

We presented a flexible and efficient statistical model designed to call differential regions of enrichment from ChIP-seq experiments with multiple replicates and multiple conditions. Our model has three main advantages over current methods tailored for differential peak detection. First, it uses an HMM-based approach that accounts for the local dependency of ChIP-seq counts and is able to precisely detect broad and short differential regions of enrichment. Second, it utilizes a GLM-based framework with model offsets that account for potential non-linear biases in the data as well as a constrained parametrization across HMM states and mixture components. Our implementation of the RCEM algorithm led to genome-wide analyses of ChIP-seq data under a computational time comparable to some of the fastest current methods and was at least 5 times faster than current HMM-based algorithms. Lastly, our method allows the simultaneous detection and classification of differential combinatorial patterns of enrichment from its embedded mixture model and the associated posterior probabilities under any number of conditions. Our software has been implemented into an R package and is available for download (Web Appendix A). % Data resulting from these assays are similar in nature but aim to identify regions that alternatively are not bound to proteins. % Such a tool could trigger, for instance, new insights on the field of epigenomics as investigators are often interested in making sense of the roles of epigenomic marks on the human genome. % Analysis of datasets from ther epigenomic assays, such as ATAC-seq, could also benefit from the presented method.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\backmatter
% \section*{Acknowledgements}
% The authors thank Professor A. Sen for some helpful suggestions,
% Dr C. R. Rangarajan for a critical reading of the original version of the
% paper, and an anonymous referee for very useful comments that improved
% the presentation of the paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{biom}\bibliography{bibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Supporting Information}
Web Appendices A-D, referenced in Sections 1-6, and a link to the implemented software
are available with this article at the Biometrics website on Wiley Online Library.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{lastpage}
\end{document}

### 2021-02-18 Latex/appendix/webappendix.Rnw
Adding latex files from appendix & paper

\documentclass[12pt,oneside]{book}

\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage[section]{placeins}
\usepackage{enumerate}
\usepackage{verbatim}
\usepackage{multirow}
\usepackage{mathtools}
\usepackage{amsfonts}
\usepackage[flushleft]{threeparttable}
\usepackage{tikz}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{csquotes}
\usepackage{float}
\usepackage{caption}
\usepackage[margin=1in]{geometry}
\usepackage{titling}
\newcommand{\subtitle}[1]{%
\posttitle{%
\par\end{center}
\begin{center}\large#1\end{center}
\vskip0.5em}%
}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\mb}[1]{\mathbf{#1}}
\usepackage{url}
\usepackage[american]{babel}
\usepackage[colorlinks]{hyperref}
\usepackage{appendix}
\AtBeginDocument{%
\hypersetup{
citecolor=black,
linkcolor=black,
urlcolor=blue}
}

\newcommand{\argmax}[1]{\underset{#1}{\operatorname{arg}\,\operatorname{max}}\;}

\captionsetup[table]{name=Web Table}
\captionsetup[figure]{name=Web Figure}

\usepackage{chngcntr}
\counterwithout{figure}{chapter}
\counterwithout{table}{chapter}

\usepackage{setspace}
\doublespacing
\AtBeginDocument{\renewcommand\appendixname{Supporting Information}}

% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}

\if1\blind{
\date{%
$^*$Department of Biostatistics, University of North Carolina at Chapel Hill, Chapel Hill, North Carolina, USA\\%
\today
}
\title{{Supporting Information for 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions' by Pedro L. Baldoni$^*$, Naim U. Rashid$^*$, and Joseph G. Ibrahim$^*$}}
} \fi

\if0\blind{
\date{%
\today
}
\title{{Supporting Information for 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions'}}
} \fi

\begin{document}
\maketitle

\appendix
\makeatletter
\renewcommand\thechapter{\Alph{chapter}}
\renewcommand\thesection{\thechapter\arabic{section}}
\renewcommand\thesubsection{\thesection.\arabic{subsection}}
\renewcommand\thesubsubsection{\thesubsection.\arabic{subsubsection}}
\renewcommand{\thetable}{\arabic{table}}
\renewcommand{\thefigure}{\arabic{figure}}
\makeatother

\chapter{}
In this appendix, we present a summary of data accession codes, data pre-processing, and parameter specifications from benchmarked methods.

\section{Data accession codes}

\begin{table}[h!]
\footnotesize
\centering
\caption{GEO sample accession codes of the analyzed data from the ENCODE Consortium}
\begin{tabular}{lccccccc}
\hline
Cell Line & H3K27me3 & H3K36me3 & EZH2 & H3K4me3 & H3K27ac & CTCF & RNA-seq\\
\hline
H1hesc & GSM733748 & GSM733725 & GSM1003524 & GSM733657 & GSM733718 & GSM733672 & GSM758566\\
HelaS3 & GSM733696 & GSM733711 & GSM1003520 & GSM733682 & GSM733684 & GSM733785 & GSM765402\\
Hepg2 & GSM733754 & GSM733685 & GSM1003487 & GSM733737 & GSM733743 & GSM733645 & GSM758575\\
Huvec & GSM733688 & GSM733757 & GSM1003518 & GSM733673 & GSM733691 & GSM733716 & GSM758563\\
\hline
\end{tabular}
\end{table}

\section{Data Processing}
First, we removed PCR duplicates from the BAM files using SAMTools \cite{li2009sequence} and converted the resulting indexed and sorted files to BED format using BEDTools \cite{quinlan2010bedtools}, as RSEG only accepts such a format as input. Then, the fragment length of each ChIP-seq experiment was estimated using csaw and its functions \textit{correlateReads} and \textit{maximizeCcf}. Finally, using the estimated fragment length, read counts from all cell lines were tabulated for their ChIP replicates using fixed-step and non overlapping windows of size 250bp, 500bp, 750bp, and 1000bp through the R package \textit{bamsignals} \cite{mammana2016package}. For all methods using window-based approaches (csaw, ChIPComp, diffReps, RSEG, THOR, and epigraHMM), we assessed their performance with different window sizes. See Section \ref{s:window} for a discussion about results with different window sizes.

All the methods considered in the data applications and simulation study outputted a set of differential genomic regions/windows that were used for benchmark purposes. THOR output a list of differential peaks in BED6+4 format (\textit{narrowPeak}) with adjusted p-values. RSEG output a WIG file with genomic windows and their posterior probabilities for differential enrichment. diffReps output an annotated TXT file with differential regions of enrichment and their adjusted p-values. DiffBind output a TXT file with differential regions of enrichment and their respective multiple testing corrected FDR. diffReps output a TXT file with differential regions of enrichment and their p-values. csaw output a TSV file with differential regions of enrichment and their FDR adjusted p-values. For a fair FDR thresholding comparison, we control the total FDR and output the differential regions of enrichment based on the set of posterior probabilities as described in the main text. For a comparison between the Viterbi and the FDR thresholding approach, see Section \ref{s:viterbi}.

The following parametrization was used when calling peaks from the benchmarked methods:
\begin{itemize}
\item THOR: \textit{rgt-THOR 'config' --name 'name' -b 'bp' --pvalue 1.0 --output-dir 'output'},
\item RSEG: \textit{rseg-diff -verbose -mode 3 -out 'output' -score 'score' -chrom 'chrom' -bin-size 'bp' -deadzones 'deadzonee' -duplicates 'sample1' 'sample2'},
\item ChIPComp: \textit{ChIPComp(makeCountSet(conf,design,filetype="bam",species="hg19",binsize=bp))},
\item diffReps: \textit{diffReps.pl --gname hg19 --report 'output' --treatment 'sample1' --control 'sample2' --btr 'control1' --bco 'control2' --window 'bp' --pval 1 --nsd 'marktype' --meth 'nb'},
\item DiffBind: \textit{dba.report(dba.analyze(dba.contrast(dba.count(dba(sampleSheet = conf)), categories=DBA\_CONDITION,minMembers=2)),th=1)},
\end{itemize}
such that $\text{bp} = \{250,500,750,100\}$ and $marktype='broad'$ if H3K27me3, H3K36me3, or EZH2, or $marktype='sharp'$ otherwise.

For DiffBind under 3 conditions (Figure 1, main text), the set of differential peaks included all peaks deemed to be differential by DiffBind under an FDR control of 0.05 simultaneously for all three pairwise contrast tests between the cell lines Helas3, Hepg2, and Huvec. In the particular genomic position shown in Figure 1, no differential peaks were reported by DiffBind.

For csaw, we used the following setup:
<<echo=TRUE,eval=FALSE,cache=TRUE>>=

# List of bam files
bam.files = list.files(path=paste0(tmpdir,'/chip'),pattern='*.bam$',
                       full.names=T)
bam.files

# Design matrix
design <- model.matrix(~factor(cell.type))
colnames(design) <- c("intercept", "cell.type")
design

# Parameters (PCR duplicates already removed and quality score filtered)
param <- readParam(dedup = F)
param

# Estimating the average fragment length (rescaling all to 200bp)
x = lapply(bam.files,correlateReads,param=param,max.dist=250)
multi.frag.lens = list(unlist(lapply(x,maximizeCcf)),200)
multi.frag.lens

# Counting reads (for a window size of 250bp, for instance)
data <- windowCounts(bam.files,width = 250,ext = multi.frag.lens,
                     param = param,filter = 20)
data

# Filtering data
data.large <- windowCounts(bam.files,width=2500,bin=T,param=param)

bin.ab <- scaledAverage(data.large, scale=median(getWidths(data.large))/
                          median(getWidths(data)))

threshold <- median(bin.ab) + log2(2)

keep.global <- aveLogCPM(asDGEList(data)) >  threshold

sum(keep.global)

# Creating filtered data
filtered.data <- data[keep.global,]

# Testing for DB (assuming composition bias is negligble,
# i.e. cell lines should exhibit a balanced number of DB regions)

y <- DGEList(assay(filtered.data), lib.size = filtered.data$totals)
y$samples$norm.factors <- 1 
y$offset <- NULL
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design, robust = TRUE)
out <- glmQLFTest(fit,contrast = contrast)
tabres <- topTags(out, nrow(out))$table
tabres <- tabres[order(as.integer(rownames(tabres))),]

merged <- mergeWindows(rowRanges(filtered.data), tol=tol,
                       max.width=max.width)
tabneg <- combineTests(merged$id, tabres)
@

For ChIPComp and DiffBind, candidate peaks were called in advance using MACS with the following syntax:
\begin{itemize}
\item MACS: \textit{macs2 callpeak -f BAM -g 2.80e+09 -B 'options' -t 'sample' -c 'control' --outdir 'output' -n 'filename'}
\end{itemize}
such that $options = \{\text{--broad --broad-cutoff 0.1}\}$ if H3K27me3, H3K36me3, or EZH2, or $options = \{\text{-q 0.01}\}$ otherwise.

\section{Software}
epigraHMM was implemented in a R package that is available on the GitHub repository \if1\blind{https://github.com/plbaldoni/epigraHMM}\fi \if0\blind{(BLINDED FOR REVIEW)}\fi.

epigraHMM is a package with a differential peak caller to detect differential enrichment regions from multiple ChIP-seq experiments with replicates. The main function of the package is \textit{epigraHMM}(). The package allows the user to specify a set of parameters that control the Expectation-Maximization (EM) algorithm. These parameters include, for instance, the convergence (and termination) criteria of the algorithm and the threshold value for the rejection controlled EM algorithm. These parameters can be defined by the function \textit{controlEM}(). Please refer to the package documentation (e.g. \textit{?epigraHMM::epigraHMM}) for additional details and the complete help manual.

\section{Code}
The necessary code to replicate the results presented in the main article and in the supplementary material can be downloaded from \if1\blind{https://github.com/plbaldoni/epigraHMMPaper}\fi \if0\blind{(BLINDED FOR REVIEW)}\fi.

\bibliographystyle{amsplain}
\bibliography{bibliography}

\end{document}

### 2021-02-18 Latex/appendix/webappendix.tex
Adding latex files from appendix & paper

--- 2021-02-18 Latex/appendix/webappendix.Rnw
@@
-\documentclass[12pt,oneside]{book}
+\documentclass[12pt,oneside]{book}\usepackage[]{graphicx}\usepackage[]{color}
+% maxwidth is the original width if it is less than linewidth
+% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 } \fi
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 For csaw, we used the following setup:
-<<echo=TRUE,eval=FALSE,cache=TRUE>>=
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# List of bam files}
+\hlstd{bam.files} \hlkwb{=} \hlkwd{list.files}\hlstd{(}\hlkwc{path}\hlstd{=}\hlkwd{paste0}\hlstd{(tmpdir,}\hlstr{'/chip'}\hlstd{),}\hlkwc{pattern}\hlstd{=}\hlstr{'*.bam$'}\hlstd{,}
+                       \hlkwc{full.names}\hlstd{=T)}
+\hlstd{bam.files}
 
-# List of bam files
-bam.files = list.files(path=paste0(tmpdir,'/chip'),pattern='*.bam$',
-                       full.names=T)
-bam.files
+\hlcom{# Design matrix}
+\hlstd{design} \hlkwb{<-} \hlkwd{model.matrix}\hlstd{(}\hlopt{~}\hlkwd{factor}\hlstd{(cell.type))}
+\hlkwd{colnames}\hlstd{(design)} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"intercept"}\hlstd{,} \hlstr{"cell.type"}\hlstd{)}
+\hlstd{design}
 
-# Design matrix
-design <- model.matrix(~factor(cell.type))
-colnames(design) <- c("intercept", "cell.type")
-design
+\hlcom{# Parameters (PCR duplicates already removed and quality score filtered)}
+\hlstd{param} \hlkwb{<-} \hlkwd{readParam}\hlstd{(}\hlkwc{dedup} \hlstd{= F)}
+\hlstd{param}
 
-# Parameters (PCR duplicates already removed and quality score filtered)
-param <- readParam(dedup = F)
-param
+\hlcom{# Estimating the average fragment length (rescaling all to 200bp)}
+\hlstd{x} \hlkwb{=} \hlkwd{lapply}\hlstd{(bam.files,correlateReads,}\hlkwc{param}\hlstd{=param,}\hlkwc{max.dist}\hlstd{=}\hlnum{250}\hlstd{)}
+\hlstd{multi.frag.lens} \hlkwb{=} \hlkwd{list}\hlstd{(}\hlkwd{unlist}\hlstd{(}\hlkwd{lapply}\hlstd{(x,maximizeCcf)),}\hlnum{200}\hlstd{)}
+\hlstd{multi.frag.lens}
 
-# Estimating the average fragment length (rescaling all to 200bp)
-x = lapply(bam.files,correlateReads,param=param,max.dist=250)
-multi.frag.lens = list(unlist(lapply(x,maximizeCcf)),200)
-multi.frag.lens
+\hlcom{# Counting reads (for a window size of 250bp, for instance)}
+\hlstd{data} \hlkwb{<-} \hlkwd{windowCounts}\hlstd{(bam.files,}\hlkwc{width} \hlstd{=} \hlnum{250}\hlstd{,}\hlkwc{ext} \hlstd{= multi.frag.lens,}
+                     \hlkwc{param} \hlstd{= param,}\hlkwc{filter} \hlstd{=} \hlnum{20}\hlstd{)}
+\hlstd{data}
 
-# Counting reads (for a window size of 250bp, for instance)
-data <- windowCounts(bam.files,width = 250,ext = multi.frag.lens,
-                     param = param,filter = 20)
-data
+\hlcom{# Filtering data}
+\hlstd{data.large} \hlkwb{<-} \hlkwd{windowCounts}\hlstd{(bam.files,}\hlkwc{width}\hlstd{=}\hlnum{2500}\hlstd{,}\hlkwc{bin}\hlstd{=T,}\hlkwc{param}\hlstd{=param)}
 
-# Filtering data
-data.large <- windowCounts(bam.files,width=2500,bin=T,param=param)
+\hlstd{bin.ab} \hlkwb{<-} \hlkwd{scaledAverage}\hlstd{(data.large,} \hlkwc{scale}\hlstd{=}\hlkwd{median}\hlstd{(}\hlkwd{getWidths}\hlstd{(data.large))}\hlopt{/}
+                          \hlkwd{median}\hlstd{(}\hlkwd{getWidths}\hlstd{(data)))}
 
-bin.ab <- scaledAverage(data.large, scale=median(getWidths(data.large))/
-                          median(getWidths(data)))
+\hlstd{threshold} \hlkwb{<-} \hlkwd{median}\hlstd{(bin.ab)} \hlopt{+} \hlkwd{log2}\hlstd{(}\hlnum{2}\hlstd{)}
 
-threshold <- median(bin.ab) + log2(2)
+\hlstd{keep.global} \hlkwb{<-} \hlkwd{aveLogCPM}\hlstd{(}\hlkwd{asDGEList}\hlstd{(data))} \hlopt{>}  \hlstd{threshold}
 
-keep.global <- aveLogCPM(asDGEList(data)) >  threshold
+\hlkwd{sum}\hlstd{(keep.global)}
 
-sum(keep.global)
+\hlcom{# Creating filtered data}
+\hlstd{filtered.data} \hlkwb{<-} \hlstd{data[keep.global,]}
 
-# Creating filtered data
-filtered.data <- data[keep.global,]
+\hlcom{# Testing for DB (assuming composition bias is negligble,}
+\hlcom{# i.e. cell lines should exhibit a balanced number of DB regions) }
 
-# Testing for DB (assuming composition bias is negligble,
-# i.e. cell lines should exhibit a balanced number of DB regions)
+\hlstd{y} \hlkwb{<-} \hlkwd{DGEList}\hlstd{(}\hlkwd{assay}\hlstd{(filtered.data),} \hlkwc{lib.size} \hlstd{= filtered.data}\hlopt{$}\hlstd{totals)}
+\hlstd{y}\hlopt{$}\hlstd{samples}\hlopt{$}\hlstd{norm.factors} \hlkwb{<-} \hlnum{1}
+\hlstd{y}\hlopt{$}\hlstd{offset} \hlkwb{<-} \hlkwa{NULL}
+\hlstd{y} \hlkwb{<-} \hlkwd{estimateDisp}\hlstd{(y, design)}
+\hlstd{fit} \hlkwb{<-} \hlkwd{glmQLFit}\hlstd{(y, design,} \hlkwc{robust} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\hlstd{out} \hlkwb{<-} \hlkwd{glmQLFTest}\hlstd{(fit,}\hlkwc{contrast} \hlstd{= contrast)}
+\hlstd{tabres} \hlkwb{<-} \hlkwd{topTags}\hlstd{(out,} \hlkwd{nrow}\hlstd{(out))}\hlopt{$}\hlstd{table}
+\hlstd{tabres} \hlkwb{<-} \hlstd{tabres[}\hlkwd{order}\hlstd{(}\hlkwd{as.integer}\hlstd{(}\hlkwd{rownames}\hlstd{(tabres))),]}
 
-y <- DGEList(assay(filtered.data), lib.size = filtered.data$totals)
-y$samples$norm.factors <- 1 
-y$offset <- NULL
-y <- estimateDisp(y, design)
-fit <- glmQLFit(y, design, robust = TRUE)
-out <- glmQLFTest(fit,contrast = contrast)
-tabres <- topTags(out, nrow(out))$table
-tabres <- tabres[order(as.integer(rownames(tabres))),]
-
-merged <- mergeWindows(rowRanges(filtered.data), tol=tol,
-                       max.width=max.width)
-tabneg <- combineTests(merged$id, tabres)
-@
+\hlstd{merged} \hlkwb{<-} \hlkwd{mergeWindows}\hlstd{(}\hlkwd{rowRanges}\hlstd{(filtered.data),} \hlkwc{tol}\hlstd{=tol,}
+                       \hlkwc{max.width}\hlstd{=max.width)}
+\hlstd{tabneg} \hlkwb{<-} \hlkwd{combineTests}\hlstd{(merged}\hlopt{$}\hlstd{id, tabres)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 

### 2021-02-18 Latex/paper/paper.tex
Adding latex files from appendix & paper

@@
 Third, the analysis of ChIP-seq data is often subject to complex biases that may vary across the genome, as differences in local read enrichment may depend on quantities such as the total read abundance in a given region or GC content \citep{teng2017accounting}.
-DPCs that disregard such complex effects or that solely rely on sample-specific global scaling factors or control subtraction methods \citep{stark2011diffbind,shen2013diffreps,chen2015novel,allhoff2016differential} may be prone to detecting spurious differences due to the lack of non-linear normalization methods \citep{lun2015csaw}.
+DPCs that disregard such complex effects or that solely rely on sample-specific global scaling factors or control subtraction methods \citep{shen2013diffreps,chen2015novel,allhoff2016differential} may be prone to detecting spurious differences due to the lack of non-linear normalization methods \citep{lun2015csaw}.
 Reflecting these limitations, a recent comparison of DPCs demonstrated that current methods tend to detect either a large number of short peaks (low sensitivity) or exhibit a high number of false positive calls (low specificity) in ChIP-seq experiments with broad regions of enrichment \citep{steinhauser2016comprehensive}.
-Moreover, few methods are able to simultaneously test for differential activity across three or more conditions \citep{chen2015novel,lun2015csaw}, or can classify specific differential combinatorial patterns.
+Moreover, only few methods are able to simultaneously test for differential activity across three or more conditions \citep{stark2011diffbind,chen2015novel,lun2015csaw}, or can classify specific differential combinatorial patterns.
 Altogether, these limitations can impact the drawing of accurate insights from modern epigenomic studies.
@@
 Third, it enables the simultaneous detection and classification of epigenomic changes under three or more conditions, a novelty not yet available in any other DPC algorithm.
-The presented method offers additional benefits over current HMM-based DPC algorithms \citep{song2011identifying,allhoff2016differential} that include a GLM-based framework with an embedded mixture model, which allows the modeling of covariates of interest as well as the inclusion of model offsets for non-linear normalization, and a fast and accurate parameter estimation scheme via rejection-controlled EM algorithm (RCEM).
+The presented method, epigraHMM (Figure~\ref{fig:fig_motivation}), offers additional benefits over current HMM-based DPC algorithms \citep{song2011identifying,allhoff2016differential} that includes a flexible GLM-based framework with an embedded mixture model, which allows the modeling of covariates of interest, the inclusion of model offsets for non-linear normalization and GC-bias adjustment, and a fast and accurate parameter estimation scheme via rejection-controlled EM algorithm (RCEM).
 
@@
 Using ChIP-seq data pertaining to histone modifications H3K27me3 and H3K36me3, and to EZH2 from the ENCODE Consortium, we find that current DPCs have difficulty in accurately detecting broad regions of differential enrichment between several common cell lines (Figure~\ref{fig:fig_motivation}).
-In line with previous findings \citep{steinhauser2016comprehensive}, we observe that even current DPCs designed for broad data \citep{song2011identifying,allhoff2016differential} tend to detect either overly fragmented differential peaks or call regions exhibiting no difference in experimental signal between conditions as differential (Figure~\ref{fig:fig_motivation}A).
+In line with previous findings \citep{steinhauser2016comprehensive}, we observe that even current DPCs designed for broad data \citep{song2011identifying,allhoff2016differential} tend to detect either overly fragmented differential peaks or call regions exhibiting no difference in experimental signal between conditions as differential (Sections \ref{sec:sim2} and \ref{sec:broad}).
 The low specificity and sensitivity of such methods may impair the biological interpretation of the resulting peak calls in downstream analyses.
@@
 The classification of such patterns would allow researchers to, for example, quantify treatment responses on the epigenomic level \citep{clouaire2014cfp1}, or identify sets of processes working together to regulate local chromatin state.
-We find that the performance of such methods exhibit low sensitivity and specificity in calling differential regions in broad marks (Figure~\ref{fig:fig_motivation}B).
+We find that the performance of such methods exhibit low sensitivity and specificity in calling differential regions in broad marks (Figure~\ref{fig:fig_motivation}).
 \begin{figure}
@@
 If $G$ conditions are of interest, there are $L = 2^G-2$ possible differential combinatorial patterns of enrichment and background across conditions at a given window.
-The emission distribution pertaining to the differential state models all $L$ possible differential combinatorial patterns via a mixture model with mixture proportions $\bs{\delta} = (\delta_{1},\ldots,\delta_{L})^\prime$, such that $\sum_{l=1}^{L}\delta_{l}=1$ (Web Figure 1).
+The emission distribution pertaining to the differential state models all $L$ possible differential combinatorial patterns via a mixture model with mixture proportions $\bs{\delta} = (\delta_{1},\ldots,\delta_{L})^\prime$, such that $\sum_{l=1}^{L}\delta_{l}=1$.
 
@@
 with $y_{hij}\in\{0,1,2,\ldots\}$, such that $\log(\mu_{(1,hij)})=\beta_{1} + u_{hij}$, $\log(\phi_{1})=\lambda_{1}$, $\log(\mu_{(3,hij)})=\beta_{1} + \beta_{3} + u_{hij}$, and $\log(\phi_{3})=\lambda_{1} + \lambda_{3}$.
-The offset $u_{hij}$ adjusts for technical artifacts and allows the non-linear normalization of the signal profile across genomic windows, conditions, and samples (Web Appendix B2).
+The offset $u_{hij}$ adjusts for technical or biological artifacts (such as GC-content bias) and allows the non-linear normalization of the signal profile across genomic windows, conditions, and samples (Web Appendix B2).
 When $u_{hij}=0$, $\beta_{1}$ and $\lambda_{1}$ represent the log-mean and log-dispersion, respectively, of read counts pertaining to consensus background state windows, whereas $\beta_{3}$ and $\lambda_{3}$ represent the difference in log-mean and log-dispersion of read counts from consensus enrichment state windows relative to consensus background state windows.
@@
 That is, $\mb{x}_{l}$ is the $G \times 1$ vector of binary indicator variables denoting which subset of conditions are enriched in pattern (mixture component) $l$.
-A graphical illustration of our proposed model is provided in Web Figure 1.
 
@@
 \subsubsection{Simulation Results}
-Table~\ref{Table:table_simulation} shows the true values and the average relative bias of parameter estimates (and the $2.5^{th}$, $97.5^{th}$ percentiles) from a hundred simulated data sets relative to the scenario of H3K27me3 with $10^5$ genomic windows (see Web Appendix C1 for additional details).
-Results are shown for different levels of SNR, number of conditions, and number of replicates per condition.
+Table~\ref{Table:table_simulation} shows the true values and the average relative bias of parameter estimates (and the $2.5^{th}$, $97.5^{th}$ percentiles) from a hundred simulated data sets relative to the scenario of H3K27me3 with $10^6$ genomic windows (see Web Appendix C1 for additional details).
+Results are shown for the SNR level estimated from the ENCODE project (mean count ratio between peak and enrichment regions of 3.21), different number of conditions, and different number of replicates per condition.
 Overall, no significant differences regarding the relative bias of parameter estimates were observed across simulations under different number of genomic windows.
@@
 In general, scenarios with higher SNR showed lower relative bias and variability of the parameter estimates in comparison to scenarios with lower SNR, regardless of the number of conditions or replicates per condition.
-In scenarios with lower SNR levels or higher number of conditions, these results also highlight the importance of experimental replicates to achieve precise parameter estimates.
+In scenarios with higher number of conditions, these results highlight the importance of experimental replicates to achieve precise parameter estimates.
 The proposed estimation approach via EM algorithm led to precise parameter estimates and was robust to a data generating mechanism that was different than the one assumed by the proposed model.
@@
 Results presented in this section pertain to the analysis of two cell lines, namely Helas3 and Hepg2.
+Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR).
+We did not observe a significant improvement in performance by accounting for input control effect or GC-content bias with epigraHMM and did not adjust our analyses for these effects.
 A discussion about the choice of the window size is presented in the Web Appendix D2.
@@
 In Figure~\ref{fig:fig_broad}F, we show examples of differential peak calls for EZH2.
-Our method was among the fastest algorithms due to our computational scheme, taking approximately 1 hour to analyze genome-wide data (Figure~\ref{fig:fig_broad}E).
+Our method was among the most efficient algorithms due to our computational scheme, taking approximately 30 minutes to analyze genome-wide data (Figure~\ref{fig:fig_broad}E).
 \begin{figure}
@@
 Posterior probabilities of the HMM differential state are shown at the bottom of each panel.
-(E): computing time of genome-wide analysis from various methods.
+(E): computing time and peak memory usage of genome-wide analysis from various methods.
 This figure appears in color in the electronic version of this article.
@@
 It has been shown that certain HMM-based approaches, including RSEG, have low accuracy under short histone modification marks and TF data \citep{hocking2016optimizing}.
-As we show, the model proposed in this article performs comparably to the evaluated DPCs known to perform best in short data (ChIPComp, DiffBind; \cite{steinhauser2016comprehensive}) and appeared to be more efficient regarding the computation time in certain scenarios.
+As we show, the model proposed in this article performs comparably to the evaluated DPCs known to perform best in short data (ChIPComp, DiffBind; \cite{steinhauser2016comprehensive}).
 
@@
 Posterior probabilities of the HMM differential state are shown at the bottom of each panel.
-(E): computing time of genome-wide analysis from various methods.
+(E): computing time and peak memory usage of genome-wide analysis from various methods.
 Results are shown under a nominal FDR control of 0.05.
@@
 Differential genomic windows were mostly representing either transcribed chromatin states or repressed chromatin states (Figure~\ref{fig:fig_multimarks}B).
-All differential combinatorial patterns expected to be rare had associated mixture proportion estimates less than 0.02 (see Web Appendix B3 for a discussion on pruning rare states).
+All differential combinatorial patterns expected to be rare had associated mixture proportion estimates less than 0.05 and were pruned using our model selection scheme via BIC presented in Web Appendix B3 (Figure~\ref{fig:fig_multimarks}C).
 %.
@@
 Genes associated with transcribed chromatin states had a significantly higher distribution of TPM counts than genes associated with repressed chromatin states (Figure~\ref{fig:fig_multimarks}C).
+In fact, all known marker genes of the Helas3 cell line \citep{morin2008profiling} were associated with differential enrichment for H3K36me3, measured by the associated average mixture posterior proportions of the overlapping differential peaks (Figure~\ref{fig:fig_multimarks}D).
 We detected broad differential regions of enrichment and the classification of differential combinatorial patterns agreed with their biological roles as well as the expression levels of associated gene bodies.
-Figure~\ref{fig:fig_multimarks}D shows an example of a genomic region with differential enrichment for the analyzed marks.
+Figure~\ref{fig:fig_multimarks}E shows an example of a genomic region with differential enrichment for the analyzed marks.
 We compare our results to ChromHMM with 3 states, a method developed for chromatin segmentation.
@@
 Lastly, our method allows the simultaneous detection and classification of differential combinatorial patterns of enrichment from its embedded mixture model and the associated posterior probabilities under any number of conditions.
+In situations where certain combinatorial patterns are rare, prunning of specific combinatorial patterns is possible via model selection via BIC.
 Our software has been implemented into an R package and is available for download (Web Appendix A).

### 2021-02-18 Supp/Figure_PR_Viterbi/Figure_PR_H3K36me3_Encode_twocells_1000bp.R
Viterbi scripts

@@ -1,0 +1,380 @@
+library(magrittr)
+library(data.table)
+library(tidyverse)
+library(tibble)
+library(RColorBrewer)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(vsn)
+library(DESeq2)
+library(ggrepel)
+library(Polychrome)
+library(ggrepel)
+
+# General parameters
+bp = 1000
+sizeanno <- 2.5
+chromosome = paste0('chr', 1:22)
+mark = 'H3K36me3'
+data = 'Encode_twocells'
+cutoff = 2
+methods <-
+  c(
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    "epigraHMM",
+    'RSEG',
+    'THOR',
+    'Genes'
+  )
+colors = c(kelly.colors(22)[c('red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'lightblue',
+                              'buff',
+                              'orange')], 'black')
+names(colors) <- methods
+
+getcov = function(gr.predicted,
+                  gr.genome,
+                  dt,
+                  threshold,
+                  method,
+                  postprob = F,
+                  viterbi = F) {
+  if (viterbi) {
+    out = list()
+    dt[, pred :=  overlapsAny(gr.genome, gr.predicted)]
+    out[['Viterbi']] = data.table(
+      Method = method,
+      Threshold = 'Viterbi',
+      TPR = sum(dt[, DE == T &
+                     pred == T]) / sum(dt[, DE == T]),
+      FPR = sum(dt[, DE == F &
+                     pred == T]) / sum(dt[, DE == F]),
+      PPV = sum(dt[, DE == T &
+                     pred == T]) / sum(dt[, pred == T]),
+      Width.Mean = mean(width(gr.predicted)),
+      Width.Median = median(width(gr.predicted)),
+      Width.SD = sd(width(gr.predicted))
+    )
+    dt[, pred := NULL]
+    out = rbindlist(out)
[…]

### 2021-02-18 Supp/Figure_PR_Viterbi/Figure_PR_H3K36me3_Encode_twocells_250bp.R

@@ -1,0 +1,380 @@
+library(magrittr)
+library(data.table)
+library(tidyverse)
+library(tibble)
+library(RColorBrewer)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(vsn)
+library(DESeq2)
+library(ggrepel)
+library(Polychrome)
+library(ggrepel)
+
+# General parameters
+bp = 250
+sizeanno <- 2.5
+chromosome = paste0('chr', 1:22)
+mark = 'H3K36me3'
+data = 'Encode_twocells'
+cutoff = 2
+methods <-
+  c(
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    "epigraHMM",
+    'RSEG',
+    'THOR',
+    'Genes'
+  )
+colors = c(kelly.colors(22)[c('red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'lightblue',
+                              'buff',
+                              'orange')], 'black')
+names(colors) <- methods
+
+getcov = function(gr.predicted,
+                  gr.genome,
+                  dt,
+                  threshold,
+                  method,
+                  postprob = F,
+                  viterbi = F) {
+  if (viterbi) {
+    out = list()
+    dt[, pred :=  overlapsAny(gr.genome, gr.predicted)]
+    out[['Viterbi']] = data.table(
+      Method = method,
+      Threshold = 'Viterbi',
+      TPR = sum(dt[, DE == T &
+                     pred == T]) / sum(dt[, DE == T]),
+      FPR = sum(dt[, DE == F &
+                     pred == T]) / sum(dt[, DE == F]),
+      PPV = sum(dt[, DE == T &
+                     pred == T]) / sum(dt[, pred == T]),
+      Width.Mean = mean(width(gr.predicted)),
+      Width.Median = median(width(gr.predicted)),
+      Width.SD = sd(width(gr.predicted))
+    )
+    dt[, pred := NULL]
+    out = rbindlist(out)
[…]

### 2021-02-18 Supp/Figure_PR_Viterbi/Figure_PR_H3K36me3_Encode_twocells_500bp.R

@@ -1,0 +1,380 @@
+library(magrittr)
+library(data.table)
+library(tidyverse)
+library(tibble)
+library(RColorBrewer)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(vsn)
+library(DESeq2)
+library(ggrepel)
+library(Polychrome)
+library(ggrepel)
+
+# General parameters
+bp = 500
+sizeanno <- 2.5
+chromosome = paste0('chr', 1:22)
+mark = 'H3K36me3'
+data = 'Encode_twocells'
+cutoff = 2
+methods <-
+  c(
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    "epigraHMM",
+    'RSEG',
+    'THOR',
+    'Genes'
+  )
+colors = c(kelly.colors(22)[c('red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'lightblue',
+                              'buff',
+                              'orange')], 'black')
+names(colors) <- methods
+
+getcov = function(gr.predicted,
+                  gr.genome,
+                  dt,
+                  threshold,
+                  method,
+                  postprob = F,
+                  viterbi = F) {
+  if (viterbi) {
+    out = list()
+    dt[, pred :=  overlapsAny(gr.genome, gr.predicted)]
+    out[['Viterbi']] = data.table(
+      Method = method,
+      Threshold = 'Viterbi',
+      TPR = sum(dt[, DE == T &
+                     pred == T]) / sum(dt[, DE == T]),
+      FPR = sum(dt[, DE == F &
+                     pred == T]) / sum(dt[, DE == F]),
+      PPV = sum(dt[, DE == T &
+                     pred == T]) / sum(dt[, pred == T]),
+      Width.Mean = mean(width(gr.predicted)),
+      Width.Median = median(width(gr.predicted)),
+      Width.SD = sd(width(gr.predicted))
+    )
+    dt[, pred := NULL]
+    out = rbindlist(out)
[…]

### 2021-02-18 Supp/Figure_PR_Viterbi/Figure_PR_H3K36me3_Encode_twocells_750bp.R

@@ -1,0 +1,380 @@
+library(magrittr)
+library(data.table)
+library(tidyverse)
+library(tibble)
+library(RColorBrewer)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(vsn)
+library(DESeq2)
+library(ggrepel)
+library(Polychrome)
+library(ggrepel)
+
+# General parameters
+bp = 750
+sizeanno <- 2.5
+chromosome = paste0('chr', 1:22)
+mark = 'H3K36me3'
+data = 'Encode_twocells'
+cutoff = 2
+methods <-
+  c(
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    "epigraHMM",
+    'RSEG',
+    'THOR',
+    'Genes'
+  )
+colors = c(kelly.colors(22)[c('red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'lightblue',
+                              'buff',
+                              'orange')], 'black')
+names(colors) <- methods
+
+getcov = function(gr.predicted,
+                  gr.genome,
+                  dt,
+                  threshold,
+                  method,
+                  postprob = F,
+                  viterbi = F) {
+  if (viterbi) {
+    out = list()
+    dt[, pred :=  overlapsAny(gr.genome, gr.predicted)]
+    out[['Viterbi']] = data.table(
+      Method = method,
+      Threshold = 'Viterbi',
+      TPR = sum(dt[, DE == T &
+                     pred == T]) / sum(dt[, DE == T]),
+      FPR = sum(dt[, DE == F &
+                     pred == T]) / sum(dt[, DE == F]),
+      PPV = sum(dt[, DE == T &
+                     pred == T]) / sum(dt[, pred == T]),
+      Width.Mean = mean(width(gr.predicted)),
+      Width.Median = median(width(gr.predicted)),
+      Width.SD = sd(width(gr.predicted))
+    )
+    dt[, pred := NULL]
+    out = rbindlist(out)
[…]

### 2021-02-18 Latex/appendix/webappendix.Rnw
Updating appendix

@@
 
-For ChIPComp and DiffBind, candidate peaks were called in advance using MACS with the following syntax:
+For ChIPComp and DiffBind, candidate peaks were called in advance using MACS2 with the following syntax:
 \begin{itemize}
-\item MACS: \textit{macs2 callpeak -f BAM -g 2.80e+09 -B 'options' -t 'sample' -c 'control' --outdir 'output' -n 'filename'}
+\item MACS2: \textit{macs2 callpeak -f BAM -g 2.80e+09 -B 'options' -t 'sample' -c 'control' --outdir 'output' -n 'filename'}
 \end{itemize}
@@
 
+\chapter{}
+In this appendix, we present additional methodological results and an overview of the EM algorithm proposed in the main article.
+
+\section{Adjustments for nuisance effects}
+\label{s:offset}
+
+\subsection{Normalization for non-linear biases via model offsets}
+
+In our analyses, we observed that the magnitude of the local differences in read counts between conditions changed with the average of local read coverage.
+Here, we accounted for these trended differences to avoid calling spurious differential peaks due to the different magnitude of library sizes across groups.
+Specifically, we implemented an approach similar to the non-linear normalization method used by csaw as follows \cite{lun2015csaw}.
+First, we create a reference sample of read counts comprised by the geometric mean of read counts from all replicates and conditions.
+Then, we fitted a loess curve on the difference between the read counts of each sample and the reference, on the average of those two quantities.
+A similar approach was first implemented by \cite{lun2015csaw} and is available in their software.
+Here, we add a continuity correction of 1 to avoid discarding genomic windows with zero counts.
+Using the smoothed curve as the model offset, we observed better results than a simple correction via either the total sum of read counts or cell-specific median log ratio.
+The rationale behind this approach is to create a reference library in which each genomic window is the geometric mean of counts across all conditions and replicates, and then read counts are properly adjusted by accounting for the smoothed differences between each individual library and the reference library.
+A useful way to evaluate the performance of this normalization method is to compare samples with respect to their adjusted read counts.
+For example, plotting the ratios between counts and the calculated offsets $y_{hij}/\exp(u_{hij})$ for all samples in the study.
+In Figure \ref{maplot} we show an example of a genomic region from three analyzed cell lines and their respective MA plot, unadjusted ChIP counts, and offset-adjusted ChIP counts.
+After accounting for the offset, the read counts from Helas3 are adjusted to its larger library size with respect to the other under sequenced cell lines.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics{../../Supp/Figure_Normalization/Figure_Normalization.pdf}}
+\end{center}
+\caption{MA plot of read counts from three distinct analyzed cell lines (top), unadjusted ChIP read counts (center), and offset-adjusted ChIP read counts (bottom) from a given genomic region on chromosome 19.
+The blue line in the MA plots shows the offset created via loess smoothing.
+\label{maplot}}
+\end{figure}
+
+\subsection{Input control adjustment in differential peak calling}
+
+Our implementation (Web Appendix A4) allows the optional inclusion of continuous covariates in the model with state-specific parametrization.
+The main purpose of the inclusion of such covariates in the model is the adjustment for input control (or any other continuous variable, such as autoregressive counts) that can be helpful in distinguishing background from enrichment signal.
+Several methods for differential peak calling allow the inclusion of input control in their computational framework \cite{stark2011diffbind,shen2013diffreps,chen2015novel,allhoff2016differential}.
+However, \cite{lun2015csaw} point out that "(...) controls are mostly irrelevant when testing for DB between ChIP samples.".
+To evaluate this claim, we ran an analysis of real data and simulated data while accounting for the input control effect.
+
+To asses whether accounting for input control effect leads to an improvement in performance, we utilized the smoothing technique proposed by \cite{chen2015novel} to account for input controls and autoregressive counts.
+Specifically, we fitted generalized additive models (GAM, instead of loess smoothing) in the data normalization step while accounting for input control (or autoregressive counts) as a covariate.
+The resulting fitted curve was then used in the analysis as model offsets.
+
+First, we analyzed real data by smoothing the input control effect and autoregressive counts with a two-step approach.
+Specifically, we first called peaks without the inclusion of extra covariates in the model, and then utilized the called differential peaks from the first step to smooth the covariates for each HMM predicted state.
+Predicted smoothing curves from the GAM approach were then passed as model offsets in a second step of analysis.
+As claimed by \cite{lun2015csaw}, we observed minor differences in the results that would justify their inclusion in the analysis.
+Results from the histone modification mark H3K36me3 are presented in Figure \ref{fig:control2}.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_ENCODE_Control/Figure_PR_H3K36me3_Encode_twocells_500bp.pdf}}
+\end{center}
+\caption{ROC curves for H3K36me3 utilizing no input controls (epigraHMM), input control only (epigraHMM + Control), autoregressive counts only (epigraHMM + AR Counts), and smoothing of both input controls and autoregressive counts (epigraHMM + Control \& AR Counts)
+\label{fig:control2}}
+\end{figure}
+
+Next, we reasoned that our approach of modeling input control effect with state-specific parametrization could not be ideal, since independent controls were available for every sample and there could exist sample-specific effects not captured by our model.
+We then attempted to verify the utility of including input control into the differential binding analysis by simulating data where ChIP-counts were generated such that their log-mean had a linear relationship with input controls (Figure \ref{fig:control3}).
+We then fitted three different models that differed regarding the inclusion of input control: a model without control, a model with control, and a model with controls where the smoothing was calculated separately for each latent HMM state.
+Again, results did not show significant improvement by including the effect of control in the analysis.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Sim/Control/Summary/Summary.pdf}}
+\end{center}
+\caption{Results from simulated data (A) where the log-means of ChIP-seq counts were generated as a linear function of input controls (B).
+Sensitivity/specificity analyses did not show significant improvement by including the effect of control in the offset scheme.
+\label{fig:control3}}
+\end{figure}
+
+Overall, we did not observe a significant improvement in performance by including input control in differential peak calling.
+Although several methods do offer the option of including controls in their analysis pipeline, we did not find that their inclusion was justifiable under our modelling assumptions.
+Our findings are in agreement with \cite{lun2015csaw}.
+
+\subsection{GC-content bias in broad marks}
+
+\subsection{Bayesian Information Criterion (BIC) for Hidden Markov Models}
+
+The BIC for hidden Markov models has been discussed by \cite{zucchini2017hidden}. For the presented three-state HMM, one can calculate the BIC as
+\begin{align}\label{eq:bic}
+BIC &= -2\log\left(\sum_{r=1}^{3}f_{Mr}^{p}\right) + (11+L)\log\left(M\sum_{h=1}^{G}n_{h}\right),
+\end{align}
+where $f_{Mr}^{p}$ is the forward probability pertaining to the $r^{th}$ state calculated at the (last) $M^{th}$ genomic window (as detailed in the Appendix of the main text), $L$ is the number of mixture components, $G$ is the number of conditions, and $n_{h}$ is the number of replicates pertaining to condition $h$.
+The number of model parameters to be estimated is $(11+L)$: 6 transition probabilities, 2 initial probabilities, 4 model coefficients pertaining to the emission distributions, and $L-1$ prior probabilities from the mixture model.
+
+As shown in the main text, the proposed HMM is robust to situations where certain combinatorial patterns are rare.
+However, if pruning rare combinatorial patterns is still of interest, such a task can be performed by making use of the BIC.
+For the analysis of $G$ experimental conditions with a given BIC threshold $\epsilon$, say $\epsilon=0.01$, and $L=2^G-2$ mixture components, one can prune rare combinatorial patterns by the following algorithm.
+
+\begin{enumerate}
+\item Fit the three-state HMM with $L$ mixture components (model $L$) and compute the model BIC, BIC$_{L}$, as in Equation \ref{eq:bic}.
+\item Fit a reduced three-state HMM with $L-1$ mixture components (model $L-1$) by excluding the component associated with the rarest combinatorial pattern of enrichment. Compute its BIC, BIC$_{L-1}$.
+\item Calculate $\Delta\text{BIC}=(\text{BIC}_{L-1}-\text{BIC}_{L})/\text{BIC}_{L}$.
+If $|\Delta\text{BIC}|\leq\epsilon$, set $L \leftarrow L-1$ and return to step 1.
+If $|\Delta\text{BIC}|>\epsilon$, stop and set the model L as the final model.
+\end{enumerate}
+
+In scenarios where the number of mixture components is smaller than $2^G-2$, the implemented method initializes the EM algorithm by clustering genomic windows with respect to the posterior probabilities of enrichment obtained from a initial run of a two-state HMM to classify genomic windows into background and enrichment windows.
+Such an initialization improves the overall computation time by reducing the time to convergence of the presented EM algorithm.
+
+We also compare the results of our method with ChromHMM, an algorithm developed for chromatin segmentation.
+In Figure \ref{fig:chromHMM}, panels A-D, we present results of ChromHMM with 3 (ideal), 4, 5, and 6 states.
+By using the BIC for model selection, one could easily choose the number of biologically relevant mixture components to be included in the model, a task that may not be as straightforward in methods such as ChromHMM (see Supplementary Figure 4 in \cite{ernst2012chromhmm}).
+Our method offers the benefit of simultaneously detecting differential peaks and classifying the combinatorial pattern of enrichment through mixture model posterior probabilities even in the context of genomic segmentation.
+Despite the choice of the number of mixture components via BIC, epigraHMM appeared to be robust in scenarios with rare combinatorial patterns.
+This is so because we utilize a constrained parametrization in which only 4 GLM-specific parameters need to be estimated regardless of the number of mixture components, and in the genome-wide analysis of ChIP-seq data one often has enough data to estimate such quantities.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_ChromHMM/Figure_ChromHMM.pdf}}
+\end{center}
+\caption{Comparative analysis of epigraHMM and ChromHMM with 3 (ideal), 4, 5, 6, 7, and 8 states (panels A-F).
+\label{fig:chromHMM}}
+\end{figure}
+
+\subsubsection{Simulation Study}
+
+\section{Study on FDR Control and the Viterbi Algorithm}
+\label{s:viterbi}
+
+In this section, we present results comparing the FDR controlling approach with the Viterbi algorithm.
+We evaluated the FDR approach using cutoffs 0.01, 0.05, 0.10, 0.15, and 0.20.
+We compared the results between the two approaches using window sizes of 250bp, 500bp, 750bp, and 1000bp.
+Overall, we observed that the Viterbi sequence of states led to similar results than the sequences based on FDR control cutoffs across all choices of window size.
+Specifically, we observed that the sensitivity and specificity of the sequence of Viterbi states were close to those from FDR control, in particular for FDR control 0.10.
+These results are shown in Figures \ref{fig:encode.viterbi.250}, \ref{fig:encode.viterbi.500}, \ref{fig:encode.viterbi.750}, and \ref{fig:encode.viterbi.1000}.
+These facts are also reflected by the length and number of called peaks.
+In Figures \ref{fig:encode.viterbi.h3k36me3} and \ref{fig:encode.viterbi.h3k27me3} we show examples of peak calls of H3K26me3 and H3K27me3, respectively, from all FDR control cutoffs and the Viterbi sequence of states.
+Overall, we observe minor differences regarding the size of peak calls of the Viterbi and FDR control sequences across different choices of window sizes.
+These differences were mainly present in the data for H3K27me3, which is known to be a histone mark that expands through broader domains than H3K36me3.
+Finally, it is worth noting that the Viterbi algorithm gives us a way to call peaks that does not depend on the choice of the FDR cutoff.
+
 \bibliographystyle{amsplain}

### 2021-02-18 Latex/appendix/webappendix.tex
Updating appendix

@@
 
+\chapter{}
+In this appendix, we present additional methodological results and an overview of the EM algorithm proposed in the main article.
+
+\section{Adjustments for nuisance effects}
+\label{s:offset}
+
+\subsection{Normalization for non-linear biases via model offsets}
+
+In our analyses, we observed that the magnitude of the local differences in read counts between conditions changed with the average of local read coverage.
+Here, we accounted for these trended differences to avoid calling spurious differential peaks due to the different magnitude of library sizes across groups.
+Specifically, we implemented an approach similar to the non-linear normalization method used by csaw as follows \cite{lun2015csaw}.
+First, we create a reference sample of read counts comprised by the geometric mean of read counts from all replicates and conditions.
+Then, we fitted a loess curve on the difference between the read counts of each sample and the reference, on the average of those two quantities.
+A similar approach was first implemented by \cite{lun2015csaw} and is available in their software.
+Here, we add a continuity correction of 1 to avoid discarding genomic windows with zero counts.
+Using the smoothed curve as the model offset, we observed better results than a simple correction via either the total sum of read counts or cell-specific median log ratio.
+The rationale behind this approach is to create a reference library in which each genomic window is the geometric mean of counts across all conditions and replicates, and then read counts are properly adjusted by accounting for the smoothed differences between each individual library and the reference library.
+A useful way to evaluate the performance of this normalization method is to compare samples with respect to their adjusted read counts.
+For example, plotting the ratios between counts and the calculated offsets $y_{hij}/\exp(u_{hij})$ for all samples in the study.
+In Figure \ref{maplot} we show an example of a genomic region from three analyzed cell lines and their respective MA plot, unadjusted ChIP counts, and offset-adjusted ChIP counts.
+After accounting for the offset, the read counts from Helas3 are adjusted to its larger library size with respect to the other under sequenced cell lines.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics{../../Supp/Figure_Normalization/Figure_Normalization.pdf}}
+\end{center}
+\caption{MA plot of read counts from three distinct analyzed cell lines (top), unadjusted ChIP read counts (center), and offset-adjusted ChIP read counts (bottom) from a given genomic region on chromosome 19.
+The blue line in the MA plots shows the offset created via loess smoothing.
+\label{maplot}}
+\end{figure}
+
+\subsection{Input control adjustment in differential peak calling}
+
+Our implementation (Web Appendix A4) allows the optional inclusion of continuous covariates in the model with state-specific parametrization.
+The main purpose of the inclusion of such covariates in the model is the adjustment for input control (or any other continuous variable, such as autoregressive counts) that can be helpful in distinguishing background from enrichment signal.
+Several methods for differential peak calling allow the inclusion of input control in their computational framework \cite{stark2011diffbind,shen2013diffreps,chen2015novel,allhoff2016differential}.
+However, \cite{lun2015csaw} point out that "(...) controls are mostly irrelevant when testing for DB between ChIP samples.".
+To evaluate this claim, we ran an analysis of real data and simulated data while accounting for the input control effect.
+
+To asses whether accounting for input control effect leads to an improvement in performance, we utilized the smoothing technique proposed by \cite{chen2015novel} to account for input controls and autoregressive counts.
+Specifically, we fitted generalized additive models (GAM, instead of loess smoothing) in the data normalization step while accounting for input control (or autoregressive counts) as a covariate.
+The resulting fitted curve was then used in the analysis as model offsets.
+
+First, we analyzed real data by smoothing the input control effect and autoregressive counts with a two-step approach.
+Specifically, we first called peaks without the inclusion of extra covariates in the model, and then utilized the called differential peaks from the first step to smooth the covariates for each HMM predicted state.
+Predicted smoothing curves from the GAM approach were then passed as model offsets in a second step of analysis.
+As claimed by \cite{lun2015csaw}, we observed minor differences in the results that would justify their inclusion in the analysis.
+Results from the histone modification mark H3K36me3 are presented in Figure \ref{fig:control2}.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_ENCODE_Control/Figure_PR_H3K36me3_Encode_twocells_500bp.pdf}}
+\end{center}
+\caption{ROC curves for H3K36me3 utilizing no input controls (epigraHMM), input control only (epigraHMM + Control), autoregressive counts only (epigraHMM + AR Counts), and smoothing of both input controls and autoregressive counts (epigraHMM + Control \& AR Counts)
+\label{fig:control2}}
+\end{figure}
+
+Next, we reasoned that our approach of modeling input control effect with state-specific parametrization could not be ideal, since independent controls were available for every sample and there could exist sample-specific effects not captured by our model.
+We then attempted to verify the utility of including input control into the differential binding analysis by simulating data where ChIP-counts were generated such that their log-mean had a linear relationship with input controls (Figure \ref{fig:control3}).
+We then fitted three different models that differed regarding the inclusion of input control: a model without control, a model with control, and a model with controls where the smoothing was calculated separately for each latent HMM state.
+Again, results did not show significant improvement by including the effect of control in the analysis.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Sim/Control/Summary/Summary.pdf}}
+\end{center}
+\caption{Results from simulated data (A) where the log-means of ChIP-seq counts were generated as a linear function of input controls (B).
+Sensitivity/specificity analyses did not show significant improvement by including the effect of control in the offset scheme.
+\label{fig:control3}}
+\end{figure}
+
+Overall, we did not observe a significant improvement in performance by including input control in differential peak calling.
+Although several methods do offer the option of including controls in their analysis pipeline, we did not find that their inclusion was justifiable under our modelling assumptions.
+Our findings are in agreement with \cite{lun2015csaw}.
+
+\section{Bayesian Information Criterion (BIC) for Hidden Markov Models}
+
+The BIC for hidden Markov models has been discussed by \cite{zucchini2017hidden}. For the presented three-state HMM, one can calculate the BIC as
+\begin{align}\label{eq:bic}
+BIC &= -2\log\left(\sum_{r=1}^{3}f_{Mr}^{p}\right) + (11+L)\log\left(M\sum_{h=1}^{G}n_{h}\right),
+\end{align}
+where $f_{Mr}^{p}$ is the forward probability pertaining to the $r^{th}$ state calculated at the (last) $M^{th}$ genomic window (as detailed in the Appendix of the main text), $L$ is the number of mixture components, $G$ is the number of conditions, and $n_{h}$ is the number of replicates pertaining to condition $h$.
+The number of model parameters to be estimated is $(11+L)$: 6 transition probabilities, 2 initial probabilities, 4 model coefficients pertaining to the emission distributions, and $L-1$ prior probabilities from the mixture model.
+
+As shown in the main text, the proposed HMM is robust to situations where certain combinatorial patterns are rare.
+However, if pruning rare combinatorial patterns is still of interest, such a task can be performed by making use of the BIC.
+For the analysis of $G$ experimental conditions with a given BIC threshold $\epsilon$, say $\epsilon=0.01$, and $L=2^G-2$ mixture components, one can prune rare combinatorial patterns by the following algorithm.
+
+\begin{enumerate}
+\item Fit the three-state HMM with $L$ mixture components (model $L$) and compute the model BIC, BIC$_{L}$, as in Equation \ref{eq:bic}.
+\item Fit a reduced three-state HMM with $L-1$ mixture components (model $L-1$) by excluding the component associated with the rarest combinatorial pattern of enrichment. Compute its BIC, BIC$_{L-1}$.
+\item Calculate $\Delta\text{BIC}=(\text{BIC}_{L-1}-\text{BIC}_{L})/\text{BIC}_{L}$.
+If $|\Delta\text{BIC}|\leq\epsilon$, set $L \leftarrow L-1$ and return to step 1.
+If $|\Delta\text{BIC}|>\epsilon$, stop and set the model L as the final model.
+\end{enumerate}
+
+In scenarios where the number of mixture components is smaller than $2^G-2$, the implemented method initializes the EM algorithm by clustering genomic windows with respect to the posterior probabilities of enrichment obtained from a initial run of a two-state HMM to classify genomic windows into background and enrichment windows.
+Such an initialization improves the overall computation time by reducing the time to convergence of the presented EM algorithm.
+
+We also compare the results of our method with ChromHMM, an algorithm developed for chromatin segmentation.
+In Figure \ref{fig:chromHMM}, panels A-D, we present results of ChromHMM with 3 (ideal), 4, 5, and 6 states.
+By using the BIC for model selection, one could easily choose the number of biologically relevant mixture components to be included in the model, a task that may not be as straightforward in methods such as ChromHMM (see Supplementary Figure 4 in \cite{ernst2012chromhmm}).
+Our method offers the benefit of simultaneously detecting differential peaks and classifying the combinatorial pattern of enrichment through mixture model posterior probabilities even in the context of genomic segmentation.
+Despite the choice of the number of mixture components via BIC, epigraHMM appeared to be robust in scenarios with rare combinatorial patterns.
+This is so because we utilize a constrained parametrization in which only 4 GLM-specific parameters need to be estimated regardless of the number of mixture components, and in the genome-wide analysis of ChIP-seq data one often has enough data to estimate such quantities.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_ChromHMM/Figure_ChromHMM.pdf}}
+\end{center}
+\caption{Comparative analysis of epigraHMM and ChromHMM with 3 (ideal), 4, 5, 6, 7, and 8 states (panels A-F).
+\label{fig:chromHMM}}
+\end{figure}
+
 \bibliographystyle{amsplain}

### 2021-02-18 Sim/Control/Analysis.R
Adding control simulation

@@ -1,0 +1,91 @@
+rm(list=ls())
+workdir <- getwd()
+
+system('mkdir ./Analysis/')
+
+dirdt <- '/pine/scr/b/a/baldoni/Rashid/Project2/Simulation_Control/Data'
+
+dirs <- list.dirs(path = dirdt,full.names = FALSE,recursive = FALSE)
+
+for(i in dirs){
+    
+    diranalysis <- paste0('./Analysis/',i)
+    
+    system(paste('mkdir',diranalysis))
+    
+    for(sim in 1:100){
+        sink(file = file.path(diranalysis,paste0('Simulation',sim,'.R')))
+        cat("
+
+library(epigraHMM)
+
+load('",file.path(dirdt,i,paste0(i,'_',sim,'.RData')),"')
+dat1 <- dat
+load('",file.path(dirdt,i,paste0(i,'_',sim+100,'.RData')),"')
+dat2 <- dat
+rm(dat)
+
+M <- nrow(dat1)
+controlEM <- epigraHMM::controlEM(quiet = F)
+
+chip <- cbind(as.matrix(dat1[,c('y.1','y.2')]),as.matrix(dat2[,c('y.1','y.2')]))
+control <- cbind(as.matrix(dat1[,c('x.1','x.2')]),as.matrix(dat2[,c('x.1','x.2')]))
+z <- dat1$z+dat2$z
+
+# Model without control
+
+object.nocontrol <- epigraHMM::epigraHMMDataSetFromMatrix(countData = chip,
+                                       colData = data.frame(condition = c('A','A','B','B'),replicate = c(1,2,1,2)),
+                                       rowRanges = GenomicRanges::GRanges('chr1',IRanges::IRanges(1+500*0:(M-1),500+500*0:(M-1))))
+
+object.nocontrol <- createOffset(object = object.nocontrol,type = 'loess',span = 1)
+
+object.nocontrol <- epigraHMM(object.nocontrol,control = controlEM,type = 'differential',dist = 'nb',random = F)
+
+
+# Model with control
+
+object.withcontrol <- epigraHMM::epigraHMMDataSetFromMatrix(countData = chip,
+                                       colData = data.frame(condition = c('A','A','B','B'),replicate = c(1,2,1,2)),
+                                       rowRanges = GenomicRanges::GRanges('chr1',IRanges::IRanges(1+500*0:(M-1),500+500*0:(M-1))))
+
[…]

### 2021-02-18 Sim/Control/Data.R

@@ -1,0 +1,115 @@
+library(Rcpp)
+library(plyr)
+
+### Simulation scnearios
+# Heterogeneity: low (L), medium (M), or high (H)
+# Sample: 2, 3, 6, or 9
+# Random: intercept (I), or slope (S)
+# Nwindow: 25,000
+# Nsim: 100
+
+scenario = expand.grid(Heterogeneity=c('L'),Sample=c(2),Random=c('I'),Nwindow=25000,Nsim=200)
+scenario$Label = with(scenario,paste0('H',Heterogeneity,'N0',Sample,'R',Random))
+print(scenario)
+
+system('mkdir Data/')
+
+## Initial probability (always begin with background) and Transition Matrix
+pi = c(1,0)
+gamma = matrix(0,2,2,byrow=T)
+diag(gamma) = c(0.995,0.99)
+gamma[1,2] = 1 - gamma[1,1]
+gamma[2,1] = 1 - gamma[2,2]
+
+## Control Parameters: mean (mux), dispersion (phix)
+# In real data (Helas3 rep 1, chr1, Control), mux = exp(2.287) =  9.845357
+# In real data (Helas3 rep 1, chr1, Control), phix = 2.970831525
+mux = 0.7560612#9.8453
+phix = 3.230474#2.9708
+
+## ChIP Parameters:
+genpar = function(random,heterogeneity){
+    if(random=='I'){
+        sigma2 = ifelse(heterogeneity=='L',1e-8,ifelse(heterogeneity=='M',1e-8,ifelse(heterogeneity=='H',1e-8,NA)))
+        # ChIP parameters
+        b1 = c(0.7560612,2.4169519)
+        b2 = c(1,0.50)
+        phi = c(3.230474,2.891908)
+    }
+    return(list('sigma2'=sigma2,'b1'=b1,'b2'=b2,'phi'=phi))
+}
+
+# Zero-inflation
+# Real data suggests p(zero inflation | background) = 0.12 and p(zero inflation | enrichment) = 0.02
+# I will use p(zero inflation | background) = 0.10 and p(zero inflation | enrichment) = 0
+# The following gives a 10% of zero inflation when colntrol=9 (mean)
+# Intercept and Slope for logit model for background only
+csi = c(-1000,-1000) 
+
+for(i in 1:nrow(scenario)){
+    scenario.i = scenario[i,]
+    
+    cmd = paste0('mkdir Data/',scenario.i$Label)
[…]

### 2021-02-18 Sim/Control/Summary.R

@@ -1,0 +1,99 @@
+rm(list=ls())
+workdir <- getwd()
+
+library(data.table)
+library(ggplot2)
+library(magrittr)
+library(plyr)
+
+dirdt <- '/pine/scr/b/a/baldoni/Rashid/Project2/Simulation_Control/Analysis'
+datdir <- '/pine/scr/b/a/baldoni/Rashid/Project2/Simulation_Control/Data'
+dirout <- 'Summary'
+idx = 7500:9500
+dt.plot <- list()
+system(paste('mkdir',dirout))
+
+dirs <- list.dirs(path = dirdt,full.names = FALSE,recursive = FALSE)
+
+res <- list()
+
+for(i in dirs){
+    subres <- list()
+    for(sim in 1:100){
+        message(i,', simulation ',sim)
+        
+        tryCatch({
+            load(file.path(dirdt,i,paste0('Simulation',sim,'.RData')))
+            
+            dt <- as.data.table(dt)[,Window := 1:.N]
+            dt <- melt(dt,id.vars = c('Window','GoldStandard'),measure.vars = c('WithoutControl','WithControl','WithGroupedControl'),variable.name = 'Method',value.name = 'Call')
+            subres[[sim]] <- dt[,.(Sensitivity = sum(Call=='D' & GoldStandard==1)/sum(GoldStandard==1),
+                                   Specificity = sum(!Call=='D' & !GoldStandard==1)/sum(!GoldStandard==1)),by = 'Method'][,Scenario := i][,Simulation := sim]
+            
+            if(sim==1){
+                load(file.path(datdir,i,paste0(i,"_",sim,".RData")))
+                dat1 <- dat
+                rm(dat)
+                load(file.path(datdir,i,paste0(i,"_",sim+100,".RData")))
+                dat2 <- dat
+                rm(dat)
+                
+                dt.plot[[paste0(i,"_",sim)]] <- rbindlist(list(as.data.table(dat1[,c('y.1','y.2','z','x.1','x.2')])[,Scenario := i][,Condition := 'A'][,Window := 1:nrow(dat1)],
+                                                               as.data.table(dat2[,c('y.1','y.2','z','x.1','x.2')])[,Scenario := i][,Condition := 'B'][,Window := 1:nrow(dat2)]))
+            }
+            
+        },error = function(cond){message(cond)})
+    }
[…]

### 2021-02-19 Latex/appendix/webappendix.Rnw
Adding supplementary files and updating tex

@@
 
-\subsection{Bayesian Information Criterion (BIC) for Hidden Markov Models}
+\section{Bayesian Information Criterion (BIC) for Hidden Markov Models}
 
@@
 
-\subsubsection{Simulation Study}
+\subsection{Simulation Study}
 
@@
 Specifically, we observed that the sensitivity and specificity of the sequence of Viterbi states were close to those from FDR control, in particular for FDR control 0.10.
-These results are shown in Figures \ref{fig:encode.viterbi.250}, \ref{fig:encode.viterbi.500}, \ref{fig:encode.viterbi.750}, and \ref{fig:encode.viterbi.1000}.
+These results are shown in Figure \ref{fig:encode.viterbi}.
 These facts are also reflected by the length and number of called peaks.
@@
 
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_PR_Viterbi/Figure_PR_Viterbi.pdf}}
+\end{center}
+\caption{FDR-based results from broad marks and Viterbi-based result from epigraHMM with 250bp (A), 500bp (B), 750bp (C), and 1000bp (D).}
+\label{fig:encode.viterbi}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_H3K36me3_Viterbi/Figure_H3K36me3_Viterbi.pdf}}
+\end{center}
+\caption{FDR- and Viterbi-based peak calls from H3K36me3 with 250bp (A), 500bp, (B), 750bp (C), and 1000bp (D).}
+\label{fig:encode.viterbi.h3k36me3}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_H3K27me3_Viterbi/Figure_H3K27me3_Viterbi.pdf}}
+\end{center}
+\caption{FDR- and Viterbi-based peak calls from H3K27me3 with 250bp (A), 500bp, (B), 750bp (C), and 1000bp (D).}
+\label{fig:encode.viterbi.h3k27me3}
+\end{figure}
+
+\section{HMM Emission Distributions}
+
+For the consensus background ($r=1$) and consensus enrichment ($r=3$) states, the emission distribution function is
+\begin{align}\label{eq:nondiffdensity}
+f_{r}(\mb{y}_{..j}|\bs{\psi}_{r}) &= \prod^{G}_{h=1}\prod^{n_{h}}_{i=1}f_{r}(y_{hij}|\bs{\psi}_{r}),\quad r\in\{1,3\}\quad\text{and}\quad y_{hij}\in\{0,1,2,\ldots\},\nonumber\\
+&=\prod^{G}_{h=1}\prod^{n_{h}}_{i=1}\text{Pr}(Y_{hij}=y_{hij}|Z_{j}=r;\bs{\psi}_{r}),\nonumber\\
+&=\prod^{G}_{h=1}\prod^{n_{h}}_{i=1}\frac{\Gamma(y_{hij}+\phi_{r})}{y_{hij}!\Gamma(\phi_{r})}\left(\frac{\phi_{r}}{\mu_{(r,hij)}+\phi_{r}}\right)^{\phi_{r}}\left(\frac{\mu_{(r,hij)}}{\mu_{(r,hij)}+\phi_{r}}\right)^{y_{hij}}.
+\end{align}
+
+For the differential state ($r=2$), the emission distribution is
+\begin{align}\label{eq:diffdensity}
+f_{2}(\mb{y}_{..j}|\mb{x};\bs{\delta},\bs{\psi}_{2}) &= \sum_{l=1}^{L}\delta_{l}f_{(2,l)}\left(\mb{y}_{..j}|\mb{x}_{l};\bs{\psi}_{(2,l)}\right),\quad y_{hij}\in\{0,1,2,\ldots\},\\
+&= \sum_{l=1}^{L}\delta_{l}\prod_{h=1}^{G}\prod_{i=1}^{n_{h}}Pr\left(Y_{hij}=y_{hij}|Z_{j}=2,\mb{x}_{l};\bs{\psi}_{(2,l)}\right),\nonumber\\
+&= \sum_{l=1}^{L}\delta_{l}\prod_{h=1}^{G}\prod_{i=1}^{n_{h}}\frac{\Gamma\left(y_{hij}+\phi_{(2,l,h)}\right)}{y_{hij}!\Gamma\left(\phi_{(2,l,h)}\right)}\left(\frac{\phi_{(2,l,h)}}{\mu_{(2,l,hij)}+\phi_{(2,l,h)}}\right)^{\phi_{(2,l,h)}}\times\nonumber\\
+&\times\left(\frac{\mu_{(2,l,hij)}}{\mu_{(2,l,hij)}+\phi_{(2,l,h)}}\right)^{y_{hij}}.\nonumber
+\end{align}
+
+Apart from the offset $u_{hij}$, we will assume that replicates from the same (different) condition share common (distinct) mean and dispersion parameters under every mixing probability distribution $f_{(2,l)}$.
+To define all possible combinations of background and enrichment across $G$ conditions, we consider the following sets of singletons $A_{1}$, pairs $A_{2}$, $\ldots$, and $(G-1)$-tuples $A_{G-1}$ such that
+\begin{align*}
+A_{1} &= \left\{a^{(1)}\left. \right\vert a^{(1)}\in\mathbb{G}_{+}\text{ and }a^{(1)}\leq G\right\},\\
+A_{2} &= \left\{(a^{(2)}_{1},a^{(2)}_{2})\left. \right\vert (a^{(2)}_{1},a^{(2)}_{2})\in\mathbb{G}^{2}_{+}\text{ and }a^{(2)}_{1}<a^{(2)}_{2}\leq G\right\},\\
+&\vdots\\
+A_{G-1} &= \left\{(a^{(G-1)}_{g})^{G-1}_{g=1}\left. \right\vert (a^{(G-1)}_{g})^{G-1}_{g=1}\in\mathbb{G}^{G-1}_{+}\text{ and }a^{(G-1)}_{1}<\ldots<a^{(G-1)}_{G-1}\leq G\right\}.
+\end{align*}
+The union of all sets $A = \cup^{G-1}_{k=1}A_{k}$ contains an exhaustive list of $L = 2^{G}-2$ elements that determines the differential pattern across $G$ conditions such that each element of $A$ indicates which of the $G$ conditions are enriched.
+For instance, if $G=3$, $A_{1} = \{1,2,3\}$ and $A_{2} = \{(1,2),(1,3),(2,3)\}$ define the six possible combinations of enrichment and background across three conditions.
+Then, we define a bijective mapping $A \rightarrow S_{1},\ldots,S_{L}$ and let $x_{hl} = I(h\in S_{l})$ indicate whether the read count of genomic window $j$ from replicate $i$ of condition $h$ is enriched in the mixture component $l$.
+We model the log-mean $\mu_{(2,l,hij)}$ and log-dispersion $\phi_{(2,l,h)}$ of mixture $l$ from the emission distribution (\ref{eq:diffdensity}) as
+\begin{align*}
+\log(\mu_{(2,l,hij)}) &= \beta_{1} + \beta_{3}x_{hl} + u_{hij},\quad \text{and}\\
+\log(\phi_{(2,l,h)}) &= \lambda_{1} + \lambda_{3}x_{hl}.
+\end{align*}
+According to this parametrization, $\beta_{1}$ and $\lambda_{1}$ are the baseline log-mean and log-dispersion parameters of the read count distribution from replicates of conditions that are not enriched under the mixing distribution $l$.
+Conversely, $\beta_{1}+\beta_{3}$ and $\lambda_{1}+\lambda_{3}$ are the baseline log-mean and log-dispersion parameters of the read count distribution from replicates of conditions enriched under the mixing distribution $l$.
+This choice of parametrization ensures that windows exhibiting differential enrichment across conditions share means and dispersions that are common between the remaining non differential HMM states.
+
+A pseudo code of the presented EM algorithm is below.
+\begin{enumerate}
+\item{Initialize $\bs{\pi}^{(0)}, \bs{\gamma}^{(0)}, \bs{\delta}^{(0)}, \beta_{1}^{(0)}, \beta_{3}^{(0)}, \lambda_{1}^{(0)}, \lambda_{3}^{(0)}$, such that $\sum_{r=1}^3\pi_{r}^{(0)}=1$ and $\sum_{s=1}^{3}\gamma_{rs}=1$.}
+\item{E step ($t\geq1$),}
+\begin{enumerate}
+\item{Calculate $Pr\left(Z_{j}=r|\mb{y},\mb{x};\bs{\Psi}^{(t-1)}\right)$ and $Pr\left(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t-1)}\right)$ for all $r$ and $s$ in $\left\{1,2,3\right\}$ and $j=1,\ldots,M$ via Forward-Backward algorithm as detailed in Appendix B of the main article}
+\item{Calculate $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t-1)})$ for all $l$ $\left\{1,\ldots,L\right\}$ and $j=1,\ldots,M$ as $f_{(2,l)}(\mb{y}_{..j}|\mb{x}_{l};\bs{\psi}_{(2,l)}^{(t-1)})\delta_{l}^{(t-1)}/\sum_{k=1}^{L}f_{(2,k)}(\mb{y}_{..j}|\mb{x}_{k};\bs{\psi}_{(2,k)}^{(t-1)})\delta_{k}^{(t-1)}$}
+\end{enumerate}
+\item{M step ($t\geq1$),}
+\begin{enumerate}
+\item Maximize Equation (5) (main text) with respect to the initial and transition probabilities to obtain for all $r$ and $s$ in $\left\{1,2,3\right\}$
+\begin{align*}
+\pi_{r}^{(t)} &= Pr\left(Z_{1}=r|\mb{y},\mb{x};\bs{\Psi}^{(t-1)}\right)\\
+\gamma_{rs}^{(t)} &= \sum_{j=2}^{M}Pr(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t-1)})/\sum_{j=2}^{M}Pr(Z_{j-1}=r|\mb{y},\mb{x};\bs{\Psi}^{(t-1)})
+\end{align*}
+\item Maximize Equation (5) with respect to $\bs{\delta}$ to obtain $\bs{\delta}^{(t)}$ such that $\sum_{l=1}^L\delta_{l}^{(t)}=1$.
+\item Conditionally upon $\bs{\delta}^{(t)}$, maximize Equation (5) with respect to $\beta_{1}, \beta_{3}, \lambda_{1}, \lambda_{3}$ to obtain $\beta_{1}^{(t)}, \beta_{3}^{(t)}, \lambda_{1}^{(t)}, \lambda_{3}^{(t)}$,
+\item Iterate between (b) and (c) until convergence.
+\end{enumerate}
+\item{Iterate between 2. and 3. until convergence.}
+\end{enumerate}
+
+\subsection{$Q$-function of the EM algorithm}
+% \label{appendix:appendix_Q}
+The $Q$-function of the EM algorithm is defined as $Q\left(\bs{\Psi}|\bs{\Psi}^{(t)}\right) = Q_{0}\left(\bs{\pi},\bs{\gamma}|\bs{\Psi}^{(t)}\right)+Q_{1}\left(\bs{\psi}_{1}|\bs{\Psi}^{(t)}\right)+Q_{2}\left(\bs{\delta},\bs{\psi}_{2}|\bs{\Psi}^{(t)}\right)+Q_{3}\left(\bs{\psi}_{3}|\bs{\Psi}^{(t)}\right)$, such that
+
+\begin{align}
+\label{eq:appendix_Q}
+Q_{0}\left(\bs{\pi},\bs{\gamma}|\bs{\Psi}^{(t)}\right) &= \sum_{r=1}^{3}\left\{Pr\left(Z_{1}=r|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)\log(\pi_{r})\right\}+\nonumber\\
+&+\sum_{j=2}^{M}\sum_{r=1}^{3}\sum_{s=1}^{3}\left\{Pr\left(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)\log(\gamma_{rs})\right\},\nonumber\\
+Q_{1}\left(\bs{\psi}_{1}|\bs{\Psi}^{(t)}\right) &= \sum_{j=1}^{M}Pr\left(Z_{j}=1|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)\log f_{1}\left(\mb{y}_{..j}|\bs{\psi}_{1}\right),\nonumber\\
+Q_{2}\left(\bs{\delta},\bs{\psi}_{2}|\bs{\Psi}^{(t)}\right) &= \sum_{j=1}^{M}Pr\left(Z_{j}=2|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)\sum_{l=1}^{L}Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})\times\nonumber\\
+&\times \log\delta_{l}f_{(2,l)}\left(\mb{y}_{..j}|\mb{x}_l;\bs{\psi}_{(2,l)}\right),\quad\text{and}\nonumber\\
+Q_{3}\left(\bs{\psi}_{3}|\bs{\Psi}^{(t)}\right) &= \sum_{j=1}^{M}Pr\left(Z_{j}=3|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)\log f_{3}\left(\mb{y}_{..j}|\bs{\psi}_{3}\right).
+\end{align}
+
+During the $t^{th}$ iteration of the EM-algorithm, one solves $\argmax{\beta_{1}^{(t)}, \beta_{3}^{(t)}, \lambda_{1}^{(t)}, \lambda_{3}^{(t)}} Q_{1}\left(\bs{\psi}_{1}|\bs{\Psi}^{(t)}\right) + Q_{2}\left(\bs{\delta},\bs{\psi}_{2}|\bs{\Psi}^{(t)}\right) + Q_{3}\left(\bs{\psi}_{3}|\bs{\Psi}^{(t)}\right)$ during the maximization step.
+This problem is analogous to estimating parameters in a weighted NB regression model, since $Q_{1}$, $Q_{2}$, and $Q_{3}$ can be seen as weighted log-likelihood functions from a NB generalized linear model with means and dispersions parametrized as described in the Section 3.1 of the main article.
+
+The forward probabilities are defined as $f_{11}^{p} = \pi_{1}f_{1}(\mb{y}_{..1}|\bs{\psi}_{1})$, $f_{12}^{p} = \pi_{2}f_{2}(\mb{y}_{..1}|\mb{x};\bs{\delta},\bs{\psi}_{2})$, $f_{13}^{p} = \pi_{3}f_{3}(\mb{y}_{..1}|\bs{\psi}_{3})$, and, for $j=1,\ldots,M$, $f_{j1}^{p} = \sum_{l=1}^{3}\gamma_{l1}f_{(j-1)l}^{p}f_{1}(\mb{y}_{..j}|\bs{\psi}_{1})$, $f_{j2}^{p} = \sum_{l=1}^{3}\gamma_{l2}f_{(j-1)l}^{p}f_{2}(\mb{y}_{..j}|\mb{x};\bs{\delta},\bs{\psi}_{2})$, and $f_{j3}^{p} = \sum_{l=1}^{3}\gamma_{l3}f_{(j-1)l}^{p}f_{3}(\mb{y}_{..j}|\bs{\psi}_{3})$.
+Conversely, the backward probabilities are defined as $b_{Mk}^{p}=1$, $\forall k=1,2,3$, and, for $j=1,\ldots,M$, $b_{j1}^{p} = \sum_{l=1}^{3}\gamma_{1l}b_{(j+1)l}^{p}f_{1}(\mb{y}_{..(j+1)}|\bs{\psi}_{1})$, $b_{j2}^{p} = \sum_{l=1}^{3}\gamma_{2l}b_{(j+1)l}^{p}f_{2}(\mb{y}_{..(j+1)}|\mb{x};\bs{\delta},\bs{\psi}_{2})$, and $b_{j3}^{p} = \sum_{l=1}^{3}\gamma_{3l}b_{(j+1)l}^{p}f_{3}(\mb{y}_{..(j+1)}|\bs{\psi}_{3})$.
+Then, we have
+
+\begin{align}
+\label{eq:appendix_FBalgorithm}
+P\left(Z_{j}=k|\mathbf{y},\mathbf{x};\bs{\Psi}\right) &= \frac{f_{jk}^{p}b_{jk}^{p}}{\sum_{l=1}^{3}f_{Ml}^{p}},\qquad \forall j=1,\ldots,M\quad\text{and}\quad k=1,2,3,\nonumber\\
+P\left(Z_{j-1}=l,Z_{j}=1|\mathbf{y},\mathbf{x};\bs{\Psi}\right) &= \frac{f_{(j-1)l}^{p}\gamma_{l1}f_{1}(\mb{y}_{..j}|\bs{\psi}_{1})b_{j1}^{p}}{\sum_{l=1}^{3}f_{Ml}^{p}},\nonumber\\
+P\left(Z_{j-1}=l,Z_{j}=2|\mathbf{y},\mathbf{x};\bs{\Psi}\right) &= \frac{f_{(j-1)l}^{p}\gamma_{l2}f_{2}(\mb{y}_{..j}|\mb{x};\bs{\delta},\bs{\psi}_{2})b_{j2}^{p}}{\sum_{l=1}^{3}f_{Ml}^{p}},\nonumber\\
+P\left(Z_{j-1}=l,Z_{j}=3|\mathbf{y},\mathbf{x};\bs{\Psi}\right) &= \frac{f_{(j-1)l}^{p}\gamma_{l3}f_{3}(\mb{y}_{..j}|\bs{\psi}_{3})b_{j3}^{p}}{\sum_{l=1}^{3}f_{Ml}^{p}},\forall j=2,\ldots,M\text{ and }l=1,2,3.
+\end{align}
+
+\chapter{}
+In this appendix, we present an overview of the simulation scenarios and additional results.
+
+\section{Read Count Simulation}
+\label{s:sim1}
+
+In this section, we present additional results from the read count based simulation.
+
+For each scenario, we simulated a hundred data sets as follows.
+First, we generated a sequence of hidden states with length $M$ from a first-order Markov chain with $2^G$ states representing every combination of background and enrichment across $G$ conditions.
+Here, we let states 1 and $2^G$ represent consensus background and consensus enrichment, respectively.
+Transition probabilities were chosen such that long sequences of enrichment regions (average length of 50 genomic windows) and equal proportions of differential states were generated, while the majority of the simulated states pertained to consensus background.
+Secondly, read counts were simulated for $n$ replicates from $G$ conditions following NB distributions with mean and dispersion parameters that led to a given SNR and respected the differential pattern of enrichment determined by the hidden state.
+For each scenario, this setup led to a simulated matrix of counts of dimension $M\times(n\times G)$.
+The aim of this study was to assess whether the presented model was able to aggregate all $2^G-2$ simulated differential hidden states into its differential HMM state while maintaining a precise parameter estimation scheme.
+The model parameters used in this simulated study were estimated from ENCODE data pertaining to H3K36me3 and H3K27me3 histone modification marks and a shown in the tables presented in this section.
+
+Overall, while results were similar regarding different number of genome sizes and histone modification marks, the number of replicates played a major role in the performance of the model.
+In scenarios with either low SNR or high number of conditions, more replicates were needed to achieve better model performance.
+However, under the observed SNR, two replicates were sufficient to achieve a small relative bias regarding the estimation of most of the model parameters throughout all the evaluated scenarios.
+This is in agreement with the ENCODE Consortium guidelines, which recommends two technical or biological replicates to achieve best performance in ChIP-seq experiments.
+
+\begin{center}
+\input{../../Supp/Table_Simulation_ReadCounts/Table_Simulation_ReadCounts_H3K27me3_1e+06.tex}
+\end{center}
+
+\begin{center}
+\input{../../Supp/Table_Simulation_ReadCounts/Table_Simulation_ReadCounts_H3K36me3_1e+06.tex}
+\end{center}
+
+Next, we assessed the sensitivity of our method to detect simulated differential regions of enrichment.
+First, differential regions were defined from the  HMM posterior probabilities pertaining to the differential state by controlling the total FDR as defined in Section 3.2 (main text).
+For different nominal FDR threshold levels, the model sensitivity was estimated as the proportion of windows correctly assigned as differential out of the total number of simulated differential windows.
+Additionally, the observed FDR was calculated as the proportion of genomic windows incorrectly called as differential out of the total number of called differential windows.
+In Figures~\ref{fig:sim1.fig1}-\ref{fig:sim1.fig3}, panel A shows the average observed true positive rate (y-axis) and the observed FDR (x-axis) for different nominal FDR levels across a hundred simulated data.
+Results are shown for different levels of SNR, number of conditions, and number of replicates per condition.
+Overall, we observed that the number of replicates per condition played a major role on the sensitivity levels of the model, in which scenarios with two and four replicates had the best results regardless of the number of conditions and SNR levels.
+For scenarios with either high number of conditions or low SNR levels, more replicates were needed to achieve higher sensitivity.
+
+Finally, we used the estimated mixture model posterior probabilities $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\hat{\bs{\Psi}})$, $j=1,\ldots,M$, to classify the differential combinatorial patterns of enrichment of detected differential windows.
+To this end, we first calculated the maximum estimated mixture model posterior probability across all $L$ components to determine the most likely differential combinatorial pattern from genomic windows assigned to be part of the differential state.
+Then, we compared the window-based classification with the true window-based simulated states from the Markov Chain (states $2,\ldots,G-1$).
+In Figures~\ref{fig:sim1.fig1}-\ref{fig:sim1.fig3}, panel B shows the confusion matrices of classified (x-axis) and simulated (y-axis) differential windows for a scenario with three conditions.
+Differential combinatorial patterns of enrichment are represented by the sequences of letters 'E' (enrichment) and 'B' (background), such that each letter corresponds to the status of a given condition.
+The number of windows (averaged over a all simulated data sets) is shown as entries of the matrices and represented by the color scale.
+Darker colors on the diagonal entries indicate better agreement between simulated and classified patterns.
+By utilizing the posterior probabilities from the mixture model, we observed a good performance when classifying the differential combinatorial pattern of enrichment from differential windows.
+Results were best under scenarios with higher number of replicates or SNR.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.7]{../../Supp/Figure_ROC_ReadCounts/Figure_ROC_ReadCounts_H3K27me3_1e+06.pdf}}
+\end{center}
+\caption{Read count simulation.
+(A): average observed TPR and FDR for different nominal FDR threshold levels for various scenarios in simulated data of H3K27me3 with $10^6$ genomic windows.
+(B): confusion matrices are shown for the 3 conditions scenario.
+On x- and y-axes the labels indicate the differential classified and simulated patterns, respectively.
+For instance, EBE denotes enrichment in conditions 1 and 3 only.
+Darker colors on the diagonal of the matrix indicate better agreement.}
+\label{fig:sim1.fig1}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.7]{../../Supp/Figure_ROC_ReadCounts/Figure_ROC_ReadCounts_H3K36me3_1e+06.pdf}}
+\end{center}
+\caption{Read count simulation.
+(A): average observed TPR and FDR for different nominal FDR threshold levels for various scenarios in simulated data of H3K36me3 with $10^5$ genomic windows.
+(B): confusion matrices are shown for the 3 conditions scenario.
+On x- and y-axes the labels indicate the differential classified and simulated patterns, respectively.
+For instance, EBE denotes enrichment in conditions 1 and 3 only.
+Darker colors on the diagonal of the matrix indicate better agreement.}
+\label{fig:sim1.fig2}
+\end{figure}
+
+\section{Sequencing Read Simulation}
+\label{s:sim2}
+In this section, we present additional results from the sequencing read simulation using different FDR control threshold for differential peak call.
+
+We performed a second simulation study aiming to compare the proposed model with the current DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, since none of them was designed to accept matrices of window-based read counts as input.
+To this purpose, we revisited the simulation study presented by \cite{lun2015csaw} where data were generated in a more general scheme without a particular read count model assumption.
+Here, sequencing reads from ChIP-seq experiments were generated for two conditions and two replicates per condition.
+For the differential peaks callers ChIPComp and DiffBind that require sets of candidate regions, we followed the analyses presented by \cite{lun2015csaw} and called peaks in advance using HOMER.
+Those were then used as input in the respective software for differential call.
+For ChIPComp and DiffBind, we attempted to call peaks in advance using MACS2.
+However, MACS2.
+exhibited issues to work on simulated data as it was not able to detect peaks.
+Here, we adapted their simulation pipeline aiming to generate broad and diffuse differential regions of enrichment.
+Specifically, a total of 4000 binding sites spaced 60-65 kbp apart were simulated such that each site was formed by three neighboring peaks 10kb wide.
+For each peak, the number of sequencing reads were simulated from a NB distribution with mean 300 and dispersions sampled from an inverse chi-squared distribution with 20 degrees of freedom.
+Reads were positioned on peak sites to form smooth binding profiles.
+To introduce differential enrichment, one or two peaks from 1000 binding sites were randomly chosen and reads pertaining to them were removed from the sequencing library in one of the two conditions.
+This process was repeated for the second condition, which led to a total of 2000 differential peaks.
+Background reads were added to the library to maintain the realistic aspects of the experiment.
+A hundred simulated data sets were generated and peaks were called by all the methods under multiple nominal FDR thresholds.
+For our method and RSEG, window-based posterior probabilities were used to control the total FDR.
+For illustration purposes, we refer to the method presented in this article as epigraHMM.
+
+Specifically, panel D from Figures \ref{fig:sim2.fig1}, \ref{fig:sim2.fig2}, \ref{fig:sim2.fig3}, and \ref{fig:sim2.fig4} show example of peak calls using FDR control cutoffs 0.01, 0.10, 0.15, and 0.20, respectively.
+Overall, results were consistent across different choices of FDR control.
+This fact is also shown by the observed sensitivity and false discovery rate from all methods (panels A).
+We observed a clear distinction of methods regarding the observed sensitivity and specificity such that RSEG, diffReps, and THOR exhibited the highest observed FDR and ChIPComp and DiffBind exhibited the lowest sensitivity.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.7]{../../Supp/Figure_Overview_SequencingReads/Figure_Simulation_SequencingReads_1.pdf}}
+\end{center}
+\caption{Sequencing read simulation.
+(A): average observed sensitivity and FDR for various methods under different nominal FDR levels.
+(B): scatter plot of average ratio of called and simulated peaks (y-axis) and average number of called peaks intersecting true differential regions (x-axis) for different nominal FDR levels.
+(C): box plot of computing time (in minutes) for various algorithms.
+(D): an example of differential peak calls under a nominal FDR threshold 0.01.
+Shaded areas indicate true differential peaks.}
+\label{fig:sim2.fig1}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.7]{../../Supp/Figure_Overview_SequencingReads/Figure_Simulation_SequencingReads_10.pdf}}
+\end{center}
+\caption{Sequencing read simulation.
+(A): average observed sensitivity and FDR for various methods under different nominal FDR levels.
+(B): scatter plot of average ratio of called and simulated peaks (y-axis) and average number of called peaks intersecting true differential regions (x-axis) for different nominal FDR levels.
+(C): box plot of computing time (in minutes) for various algorithms.
+(D): an example of differential peak calls under a nominal FDR threshold 0.10.
+Shaded areas indicate true differential peaks.}
+\label{fig:sim2.fig2}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.7]{../../Supp/Figure_Overview_SequencingReads/Figure_Simulation_SequencingReads_15.pdf}}
+\end{center}
+\caption{Sequencing read simulation.
+(A): average observed sensitivity and FDR for various methods under different nominal FDR levels.
+(B): scatter plot of average ratio of called and simulated peaks (y-axis) and average number of called peaks intersecting true differential regions (x-axis) for different nominal FDR levels.
+(C): box plot of computing time (in minutes) for various algorithms.
+(D): an example of differential peak calls under a nominal FDR threshold 0.15.
+Shaded areas indicate true differential peaks.}
+\label{fig:sim2.fig3}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.7]{../../Supp/Figure_Overview_SequencingReads/Figure_Simulation_SequencingReads_20.pdf}}
+\end{center}
+\caption{Sequencing read simulation.
+(A): average observed sensitivity and FDR for various methods under different nominal FDR levels.
+(B): scatter plot of average ratio of called and simulated peaks (y-axis) and average number of called peaks intersecting true differential regions (x-axis) for different nominal FDR levels.
+(C): box plot of computing time (in minutes) for various algorithms.
+(D): an example of differential peak calls under a nominal FDR threshold 0.20.
+Shaded areas indicate true differential peaks.}
+\label{fig:sim2.fig4}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_Overview_SequencingReads/Figure_SequencingReads_Example.pdf}}
+\end{center}
+\caption{Sequencing read simulation. Examples of differential called peaks from the various methods under an FDR control of 0.05.}
+\label{fig:sim2.fig5}
+\end{figure}
+
+\chapter{}
+In this appendix, we present an overview of the application on data from the ENCODE Consortium.
+
+\section{RNA-seq data and gene transcription}
+\label{s:rnaseq}
+Under the three cell lines scenario (presented in the main text), we used RNA-seq experimental data from the ENCODE Consortium on Helas3, Hepg2, and Huvec human cells to quantify the expression levels of genes as follows.
+First, we used Salmon \cite{patro2017salmon} to quantify transcript expression from cell-specific RNA-seq experiments.
+We then calculated, using the R package $tximport$ \cite{soneson2015differential}, estimated counts using abundance estimates (transcripts per million, TPM) scaled up to the average transcript length over samples and library size.
+This step ensures that counts computed from Salmon are not correlated with the average transcript length.
+Secondly, we calculated the number of ChIP-seq reads from H3K4me3 and H3K27ac overlapping gene promoters.
+Promoter regions extend around the transcription start site 2000 base pairs upstream and 200 base pair downstream.
+Finally, read counts from RNA-seq and ChIP-seq were normalized for sequencing depth using DESeq2.
+
+For the two cell line scenario, differentially transcribed genes were defined through log2 fold changes of H3K36me3 ChIP-seq read counts.
+We observed several cases in which differentially expressed genes (defined by RNA-seq data) did not exhibit differential enrichment for H3K36me3.
+However, due to the activating roles of H3K36me3 on gene transcription, we follow the ideas presented by \cite{steinhauser2016comprehensive} and \cite{ji2013differential} and defined differentially transcribed genes based on log2 fold changes of H3K36me3 ChIP-seq read counts.
+
+Sensitivity and specificity metrics were calculated on the window-level as follows.
+For non-overlapping genomic windows $b_{1},\ldots,b_{M}$, let $g_{j}=I(b_{j}\in\text{ differentially transcribed gene})$ and $d_{j}=I(b_{j}\in\text{ differential peak})$ denote the indicators of genomic windows being associated with either differential gene bodies or differential peaks, respectively, for $j=1,\ldots,M$, for a given method and nominal FDR level.
+Then, the observed sensitivity (TPR), specificity (1-FPR), and FDR were calculated as follows:
+\begin{align}
+\text{Sensitivity} &= \frac{\sum_{j=1}^{M}{g_{j}d_{j}}}{\sum_{j=1}^{M}{g_{j}}}\nonumber\\
+\text{Specificity} &= \frac{\sum_{j=1}^{M}{(1-g_{j})(1-d_{j})}}{\sum_{j=1}^{M}{(1-g_{j})}}\nonumber\\
+\text{FDR} &= \frac{\sum_{j=1}^{M}{(1-g_{j})d_{j}}}{\sum_{j=1}^{M}{d_{j}}}\nonumber
+\end{align}
+
+We present additional results using log2 FC 1, 1.5, 2.5, and 3 in Figure~\ref{fig:LFC}.
+Overall the presented model outperformed all current methods regardless of the choice of FC cutoff, as we maintained the highest sensitivity for a limited observed false discovery rate.
+RSEG showed significant higher observed false discovery rate in all the scenarios.
+
+\begin{figure}
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_PR_LFC/Figure_PR_LFC.pdf}}
+\end{center}
+\caption{ROC curves for H3K36me3 differential peaks covering differentially transcribed gene bodies defined if having absolute $LFC>1$ (A), $LFC>1.5$ (B), $LFC>2.5$ (C), or $LFC>3$ (D).}
+\label{fig:LFC}
+\end{figure}
+
+\section{Study on Window Sizes}
+\label{s:window}
+
+We compared methods regarding the metrics used in this paper to analyze broad marks using window sizes of 250bp, 500bp , 750bp, and 1000bp.
+Overall, results were consistent across different window sizes.
+The HMM-based methods RSEG and THOR appeared to be highly sensitive to the window size and showed better results for 1000bp.
+However, the presented model appeared to be robust to all scenarios as it consistently called a small number of broad peaks overall (Figures \ref{fig:encode.broad.250}, \ref{fig:encode.broad.750}, and \ref{fig:encode.broad.1000}, panels C).
+The methods csaw, ChIPComp, diffReps, and DiffBind exhibited fragmented and narrow differential peaks regardless of the window size.
+The performance of all methods appeared to be consistent to the choice of window size under sharp data set.
+The presented model was robust in all the scenarios as it was able to detect large differences between read counts of the analyzed cell lines.
+
+\begin{figure}
+\begin{center}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Broad_250bp/Figure_Broad.pdf}}
+\end{center}
+\caption{Results from broad marks (250bp).
+(A): ROC curves of differentially transcribed gene coverage based on H3K36me3 differential peak calls.
+(C): average number (y-axis) and size (x-axis) of H3K27me3 called peaks for various methods and different nominal FDR thresholds.
+(B), (D), and (F): example of peak calls from H3K36me3, H3K27me3, and EZH2, respectively.
+(E): computing time of genome-wide analysis of broad marks from various methods.}
+\label{fig:encode.broad.250}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Broad_750bp/Figure_Broad.pdf}}
+\end{center}
+\caption{Results from broad marks (750bp).
+(A): ROC curves of differentially transcribed gene coverage based on H3K36me3 differential peak calls.
+(C): average number (y-axis) and size (x-axis) of H3K27me3 called peaks for various methods and different nominal FDR thresholds.
+(B), (D), and (F): example of peak calls from H3K36me3, H3K27me3, and EZH2, respectively.
+(E): computing time of genome-wide analysis of broad marks from various methods.}
+\label{fig:encode.broad.750}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Broad_1000bp/Figure_Broad.pdf}}
+\end{center}
+\caption{Results from broad marks (1000bp).
+(A): ROC curves of differentially transcribed gene coverage based on H3K36me3 differential peak calls.
+(C): average number (y-axis) and size (x-axis) of H3K27me3 called peaks for various methods and different nominal FDR thresholds.
+(B), (D), and (F): example of peak calls from H3K36me3, H3K27me3, and EZH2, respectively.
+(E): computing time of genome-wide analysis of broad marks from various methods.}
+\label{fig:encode.broad.1000}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Narrow_500bp/Figure_Narrow.pdf}}
+\caption{Results from sharp marks (500bp).
+(A) and (C): median $\log_{2}FC$ and correlation between cell lines of ChIP-seq counts from differential called peaks (FDR 0.05, sorted by the absolute LFC) for CTCF and H3K4me3, respectively.
+(B), (D), and (F): example of broad peak calls from CTCF, H3K4me3, and H3K27ac, respectively.
+(E): computing time of genome-wide analysis of sharp marks from various methods.}
+\label{fig:encode.sharp.500}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Narrow_750bp/Figure_Narrow.pdf}}
+\caption{Results from sharp marks (750bp).
+(A) and (C): median $\log_{2}FC$ and correlation between cell lines of ChIP-seq counts from differential called peaks (FDR 0.05, sorted by the absolute LFC) for CTCF and H3K4me3, respectively.
+(B), (D), and (F): example of broad peak calls from CTCF, H3K4me3, and H3K27ac, respectively.
+(E): computing time of genome-wide analysis of sharp marks from various methods.}
+\label{fig:encode.sharp.750}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Narrow_1000bp/Figure_Narrow.pdf}}
+\caption{Results from sharp marks (1000bp).
+(A) and (C): median $\log_{2}FC$ and correlation between cell lines of ChIP-seq counts from differential called peaks (FDR 0.05, sorted by the absolute LFC) for CTCF and H3K4me3, respectively.
+(B), (D), and (F): example of broad peak calls from CTCF, H3K4me3, and H3K27ac, respectively.
+(E): computing time of genome-wide analysis of sharp marks from various methods.}
+\label{fig:encode.sharp.1000}
+\end{figure}
+
+\section{Classification of Combinatorial Patterns of Protein-DNA Binding Activity and Their Associations With Gene Expression}
+\label{s:patterns}
+
+Lastly, we jointly analyzed data from cell lines Helas3, Hepg2, and Huvec to assess the performance and biological relevance of the classification of differential combinatorial patterns of enrichment from H3K4me3 and H3K27ac.
+These epigenomic marks have been associated with gene transcription and are often deposited on promoter regions of actively transcribed gene bodies.
+First, we evaluated the classification performance of differential combinatorial patterns of enrichment across cell lines by making use of the estimated mixture model posterior probabilities.
+Given the differential HMM state, these posterior probabilities allow us assess whether read counts from genomic windows are likely to be generated from a particular combinatorial pattern of enrichment.
+Secondly, we compared the biological role of the classified patterns with gene expression data.
+To this end, RNA-seq and ChIP-seq read counts were mapped onto gene bodies and gene promoters, respectively.
+For the former, we used counts from abundance estimates scaled using the average transcript length over samples \cite{soneson2015differential}.
+Using DESeq2 \cite{love2014moderated}, normalization between samples was performed to avoid spurious differences due to sequencing depth in both RNA-seq and ChIP-seq data.
+Then, pairwise LFC of read counts between cell lines were calculated.
+Genes with zero counts or with low total read count across cell lines were excluded from the analysis (25\% of the protein-coding genes).
+Finally, we used the maximum window-based posterior probability of the presented mixture model to classify the combinatorial patterns of enrichment from the resulting genes whose promoters overlapped differential peaks.
+
+Figure~\ref{fig:fig_tern} shows the main results from our analyses under a nominal FDR control of 0.05.
+At the top, we present ternary density plots of differential \textit{peaks} and their associated ChIP-seq read counts.
+The colors indicate the classified differential pattern of enrichment of peaks based on window-based posterior probabilities of the mixture model.
+We observed that the classification pattern agreed with the observed direction of enrichment from ChIP-seq data (vertices and opposite sides of the triangle vertices).
+At the bottom, we show the scatter plots of LFCs between cell lines of \textit{gene} bodies (RNA-seq counts) and \textit{gene} promoters (ChIP-seq counts), along with their Pearson correlation, for some of the classified combinatorial patterns of enrichment.
+As expected from the activating roles of H3K27ac and H3K4me3, we observed a high agreement between ChIP-seq and RNA-seq LFCs.
+Here, the colors indicate the classified differential pattern of activation of genes overlapping differential called peaks.
+These results show that the classified pattern of epigenomic change across the three cell lines agreed with the observed direction of gene expression and highlight the novelties of the presented model.
+We were able to accurately classify the differential pattern of enrichment of both called peaks (ternary plots) and genes overlapping called peaks (scatter plots) under more than two conditions.
+The embedded mixture model and its posterior probabilities generalizes the problem of differential peak calling for multiple conditions and provide a tool that is not yet available in any other method presented in the literature.
+
+\begin{figure}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Ternary/Figure_Ternary.pdf}}
+\caption{Ternary plots (top) of ChIP-seq counts mapped on differential peaks and scatter plots (bottom) of $\log_{2}FC$s (LFC) between cell lines of ChIP-seq (x-axis) and RNA-seq counts (y-axis) mapped on genes intersecting differential peaks.
+Colors indicate the classified combinatorial pattern of peak enrichment (ternary plots) and gene activation (scatter plots).
+Correlations indicate high agreement between ChIP-seq and RNA-seq $\log_{2}FC$s for these activating marks.
+\label{fig:fig_tern}}
+\end{figure}
+
+\section{Genomic Segmentation Analysis of H3K36me3 and CTCF}
+\label{s:segment}
+\begin{figure}
+\centerline{\includegraphics{../../Supp/Figure_MultiMarks/Figure_MultiMarks.pdf}}
+\caption{Genomic segmentation analysis of H3K36me3 and CTCF in Helas3 cell line.
+The chosen model parametrization and the normalization for non-linear biases via model offsets allow the segmentation of highly diverse epigenomic marks.
+The implemented hidden Markov model is able to properly account for the differences in length of enrichment regions between CTCF (short) and H3K36me3 (broad).
+\label{fig:fig_tern}}
+\end{figure}
+
 \bibliographystyle{amsplain}

### 2021-02-19 Latex/appendix/webappendix.tex
Adding supplementary files and updating tex

@@
 
+\subsection{GC-content bias in broad marks}
+
 \section{Bayesian Information Criterion (BIC) for Hidden Markov Models}
@@
 
+\subsection{Simulation Study}
+
+\section{Study on FDR Control and the Viterbi Algorithm}
+\label{s:viterbi}
+
+In this section, we present results comparing the FDR controlling approach with the Viterbi algorithm.
+We evaluated the FDR approach using cutoffs 0.01, 0.05, 0.10, 0.15, and 0.20.
+We compared the results between the two approaches using window sizes of 250bp, 500bp, 750bp, and 1000bp.
+Overall, we observed that the Viterbi sequence of states led to similar results than the sequences based on FDR control cutoffs across all choices of window size.
+Specifically, we observed that the sensitivity and specificity of the sequence of Viterbi states were close to those from FDR control, in particular for FDR control 0.10.
+These results are shown in Figure \ref{fig:encode.viterbi}.
+These facts are also reflected by the length and number of called peaks.
+In Figures \ref{fig:encode.viterbi.h3k36me3} and \ref{fig:encode.viterbi.h3k27me3} we show examples of peak calls of H3K26me3 and H3K27me3, respectively, from all FDR control cutoffs and the Viterbi sequence of states.
+Overall, we observe minor differences regarding the size of peak calls of the Viterbi and FDR control sequences across different choices of window sizes.
+These differences were mainly present in the data for H3K27me3, which is known to be a histone mark that expands through broader domains than H3K36me3.
+Finally, it is worth noting that the Viterbi algorithm gives us a way to call peaks that does not depend on the choice of the FDR cutoff.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_PR_Viterbi/Figure_PR_Viterbi.pdf}}
+\end{center}
+\caption{FDR-based results from broad marks and Viterbi-based result from epigraHMM with 250bp (A), 500bp (B), 750bp (C), and 1000bp (D).}
+\label{fig:encode.viterbi}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_H3K36me3_Viterbi/Figure_H3K36me3_Viterbi.pdf}}
+\end{center}
+\caption{FDR- and Viterbi-based peak calls from H3K36me3 with 250bp (A), 500bp, (B), 750bp (C), and 1000bp (D).}
+\label{fig:encode.viterbi.h3k36me3}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_H3K27me3_Viterbi/Figure_H3K27me3_Viterbi.pdf}}
+\end{center}
+\caption{FDR- and Viterbi-based peak calls from H3K27me3 with 250bp (A), 500bp, (B), 750bp (C), and 1000bp (D).}
+\label{fig:encode.viterbi.h3k27me3}
+\end{figure}
+
+\section{HMM Emission Distributions}
+
+For the consensus background ($r=1$) and consensus enrichment ($r=3$) states, the emission distribution function is
+\begin{align}\label{eq:nondiffdensity}
+f_{r}(\mb{y}_{..j}|\bs{\psi}_{r}) &= \prod^{G}_{h=1}\prod^{n_{h}}_{i=1}f_{r}(y_{hij}|\bs{\psi}_{r}),\quad r\in\{1,3\}\quad\text{and}\quad y_{hij}\in\{0,1,2,\ldots\},\nonumber\\
+&=\prod^{G}_{h=1}\prod^{n_{h}}_{i=1}\text{Pr}(Y_{hij}=y_{hij}|Z_{j}=r;\bs{\psi}_{r}),\nonumber\\
+&=\prod^{G}_{h=1}\prod^{n_{h}}_{i=1}\frac{\Gamma(y_{hij}+\phi_{r})}{y_{hij}!\Gamma(\phi_{r})}\left(\frac{\phi_{r}}{\mu_{(r,hij)}+\phi_{r}}\right)^{\phi_{r}}\left(\frac{\mu_{(r,hij)}}{\mu_{(r,hij)}+\phi_{r}}\right)^{y_{hij}}.
+\end{align}
+
+For the differential state ($r=2$), the emission distribution is
+\begin{align}\label{eq:diffdensity}
+f_{2}(\mb{y}_{..j}|\mb{x};\bs{\delta},\bs{\psi}_{2}) &= \sum_{l=1}^{L}\delta_{l}f_{(2,l)}\left(\mb{y}_{..j}|\mb{x}_{l};\bs{\psi}_{(2,l)}\right),\quad y_{hij}\in\{0,1,2,\ldots\},\\
+&= \sum_{l=1}^{L}\delta_{l}\prod_{h=1}^{G}\prod_{i=1}^{n_{h}}Pr\left(Y_{hij}=y_{hij}|Z_{j}=2,\mb{x}_{l};\bs{\psi}_{(2,l)}\right),\nonumber\\
+&= \sum_{l=1}^{L}\delta_{l}\prod_{h=1}^{G}\prod_{i=1}^{n_{h}}\frac{\Gamma\left(y_{hij}+\phi_{(2,l,h)}\right)}{y_{hij}!\Gamma\left(\phi_{(2,l,h)}\right)}\left(\frac{\phi_{(2,l,h)}}{\mu_{(2,l,hij)}+\phi_{(2,l,h)}}\right)^{\phi_{(2,l,h)}}\times\nonumber\\
+&\times\left(\frac{\mu_{(2,l,hij)}}{\mu_{(2,l,hij)}+\phi_{(2,l,h)}}\right)^{y_{hij}}.\nonumber
+\end{align}
+
+Apart from the offset $u_{hij}$, we will assume that replicates from the same (different) condition share common (distinct) mean and dispersion parameters under every mixing probability distribution $f_{(2,l)}$.
+To define all possible combinations of background and enrichment across $G$ conditions, we consider the following sets of singletons $A_{1}$, pairs $A_{2}$, $\ldots$, and $(G-1)$-tuples $A_{G-1}$ such that
+\begin{align*}
+A_{1} &= \left\{a^{(1)}\left. \right\vert a^{(1)}\in\mathbb{G}_{+}\text{ and }a^{(1)}\leq G\right\},\\
+A_{2} &= \left\{(a^{(2)}_{1},a^{(2)}_{2})\left. \right\vert (a^{(2)}_{1},a^{(2)}_{2})\in\mathbb{G}^{2}_{+}\text{ and }a^{(2)}_{1}<a^{(2)}_{2}\leq G\right\},\\
+&\vdots\\
+A_{G-1} &= \left\{(a^{(G-1)}_{g})^{G-1}_{g=1}\left. \right\vert (a^{(G-1)}_{g})^{G-1}_{g=1}\in\mathbb{G}^{G-1}_{+}\text{ and }a^{(G-1)}_{1}<\ldots<a^{(G-1)}_{G-1}\leq G\right\}.
+\end{align*}
+The union of all sets $A = \cup^{G-1}_{k=1}A_{k}$ contains an exhaustive list of $L = 2^{G}-2$ elements that determines the differential pattern across $G$ conditions such that each element of $A$ indicates which of the $G$ conditions are enriched.
+For instance, if $G=3$, $A_{1} = \{1,2,3\}$ and $A_{2} = \{(1,2),(1,3),(2,3)\}$ define the six possible combinations of enrichment and background across three conditions.
+Then, we define a bijective mapping $A \rightarrow S_{1},\ldots,S_{L}$ and let $x_{hl} = I(h\in S_{l})$ indicate whether the read count of genomic window $j$ from replicate $i$ of condition $h$ is enriched in the mixture component $l$.
+We model the log-mean $\mu_{(2,l,hij)}$ and log-dispersion $\phi_{(2,l,h)}$ of mixture $l$ from the emission distribution (\ref{eq:diffdensity}) as
+\begin{align*}
+\log(\mu_{(2,l,hij)}) &= \beta_{1} + \beta_{3}x_{hl} + u_{hij},\quad \text{and}\\
+\log(\phi_{(2,l,h)}) &= \lambda_{1} + \lambda_{3}x_{hl}.
+\end{align*}
+According to this parametrization, $\beta_{1}$ and $\lambda_{1}$ are the baseline log-mean and log-dispersion parameters of the read count distribution from replicates of conditions that are not enriched under the mixing distribution $l$.
+Conversely, $\beta_{1}+\beta_{3}$ and $\lambda_{1}+\lambda_{3}$ are the baseline log-mean and log-dispersion parameters of the read count distribution from replicates of conditions enriched under the mixing distribution $l$.
+This choice of parametrization ensures that windows exhibiting differential enrichment across conditions share means and dispersions that are common between the remaining non differential HMM states.
+
+A pseudo code of the presented EM algorithm is below.
+\begin{enumerate}
+\item{Initialize $\bs{\pi}^{(0)}, \bs{\gamma}^{(0)}, \bs{\delta}^{(0)}, \beta_{1}^{(0)}, \beta_{3}^{(0)}, \lambda_{1}^{(0)}, \lambda_{3}^{(0)}$, such that $\sum_{r=1}^3\pi_{r}^{(0)}=1$ and $\sum_{s=1}^{3}\gamma_{rs}=1$.}
+\item{E step ($t\geq1$),}
+\begin{enumerate}
+\item{Calculate $Pr\left(Z_{j}=r|\mb{y},\mb{x};\bs{\Psi}^{(t-1)}\right)$ and $Pr\left(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t-1)}\right)$ for all $r$ and $s$ in $\left\{1,2,3\right\}$ and $j=1,\ldots,M$ via Forward-Backward algorithm as detailed in Appendix B of the main article}
+\item{Calculate $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t-1)})$ for all $l$ $\left\{1,\ldots,L\right\}$ and $j=1,\ldots,M$ as $f_{(2,l)}(\mb{y}_{..j}|\mb{x}_{l};\bs{\psi}_{(2,l)}^{(t-1)})\delta_{l}^{(t-1)}/\sum_{k=1}^{L}f_{(2,k)}(\mb{y}_{..j}|\mb{x}_{k};\bs{\psi}_{(2,k)}^{(t-1)})\delta_{k}^{(t-1)}$}
+\end{enumerate}
+\item{M step ($t\geq1$),}
+\begin{enumerate}
+\item Maximize Equation (5) (main text) with respect to the initial and transition probabilities to obtain for all $r$ and $s$ in $\left\{1,2,3\right\}$
+\begin{align*}
+\pi_{r}^{(t)} &= Pr\left(Z_{1}=r|\mb{y},\mb{x};\bs{\Psi}^{(t-1)}\right)\\
+\gamma_{rs}^{(t)} &= \sum_{j=2}^{M}Pr(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t-1)})/\sum_{j=2}^{M}Pr(Z_{j-1}=r|\mb{y},\mb{x};\bs{\Psi}^{(t-1)})
+\end{align*}
+\item Maximize Equation (5) with respect to $\bs{\delta}$ to obtain $\bs{\delta}^{(t)}$ such that $\sum_{l=1}^L\delta_{l}^{(t)}=1$.
+\item Conditionally upon $\bs{\delta}^{(t)}$, maximize Equation (5) with respect to $\beta_{1}, \beta_{3}, \lambda_{1}, \lambda_{3}$ to obtain $\beta_{1}^{(t)}, \beta_{3}^{(t)}, \lambda_{1}^{(t)}, \lambda_{3}^{(t)}$,
+\item Iterate between (b) and (c) until convergence.
+\end{enumerate}
+\item{Iterate between 2. and 3. until convergence.}
+\end{enumerate}
+
+\subsection{$Q$-function of the EM algorithm}
+% \label{appendix:appendix_Q}
+The $Q$-function of the EM algorithm is defined as $Q\left(\bs{\Psi}|\bs{\Psi}^{(t)}\right) = Q_{0}\left(\bs{\pi},\bs{\gamma}|\bs{\Psi}^{(t)}\right)+Q_{1}\left(\bs{\psi}_{1}|\bs{\Psi}^{(t)}\right)+Q_{2}\left(\bs{\delta},\bs{\psi}_{2}|\bs{\Psi}^{(t)}\right)+Q_{3}\left(\bs{\psi}_{3}|\bs{\Psi}^{(t)}\right)$, such that
+
+\begin{align}
+\label{eq:appendix_Q}
+Q_{0}\left(\bs{\pi},\bs{\gamma}|\bs{\Psi}^{(t)}\right) &= \sum_{r=1}^{3}\left\{Pr\left(Z_{1}=r|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)\log(\pi_{r})\right\}+\nonumber\\
+&+\sum_{j=2}^{M}\sum_{r=1}^{3}\sum_{s=1}^{3}\left\{Pr\left(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)\log(\gamma_{rs})\right\},\nonumber\\
+Q_{1}\left(\bs{\psi}_{1}|\bs{\Psi}^{(t)}\right) &= \sum_{j=1}^{M}Pr\left(Z_{j}=1|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)\log f_{1}\left(\mb{y}_{..j}|\bs{\psi}_{1}\right),\nonumber\\
+Q_{2}\left(\bs{\delta},\bs{\psi}_{2}|\bs{\Psi}^{(t)}\right) &= \sum_{j=1}^{M}Pr\left(Z_{j}=2|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)\sum_{l=1}^{L}Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})\times\nonumber\\
+&\times \log\delta_{l}f_{(2,l)}\left(\mb{y}_{..j}|\mb{x}_l;\bs{\psi}_{(2,l)}\right),\quad\text{and}\nonumber\\
+Q_{3}\left(\bs{\psi}_{3}|\bs{\Psi}^{(t)}\right) &= \sum_{j=1}^{M}Pr\left(Z_{j}=3|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)\log f_{3}\left(\mb{y}_{..j}|\bs{\psi}_{3}\right).
+\end{align}
+
+During the $t^{th}$ iteration of the EM-algorithm, one solves $\argmax{\beta_{1}^{(t)}, \beta_{3}^{(t)}, \lambda_{1}^{(t)}, \lambda_{3}^{(t)}} Q_{1}\left(\bs{\psi}_{1}|\bs{\Psi}^{(t)}\right) + Q_{2}\left(\bs{\delta},\bs{\psi}_{2}|\bs{\Psi}^{(t)}\right) + Q_{3}\left(\bs{\psi}_{3}|\bs{\Psi}^{(t)}\right)$ during the maximization step.
+This problem is analogous to estimating parameters in a weighted NB regression model, since $Q_{1}$, $Q_{2}$, and $Q_{3}$ can be seen as weighted log-likelihood functions from a NB generalized linear model with means and dispersions parametrized as described in the Section 3.1 of the main article.
+
+The forward probabilities are defined as $f_{11}^{p} = \pi_{1}f_{1}(\mb{y}_{..1}|\bs{\psi}_{1})$, $f_{12}^{p} = \pi_{2}f_{2}(\mb{y}_{..1}|\mb{x};\bs{\delta},\bs{\psi}_{2})$, $f_{13}^{p} = \pi_{3}f_{3}(\mb{y}_{..1}|\bs{\psi}_{3})$, and, for $j=1,\ldots,M$, $f_{j1}^{p} = \sum_{l=1}^{3}\gamma_{l1}f_{(j-1)l}^{p}f_{1}(\mb{y}_{..j}|\bs{\psi}_{1})$, $f_{j2}^{p} = \sum_{l=1}^{3}\gamma_{l2}f_{(j-1)l}^{p}f_{2}(\mb{y}_{..j}|\mb{x};\bs{\delta},\bs{\psi}_{2})$, and $f_{j3}^{p} = \sum_{l=1}^{3}\gamma_{l3}f_{(j-1)l}^{p}f_{3}(\mb{y}_{..j}|\bs{\psi}_{3})$.
+Conversely, the backward probabilities are defined as $b_{Mk}^{p}=1$, $\forall k=1,2,3$, and, for $j=1,\ldots,M$, $b_{j1}^{p} = \sum_{l=1}^{3}\gamma_{1l}b_{(j+1)l}^{p}f_{1}(\mb{y}_{..(j+1)}|\bs{\psi}_{1})$, $b_{j2}^{p} = \sum_{l=1}^{3}\gamma_{2l}b_{(j+1)l}^{p}f_{2}(\mb{y}_{..(j+1)}|\mb{x};\bs{\delta},\bs{\psi}_{2})$, and $b_{j3}^{p} = \sum_{l=1}^{3}\gamma_{3l}b_{(j+1)l}^{p}f_{3}(\mb{y}_{..(j+1)}|\bs{\psi}_{3})$.
+Then, we have
+
+\begin{align}
+\label{eq:appendix_FBalgorithm}
+P\left(Z_{j}=k|\mathbf{y},\mathbf{x};\bs{\Psi}\right) &= \frac{f_{jk}^{p}b_{jk}^{p}}{\sum_{l=1}^{3}f_{Ml}^{p}},\qquad \forall j=1,\ldots,M\quad\text{and}\quad k=1,2,3,\nonumber\\
+P\left(Z_{j-1}=l,Z_{j}=1|\mathbf{y},\mathbf{x};\bs{\Psi}\right) &= \frac{f_{(j-1)l}^{p}\gamma_{l1}f_{1}(\mb{y}_{..j}|\bs{\psi}_{1})b_{j1}^{p}}{\sum_{l=1}^{3}f_{Ml}^{p}},\nonumber\\
+P\left(Z_{j-1}=l,Z_{j}=2|\mathbf{y},\mathbf{x};\bs{\Psi}\right) &= \frac{f_{(j-1)l}^{p}\gamma_{l2}f_{2}(\mb{y}_{..j}|\mb{x};\bs{\delta},\bs{\psi}_{2})b_{j2}^{p}}{\sum_{l=1}^{3}f_{Ml}^{p}},\nonumber\\
+P\left(Z_{j-1}=l,Z_{j}=3|\mathbf{y},\mathbf{x};\bs{\Psi}\right) &= \frac{f_{(j-1)l}^{p}\gamma_{l3}f_{3}(\mb{y}_{..j}|\bs{\psi}_{3})b_{j3}^{p}}{\sum_{l=1}^{3}f_{Ml}^{p}},\forall j=2,\ldots,M\text{ and }l=1,2,3.
+\end{align}
+
+\chapter{}
+In this appendix, we present an overview of the simulation scenarios and additional results.
+
+\section{Read Count Simulation}
+\label{s:sim1}
+
+In this section, we present additional results from the read count based simulation.
+
+For each scenario, we simulated a hundred data sets as follows.
+First, we generated a sequence of hidden states with length $M$ from a first-order Markov chain with $2^G$ states representing every combination of background and enrichment across $G$ conditions.
+Here, we let states 1 and $2^G$ represent consensus background and consensus enrichment, respectively.
+Transition probabilities were chosen such that long sequences of enrichment regions (average length of 50 genomic windows) and equal proportions of differential states were generated, while the majority of the simulated states pertained to consensus background.
+Secondly, read counts were simulated for $n$ replicates from $G$ conditions following NB distributions with mean and dispersion parameters that led to a given SNR and respected the differential pattern of enrichment determined by the hidden state.
+For each scenario, this setup led to a simulated matrix of counts of dimension $M\times(n\times G)$.
+The aim of this study was to assess whether the presented model was able to aggregate all $2^G-2$ simulated differential hidden states into its differential HMM state while maintaining a precise parameter estimation scheme.
+The model parameters used in this simulated study were estimated from ENCODE data pertaining to H3K36me3 and H3K27me3 histone modification marks and a shown in the tables presented in this section.
+
+Overall, while results were similar regarding different number of genome sizes and histone modification marks, the number of replicates played a major role in the performance of the model.
+In scenarios with either low SNR or high number of conditions, more replicates were needed to achieve better model performance.
+However, under the observed SNR, two replicates were sufficient to achieve a small relative bias regarding the estimation of most of the model parameters throughout all the evaluated scenarios.
+This is in agreement with the ENCODE Consortium guidelines, which recommends two technical or biological replicates to achieve best performance in ChIP-seq experiments.
+
+\begin{center}
+\input{../../Supp/Table_Simulation_ReadCounts/Table_Simulation_ReadCounts_H3K27me3_1e+06.tex}
+\end{center}
+
+\begin{center}
+\input{../../Supp/Table_Simulation_ReadCounts/Table_Simulation_ReadCounts_H3K36me3_1e+06.tex}
+\end{center}
+
+Next, we assessed the sensitivity of our method to detect simulated differential regions of enrichment.
+First, differential regions were defined from the  HMM posterior probabilities pertaining to the differential state by controlling the total FDR as defined in Section 3.2 (main text).
+For different nominal FDR threshold levels, the model sensitivity was estimated as the proportion of windows correctly assigned as differential out of the total number of simulated differential windows.
+Additionally, the observed FDR was calculated as the proportion of genomic windows incorrectly called as differential out of the total number of called differential windows.
+In Figures~\ref{fig:sim1.fig1}-\ref{fig:sim1.fig3}, panel A shows the average observed true positive rate (y-axis) and the observed FDR (x-axis) for different nominal FDR levels across a hundred simulated data.
+Results are shown for different levels of SNR, number of conditions, and number of replicates per condition.
+Overall, we observed that the number of replicates per condition played a major role on the sensitivity levels of the model, in which scenarios with two and four replicates had the best results regardless of the number of conditions and SNR levels.
+For scenarios with either high number of conditions or low SNR levels, more replicates were needed to achieve higher sensitivity.
+
+Finally, we used the estimated mixture model posterior probabilities $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\hat{\bs{\Psi}})$, $j=1,\ldots,M$, to classify the differential combinatorial patterns of enrichment of detected differential windows.
+To this end, we first calculated the maximum estimated mixture model posterior probability across all $L$ components to determine the most likely differential combinatorial pattern from genomic windows assigned to be part of the differential state.
+Then, we compared the window-based classification with the true window-based simulated states from the Markov Chain (states $2,\ldots,G-1$).
+In Figures~\ref{fig:sim1.fig1}-\ref{fig:sim1.fig3}, panel B shows the confusion matrices of classified (x-axis) and simulated (y-axis) differential windows for a scenario with three conditions.
+Differential combinatorial patterns of enrichment are represented by the sequences of letters 'E' (enrichment) and 'B' (background), such that each letter corresponds to the status of a given condition.
+The number of windows (averaged over a all simulated data sets) is shown as entries of the matrices and represented by the color scale.
+Darker colors on the diagonal entries indicate better agreement between simulated and classified patterns.
+By utilizing the posterior probabilities from the mixture model, we observed a good performance when classifying the differential combinatorial pattern of enrichment from differential windows.
+Results were best under scenarios with higher number of replicates or SNR.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.7]{../../Supp/Figure_ROC_ReadCounts/Figure_ROC_ReadCounts_H3K27me3_1e+06.pdf}}
+\end{center}
+\caption{Read count simulation.
+(A): average observed TPR and FDR for different nominal FDR threshold levels for various scenarios in simulated data of H3K27me3 with $10^6$ genomic windows.
+(B): confusion matrices are shown for the 3 conditions scenario.
+On x- and y-axes the labels indicate the differential classified and simulated patterns, respectively.
+For instance, EBE denotes enrichment in conditions 1 and 3 only.
+Darker colors on the diagonal of the matrix indicate better agreement.}
+\label{fig:sim1.fig1}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.7]{../../Supp/Figure_ROC_ReadCounts/Figure_ROC_ReadCounts_H3K36me3_1e+06.pdf}}
+\end{center}
+\caption{Read count simulation.
+(A): average observed TPR and FDR for different nominal FDR threshold levels for various scenarios in simulated data of H3K36me3 with $10^5$ genomic windows.
+(B): confusion matrices are shown for the 3 conditions scenario.
+On x- and y-axes the labels indicate the differential classified and simulated patterns, respectively.
+For instance, EBE denotes enrichment in conditions 1 and 3 only.
+Darker colors on the diagonal of the matrix indicate better agreement.}
+\label{fig:sim1.fig2}
+\end{figure}
+
+\section{Sequencing Read Simulation}
+\label{s:sim2}
+In this section, we present additional results from the sequencing read simulation using different FDR control threshold for differential peak call.
+
+We performed a second simulation study aiming to compare the proposed model with the current DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, since none of them was designed to accept matrices of window-based read counts as input.
+To this purpose, we revisited the simulation study presented by \cite{lun2015csaw} where data were generated in a more general scheme without a particular read count model assumption.
+Here, sequencing reads from ChIP-seq experiments were generated for two conditions and two replicates per condition.
+For the differential peaks callers ChIPComp and DiffBind that require sets of candidate regions, we followed the analyses presented by \cite{lun2015csaw} and called peaks in advance using HOMER.
+Those were then used as input in the respective software for differential call.
+For ChIPComp and DiffBind, we attempted to call peaks in advance using MACS2.
+However, MACS2.
+exhibited issues to work on simulated data as it was not able to detect peaks.
+Here, we adapted their simulation pipeline aiming to generate broad and diffuse differential regions of enrichment.
+Specifically, a total of 4000 binding sites spaced 60-65 kbp apart were simulated such that each site was formed by three neighboring peaks 10kb wide.
+For each peak, the number of sequencing reads were simulated from a NB distribution with mean 300 and dispersions sampled from an inverse chi-squared distribution with 20 degrees of freedom.
+Reads were positioned on peak sites to form smooth binding profiles.
+To introduce differential enrichment, one or two peaks from 1000 binding sites were randomly chosen and reads pertaining to them were removed from the sequencing library in one of the two conditions.
+This process was repeated for the second condition, which led to a total of 2000 differential peaks.
+Background reads were added to the library to maintain the realistic aspects of the experiment.
+A hundred simulated data sets were generated and peaks were called by all the methods under multiple nominal FDR thresholds.
+For our method and RSEG, window-based posterior probabilities were used to control the total FDR.
+For illustration purposes, we refer to the method presented in this article as epigraHMM.
+
+Specifically, panel D from Figures \ref{fig:sim2.fig1}, \ref{fig:sim2.fig2}, \ref{fig:sim2.fig3}, and \ref{fig:sim2.fig4} show example of peak calls using FDR control cutoffs 0.01, 0.10, 0.15, and 0.20, respectively.
+Overall, results were consistent across different choices of FDR control.
+This fact is also shown by the observed sensitivity and false discovery rate from all methods (panels A).
+We observed a clear distinction of methods regarding the observed sensitivity and specificity such that RSEG, diffReps, and THOR exhibited the highest observed FDR and ChIPComp and DiffBind exhibited the lowest sensitivity.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.7]{../../Supp/Figure_Overview_SequencingReads/Figure_Simulation_SequencingReads_1.pdf}}
+\end{center}
+\caption{Sequencing read simulation.
+(A): average observed sensitivity and FDR for various methods under different nominal FDR levels.
+(B): scatter plot of average ratio of called and simulated peaks (y-axis) and average number of called peaks intersecting true differential regions (x-axis) for different nominal FDR levels.
+(C): box plot of computing time (in minutes) for various algorithms.
+(D): an example of differential peak calls under a nominal FDR threshold 0.01.
+Shaded areas indicate true differential peaks.}
+\label{fig:sim2.fig1}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.7]{../../Supp/Figure_Overview_SequencingReads/Figure_Simulation_SequencingReads_10.pdf}}
+\end{center}
+\caption{Sequencing read simulation.
+(A): average observed sensitivity and FDR for various methods under different nominal FDR levels.
+(B): scatter plot of average ratio of called and simulated peaks (y-axis) and average number of called peaks intersecting true differential regions (x-axis) for different nominal FDR levels.
+(C): box plot of computing time (in minutes) for various algorithms.
+(D): an example of differential peak calls under a nominal FDR threshold 0.10.
+Shaded areas indicate true differential peaks.}
+\label{fig:sim2.fig2}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.7]{../../Supp/Figure_Overview_SequencingReads/Figure_Simulation_SequencingReads_15.pdf}}
+\end{center}
+\caption{Sequencing read simulation.
+(A): average observed sensitivity and FDR for various methods under different nominal FDR levels.
+(B): scatter plot of average ratio of called and simulated peaks (y-axis) and average number of called peaks intersecting true differential regions (x-axis) for different nominal FDR levels.
+(C): box plot of computing time (in minutes) for various algorithms.
+(D): an example of differential peak calls under a nominal FDR threshold 0.15.
+Shaded areas indicate true differential peaks.}
+\label{fig:sim2.fig3}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.7]{../../Supp/Figure_Overview_SequencingReads/Figure_Simulation_SequencingReads_20.pdf}}
+\end{center}
+\caption{Sequencing read simulation.
+(A): average observed sensitivity and FDR for various methods under different nominal FDR levels.
+(B): scatter plot of average ratio of called and simulated peaks (y-axis) and average number of called peaks intersecting true differential regions (x-axis) for different nominal FDR levels.
+(C): box plot of computing time (in minutes) for various algorithms.
+(D): an example of differential peak calls under a nominal FDR threshold 0.20.
+Shaded areas indicate true differential peaks.}
+\label{fig:sim2.fig4}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_Overview_SequencingReads/Figure_SequencingReads_Example.pdf}}
+\end{center}
+\caption{Sequencing read simulation. Examples of differential called peaks from the various methods under an FDR control of 0.05.}
+\label{fig:sim2.fig5}
+\end{figure}
+
+\chapter{}
+In this appendix, we present an overview of the application on data from the ENCODE Consortium.
+
+\section{RNA-seq data and gene transcription}
+\label{s:rnaseq}
+Under the three cell lines scenario (presented in the main text), we used RNA-seq experimental data from the ENCODE Consortium on Helas3, Hepg2, and Huvec human cells to quantify the expression levels of genes as follows.
+First, we used Salmon \cite{patro2017salmon} to quantify transcript expression from cell-specific RNA-seq experiments.
+We then calculated, using the R package $tximport$ \cite{soneson2015differential}, estimated counts using abundance estimates (transcripts per million, TPM) scaled up to the average transcript length over samples and library size.
+This step ensures that counts computed from Salmon are not correlated with the average transcript length.
+Secondly, we calculated the number of ChIP-seq reads from H3K4me3 and H3K27ac overlapping gene promoters.
+Promoter regions extend around the transcription start site 2000 base pairs upstream and 200 base pair downstream.
+Finally, read counts from RNA-seq and ChIP-seq were normalized for sequencing depth using DESeq2.
+
+For the two cell line scenario, differentially transcribed genes were defined through log2 fold changes of H3K36me3 ChIP-seq read counts.
+We observed several cases in which differentially expressed genes (defined by RNA-seq data) did not exhibit differential enrichment for H3K36me3.
+However, due to the activating roles of H3K36me3 on gene transcription, we follow the ideas presented by \cite{steinhauser2016comprehensive} and \cite{ji2013differential} and defined differentially transcribed genes based on log2 fold changes of H3K36me3 ChIP-seq read counts.
+
+Sensitivity and specificity metrics were calculated on the window-level as follows.
+For non-overlapping genomic windows $b_{1},\ldots,b_{M}$, let $g_{j}=I(b_{j}\in\text{ differentially transcribed gene})$ and $d_{j}=I(b_{j}\in\text{ differential peak})$ denote the indicators of genomic windows being associated with either differential gene bodies or differential peaks, respectively, for $j=1,\ldots,M$, for a given method and nominal FDR level.
+Then, the observed sensitivity (TPR), specificity (1-FPR), and FDR were calculated as follows:
+\begin{align}
+\text{Sensitivity} &= \frac{\sum_{j=1}^{M}{g_{j}d_{j}}}{\sum_{j=1}^{M}{g_{j}}}\nonumber\\
+\text{Specificity} &= \frac{\sum_{j=1}^{M}{(1-g_{j})(1-d_{j})}}{\sum_{j=1}^{M}{(1-g_{j})}}\nonumber\\
+\text{FDR} &= \frac{\sum_{j=1}^{M}{(1-g_{j})d_{j}}}{\sum_{j=1}^{M}{d_{j}}}\nonumber
+\end{align}
+
+We present additional results using log2 FC 1, 1.5, 2.5, and 3 in Figure~\ref{fig:LFC}.
+Overall the presented model outperformed all current methods regardless of the choice of FC cutoff, as we maintained the highest sensitivity for a limited observed false discovery rate.
+RSEG showed significant higher observed false discovery rate in all the scenarios.
+
+\begin{figure}
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_PR_LFC/Figure_PR_LFC.pdf}}
+\end{center}
+\caption{ROC curves for H3K36me3 differential peaks covering differentially transcribed gene bodies defined if having absolute $LFC>1$ (A), $LFC>1.5$ (B), $LFC>2.5$ (C), or $LFC>3$ (D).}
+\label{fig:LFC}
+\end{figure}
+
+\section{Study on Window Sizes}
+\label{s:window}
+
+We compared methods regarding the metrics used in this paper to analyze broad marks using window sizes of 250bp, 500bp , 750bp, and 1000bp.
+Overall, results were consistent across different window sizes.
+The HMM-based methods RSEG and THOR appeared to be highly sensitive to the window size and showed better results for 1000bp.
+However, the presented model appeared to be robust to all scenarios as it consistently called a small number of broad peaks overall (Figures \ref{fig:encode.broad.250}, \ref{fig:encode.broad.750}, and \ref{fig:encode.broad.1000}, panels C).
+The methods csaw, ChIPComp, diffReps, and DiffBind exhibited fragmented and narrow differential peaks regardless of the window size.
+The performance of all methods appeared to be consistent to the choice of window size under sharp data set.
+The presented model was robust in all the scenarios as it was able to detect large differences between read counts of the analyzed cell lines.
+
+\begin{figure}
+\begin{center}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Broad_250bp/Figure_Broad.pdf}}
+\end{center}
+\caption{Results from broad marks (250bp).
+(A): ROC curves of differentially transcribed gene coverage based on H3K36me3 differential peak calls.
+(C): average number (y-axis) and size (x-axis) of H3K27me3 called peaks for various methods and different nominal FDR thresholds.
+(B), (D), and (F): example of peak calls from H3K36me3, H3K27me3, and EZH2, respectively.
+(E): computing time of genome-wide analysis of broad marks from various methods.}
+\label{fig:encode.broad.250}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Broad_750bp/Figure_Broad.pdf}}
+\end{center}
+\caption{Results from broad marks (750bp).
+(A): ROC curves of differentially transcribed gene coverage based on H3K36me3 differential peak calls.
+(C): average number (y-axis) and size (x-axis) of H3K27me3 called peaks for various methods and different nominal FDR thresholds.
+(B), (D), and (F): example of peak calls from H3K36me3, H3K27me3, and EZH2, respectively.
+(E): computing time of genome-wide analysis of broad marks from various methods.}
+\label{fig:encode.broad.750}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Broad_1000bp/Figure_Broad.pdf}}
+\end{center}
+\caption{Results from broad marks (1000bp).
+(A): ROC curves of differentially transcribed gene coverage based on H3K36me3 differential peak calls.
+(C): average number (y-axis) and size (x-axis) of H3K27me3 called peaks for various methods and different nominal FDR thresholds.
+(B), (D), and (F): example of peak calls from H3K36me3, H3K27me3, and EZH2, respectively.
+(E): computing time of genome-wide analysis of broad marks from various methods.}
+\label{fig:encode.broad.1000}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Narrow_500bp/Figure_Narrow.pdf}}
+\caption{Results from sharp marks (500bp).
+(A) and (C): median $\log_{2}FC$ and correlation between cell lines of ChIP-seq counts from differential called peaks (FDR 0.05, sorted by the absolute LFC) for CTCF and H3K4me3, respectively.
+(B), (D), and (F): example of broad peak calls from CTCF, H3K4me3, and H3K27ac, respectively.
+(E): computing time of genome-wide analysis of sharp marks from various methods.}
+\label{fig:encode.sharp.500}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Narrow_750bp/Figure_Narrow.pdf}}
+\caption{Results from sharp marks (750bp).
+(A) and (C): median $\log_{2}FC$ and correlation between cell lines of ChIP-seq counts from differential called peaks (FDR 0.05, sorted by the absolute LFC) for CTCF and H3K4me3, respectively.
+(B), (D), and (F): example of broad peak calls from CTCF, H3K4me3, and H3K27ac, respectively.
+(E): computing time of genome-wide analysis of sharp marks from various methods.}
+\label{fig:encode.sharp.750}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Narrow_1000bp/Figure_Narrow.pdf}}
+\caption{Results from sharp marks (1000bp).
+(A) and (C): median $\log_{2}FC$ and correlation between cell lines of ChIP-seq counts from differential called peaks (FDR 0.05, sorted by the absolute LFC) for CTCF and H3K4me3, respectively.
+(B), (D), and (F): example of broad peak calls from CTCF, H3K4me3, and H3K27ac, respectively.
+(E): computing time of genome-wide analysis of sharp marks from various methods.}
+\label{fig:encode.sharp.1000}
+\end{figure}
+
+\section{Classification of Combinatorial Patterns of Protein-DNA Binding Activity and Their Associations With Gene Expression}
+\label{s:patterns}
+
+Lastly, we jointly analyzed data from cell lines Helas3, Hepg2, and Huvec to assess the performance and biological relevance of the classification of differential combinatorial patterns of enrichment from H3K4me3 and H3K27ac.
+These epigenomic marks have been associated with gene transcription and are often deposited on promoter regions of actively transcribed gene bodies.
+First, we evaluated the classification performance of differential combinatorial patterns of enrichment across cell lines by making use of the estimated mixture model posterior probabilities.
+Given the differential HMM state, these posterior probabilities allow us assess whether read counts from genomic windows are likely to be generated from a particular combinatorial pattern of enrichment.
+Secondly, we compared the biological role of the classified patterns with gene expression data.
+To this end, RNA-seq and ChIP-seq read counts were mapped onto gene bodies and gene promoters, respectively.
+For the former, we used counts from abundance estimates scaled using the average transcript length over samples \cite{soneson2015differential}.
+Using DESeq2 \cite{love2014moderated}, normalization between samples was performed to avoid spurious differences due to sequencing depth in both RNA-seq and ChIP-seq data.
+Then, pairwise LFC of read counts between cell lines were calculated.
+Genes with zero counts or with low total read count across cell lines were excluded from the analysis (25\% of the protein-coding genes).
+Finally, we used the maximum window-based posterior probability of the presented mixture model to classify the combinatorial patterns of enrichment from the resulting genes whose promoters overlapped differential peaks.
+
+Figure~\ref{fig:fig_tern} shows the main results from our analyses under a nominal FDR control of 0.05.
+At the top, we present ternary density plots of differential \textit{peaks} and their associated ChIP-seq read counts.
+The colors indicate the classified differential pattern of enrichment of peaks based on window-based posterior probabilities of the mixture model.
+We observed that the classification pattern agreed with the observed direction of enrichment from ChIP-seq data (vertices and opposite sides of the triangle vertices).
+At the bottom, we show the scatter plots of LFCs between cell lines of \textit{gene} bodies (RNA-seq counts) and \textit{gene} promoters (ChIP-seq counts), along with their Pearson correlation, for some of the classified combinatorial patterns of enrichment.
+As expected from the activating roles of H3K27ac and H3K4me3, we observed a high agreement between ChIP-seq and RNA-seq LFCs.
+Here, the colors indicate the classified differential pattern of activation of genes overlapping differential called peaks.
+These results show that the classified pattern of epigenomic change across the three cell lines agreed with the observed direction of gene expression and highlight the novelties of the presented model.
+We were able to accurately classify the differential pattern of enrichment of both called peaks (ternary plots) and genes overlapping called peaks (scatter plots) under more than two conditions.
+The embedded mixture model and its posterior probabilities generalizes the problem of differential peak calling for multiple conditions and provide a tool that is not yet available in any other method presented in the literature.
+
+\begin{figure}
+\centerline{\includegraphics[scale = 0.65]{../../Supp/Figure_Ternary/Figure_Ternary.pdf}}
+\caption{Ternary plots (top) of ChIP-seq counts mapped on differential peaks and scatter plots (bottom) of $\log_{2}FC$s (LFC) between cell lines of ChIP-seq (x-axis) and RNA-seq counts (y-axis) mapped on genes intersecting differential peaks.
+Colors indicate the classified combinatorial pattern of peak enrichment (ternary plots) and gene activation (scatter plots).
+Correlations indicate high agreement between ChIP-seq and RNA-seq $\log_{2}FC$s for these activating marks.
+\label{fig:fig_tern}}
+\end{figure}
+
+\section{Genomic Segmentation Analysis of H3K36me3 and CTCF}
+\label{s:segment}
+\begin{figure}
+\centerline{\includegraphics{../../Supp/Figure_MultiMarks/Figure_MultiMarks.pdf}}
+\caption{Genomic segmentation analysis of H3K36me3 and CTCF in Helas3 cell line.
+The chosen model parametrization and the normalization for non-linear biases via model offsets allow the segmentation of highly diverse epigenomic marks.
+The implemented hidden Markov model is able to properly account for the differences in length of enrichment regions between CTCF (short) and H3K36me3 (broad).
+\label{fig:fig_tern}}
+\end{figure}
+
 \bibliographystyle{amsplain}

### 2021-02-19 Supp/Figure_Broad_1000bp/Figure_Broad.R

@@ -1,0 +1,22 @@
+library(ggpubr)
+
+load('Figure_PR_H3K36me3_Encode_twocells_1000bp.RData')
+load('Figure_PeakCalls_H3K36me3_Encode_twocells_1000bp.RData')
+load('Figure_PeakCalls_H3K27me3_Encode_twocells_1000bp.RData')
+load('Figure_PeakCalls_EZH2_Encode_twocells_1000bp.RData')
+load('Figure_Memory_Time.RData')
+
+
+# Fig Peak Calls
+fig.calls = ggarrange(fig.h3k36me3,fig.h3k27me3,fig.ezh2,nrow = 3,ncol = 1,legend = 'none',labels = list('B','D','F'))
+
+# Figure Metrics
+fig.metrics = ggarrange(fig.ROC,fig.TPR,fig_memory_time,ncol=1,nrow=3,legend = 'none',labels = list('A','C','E'))
+
+# Put them all together
+
+fig.final = ggarrange(fig.metrics,fig.calls,nrow=1,ncol=2)
+fig.final.label = ggarrange(get_legend(fig.ROC + theme(legend.position = 'top',legend.direction = 'horizontal',legend.box = 'vertical',legend.spacing.y = unit(-0.25, "cm")) + guides(col = guide_legend(nrow = 1,order = 1),shape = guide_legend(nrow = 1,order = 2,title = 'Nominal FDR'))),
+                            fig.final,nrow=2,heights = c(0.05,0.95))
+
+ggsave(filename = 'Figure_Broad.pdf',plot = fig.final.label,height = 11.5,width = 9,dpi = 'retina')

### 2021-02-19 Supp/Figure_Broad_1000bp/Figure_Memory_Time.R

@@ -1,0 +1,167 @@
+# Libraries
+
+library(data.table)
+library(Polychrome)
+library(magrittr)
+library(ggplot2)
+library(microbenchmark)
+
+# Setup
+
+colors = c(kelly.colors(22)[c('lightblue',
+                              'red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'buff',
+                              'orange')])
+methods <-
+  c(
+    "epigraHMM",
+    "ChIPComp +\nMACS2",
+    "csaw",
+    "DiffBind +\nMACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+names(colors) <- methods
+
+################################################################################
+### Memory
+################################################################################
+
+files.csv <-
+  list.files('../../Public',
+             'slurm*.*.csv$',
+             recursive = TRUE,
+             full.names = TRUE)
+
+# Filtering
+
+files.csv <- files.csv[grep('Encode_twocells|MACS2', files.csv)]
+
+# Looping through files
+
+mem.dt <- rbindlist(lapply(strsplit(files.csv, '/'), function(x) {
+  # Does not run for MACS2
+  if (length(grep('MACS2', x)) == 0) {
+    subx <- strsplit(x[7], '_|bp')[[1]]
+    
+    csvx <- fread(paste(x, collapse = '/'))[2, ]$MaxRSS
+    
+    # Do not proceed if job is not completed
+    if (!is.na(csvx)) {
+      # If ChIPComp or DiffBind, need to check MACS2
+      if (x[4] %in% c('ChIPComp', 'DiffBind')) {
+        csvmacs2 <-
+          rbindlist(lapply(files.csv[intersect(grep(paste0('MACS2/', x[5]), files.csv), grep(paste0('Helas3|Hepg2'), files.csv))], function(y) {
+            fread(y)[JobName == 'batch', ]
+          }))$MaxRSS
+        
+        mem <-
+          max(as.numeric(strsplit(csvx, 'K')[[1]]), max(as.numeric(unlist(
+            strsplit(as.character(csvmacs2), 'K')
+          ))))
+      } else{
+        mem <- as.numeric(strsplit(csvx, 'K')[[1]])
+      }
+      
+      return(data.table(
+        Method = x[4],
+        Mark = x[5],
+        Data = x[6],
+        BP = subx[5],
+        Memory = mem * 1e-6
+      )) # From Kb to Gb
+    }
[…]

### 2021-02-19 Supp/Figure_Broad_1000bp/Figure_PR_H3K36me3_Encode_twocells_1000bp.R

@@ -12,7 +12,7 @@ library(Polychrome)
 library(ggrepel)
 
 # General parameters
-bp = 750
+bp = 1000
 sizeanno <- 2.5
 chromosome = paste0('chr', 1:22)
 mark = 'H3K36me3'
@@ -43,59 +43,37 @@ getcov = function(gr.predicted,
                   dt,
                   threshold,
                   method,
-                  postprob = F,
-                  viterbi = F) {
-  if (viterbi) {
-    out = list()
-    dt[, pred :=  overlapsAny(gr.genome, gr.predicted)]
-    out[['Viterbi']] = data.table(
+                  postprob = F) {
+  out = list()
+  for (x in 1:length(threshold)) {
+    if (postprob) {
+      gr.predicted.subset <-
+        GenomicRanges::reduce(gr.predicted[epigraHMM:::fdrControl(gr.predicted$FDR, fdr = threshold[x])], ignore.strand =
+                                T) # Filtering significant windows
+    } else{
+      gr.predicted.subset = GenomicRanges::reduce(gr.predicted[gr.predicted$FDR <=
+                                                                 threshold[x]], ignore.strand = T)
+    }
+    
+    dt[, pred := overlapsAny(gr.genome, gr.predicted.subset)]
+    
+    out[[x]] = data.table(
       Method = method,
-      Threshold = 'Viterbi',
+      Threshold = threshold[x],
+      TP = sum(dt[, DE == T & pred == T]),
+      Median_Size = median(width(gr.predicted.subset)),
+      Mean_Size = mean(width(gr.predicted.subset)),
       TPR = sum(dt[, DE == T &
                      pred == T]) / sum(dt[, DE == T]),
       FPR = sum(dt[, DE == F &
                      pred == T]) / sum(dt[, DE == F]),
       PPV = sum(dt[, DE == T &
-                     pred == T]) / sum(dt[, pred == T]),
-      Width.Mean = mean(width(gr.predicted)),
-      Width.Median = median(width(gr.predicted)),
-      Width.SD = sd(width(gr.predicted))
[…]

### 2021-02-19 Supp/Figure_Broad_1000bp/Figure_PeakCalls_EZH2_Encode_twocells_1000bp.R

@@ -1,0 +1,426 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 1000
+chromosome = 'chr22'
+ngroups = 2
+mark = 'EZH2'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3Ezh239875AlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3Ezh239875AlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2Ezh239875AlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
[…]

### 2021-02-19 Supp/Figure_Broad_1000bp/Figure_PeakCalls_H3K27me3_Encode_twocells_1000bp.R

@@ -1,0 +1,426 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 1000
+chromosome = 'chr20'
+ngroups = 2
+mark = 'H3K27me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2H3k27me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
[…]

### 2021-02-19 Supp/Figure_Broad_1000bp/Figure_PeakCalls_H3K36me3_Encode_twocells_1000bp.R

@@ -1,0 +1,433 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 1000
+chromosome = 'chr19'
+ngroups = 2
+mark = 'H3K36me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2H3k36me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
[…]

### 2021-02-19 Supp/Figure_Broad_250bp/Figure_Broad.R

@@ -1,0 +1,22 @@
+library(ggpubr)
+
+load('Figure_PR_H3K36me3_Encode_twocells_250bp.RData')
+load('Figure_PeakCalls_H3K36me3_Encode_twocells_250bp.RData')
+load('Figure_PeakCalls_H3K27me3_Encode_twocells_250bp.RData')
+load('Figure_PeakCalls_EZH2_Encode_twocells_250bp.RData')
+load('Figure_Memory_Time.RData')
+
+
+# Fig Peak Calls
+fig.calls = ggarrange(fig.h3k36me3,fig.h3k27me3,fig.ezh2,nrow = 3,ncol = 1,legend = 'none',labels = list('B','D','F'))
+
+# Figure Metrics
+fig.metrics = ggarrange(fig.ROC,fig.TPR,fig_memory_time,ncol=1,nrow=3,legend = 'none',labels = list('A','C','E'))
+
+# Put them all together
+
+fig.final = ggarrange(fig.metrics,fig.calls,nrow=1,ncol=2)
+fig.final.label = ggarrange(get_legend(fig.ROC + theme(legend.position = 'top',legend.direction = 'horizontal',legend.box = 'vertical',legend.spacing.y = unit(-0.25, "cm")) + guides(col = guide_legend(nrow = 1,order = 1),shape = guide_legend(nrow = 1,order = 2,title = 'Nominal FDR'))),
+                            fig.final,nrow=2,heights = c(0.05,0.95))
+
+ggsave(filename = 'Figure_Broad.pdf',plot = fig.final.label,height = 11.5,width = 9,dpi = 'retina')

### 2021-02-19 Supp/Figure_Broad_250bp/Figure_Memory_Time.R

@@ -1,0 +1,167 @@
+# Libraries
+
+library(data.table)
+library(Polychrome)
+library(magrittr)
+library(ggplot2)
+library(microbenchmark)
+
+# Setup
+
+colors = c(kelly.colors(22)[c('lightblue',
+                              'red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'buff',
+                              'orange')])
+methods <-
+  c(
+    "epigraHMM",
+    "ChIPComp +\nMACS2",
+    "csaw",
+    "DiffBind +\nMACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+names(colors) <- methods
+
+################################################################################
+### Memory
+################################################################################
+
+files.csv <-
+  list.files('../../Public',
+             'slurm*.*.csv$',
+             recursive = TRUE,
+             full.names = TRUE)
+
+# Filtering
+
+files.csv <- files.csv[grep('Encode_twocells|MACS2', files.csv)]
+
+# Looping through files
+
+mem.dt <- rbindlist(lapply(strsplit(files.csv, '/'), function(x) {
+  # Does not run for MACS2
+  if (length(grep('MACS2', x)) == 0) {
+    subx <- strsplit(x[7], '_|bp')[[1]]
+    
+    csvx <- fread(paste(x, collapse = '/'))[2, ]$MaxRSS
+    
+    # Do not proceed if job is not completed
+    if (!is.na(csvx)) {
+      # If ChIPComp or DiffBind, need to check MACS2
+      if (x[4] %in% c('ChIPComp', 'DiffBind')) {
+        csvmacs2 <-
+          rbindlist(lapply(files.csv[intersect(grep(paste0('MACS2/', x[5]), files.csv), grep(paste0('Helas3|Hepg2'), files.csv))], function(y) {
+            fread(y)[JobName == 'batch', ]
+          }))$MaxRSS
+        
+        mem <-
+          max(as.numeric(strsplit(csvx, 'K')[[1]]), max(as.numeric(unlist(
+            strsplit(as.character(csvmacs2), 'K')
+          ))))
+      } else{
+        mem <- as.numeric(strsplit(csvx, 'K')[[1]])
+      }
+      
+      return(data.table(
+        Method = x[4],
+        Mark = x[5],
+        Data = x[6],
+        BP = subx[5],
+        Memory = mem * 1e-6
+      )) # From Kb to Gb
+    }
[…]

### 2021-02-19 Supp/Figure_Broad_250bp/Figure_PR_H3K36me3_Encode_twocells_250bp.R

@@ -12,7 +12,7 @@ library(Polychrome)
 library(ggrepel)
 
 # General parameters
-bp = 750
+bp = 250
 sizeanno <- 2.5
 chromosome = paste0('chr', 1:22)
 mark = 'H3K36me3'
@@ -43,59 +43,37 @@ getcov = function(gr.predicted,
                   dt,
                   threshold,
                   method,
-                  postprob = F,
-                  viterbi = F) {
-  if (viterbi) {
-    out = list()
-    dt[, pred :=  overlapsAny(gr.genome, gr.predicted)]
-    out[['Viterbi']] = data.table(
+                  postprob = F) {
+  out = list()
+  for (x in 1:length(threshold)) {
+    if (postprob) {
+      gr.predicted.subset <-
+        GenomicRanges::reduce(gr.predicted[epigraHMM:::fdrControl(gr.predicted$FDR, fdr = threshold[x])], ignore.strand =
+                                T) # Filtering significant windows
+    } else{
+      gr.predicted.subset = GenomicRanges::reduce(gr.predicted[gr.predicted$FDR <=
+                                                                 threshold[x]], ignore.strand = T)
+    }
+    
+    dt[, pred := overlapsAny(gr.genome, gr.predicted.subset)]
+    
+    out[[x]] = data.table(
       Method = method,
-      Threshold = 'Viterbi',
+      Threshold = threshold[x],
+      TP = sum(dt[, DE == T & pred == T]),
+      Median_Size = median(width(gr.predicted.subset)),
+      Mean_Size = mean(width(gr.predicted.subset)),
       TPR = sum(dt[, DE == T &
                      pred == T]) / sum(dt[, DE == T]),
       FPR = sum(dt[, DE == F &
                      pred == T]) / sum(dt[, DE == F]),
       PPV = sum(dt[, DE == T &
-                     pred == T]) / sum(dt[, pred == T]),
-      Width.Mean = mean(width(gr.predicted)),
-      Width.Median = median(width(gr.predicted)),
-      Width.SD = sd(width(gr.predicted))
[…]

### 2021-02-19 Supp/Figure_Broad_250bp/Figure_PeakCalls_EZH2_Encode_twocells_250bp.R

@@ -1,0 +1,426 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 250
+chromosome = 'chr22'
+ngroups = 2
+mark = 'EZH2'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3Ezh239875AlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3Ezh239875AlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2Ezh239875AlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
[…]

### 2021-02-19 Supp/Figure_Broad_250bp/Figure_PeakCalls_H3K27me3_Encode_twocells_250bp.R

@@ -1,0 +1,426 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 250
+chromosome = 'chr20'
+ngroups = 2
+mark = 'H3K27me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2H3k27me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
[…]

### 2021-02-19 Supp/Figure_Broad_250bp/Figure_PeakCalls_H3K36me3_Encode_twocells_250bp.R

@@ -1,0 +1,433 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 250
+chromosome = 'chr19'
+ngroups = 2
+mark = 'H3K36me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2H3k36me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
[…]

### 2021-02-19 Supp/Figure_Broad_750bp/Figure_Broad.R

@@ -1,0 +1,22 @@
+library(ggpubr)
+
+load('Figure_PR_H3K36me3_Encode_twocells_750bp.RData')
+load('Figure_PeakCalls_H3K36me3_Encode_twocells_750bp.RData')
+load('Figure_PeakCalls_H3K27me3_Encode_twocells_750bp.RData')
+load('Figure_PeakCalls_EZH2_Encode_twocells_750bp.RData')
+load('Figure_Memory_Time.RData')
+
+
+# Fig Peak Calls
+fig.calls = ggarrange(fig.h3k36me3,fig.h3k27me3,fig.ezh2,nrow = 3,ncol = 1,legend = 'none',labels = list('B','D','F'))
+
+# Figure Metrics
+fig.metrics = ggarrange(fig.ROC,fig.TPR,fig_memory_time,ncol=1,nrow=3,legend = 'none',labels = list('A','C','E'))
+
+# Put them all together
+
+fig.final = ggarrange(fig.metrics,fig.calls,nrow=1,ncol=2)
+fig.final.label = ggarrange(get_legend(fig.ROC + theme(legend.position = 'top',legend.direction = 'horizontal',legend.box = 'vertical',legend.spacing.y = unit(-0.25, "cm")) + guides(col = guide_legend(nrow = 1,order = 1),shape = guide_legend(nrow = 1,order = 2,title = 'Nominal FDR'))),
+                            fig.final,nrow=2,heights = c(0.05,0.95))
+
+ggsave(filename = 'Figure_Broad.pdf',plot = fig.final.label,height = 11.5,width = 9,dpi = 'retina')

### 2021-02-19 Supp/Figure_Broad_750bp/Figure_Memory_Time.R

@@ -1,0 +1,167 @@
+# Libraries
+
+library(data.table)
+library(Polychrome)
+library(magrittr)
+library(ggplot2)
+library(microbenchmark)
+
+# Setup
+
+colors = c(kelly.colors(22)[c('lightblue',
+                              'red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'buff',
+                              'orange')])
+methods <-
+  c(
+    "epigraHMM",
+    "ChIPComp +\nMACS2",
+    "csaw",
+    "DiffBind +\nMACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+names(colors) <- methods
+
+################################################################################
+### Memory
+################################################################################
+
+files.csv <-
+  list.files('../../Public',
+             'slurm*.*.csv$',
+             recursive = TRUE,
+             full.names = TRUE)
+
+# Filtering
+
+files.csv <- files.csv[grep('Encode_twocells|MACS2', files.csv)]
+
+# Looping through files
+
+mem.dt <- rbindlist(lapply(strsplit(files.csv, '/'), function(x) {
+  # Does not run for MACS2
+  if (length(grep('MACS2', x)) == 0) {
+    subx <- strsplit(x[7], '_|bp')[[1]]
+    
+    csvx <- fread(paste(x, collapse = '/'))[2, ]$MaxRSS
+    
+    # Do not proceed if job is not completed
+    if (!is.na(csvx)) {
+      # If ChIPComp or DiffBind, need to check MACS2
+      if (x[4] %in% c('ChIPComp', 'DiffBind')) {
+        csvmacs2 <-
+          rbindlist(lapply(files.csv[intersect(grep(paste0('MACS2/', x[5]), files.csv), grep(paste0('Helas3|Hepg2'), files.csv))], function(y) {
+            fread(y)[JobName == 'batch', ]
+          }))$MaxRSS
+        
+        mem <-
+          max(as.numeric(strsplit(csvx, 'K')[[1]]), max(as.numeric(unlist(
+            strsplit(as.character(csvmacs2), 'K')
+          ))))
+      } else{
+        mem <- as.numeric(strsplit(csvx, 'K')[[1]])
+      }
+      
+      return(data.table(
+        Method = x[4],
+        Mark = x[5],
+        Data = x[6],
+        BP = subx[5],
+        Memory = mem * 1e-6
+      )) # From Kb to Gb
+    }
[…]

### 2021-02-19 Supp/Figure_Broad_750bp/Figure_PR_H3K36me3_Encode_twocells_750bp.R

@@ -43,59 +43,37 @@ getcov = function(gr.predicted,
                   dt,
                   threshold,
                   method,
-                  postprob = F,
-                  viterbi = F) {
-  if (viterbi) {
-    out = list()
-    dt[, pred :=  overlapsAny(gr.genome, gr.predicted)]
-    out[['Viterbi']] = data.table(
+                  postprob = F) {
+  out = list()
+  for (x in 1:length(threshold)) {
+    if (postprob) {
+      gr.predicted.subset <-
+        GenomicRanges::reduce(gr.predicted[epigraHMM:::fdrControl(gr.predicted$FDR, fdr = threshold[x])], ignore.strand =
+                                T) # Filtering significant windows
+    } else{
+      gr.predicted.subset = GenomicRanges::reduce(gr.predicted[gr.predicted$FDR <=
+                                                                 threshold[x]], ignore.strand = T)
+    }
+    
+    dt[, pred := overlapsAny(gr.genome, gr.predicted.subset)]
+    
+    out[[x]] = data.table(
       Method = method,
-      Threshold = 'Viterbi',
+      Threshold = threshold[x],
+      TP = sum(dt[, DE == T & pred == T]),
+      Median_Size = median(width(gr.predicted.subset)),
+      Mean_Size = mean(width(gr.predicted.subset)),
       TPR = sum(dt[, DE == T &
                      pred == T]) / sum(dt[, DE == T]),
       FPR = sum(dt[, DE == F &
                      pred == T]) / sum(dt[, DE == F]),
       PPV = sum(dt[, DE == T &
-                     pred == T]) / sum(dt[, pred == T]),
-      Width.Mean = mean(width(gr.predicted)),
-      Width.Median = median(width(gr.predicted)),
-      Width.SD = sd(width(gr.predicted))
+                     pred == T]) / sum(dt[, pred == T])
     )
-    dt[, pred := NULL]
-    out = rbindlist(out)
-    return(out)
-  } else{
-    out = list()
-    for (x in 1:length(threshold)) {
-      if (postprob) {
[…]

### 2021-02-19 Supp/Figure_Broad_750bp/Figure_PeakCalls_EZH2_Encode_twocells_750bp.R

@@ -1,0 +1,426 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 750
+chromosome = 'chr22'
+ngroups = 2
+mark = 'EZH2'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3Ezh239875AlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3Ezh239875AlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2Ezh239875AlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
[…]

### 2021-02-19 Supp/Figure_Broad_750bp/Figure_PeakCalls_H3K27me3_Encode_twocells_750bp.R

@@ -1,0 +1,426 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 750
+chromosome = 'chr20'
+ngroups = 2
+mark = 'H3K27me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2H3k27me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
[…]

### 2021-02-19 Supp/Figure_Broad_750bp/Figure_PeakCalls_H3K36me3_Encode_twocells_750bp.R

@@ -1,0 +1,433 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 750
+chromosome = 'chr19'
+ngroups = 2
+mark = 'H3K36me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2H3k36me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
[…]

### 2021-02-19 Supp/Figure_ChromHMM/Figure_ChromHMM.R

@@ -1,0 +1,251 @@
+# Figure F: genomic segmentation with Viterbi path
+
+# Loading libraries
+library(epigraHMM)
+library(data.table)
+library(magrittr)
+library(plyr)
+library(ggplot2)
+library(ggrepel)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(tidyverse)
+library(ggnewscale)
+
+# Some general parameters and functions
+
+bp <- 500
+chromosome <- 'chr1'#'chr19'
+idx.genome <- c(161626774,161943334)#c(56588768,56892457)
+cellline <- 'Helas3'
+size.text <- 2
+size.axis <- 6
+h.track = c(75,62.5,50)
+h.txt = c(75,62.5,50)
+
+Modes <- function(x) {
+  ux <- unique(x)
+  tab <- tabulate(match(x, ux))
+  if (length(ux[tab == max(tab)]) == 1) {
+    return(ux[tab == max(tab)])
+  } else{
+    out = ux[tab == max(tab)]
+    return(max(out))
+  }
+}
+
+pat.peaks = function(x,gr.peaks,gr.genome){
+  ovp = findOverlaps(gr.peaks,gr.genome,ignore.strand = T)
+  out = rbindlist(lapply(unique(queryHits(ovp)), FUN = function(i){data.table(Peak = i,Mode = Modes(x[subjectHits(ovp)[queryHits(ovp) == i],]$Pattern))}))
+  Pattern = rep(NA,length(gr.peaks))
+  Pattern[out$Peak] = out$Mode
+  return(Pattern)
+}
+
+# Loading data
+
+load(paste0('../../Data/Encode_helas3/H3K36me3/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep1.markdup.q10.sorted.RData'))
+h3k36me3.rep1 = counts[[paste0(bp)]];rm(counts)
+load(paste0('../../Data/Encode_helas3/H3K36me3/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep2.markdup.q10.sorted.RData'))
+h3k36me3.rep2 = counts[[paste0(bp)]];rm(counts)
+
+load(paste0('../../Data/Encode_helas3/H3K27me3/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep1.markdup.q10.sorted.RData'))
+h3k27me3.rep1 = counts[[paste0(bp)]];rm(counts)
[…]

### 2021-02-19 Supp/Figure_ENCODE_Control/Figure_PR_H3K36me3_Encode_twocells_500bp.R

@@ -12,7 +12,7 @@ library(Polychrome)
 library(ggrepel)
 
 # General parameters
-bp = 750
+bp = 500
 sizeanno <- 2.5
 chromosome = paste0('chr', 1:22)
 mark = 'H3K36me3'
@@ -25,6 +25,9 @@ methods <-
     "DiffBind + MACS2",
     "diffReps",
     "epigraHMM",
+    'epigraHMM + Control',
+    'epigraHMM + AR Counts',
+    'epigraHMM + Control & AR Counts',
     'RSEG',
     'THOR',
     'Genes'
@@ -34,6 +37,9 @@ colors = c(kelly.colors(22)[c('red',
                               'purplishpink',
                               'yellowgreen',
                               'lightblue',
+                              'purple',
+                              'purplishpink',
+                              'violet',
                               'buff',
                               'orange')], 'black')
 names(colors) <- methods
@@ -43,59 +49,37 @@ getcov = function(gr.predicted,
                   dt,
                   threshold,
                   method,
-                  postprob = F,
-                  viterbi = F) {
-  if (viterbi) {
-    out = list()
-    dt[, pred :=  overlapsAny(gr.genome, gr.predicted)]
-    out[['Viterbi']] = data.table(
+                  postprob = F) {
+  out = list()
+  for (x in 1:length(threshold)) {
+    if (postprob) {
+      gr.predicted.subset <-
+        GenomicRanges::reduce(gr.predicted[epigraHMM:::fdrControl(gr.predicted$FDR, fdr = threshold[x])], ignore.strand =
+                                T) # Filtering significant windows
+    } else{
+      gr.predicted.subset = GenomicRanges::reduce(gr.predicted[gr.predicted$FDR <=
+                                                                 threshold[x]], ignore.strand = T)
+    }
+    
+    dt[, pred := overlapsAny(gr.genome, gr.predicted.subset)]
+    
+    out[[x]] = data.table(
       Method = method,
[…]

### 2021-02-19 Supp/Figure_H3K27me3_Viterbi/Figure_H3K27me3_Viterbi.R

@@ -1,0 +1,10 @@
+load('Figure_PeakCalls_H3K27me3_Encode_twocells_250bp.RData')
+load('Figure_PeakCalls_H3K27me3_Encode_twocells_500bp.RData')
+load('Figure_PeakCalls_H3K27me3_Encode_twocells_750bp.RData')
+load('Figure_PeakCalls_H3K27me3_Encode_twocells_1000bp.RData')
+
+fig.calls = ggarrange(fig.250,fig.500,fig.750,fig.1000,
+                      nrow = 2,ncol = 2,legend = 'top',common.legend = TRUE,
+                      labels = list('A','B','C','D'))
+
+ggsave(fig.calls,filename = 'Figure_H3K27me3_Viterbi.pdf',height = 11.5,width = 9)

### 2021-02-19 Supp/Figure_H3K27me3_Viterbi/Figure_PeakCalls_H3K27me3_Encode_twocells_1000bp.R

@@ -1,0 +1,202 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+library(epigraHMM)
+
+bp = 1000
+chromosome = 'chr20'
+ngroups = 2
+mark = 'H3K27me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = c(0.01,0.05,0.10,0.15,0.20)
+methods <- c('Genes','0.01','0.05','0.10','0.15','0.20','Viterbi')
+colors = c('#000000',brewer.pal(6,'Dark2'))
+names(colors) <- methods 
+
+peakpos = c(1.1,1.1+seq(0.1,by = 0.1,length.out = length(fdr)+1))
+names(peakpos) <- methods 
+
+# Loading data
+load(paste0('../../Data/Encode_helas3/H3K27me3/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep1.markdup.q10.sorted.RData'))
+helas31 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_helas3/H3K27me3/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep2.markdup.q10.sorted.RData'))
+helas32 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+
+load(paste0('../../Data/Encode_hepg2/H3K27me3/wgEncodeBroadHistoneHepg2H3k27me3StdAlnRep1.markdup.q10.sorted.RData'))
+hepg21 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_hepg2/H3K27me3/wgEncodeBroadHistoneHepg2H3k27me3StdAlnRep2.markdup.q10.sorted.RData'))
+hepg22 = subset(counts[[paste0(bp)]],chr==chromosome)
+
+rawcounts = as.data.table(counts[[paste0(bp)]])
+rawcounts = rawcounts[chr%in%paste0('chr',1:22),]
+gr.rawcounts <- rawcounts[,GRanges(chr, IRanges(start=start, end=stop))]
+
+idx.chr = (counts[[paste0(bp)]]$chr==chromosome)
+
[…]

### 2021-02-19 Supp/Figure_H3K27me3_Viterbi/Figure_PeakCalls_H3K27me3_Encode_twocells_250bp.R

@@ -1,0 +1,202 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+library(epigraHMM)
+
+bp = 250
+chromosome = 'chr20'
+ngroups = 2
+mark = 'H3K27me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = c(0.01,0.05,0.10,0.15,0.20)
+methods <- c('Genes','0.01','0.05','0.10','0.15','0.20','Viterbi')
+colors = c('#000000',brewer.pal(6,'Dark2'))
+names(colors) <- methods 
+
+peakpos = c(1.1,1.1+seq(0.1,by = 0.1,length.out = length(fdr)+1))
+names(peakpos) <- methods 
+
+# Loading data
+load(paste0('../../Data/Encode_helas3/H3K27me3/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep1.markdup.q10.sorted.RData'))
+helas31 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_helas3/H3K27me3/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep2.markdup.q10.sorted.RData'))
+helas32 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+
+load(paste0('../../Data/Encode_hepg2/H3K27me3/wgEncodeBroadHistoneHepg2H3k27me3StdAlnRep1.markdup.q10.sorted.RData'))
+hepg21 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_hepg2/H3K27me3/wgEncodeBroadHistoneHepg2H3k27me3StdAlnRep2.markdup.q10.sorted.RData'))
+hepg22 = subset(counts[[paste0(bp)]],chr==chromosome)
+
+rawcounts = as.data.table(counts[[paste0(bp)]])
+rawcounts = rawcounts[chr%in%paste0('chr',1:22),]
+gr.rawcounts <- rawcounts[,GRanges(chr, IRanges(start=start, end=stop))]
+
+idx.chr = (counts[[paste0(bp)]]$chr==chromosome)
+
[…]

### 2021-02-19 Supp/Figure_H3K27me3_Viterbi/Figure_PeakCalls_H3K27me3_Encode_twocells_500bp.R

@@ -1,0 +1,202 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+library(epigraHMM)
+
+bp = 500
+chromosome = 'chr20'
+ngroups = 2
+mark = 'H3K27me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = c(0.01,0.05,0.10,0.15,0.20)
+methods <- c('Genes','0.01','0.05','0.10','0.15','0.20','Viterbi')
+colors = c('#000000',brewer.pal(6,'Dark2'))
+names(colors) <- methods 
+
+peakpos = c(1.1,1.1+seq(0.1,by = 0.1,length.out = length(fdr)+1))
+names(peakpos) <- methods 
+
+# Loading data
+load(paste0('../../Data/Encode_helas3/H3K27me3/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep1.markdup.q10.sorted.RData'))
+helas31 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_helas3/H3K27me3/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep2.markdup.q10.sorted.RData'))
+helas32 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+
+load(paste0('../../Data/Encode_hepg2/H3K27me3/wgEncodeBroadHistoneHepg2H3k27me3StdAlnRep1.markdup.q10.sorted.RData'))
+hepg21 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_hepg2/H3K27me3/wgEncodeBroadHistoneHepg2H3k27me3StdAlnRep2.markdup.q10.sorted.RData'))
+hepg22 = subset(counts[[paste0(bp)]],chr==chromosome)
+
+rawcounts = as.data.table(counts[[paste0(bp)]])
+rawcounts = rawcounts[chr%in%paste0('chr',1:22),]
+gr.rawcounts <- rawcounts[,GRanges(chr, IRanges(start=start, end=stop))]
+
+idx.chr = (counts[[paste0(bp)]]$chr==chromosome)
+
[…]

### 2021-02-19 Supp/Figure_H3K27me3_Viterbi/Figure_PeakCalls_H3K27me3_Encode_twocells_750bp.R

@@ -1,0 +1,202 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+library(epigraHMM)
+
+bp = 750
+chromosome = 'chr20'
+ngroups = 2
+mark = 'H3K27me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = c(0.01,0.05,0.10,0.15,0.20)
+methods <- c('Genes','0.01','0.05','0.10','0.15','0.20','Viterbi')
+colors = c('#000000',brewer.pal(6,'Dark2'))
+names(colors) <- methods 
+
+peakpos = c(1.1,1.1+seq(0.1,by = 0.1,length.out = length(fdr)+1))
+names(peakpos) <- methods 
+
+# Loading data
+load(paste0('../../Data/Encode_helas3/H3K27me3/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep1.markdup.q10.sorted.RData'))
+helas31 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_helas3/H3K27me3/wgEncodeBroadHistoneHelas3H3k27me3StdAlnRep2.markdup.q10.sorted.RData'))
+helas32 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+
+load(paste0('../../Data/Encode_hepg2/H3K27me3/wgEncodeBroadHistoneHepg2H3k27me3StdAlnRep1.markdup.q10.sorted.RData'))
+hepg21 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_hepg2/H3K27me3/wgEncodeBroadHistoneHepg2H3k27me3StdAlnRep2.markdup.q10.sorted.RData'))
+hepg22 = subset(counts[[paste0(bp)]],chr==chromosome)
+
+rawcounts = as.data.table(counts[[paste0(bp)]])
+rawcounts = rawcounts[chr%in%paste0('chr',1:22),]
+gr.rawcounts <- rawcounts[,GRanges(chr, IRanges(start=start, end=stop))]
+
+idx.chr = (counts[[paste0(bp)]]$chr==chromosome)
+
[…]

### 2021-02-19 Supp/Figure_H3K36me3_Viterbi/Figure_H3K36me3_Viterbi.R

@@ -1,0 +1,10 @@
+load('Figure_PeakCalls_H3K36me3_Encode_twocells_250bp.RData')
+load('Figure_PeakCalls_H3K36me3_Encode_twocells_500bp.RData')
+load('Figure_PeakCalls_H3K36me3_Encode_twocells_750bp.RData')
+load('Figure_PeakCalls_H3K36me3_Encode_twocells_1000bp.RData')
+
+fig.calls = ggarrange(fig.250,fig.500,fig.750,fig.1000,
+                      nrow = 2,ncol = 2,legend = 'top',common.legend = TRUE,
+                      labels = list('A','B','C','D'))
+
+ggsave(fig.calls,filename = 'Figure_H3K36me3_Viterbi.pdf',height = 11.5,width = 9)

### 2021-02-19 Supp/Figure_H3K36me3_Viterbi/Figure_PeakCalls_H3K36me3_Encode_twocells_1000bp.R

@@ -1,0 +1,202 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+library(epigraHMM)
+
+bp = 1000
+chromosome = 'chr19'
+ngroups = 2
+mark = 'H3K36me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = c(0.01,0.05,0.10,0.15,0.20)
+methods <- c('Genes','0.01','0.05','0.10','0.15','0.20','Viterbi')
+colors = c('#000000',brewer.pal(6,'Dark2'))
+names(colors) <- methods 
+
+peakpos = c(1.1,1.1+seq(0.1,by = 0.1,length.out = length(fdr)+1))
+names(peakpos) <- methods 
+
+# Loading data
+load(paste0('../../Data/Encode_helas3/H3K36me3/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep1.markdup.q10.sorted.RData'))
+helas31 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_helas3/H3K36me3/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep2.markdup.q10.sorted.RData'))
+helas32 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+
+load(paste0('../../Data/Encode_hepg2/H3K36me3/wgEncodeBroadHistoneHepg2H3k36me3StdAlnRep1.markdup.q10.sorted.RData'))
+hepg21 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_hepg2/H3K36me3/wgEncodeBroadHistoneHepg2H3k36me3StdAlnRep2.markdup.q10.sorted.RData'))
+hepg22 = subset(counts[[paste0(bp)]],chr==chromosome)
+
+rawcounts = as.data.table(counts[[paste0(bp)]])
+rawcounts = rawcounts[chr%in%paste0('chr',1:22),]
+gr.rawcounts <- rawcounts[,GRanges(chr, IRanges(start=start, end=stop))]
+
+idx.chr = (counts[[paste0(bp)]]$chr==chromosome)
+
[…]

### 2021-02-19 Supp/Figure_H3K36me3_Viterbi/Figure_PeakCalls_H3K36me3_Encode_twocells_250bp.R

@@ -1,0 +1,202 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+library(epigraHMM)
+
+bp = 250
+chromosome = 'chr19'
+ngroups = 2
+mark = 'H3K36me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = c(0.01,0.05,0.10,0.15,0.20)
+methods <- c('Genes','0.01','0.05','0.10','0.15','0.20','Viterbi')
+colors = c('#000000',brewer.pal(6,'Dark2'))
+names(colors) <- methods 
+
+peakpos = c(1.1,1.1+seq(0.1,by = 0.1,length.out = length(fdr)+1))
+names(peakpos) <- methods 
+
+# Loading data
+load(paste0('../../Data/Encode_helas3/H3K36me3/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep1.markdup.q10.sorted.RData'))
+helas31 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_helas3/H3K36me3/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep2.markdup.q10.sorted.RData'))
+helas32 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+
+load(paste0('../../Data/Encode_hepg2/H3K36me3/wgEncodeBroadHistoneHepg2H3k36me3StdAlnRep1.markdup.q10.sorted.RData'))
+hepg21 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_hepg2/H3K36me3/wgEncodeBroadHistoneHepg2H3k36me3StdAlnRep2.markdup.q10.sorted.RData'))
+hepg22 = subset(counts[[paste0(bp)]],chr==chromosome)
+
+rawcounts = as.data.table(counts[[paste0(bp)]])
+rawcounts = rawcounts[chr%in%paste0('chr',1:22),]
+gr.rawcounts <- rawcounts[,GRanges(chr, IRanges(start=start, end=stop))]
+
+idx.chr = (counts[[paste0(bp)]]$chr==chromosome)
+
[…]

### 2021-02-19 Supp/Figure_H3K36me3_Viterbi/Figure_PeakCalls_H3K36me3_Encode_twocells_500bp.R

@@ -1,0 +1,202 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+library(epigraHMM)
+
+bp = 500
+chromosome = 'chr19'
+ngroups = 2
+mark = 'H3K36me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = c(0.01,0.05,0.10,0.15,0.20)
+methods <- c('Genes','0.01','0.05','0.10','0.15','0.20','Viterbi')
+colors = c('#000000',brewer.pal(6,'Dark2'))
+names(colors) <- methods 
+
+peakpos = c(1.1,1.1+seq(0.1,by = 0.1,length.out = length(fdr)+1))
+names(peakpos) <- methods 
+
+# Loading data
+load(paste0('../../Data/Encode_helas3/H3K36me3/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep1.markdup.q10.sorted.RData'))
+helas31 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_helas3/H3K36me3/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep2.markdup.q10.sorted.RData'))
+helas32 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+
+load(paste0('../../Data/Encode_hepg2/H3K36me3/wgEncodeBroadHistoneHepg2H3k36me3StdAlnRep1.markdup.q10.sorted.RData'))
+hepg21 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_hepg2/H3K36me3/wgEncodeBroadHistoneHepg2H3k36me3StdAlnRep2.markdup.q10.sorted.RData'))
+hepg22 = subset(counts[[paste0(bp)]],chr==chromosome)
+
+rawcounts = as.data.table(counts[[paste0(bp)]])
+rawcounts = rawcounts[chr%in%paste0('chr',1:22),]
+gr.rawcounts <- rawcounts[,GRanges(chr, IRanges(start=start, end=stop))]
+
+idx.chr = (counts[[paste0(bp)]]$chr==chromosome)
+
[…]

### 2021-02-19 Supp/Figure_H3K36me3_Viterbi/Figure_PeakCalls_H3K36me3_Encode_twocells_750bp.R

@@ -1,0 +1,202 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+library(epigraHMM)
+
+bp = 750
+chromosome = 'chr19'
+ngroups = 2
+mark = 'H3K36me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = c(0.01,0.05,0.10,0.15,0.20)
+methods <- c('Genes','0.01','0.05','0.10','0.15','0.20','Viterbi')
+colors = c('#000000',brewer.pal(6,'Dark2'))
+names(colors) <- methods 
+
+peakpos = c(1.1,1.1+seq(0.1,by = 0.1,length.out = length(fdr)+1))
+names(peakpos) <- methods 
+
+# Loading data
+load(paste0('../../Data/Encode_helas3/H3K36me3/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep1.markdup.q10.sorted.RData'))
+helas31 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_helas3/H3K36me3/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep2.markdup.q10.sorted.RData'))
+helas32 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+
+load(paste0('../../Data/Encode_hepg2/H3K36me3/wgEncodeBroadHistoneHepg2H3k36me3StdAlnRep1.markdup.q10.sorted.RData'))
+hepg21 = subset(counts[[paste0(bp)]],chr==chromosome);rm(counts)
+load(paste0('../../Data/Encode_hepg2/H3K36me3/wgEncodeBroadHistoneHepg2H3k36me3StdAlnRep2.markdup.q10.sorted.RData'))
+hepg22 = subset(counts[[paste0(bp)]],chr==chromosome)
+
+rawcounts = as.data.table(counts[[paste0(bp)]])
+rawcounts = rawcounts[chr%in%paste0('chr',1:22),]
+gr.rawcounts <- rawcounts[,GRanges(chr, IRanges(start=start, end=stop))]
+
+idx.chr = (counts[[paste0(bp)]]$chr==chromosome)
+
[…]

### 2021-02-19 Supp/Figure_MultiMarks/Figure_MultiMarks.R

@@ -1,0 +1,213 @@
+# Figure F: genomic segmentation with Viterbi path
+
+# Loading libraries
+library(epigraHMM)
+library(data.table)
+library(magrittr)
+library(plyr)
+library(ggplot2)
+library(ggrepel)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(tidyverse)
+library(ggnewscale)
+
+# Some general parameters and functions
+
+bp <- 500
+chromosome <- 'chr19'
+idx.genome <- c(21251806,21417266)
+cellline <- 'Helas3'
+size.text <- 2
+size.axis <- 6
+h.track = c(50,40)
+h.txt = c(50,40)
+
+Modes <- function(x) {
+  ux <- unique(x)
+  tab <- tabulate(match(x, ux))
+  if (length(ux[tab == max(tab)]) == 1) {
+    return(ux[tab == max(tab)])
+  } else{
+    out = ux[tab == max(tab)]
+    return(max(out))
+  }
+}
+
+pat.peaks = function(x,gr.peaks,gr.genome){
+  ovp = findOverlaps(gr.peaks,gr.genome,ignore.strand = T)
+  out = rbindlist(lapply(unique(queryHits(ovp)), FUN = function(i){data.table(Peak = i,Mode = Modes(x[subjectHits(ovp)[queryHits(ovp) == i],]$Pattern))}))
+  Pattern = rep(NA,length(gr.peaks))
+  Pattern[out$Peak] = out$Mode
+  return(Pattern)
+}
+
+# Loading data
+
+load(paste0('../../Data/Encode_helas3/CTCF/wgEncodeBroadHistoneHelas3CtcfStdAlnRep1.markdup.q10.sorted.RData'))
+ctcf.rep1 = counts[[paste0(bp)]];rm(counts)
+load(paste0('../../Data/Encode_helas3/CTCF/wgEncodeBroadHistoneHelas3CtcfStdAlnRep2.markdup.q10.sorted.RData'))
+ctcf.rep2 = counts[[paste0(bp)]];rm(counts)
+
+load(paste0('../../Data/Encode_helas3/H3K36me3/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep1.markdup.q10.sorted.RData'))
+h3k36me3.rep1 = counts[[paste0(bp)]];rm(counts)
[…]

### 2021-02-19 Supp/Figure_Narrow_1000bp/Figure_Correlation_CTCF_Encode_twocells_1000bp.R

@@ -1,0 +1,270 @@
+library(magrittr)
+library(data.table)
+library(tidyverse)
+library(tibble)
+library(RColorBrewer)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(vsn)
+library(DESeq2)
+library(ggpubr)
+library(Polychrome)
+
+# General parameters
+bp = 1000
+chromosome = paste0('chr',1:22)
+mark = 'CTCF'
+data = 'Encode_twocells'
+fdr = 0.05
+sizetext = 7
+sizetitle = 9
+
+agg = function(x,gr.peak,gr.genome,cells = c('Helas3.adj','Hepg2.adj')){
+  gr.summarized = as.data.frame(GenomicRanges::findOverlaps(gr.peak,gr.genome))
+  gr.summarized = as.data.table(cbind(gr.summarized,x[gr.summarized$subjectHits,cells,with=F]));colnames(gr.summarized)=c('Peak','Genome',cells)
+  gr.summarized[,Width :=  width(gr.peak[Peak])]
+  gr.summarized <- gr.summarized[,.(Summary_Helas3.adj = sum(Helas3.adj,na.rm = T),
+                                    Summary_Hepg2.adj = sum(Hepg2.adj,na.rm = T),
+                                    Summary_Width = unique(Width)),by=Peak]
+  
+  return(gr.summarized)
+}
+
+getcor <- function(gr,p){
+  newgr <- gr[order(-abs(log2(Summary_Hepg2.adj+1)-log2(Summary_Helas3.adj+1))),][1:round(.N*p),]
+  newgr.UP = newgr[(Summary_Hepg2.adj>=Summary_Helas3.adj),]
+  newgr.DOWN = newgr[(Summary_Hepg2.adj<Summary_Helas3.adj),]
+  
+  return(data.table(Up.LFC = ifelse(newgr.UP[,.N]>1,newgr.UP[,median(log2((Summary_Hepg2.adj+1)/(Summary_Helas3.adj+1)))],NA),
+                    Down.LFC = ifelse(newgr.DOWN[,.N]>1,newgr.DOWN[,median(log2((Summary_Hepg2.adj+1)/(Summary_Helas3.adj+1)))],NA),
+                    Cor = newgr[,cor(log2(Summary_Helas3.adj+1),log2(Summary_Hepg2.adj+1),method = 'spearman')]))
+}
+
[…]

### 2021-02-19 Supp/Figure_Narrow_1000bp/Figure_Correlation_H3K4me3_Encode_twocells_1000bp.R

@@ -1,0 +1,401 @@
+library(magrittr)
+library(data.table)
+library(tidyverse)
+library(tibble)
+library(RColorBrewer)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(vsn)
+library(DESeq2)
+library(ggpubr)
+library(Polychrome)
+
+# General parameters
+bp = 1000
+chromosome = paste0('chr', 1:22)
+mark = 'H3K4me3'
+data = 'Encode_twocells'
+fdr = 0.05
+sizetext = 7
+sizetitle = 9
+
+agg = function(x,
+               gr.peak,
+               gr.genome,
+               cells = c('Helas3.adj', 'Hepg2.adj')) {
+  gr.summarized = as.data.frame(GenomicRanges::findOverlaps(gr.peak, gr.genome))
+  gr.summarized = as.data.table(cbind(gr.summarized, x[gr.summarized$subjectHits, cells, with =
+                                                         F]))
+  colnames(gr.summarized) = c('Peak', 'Genome', cells)
+  gr.summarized[, Width :=  width(gr.peak[Peak])]
+  gr.summarized <-
+    gr.summarized[, .(
+      Summary_Helas3.adj = sum(Helas3.adj, na.rm = T),
+      Summary_Hepg2.adj = sum(Hepg2.adj, na.rm = T),
+      Summary_Width = unique(Width)
+    ), by = Peak]
+  
+  return(gr.summarized)
+}
+
+getcor <- function(gr, p) {
+  newgr <-
+    gr[order(-abs(log2(Summary_Hepg2.adj + 1) - log2(Summary_Helas3.adj + 1))), ][1:round(.N *
+                                                                                            p), ]
+  newgr.UP = newgr[(Summary_Hepg2.adj >= Summary_Helas3.adj), ]
+  newgr.DOWN = newgr[(Summary_Hepg2.adj < Summary_Helas3.adj), ]
+  
+  return(data.table(
+    Up.LFC = ifelse(newgr.UP[, .N] > 1, newgr.UP[, median(log2((Summary_Hepg2.adj +
+                                                                  1) / (Summary_Helas3.adj + 1)))], NA),
[…]

### 2021-02-19 Supp/Figure_Narrow_1000bp/Figure_Memory_Time.R

@@ -1,0 +1,174 @@
+# Libraries
+
+library(data.table)
+library(Polychrome)
+library(magrittr)
+library(ggplot2)
+library(microbenchmark)
+
+# Setup
+
+colors = c(kelly.colors(22)[c('lightblue',
+                              'red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'buff',
+                              'orange')])
+methods <-
+  c(
+    "epigraHMM",
+    "ChIPComp +\nMACS2",
+    "csaw",
+    "DiffBind +\nMACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+names(colors) <- methods
+
+################################################################################
+### Memory
+################################################################################
+
+files.csv <-
+  list.files('../../Public',
+             'slurm*.*.csv$',
+             recursive = TRUE,
+             full.names = TRUE)
+
+# Filtering
+
+files.csv <- files.csv[grep('Encode_twocells|MACS2', files.csv)]
+
+# Looping through files
+
+mem.dt <- rbindlist(lapply(strsplit(files.csv, '/'), function(x) {
+  # Does not run for MACS2
+  if (length(grep('MACS2', x)) == 0) {
+    subx <- strsplit(x[7], '_|bp')[[1]]
+    
+    csvx <- fread(paste(x, collapse = '/'))[2, ]$MaxRSS
+    
+    # Do not proceed if job is not completed
+    if (!is.na(csvx)) {
+      # If ChIPComp or DiffBind, need to check MACS2
+      if (x[4] %in% c('ChIPComp', 'DiffBind')) {
+        csvmacs2 <-
+          rbindlist(lapply(files.csv[intersect(grep(paste0('MACS2/', x[5]), files.csv), grep(paste0('Helas3|Hepg2'), files.csv))], function(y) {
+            fread(y)[JobName == 'batch', ]
+          }))$MaxRSS
+        
+        mem <-
+          max(as.numeric(strsplit(csvx, 'K')[[1]]), max(as.numeric(unlist(
+            strsplit(as.character(csvmacs2), 'K')
+          ))))
+      } else{
+        mem <- as.numeric(strsplit(csvx, 'K')[[1]])
+      }
+      
+      return(data.table(
+        Method = x[4],
+        Mark = x[5],
+        Data = x[6],
+        BP = subx[5],
[…]

### 2021-02-19 Supp/Figure_Narrow_1000bp/Figure_Narrow.R

@@ -1,0 +1,22 @@
+library(ggpubr)
+
+load('Figure_Correlation_CTCF_Encode_twocells_1000bp.RData')
+load('Figure_Correlation_H3K4me3_Encode_twocells_1000bp.RData')
+load('Figure_PeakCalls_CTCF_Encode_twocells_1000bp.RData')
+load('Figure_PeakCalls_H3K4me3_Encode_twocells_1000bp.RData')
+load('Figure_PeakCalls_H3K27ac_Encode_twocells_1000bp.RData')
+load('Figure_Memory_Time.RData')
+
+# Fig Peak Calls
+fig.calls = ggarrange(fig.ctcf,fig.h3k4me3,fig.h3k27ac,nrow = 3,ncol = 1,legend = 'none',labels = list('B','D','F'))
+
+# Figure Metrics
+fig.metrics = ggarrange(fig.lfc,fig.cor,fig_memory_time,ncol=1,nrow=3,legend = 'none',labels = list('A','C','E'))
+
+# Put them all together
+
+fig.final = ggarrange(fig.metrics,fig.calls,nrow=1,ncol=2)
+fig.final.label = ggarrange(get_legend(fig.lfc + theme(legend.position = 'top',legend.direction = 'horizontal',legend.box = 'vertical',legend.spacing.y = unit(-0.25, "cm")) + guides(col = guide_legend(nrow = 1,order = 1),linetype = guide_legend(nrow = 1,order = 2,override.aes = list(col = 'black'),title = 'Observed Enrichment'))),
+                            fig.final,nrow=2,heights = c(0.05,0.95))
+
+ggsave(filename = 'Figure_Narrow.pdf',plot = fig.final.label,height = 11.5,width = 9,dpi = 'retina')

### 2021-02-19 Supp/Figure_Narrow_1000bp/Figure_PeakCalls_CTCF_Encode_twocells_1000bp.R

@@ -1,0 +1,434 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 1000
+chromosome = 'chr8'
+ngroups = 2
+mark = 'CTCF'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3CtcfStdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3CtcfStdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2CtcfStdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
[…]

### 2021-02-19 Supp/Figure_Narrow_1000bp/Figure_PeakCalls_H3K27ac_Encode_twocells_1000bp.R

@@ -1,0 +1,434 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 1000
+chromosome = 'chr19'
+ngroups = 2
+mark = 'H3K27ac'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k27acStdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k27acStdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2H3k27acStdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
[…]

### 2021-02-19 Supp/Figure_Narrow_1000bp/Figure_PeakCalls_H3K4me3_Encode_twocells_1000bp.R

@@ -1,0 +1,432 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 1000
+chromosome = 'chr10'
+ngroups = 2
+mark = 'H3K4me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k4me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k4me3StdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2H3k4me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
[…]

### 2021-02-19 Supp/Figure_Narrow_500bp/Figure_Correlation_CTCF_Encode_twocells_500bp.R

@@ -1,0 +1,270 @@
+library(magrittr)
+library(data.table)
+library(tidyverse)
+library(tibble)
+library(RColorBrewer)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(vsn)
+library(DESeq2)
+library(ggpubr)
+library(Polychrome)
+
+# General parameters
+bp = 500
+chromosome = paste0('chr',1:22)
+mark = 'CTCF'
+data = 'Encode_twocells'
+fdr = 0.05
+sizetext = 7
+sizetitle = 9
+
+agg = function(x,gr.peak,gr.genome,cells = c('Helas3.adj','Hepg2.adj')){
+  gr.summarized = as.data.frame(GenomicRanges::findOverlaps(gr.peak,gr.genome))
+  gr.summarized = as.data.table(cbind(gr.summarized,x[gr.summarized$subjectHits,cells,with=F]));colnames(gr.summarized)=c('Peak','Genome',cells)
+  gr.summarized[,Width :=  width(gr.peak[Peak])]
+  gr.summarized <- gr.summarized[,.(Summary_Helas3.adj = sum(Helas3.adj,na.rm = T),
+                                    Summary_Hepg2.adj = sum(Hepg2.adj,na.rm = T),
+                                    Summary_Width = unique(Width)),by=Peak]
+  
+  return(gr.summarized)
+}
+
+getcor <- function(gr,p){
+  newgr <- gr[order(-abs(log2(Summary_Hepg2.adj+1)-log2(Summary_Helas3.adj+1))),][1:round(.N*p),]
+  newgr.UP = newgr[(Summary_Hepg2.adj>=Summary_Helas3.adj),]
+  newgr.DOWN = newgr[(Summary_Hepg2.adj<Summary_Helas3.adj),]
+  
+  return(data.table(Up.LFC = ifelse(newgr.UP[,.N]>1,newgr.UP[,median(log2((Summary_Hepg2.adj+1)/(Summary_Helas3.adj+1)))],NA),
+                    Down.LFC = ifelse(newgr.DOWN[,.N]>1,newgr.DOWN[,median(log2((Summary_Hepg2.adj+1)/(Summary_Helas3.adj+1)))],NA),
+                    Cor = newgr[,cor(log2(Summary_Helas3.adj+1),log2(Summary_Hepg2.adj+1),method = 'spearman')]))
+}
+
[…]

### 2021-02-19 Supp/Figure_Narrow_500bp/Figure_Correlation_H3K4me3_Encode_twocells_500bp.R

@@ -1,0 +1,401 @@
+library(magrittr)
+library(data.table)
+library(tidyverse)
+library(tibble)
+library(RColorBrewer)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(vsn)
+library(DESeq2)
+library(ggpubr)
+library(Polychrome)
+
+# General parameters
+bp = 500
+chromosome = paste0('chr', 1:22)
+mark = 'H3K4me3'
+data = 'Encode_twocells'
+fdr = 0.05
+sizetext = 7
+sizetitle = 9
+
+agg = function(x,
+               gr.peak,
+               gr.genome,
+               cells = c('Helas3.adj', 'Hepg2.adj')) {
+  gr.summarized = as.data.frame(GenomicRanges::findOverlaps(gr.peak, gr.genome))
+  gr.summarized = as.data.table(cbind(gr.summarized, x[gr.summarized$subjectHits, cells, with =
+                                                         F]))
+  colnames(gr.summarized) = c('Peak', 'Genome', cells)
+  gr.summarized[, Width :=  width(gr.peak[Peak])]
+  gr.summarized <-
+    gr.summarized[, .(
+      Summary_Helas3.adj = sum(Helas3.adj, na.rm = T),
+      Summary_Hepg2.adj = sum(Hepg2.adj, na.rm = T),
+      Summary_Width = unique(Width)
+    ), by = Peak]
+  
+  return(gr.summarized)
+}
+
+getcor <- function(gr, p) {
+  newgr <-
+    gr[order(-abs(log2(Summary_Hepg2.adj + 1) - log2(Summary_Helas3.adj + 1))), ][1:round(.N *
+                                                                                            p), ]
+  newgr.UP = newgr[(Summary_Hepg2.adj >= Summary_Helas3.adj), ]
+  newgr.DOWN = newgr[(Summary_Hepg2.adj < Summary_Helas3.adj), ]
+  
+  return(data.table(
+    Up.LFC = ifelse(newgr.UP[, .N] > 1, newgr.UP[, median(log2((Summary_Hepg2.adj +
+                                                                  1) / (Summary_Helas3.adj + 1)))], NA),
[…]

### 2021-02-19 Supp/Figure_Narrow_500bp/Figure_Memory_Time.R

@@ -1,0 +1,174 @@
+# Libraries
+
+library(data.table)
+library(Polychrome)
+library(magrittr)
+library(ggplot2)
+library(microbenchmark)
+
+# Setup
+
+colors = c(kelly.colors(22)[c('lightblue',
+                              'red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'buff',
+                              'orange')])
+methods <-
+  c(
+    "epigraHMM",
+    "ChIPComp +\nMACS2",
+    "csaw",
+    "DiffBind +\nMACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+names(colors) <- methods
+
+################################################################################
+### Memory
+################################################################################
+
+files.csv <-
+  list.files('../../Public',
+             'slurm*.*.csv$',
+             recursive = TRUE,
+             full.names = TRUE)
+
+# Filtering
+
+files.csv <- files.csv[grep('Encode_twocells|MACS2', files.csv)]
+
+# Looping through files
+
+mem.dt <- rbindlist(lapply(strsplit(files.csv, '/'), function(x) {
+  # Does not run for MACS2
+  if (length(grep('MACS2', x)) == 0) {
+    subx <- strsplit(x[7], '_|bp')[[1]]
+    
+    csvx <- fread(paste(x, collapse = '/'))[2, ]$MaxRSS
+    
+    # Do not proceed if job is not completed
+    if (!is.na(csvx)) {
+      # If ChIPComp or DiffBind, need to check MACS2
+      if (x[4] %in% c('ChIPComp', 'DiffBind')) {
+        csvmacs2 <-
+          rbindlist(lapply(files.csv[intersect(grep(paste0('MACS2/', x[5]), files.csv), grep(paste0('Helas3|Hepg2'), files.csv))], function(y) {
+            fread(y)[JobName == 'batch', ]
+          }))$MaxRSS
+        
+        mem <-
+          max(as.numeric(strsplit(csvx, 'K')[[1]]), max(as.numeric(unlist(
+            strsplit(as.character(csvmacs2), 'K')
+          ))))
+      } else{
+        mem <- as.numeric(strsplit(csvx, 'K')[[1]])
+      }
+      
+      return(data.table(
+        Method = x[4],
+        Mark = x[5],
+        Data = x[6],
+        BP = subx[5],
[…]

### 2021-02-19 Supp/Figure_Narrow_500bp/Figure_Narrow.R

@@ -1,0 +1,22 @@
+library(ggpubr)
+
+load('Figure_Correlation_CTCF_Encode_twocells_500bp.RData')
+load('Figure_Correlation_H3K4me3_Encode_twocells_500bp.RData')
+load('Figure_PeakCalls_CTCF_Encode_twocells_500bp.RData')
+load('Figure_PeakCalls_H3K4me3_Encode_twocells_500bp.RData')
+load('Figure_PeakCalls_H3K27ac_Encode_twocells_500bp.RData')
+load('Figure_Memory_Time.RData')
+
+# Fig Peak Calls
+fig.calls = ggarrange(fig.ctcf,fig.h3k4me3,fig.h3k27ac,nrow = 3,ncol = 1,legend = 'none',labels = list('B','D','F'))
+
+# Figure Metrics
+fig.metrics = ggarrange(fig.lfc,fig.cor,fig_memory_time,ncol=1,nrow=3,legend = 'none',labels = list('A','C','E'))
+
+# Put them all together
+
+fig.final = ggarrange(fig.metrics,fig.calls,nrow=1,ncol=2)
+fig.final.label = ggarrange(get_legend(fig.lfc + theme(legend.position = 'top',legend.direction = 'horizontal',legend.box = 'vertical',legend.spacing.y = unit(-0.25, "cm")) + guides(col = guide_legend(nrow = 1,order = 1),linetype = guide_legend(nrow = 1,order = 2,override.aes = list(col = 'black'),title = 'Observed Enrichment'))),
+                            fig.final,nrow=2,heights = c(0.05,0.95))
+
+ggsave(filename = 'Figure_Narrow.pdf',plot = fig.final.label,height = 11.5,width = 9,dpi = 'retina')

### 2021-02-19 Supp/Figure_Narrow_500bp/Figure_PeakCalls_CTCF_Encode_twocells_500bp.R

@@ -1,0 +1,434 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 500
+chromosome = 'chr8'
+ngroups = 2
+mark = 'CTCF'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3CtcfStdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3CtcfStdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2CtcfStdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
[…]

### 2021-02-19 Supp/Figure_Narrow_500bp/Figure_PeakCalls_H3K27ac_Encode_twocells_500bp.R

@@ -1,0 +1,434 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 500
+chromosome = 'chr19'
+ngroups = 2
+mark = 'H3K27ac'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k27acStdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k27acStdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2H3k27acStdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
[…]

### 2021-02-19 Supp/Figure_Narrow_500bp/Figure_PeakCalls_H3K4me3_Encode_twocells_500bp.R

@@ -1,0 +1,432 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 500
+chromosome = 'chr10'
+ngroups = 2
+mark = 'H3K4me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k4me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k4me3StdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2H3k4me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
[…]

### 2021-02-19 Supp/Figure_Narrow_750bp/Figure_Correlation_CTCF_Encode_twocells_750bp.R

@@ -1,0 +1,270 @@
+library(magrittr)
+library(data.table)
+library(tidyverse)
+library(tibble)
+library(RColorBrewer)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(vsn)
+library(DESeq2)
+library(ggpubr)
+library(Polychrome)
+
+# General parameters
+bp = 750
+chromosome = paste0('chr',1:22)
+mark = 'CTCF'
+data = 'Encode_twocells'
+fdr = 0.05
+sizetext = 7
+sizetitle = 9
+
+agg = function(x,gr.peak,gr.genome,cells = c('Helas3.adj','Hepg2.adj')){
+  gr.summarized = as.data.frame(GenomicRanges::findOverlaps(gr.peak,gr.genome))
+  gr.summarized = as.data.table(cbind(gr.summarized,x[gr.summarized$subjectHits,cells,with=F]));colnames(gr.summarized)=c('Peak','Genome',cells)
+  gr.summarized[,Width :=  width(gr.peak[Peak])]
+  gr.summarized <- gr.summarized[,.(Summary_Helas3.adj = sum(Helas3.adj,na.rm = T),
+                                    Summary_Hepg2.adj = sum(Hepg2.adj,na.rm = T),
+                                    Summary_Width = unique(Width)),by=Peak]
+  
+  return(gr.summarized)
+}
+
+getcor <- function(gr,p){
+  newgr <- gr[order(-abs(log2(Summary_Hepg2.adj+1)-log2(Summary_Helas3.adj+1))),][1:round(.N*p),]
+  newgr.UP = newgr[(Summary_Hepg2.adj>=Summary_Helas3.adj),]
+  newgr.DOWN = newgr[(Summary_Hepg2.adj<Summary_Helas3.adj),]
+  
+  return(data.table(Up.LFC = ifelse(newgr.UP[,.N]>1,newgr.UP[,median(log2((Summary_Hepg2.adj+1)/(Summary_Helas3.adj+1)))],NA),
+                    Down.LFC = ifelse(newgr.DOWN[,.N]>1,newgr.DOWN[,median(log2((Summary_Hepg2.adj+1)/(Summary_Helas3.adj+1)))],NA),
+                    Cor = newgr[,cor(log2(Summary_Helas3.adj+1),log2(Summary_Hepg2.adj+1),method = 'spearman')]))
+}
+
[…]

### 2021-02-19 Supp/Figure_Narrow_750bp/Figure_Correlation_H3K4me3_Encode_twocells_750bp.R

@@ -1,0 +1,401 @@
+library(magrittr)
+library(data.table)
+library(tidyverse)
+library(tibble)
+library(RColorBrewer)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(vsn)
+library(DESeq2)
+library(ggpubr)
+library(Polychrome)
+
+# General parameters
+bp = 750
+chromosome = paste0('chr', 1:22)
+mark = 'H3K4me3'
+data = 'Encode_twocells'
+fdr = 0.05
+sizetext = 7
+sizetitle = 9
+
+agg = function(x,
+               gr.peak,
+               gr.genome,
+               cells = c('Helas3.adj', 'Hepg2.adj')) {
+  gr.summarized = as.data.frame(GenomicRanges::findOverlaps(gr.peak, gr.genome))
+  gr.summarized = as.data.table(cbind(gr.summarized, x[gr.summarized$subjectHits, cells, with =
+                                                         F]))
+  colnames(gr.summarized) = c('Peak', 'Genome', cells)
+  gr.summarized[, Width :=  width(gr.peak[Peak])]
+  gr.summarized <-
+    gr.summarized[, .(
+      Summary_Helas3.adj = sum(Helas3.adj, na.rm = T),
+      Summary_Hepg2.adj = sum(Hepg2.adj, na.rm = T),
+      Summary_Width = unique(Width)
+    ), by = Peak]
+  
+  return(gr.summarized)
+}
+
+getcor <- function(gr, p) {
+  newgr <-
+    gr[order(-abs(log2(Summary_Hepg2.adj + 1) - log2(Summary_Helas3.adj + 1))), ][1:round(.N *
+                                                                                            p), ]
+  newgr.UP = newgr[(Summary_Hepg2.adj >= Summary_Helas3.adj), ]
+  newgr.DOWN = newgr[(Summary_Hepg2.adj < Summary_Helas3.adj), ]
+  
+  return(data.table(
+    Up.LFC = ifelse(newgr.UP[, .N] > 1, newgr.UP[, median(log2((Summary_Hepg2.adj +
+                                                                  1) / (Summary_Helas3.adj + 1)))], NA),
[…]

### 2021-02-19 Supp/Figure_Narrow_750bp/Figure_Memory_Time.R

@@ -1,0 +1,174 @@
+# Libraries
+
+library(data.table)
+library(Polychrome)
+library(magrittr)
+library(ggplot2)
+library(microbenchmark)
+
+# Setup
+
+colors = c(kelly.colors(22)[c('lightblue',
+                              'red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'buff',
+                              'orange')])
+methods <-
+  c(
+    "epigraHMM",
+    "ChIPComp +\nMACS2",
+    "csaw",
+    "DiffBind +\nMACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+names(colors) <- methods
+
+################################################################################
+### Memory
+################################################################################
+
+files.csv <-
+  list.files('../../Public',
+             'slurm*.*.csv$',
+             recursive = TRUE,
+             full.names = TRUE)
+
+# Filtering
+
+files.csv <- files.csv[grep('Encode_twocells|MACS2', files.csv)]
+
+# Looping through files
+
+mem.dt <- rbindlist(lapply(strsplit(files.csv, '/'), function(x) {
+  # Does not run for MACS2
+  if (length(grep('MACS2', x)) == 0) {
+    subx <- strsplit(x[7], '_|bp')[[1]]
+    
+    csvx <- fread(paste(x, collapse = '/'))[2, ]$MaxRSS
+    
+    # Do not proceed if job is not completed
+    if (!is.na(csvx)) {
+      # If ChIPComp or DiffBind, need to check MACS2
+      if (x[4] %in% c('ChIPComp', 'DiffBind')) {
+        csvmacs2 <-
+          rbindlist(lapply(files.csv[intersect(grep(paste0('MACS2/', x[5]), files.csv), grep(paste0('Helas3|Hepg2'), files.csv))], function(y) {
+            fread(y)[JobName == 'batch', ]
+          }))$MaxRSS
+        
+        mem <-
+          max(as.numeric(strsplit(csvx, 'K')[[1]]), max(as.numeric(unlist(
+            strsplit(as.character(csvmacs2), 'K')
+          ))))
+      } else{
+        mem <- as.numeric(strsplit(csvx, 'K')[[1]])
+      }
+      
+      return(data.table(
+        Method = x[4],
+        Mark = x[5],
+        Data = x[6],
+        BP = subx[5],
[…]

### 2021-02-19 Supp/Figure_Narrow_750bp/Figure_Narrow.R

@@ -1,0 +1,22 @@
+library(ggpubr)
+
+load('Figure_Correlation_CTCF_Encode_twocells_750bp.RData')
+load('Figure_Correlation_H3K4me3_Encode_twocells_750bp.RData')
+load('Figure_PeakCalls_CTCF_Encode_twocells_750bp.RData')
+load('Figure_PeakCalls_H3K4me3_Encode_twocells_750bp.RData')
+load('Figure_PeakCalls_H3K27ac_Encode_twocells_750bp.RData')
+load('Figure_Memory_Time.RData')
+
+# Fig Peak Calls
+fig.calls = ggarrange(fig.ctcf,fig.h3k4me3,fig.h3k27ac,nrow = 3,ncol = 1,legend = 'none',labels = list('B','D','F'))
+
+# Figure Metrics
+fig.metrics = ggarrange(fig.lfc,fig.cor,fig_memory_time,ncol=1,nrow=3,legend = 'none',labels = list('A','C','E'))
+
+# Put them all together
+
+fig.final = ggarrange(fig.metrics,fig.calls,nrow=1,ncol=2)
+fig.final.label = ggarrange(get_legend(fig.lfc + theme(legend.position = 'top',legend.direction = 'horizontal',legend.box = 'vertical',legend.spacing.y = unit(-0.25, "cm")) + guides(col = guide_legend(nrow = 1,order = 1),linetype = guide_legend(nrow = 1,order = 2,override.aes = list(col = 'black'),title = 'Observed Enrichment'))),
+                            fig.final,nrow=2,heights = c(0.05,0.95))
+
+ggsave(filename = 'Figure_Narrow.pdf',plot = fig.final.label,height = 11.5,width = 9,dpi = 'retina')

### 2021-02-19 Supp/Figure_Narrow_750bp/Figure_PeakCalls_CTCF_Encode_twocells_750bp.R

@@ -1,0 +1,434 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 750
+chromosome = 'chr8'
+ngroups = 2
+mark = 'CTCF'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3CtcfStdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3CtcfStdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2CtcfStdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
[…]

### 2021-02-19 Supp/Figure_Narrow_750bp/Figure_PeakCalls_H3K27ac_Encode_twocells_750bp.R

@@ -1,0 +1,434 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 750
+chromosome = 'chr19'
+ngroups = 2
+mark = 'H3K27ac'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k27acStdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k27acStdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2H3k27acStdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
[…]

### 2021-02-19 Supp/Figure_Narrow_750bp/Figure_PeakCalls_H3K4me3_Encode_twocells_750bp.R

@@ -1,0 +1,432 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 750
+chromosome = 'chr10'
+ngroups = 2
+mark = 'H3K4me3'
+data = 'Encode_twocells'
+size = 1.25
+fdr = 0.05
+methods <-
+  c(
+    'Genes',
+    "epigraHMM",
+    "ChIPComp + MACS2",
+    "csaw",
+    "DiffBind + MACS2",
+    "diffReps",
+    'RSEG',
+    'THOR'
+  )
+colors = c('black', kelly.colors(22)[c('lightblue',
+                                       'red',
+                                       'yellow',
+                                       'purplishpink',
+                                       'yellowgreen',
+                                       'buff',
+                                       'orange')])
+names(colors) <- methods
+
+peakpos = c(1.1, seq(1.2, by = 0.1, length.out = 7))
+names(peakpos) <- methods
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k4me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3H3k4me3StdAlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2H3k4me3StdAlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
[…]

### 2021-02-19 Supp/Figure_Normalization/Figure_Normalization.R

@@ -1,0 +1,148 @@
+#This creates a figure that shows MA plots and adjusted read counts
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(edgeR)
+library(epigraHMM)
+
+bp = 500
+chromosome = 'chr19'
+ngroups=3
+chipnames = c('h1hesc_rep1','h1hesc_rep2','helas3_rep1','helas3_rep2','hepg2_rep1','hepg2_rep2')
+
+# Loading data
+
+## Group 1: H1hesc
+load('../../Data/Encode_h1hesc/H3K36me3/wgEncodeBroadHistoneH1hescH3k36me3StdAlnRep1.markdup.q10.sorted.RData')
+h1hesc_1 = subset(counts[[paste0(bp)]],chr==chromosome)$counts;rm(counts)
+load('../../Data/Encode_h1hesc/H3K36me3/wgEncodeBroadHistoneH1hescH3k36me3StdAlnRep1.markdup.q10.sorted.RData')
+h1hesc_2 = subset(counts[[paste0(bp)]],chr==chromosome)$counts;rm(counts)
+
+## Group 2: Helas3
+load('../../Data/Encode_helas3/H3K36me3/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep1.markdup.q10.sorted.RData')
+helas3_1 = subset(counts[[paste0(bp)]],chr==chromosome)$counts;rm(counts)
+load('../../Data/Encode_helas3/H3K36me3/wgEncodeBroadHistoneHelas3H3k36me3StdAlnRep2.markdup.q10.sorted.RData')
+helas3_2 = subset(counts[[paste0(bp)]],chr==chromosome)$counts;rm(counts)
+
+## Group 2: Hepg2
+load('../../Data/Encode_hepg2/H3K36me3/wgEncodeBroadHistoneHepg2H3k36me3StdAlnRep1.markdup.q10.sorted.RData')
+hepg2_1 = subset(counts[[paste0(bp)]],chr==chromosome)$counts;rm(counts)
+load('../../Data/Encode_hepg2/H3K36me3/wgEncodeBroadHistoneHepg2H3k36me3StdAlnRep2.markdup.q10.sorted.RData')
+hepg2_2 = subset(counts[[paste0(bp)]],chr==chromosome)$counts
+
[…]

### 2021-02-19 Supp/Figure_Overview_SequencingReads/Figure_SequencingReads_Example.R

@@ -1,0 +1,118 @@
+library(data.table)
+library(ggplot2)
+library(Polychrome)
+library(magrittr)
+library(forcats)
+library(plyr)
+library(GenomicRanges)
+library(ggpubr)
+
+size = 1.25
+cutoff <- 0.05
+methods <- c("epigraHMM","ChIPComp + HOMER","csaw", "DiffBind + HOMER","diffReps",'RSEG','THOR')
+heights = 1+seq(0.15,by = 0.15,length.out = 7)
+names(heights) = c(methods)
+colors = c(kelly.colors(22)[c('lightblue','red','yellow','purplishpink','yellowgreen','buff','orange')])
+names(colors) <- methods 
+
+gr.true = with(read.table('../../Sim/SequencingReads/autosim1//hist1_log.txt',header=T),GRanges(chr,IRanges(start,end)))
+
+gr.tile = with(fread('../../Sim/SequencingReads/autosim1/results/epigraHMM_ranges.txt'),GRanges(seqnames,IRanges(start,end)))
+
+dt.tile = data.table(chr='chrA',start=start(gr.tile),
+                     stop=end(gr.tile),
+                     window = 1:length(start(gr.tile)),
+                     counts1=bamsignals::bamCount(bampath='../../Sim/SequencingReads/autosim1/hist1_out_1.bam',gr=gr.tile,verbose=T,shift=0),
+                     counts2=bamsignals::bamCount(bampath='../../Sim/SequencingReads/autosim1/hist1_out_2.bam',gr=gr.tile,verbose=T,shift=0),
+                     counts3=bamsignals::bamCount(bampath='../../Sim/SequencingReads/autosim1/hist1_out_3.bam',gr=gr.tile,verbose=T,shift=0),
+                     counts4=bamsignals::bamCount(bampath='../../Sim/SequencingReads/autosim1/hist1_out_4.bam',gr=gr.tile,verbose=T,shift=0),
+                     control1=bamsignals::bamCount(bampath='../../Sim/SequencingReads/autosim1/control1_out_1.bam',gr=gr.tile,verbose=T,shift=0),
+                     control2=bamsignals::bamCount(bampath='../../Sim/SequencingReads/autosim1/control1_out_2.bam',gr=gr.tile,verbose=T,shift=0),
[…]

### 2021-02-19 Supp/Figure_Overview_SequencingReads/Figure_Simulation_SequencingReads_1.R

@@ -1,0 +1,501 @@
+library(data.table)
+library(ggplot2)
+library(Polychrome)
+library(magrittr)
+library(forcats)
+library(plyr)
+library(GenomicRanges)
+library(ggpubr)
+
+# This plots the specificity and sensitivity values relevant to a given *.tsv file.
+
+prefix = 'hist'
+incoming = paste0('../../Sim/SequencingReads/', prefix, "_result.tsv")
+methods <-
+  c("epigraHMM",
+    "ChIPComp + HOMER",
+    "csaw",
+    "DiffBind + HOMER",
+    "diffReps",
+    'RSEG',
+    'THOR')
+colors = c(kelly.colors(22)[c('lightblue',
+                              'red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'buff',
+                              'orange')])
+names(colors) <- methods
+
+cutoff <- 0.01
+sizeanno <- 2.5
+units = 1e3
+heights = 1 + seq(0.15, by = 0.15, length.out = 7)
+names(heights) = c(methods)
+size = 1.25
+
+# Loading in the data.
+ided <- fread(incoming)
+
+# How many times RSEG failed out of 100 simulated data sets?
+ided[Method == 'RSEG', .(Error = sum(Error == 1)), by = c('It')][, sum(Error >
+                                                                         0)] #26
+
+# Out of the number of times that RSEG did not fail, how many times it called the entire genome as differential?
+ided[Method == 'RSEG' &
+       Error == 0, ][, sum(CallSize > 1e6), by = 'Cutoff'] #3
+
+# Plotting Precision/Recall curve
+
+summarized.roc.ided <- ided[, .(
+  TPR = mean(TPR, na.rm = T),
+  FPR = mean(FPR, na.rm = T),
+  PPV = mean(PPV, na.rm = T)
+), by = c('Method', 'Cutoff')]
+summarized.roc.ided$Cutoff %<>% as.character()
+
+figA = ggplot(data = summarized.roc.ided, aes(x = (1 - PPV), y = TPR)) +
+  geom_line(aes(color = Method), size = 1.25) +
+  geom_point(aes(
+    x = (1 - PPV),
+    y = TPR,
+    fill = Method,
[…]

### 2021-02-19 Supp/Figure_Overview_SequencingReads/Figure_Simulation_SequencingReads_10.R

@@ -1,0 +1,501 @@
+library(data.table)
+library(ggplot2)
+library(Polychrome)
+library(magrittr)
+library(forcats)
+library(plyr)
+library(GenomicRanges)
+library(ggpubr)
+
+# This plots the specificity and sensitivity values relevant to a given *.tsv file.
+
+prefix = 'hist'
+incoming = paste0('../../Sim/SequencingReads/', prefix, "_result.tsv")
+methods <-
+  c("epigraHMM",
+    "ChIPComp + HOMER",
+    "csaw",
+    "DiffBind + HOMER",
+    "diffReps",
+    'RSEG',
+    'THOR')
+colors = c(kelly.colors(22)[c('lightblue',
+                              'red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'buff',
+                              'orange')])
+names(colors) <- methods
+
+cutoff <- 0.10
+sizeanno <- 2.5
+units = 1e3
+heights = 1 + seq(0.15, by = 0.15, length.out = 7)
+names(heights) = c(methods)
+size = 1.25
+
+# Loading in the data.
+ided <- fread(incoming)
+
+# How many times RSEG failed out of 100 simulated data sets?
+ided[Method == 'RSEG', .(Error = sum(Error == 1)), by = c('It')][, sum(Error >
+                                                                         0)] #26
+
+# Out of the number of times that RSEG did not fail, how many times it called the entire genome as differential?
+ided[Method == 'RSEG' &
+       Error == 0, ][, sum(CallSize > 1e6), by = 'Cutoff'] #3
+
+# Plotting Precision/Recall curve
+
+summarized.roc.ided <- ided[, .(
+  TPR = mean(TPR, na.rm = T),
+  FPR = mean(FPR, na.rm = T),
+  PPV = mean(PPV, na.rm = T)
+), by = c('Method', 'Cutoff')]
+summarized.roc.ided$Cutoff %<>% as.character()
+
+figA = ggplot(data = summarized.roc.ided, aes(x = (1 - PPV), y = TPR)) +
+  geom_line(aes(color = Method), size = 1.25) +
+  geom_point(aes(
+    x = (1 - PPV),
+    y = TPR,
+    fill = Method,
[…]

### 2021-02-19 Supp/Figure_Overview_SequencingReads/Figure_Simulation_SequencingReads_15.R

@@ -1,0 +1,501 @@
+library(data.table)
+library(ggplot2)
+library(Polychrome)
+library(magrittr)
+library(forcats)
+library(plyr)
+library(GenomicRanges)
+library(ggpubr)
+
+# This plots the specificity and sensitivity values relevant to a given *.tsv file.
+
+prefix = 'hist'
+incoming = paste0('../../Sim/SequencingReads/', prefix, "_result.tsv")
+methods <-
+  c("epigraHMM",
+    "ChIPComp + HOMER",
+    "csaw",
+    "DiffBind + HOMER",
+    "diffReps",
+    'RSEG',
+    'THOR')
+colors = c(kelly.colors(22)[c('lightblue',
+                              'red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'buff',
+                              'orange')])
+names(colors) <- methods
+
+cutoff <- 0.15
+sizeanno <- 2.5
+units = 1e3
+heights = 1 + seq(0.15, by = 0.15, length.out = 7)
+names(heights) = c(methods)
+size = 1.25
+
+# Loading in the data.
+ided <- fread(incoming)
+
+# How many times RSEG failed out of 100 simulated data sets?
+ided[Method == 'RSEG', .(Error = sum(Error == 1)), by = c('It')][, sum(Error >
+                                                                         0)] #26
+
+# Out of the number of times that RSEG did not fail, how many times it called the entire genome as differential?
+ided[Method == 'RSEG' &
+       Error == 0, ][, sum(CallSize > 1e6), by = 'Cutoff'] #3
+
+# Plotting Precision/Recall curve
+
+summarized.roc.ided <- ided[, .(
+  TPR = mean(TPR, na.rm = T),
+  FPR = mean(FPR, na.rm = T),
+  PPV = mean(PPV, na.rm = T)
+), by = c('Method', 'Cutoff')]
+summarized.roc.ided$Cutoff %<>% as.character()
+
+figA = ggplot(data = summarized.roc.ided, aes(x = (1 - PPV), y = TPR)) +
+  geom_line(aes(color = Method), size = 1.25) +
+  geom_point(aes(
+    x = (1 - PPV),
+    y = TPR,
+    fill = Method,
[…]

### 2021-02-19 Supp/Figure_Overview_SequencingReads/Figure_Simulation_SequencingReads_20.R

@@ -1,0 +1,501 @@
+library(data.table)
+library(ggplot2)
+library(Polychrome)
+library(magrittr)
+library(forcats)
+library(plyr)
+library(GenomicRanges)
+library(ggpubr)
+
+# This plots the specificity and sensitivity values relevant to a given *.tsv file.
+
+prefix = 'hist'
+incoming = paste0('../../Sim/SequencingReads/', prefix, "_result.tsv")
+methods <-
+  c("epigraHMM",
+    "ChIPComp + HOMER",
+    "csaw",
+    "DiffBind + HOMER",
+    "diffReps",
+    'RSEG',
+    'THOR')
+colors = c(kelly.colors(22)[c('lightblue',
+                              'red',
+                              'yellow',
+                              'purplishpink',
+                              'yellowgreen',
+                              'buff',
+                              'orange')])
+names(colors) <- methods
+
+cutoff <- 0.20
+sizeanno <- 2.5
+units = 1e3
+heights = 1 + seq(0.15, by = 0.15, length.out = 7)
+names(heights) = c(methods)
+size = 1.25
+
+# Loading in the data.
+ided <- fread(incoming)
+
+# How many times RSEG failed out of 100 simulated data sets?
+ided[Method == 'RSEG', .(Error = sum(Error == 1)), by = c('It')][, sum(Error >
+                                                                         0)] #26
+
+# Out of the number of times that RSEG did not fail, how many times it called the entire genome as differential?
+ided[Method == 'RSEG' &
+       Error == 0, ][, sum(CallSize > 1e6), by = 'Cutoff'] #3
+
+# Plotting Precision/Recall curve
+
+summarized.roc.ided <- ided[, .(
+  TPR = mean(TPR, na.rm = T),
+  FPR = mean(FPR, na.rm = T),
+  PPV = mean(PPV, na.rm = T)
+), by = c('Method', 'Cutoff')]
+summarized.roc.ided$Cutoff %<>% as.character()
+
+figA = ggplot(data = summarized.roc.ided, aes(x = (1 - PPV), y = TPR)) +
+  geom_line(aes(color = Method), size = 1.25) +
+  geom_point(aes(
+    x = (1 - PPV),
+    y = TPR,
+    fill = Method,
[…]

### 2021-02-19 Supp/Figure_PR_LFC/Figure_PR_H3K36me3_Encode_twocells_500bp_10LFC.R

@@ -12,12 +12,12 @@ library(Polychrome)
 library(ggrepel)
 
 # General parameters
-bp = 750
+bp = 500
 sizeanno <- 2.5
 chromosome = paste0('chr', 1:22)
 mark = 'H3K36me3'
 data = 'Encode_twocells'
-cutoff = 2
+cutoff = 1
 methods <-
   c(
     "ChIPComp + MACS2",
@@ -208,16 +208,6 @@ gr.epigrahmm = rowRanges(get(paste(
 )))
 gr.epigrahmm$FDR = pp.epigrahmm[, 2]
 
-# epigraHMM + Viterbi
-viterbi <-
-  rhdf5::h5read(metadata(get(
-    paste('epigraHMM', mark, data, 'Output', paste0(bp, 'bp'), sep = '_')
-  ))$output, "viterbi")[, 1]
-gr.viterbi <-
-  rowRanges(get(paste(
-    'epigraHMM', mark, data, 'Output', paste0(bp, 'bp'), sep = '_'
-  )))[viterbi == 1]
-
 # Loading peaks: csaw
 csaw <-
   fread(list.files(
@@ -256,16 +246,16 @@ gr.rseg <-
 gr.rseg <-
   gr.rseg[seqnames(gr.rseg) %in% chromosome] # Selecting only chromosomes of interest
 
-# # Loading peaks: diffReps (it failed)
-# diffreps = fread(list.files(
-#   file.path('../../Public/diffReps', mark, data, paste0('Output', bp)),
-#   '.annotated',
-#   full.names = TRUE
-# ),
-# header = T)
-# gr.diffreps = with(diffreps, GenomicRanges::GRanges(Chrom, IRanges(Start, End)))
-# gr.diffreps$FDR = diffreps$padj #It already gives me adjusted p-values
-# gr.diffreps <- gr.diffreps[seqnames(gr.diffreps) %in% chromosome]
+# # Loading peaks: diffReps
+diffreps = fread(list.files(
+  file.path('../../Public/diffReps', mark, data, paste0('Output', bp)),
+  '.annotated',
+  full.names = TRUE
+),
+header = T)
+gr.diffreps = with(diffreps, GenomicRanges::GRanges(Chrom, IRanges(Start, End)))
+gr.diffreps$FDR = diffreps$padj #It already gives me adjusted p-values
[…]

### 2021-02-19 Supp/Figure_PR_LFC/Figure_PR_H3K36me3_Encode_twocells_500bp_15LFC.R

@@ -12,12 +12,12 @@ library(Polychrome)
 library(ggrepel)
 
 # General parameters
-bp = 750
+bp = 500
 sizeanno <- 2.5
 chromosome = paste0('chr', 1:22)
 mark = 'H3K36me3'
 data = 'Encode_twocells'
-cutoff = 2
+cutoff = 1.5
 methods <-
   c(
     "ChIPComp + MACS2",
@@ -208,16 +208,6 @@ gr.epigrahmm = rowRanges(get(paste(
 )))
 gr.epigrahmm$FDR = pp.epigrahmm[, 2]
 
-# epigraHMM + Viterbi
-viterbi <-
-  rhdf5::h5read(metadata(get(
-    paste('epigraHMM', mark, data, 'Output', paste0(bp, 'bp'), sep = '_')
-  ))$output, "viterbi")[, 1]
-gr.viterbi <-
-  rowRanges(get(paste(
-    'epigraHMM', mark, data, 'Output', paste0(bp, 'bp'), sep = '_'
-  )))[viterbi == 1]
-
 # Loading peaks: csaw
 csaw <-
   fread(list.files(
@@ -256,16 +246,16 @@ gr.rseg <-
 gr.rseg <-
   gr.rseg[seqnames(gr.rseg) %in% chromosome] # Selecting only chromosomes of interest
 
-# # Loading peaks: diffReps (it failed)
-# diffreps = fread(list.files(
-#   file.path('../../Public/diffReps', mark, data, paste0('Output', bp)),
-#   '.annotated',
-#   full.names = TRUE
-# ),
-# header = T)
-# gr.diffreps = with(diffreps, GenomicRanges::GRanges(Chrom, IRanges(Start, End)))
-# gr.diffreps$FDR = diffreps$padj #It already gives me adjusted p-values
-# gr.diffreps <- gr.diffreps[seqnames(gr.diffreps) %in% chromosome]
+# # Loading peaks: diffReps
+diffreps = fread(list.files(
+  file.path('../../Public/diffReps', mark, data, paste0('Output', bp)),
+  '.annotated',
+  full.names = TRUE
+),
+header = T)
+gr.diffreps = with(diffreps, GenomicRanges::GRanges(Chrom, IRanges(Start, End)))
+gr.diffreps$FDR = diffreps$padj #It already gives me adjusted p-values
[…]

### 2021-02-19 Supp/Figure_PR_LFC/Figure_PR_H3K36me3_Encode_twocells_500bp_25LFC.R

@@ -12,12 +12,12 @@ library(Polychrome)
 library(ggrepel)
 
 # General parameters
-bp = 750
+bp = 500
 sizeanno <- 2.5
 chromosome = paste0('chr', 1:22)
 mark = 'H3K36me3'
 data = 'Encode_twocells'
-cutoff = 2
+cutoff = 2.5
 methods <-
   c(
     "ChIPComp + MACS2",
@@ -208,16 +208,6 @@ gr.epigrahmm = rowRanges(get(paste(
 )))
 gr.epigrahmm$FDR = pp.epigrahmm[, 2]
 
-# epigraHMM + Viterbi
-viterbi <-
-  rhdf5::h5read(metadata(get(
-    paste('epigraHMM', mark, data, 'Output', paste0(bp, 'bp'), sep = '_')
-  ))$output, "viterbi")[, 1]
-gr.viterbi <-
-  rowRanges(get(paste(
-    'epigraHMM', mark, data, 'Output', paste0(bp, 'bp'), sep = '_'
-  )))[viterbi == 1]
-
 # Loading peaks: csaw
 csaw <-
   fread(list.files(
@@ -256,16 +246,16 @@ gr.rseg <-
 gr.rseg <-
   gr.rseg[seqnames(gr.rseg) %in% chromosome] # Selecting only chromosomes of interest
 
-# # Loading peaks: diffReps (it failed)
-# diffreps = fread(list.files(
-#   file.path('../../Public/diffReps', mark, data, paste0('Output', bp)),
-#   '.annotated',
-#   full.names = TRUE
-# ),
-# header = T)
-# gr.diffreps = with(diffreps, GenomicRanges::GRanges(Chrom, IRanges(Start, End)))
-# gr.diffreps$FDR = diffreps$padj #It already gives me adjusted p-values
-# gr.diffreps <- gr.diffreps[seqnames(gr.diffreps) %in% chromosome]
+# # Loading peaks: diffReps
+diffreps = fread(list.files(
+  file.path('../../Public/diffReps', mark, data, paste0('Output', bp)),
+  '.annotated',
+  full.names = TRUE
+),
+header = T)
+gr.diffreps = with(diffreps, GenomicRanges::GRanges(Chrom, IRanges(Start, End)))
+gr.diffreps$FDR = diffreps$padj #It already gives me adjusted p-values
[…]

### 2021-02-19 Supp/Figure_PR_LFC/Figure_PR_H3K36me3_Encode_twocells_500bp_30LFC.R

@@ -12,12 +12,12 @@ library(Polychrome)
 library(ggrepel)
 
 # General parameters
-bp = 750
+bp = 500
 sizeanno <- 2.5
 chromosome = paste0('chr', 1:22)
 mark = 'H3K36me3'
 data = 'Encode_twocells'
-cutoff = 2
+cutoff = 3
 methods <-
   c(
     "ChIPComp + MACS2",
@@ -208,16 +208,6 @@ gr.epigrahmm = rowRanges(get(paste(
 )))
 gr.epigrahmm$FDR = pp.epigrahmm[, 2]
 
-# epigraHMM + Viterbi
-viterbi <-
-  rhdf5::h5read(metadata(get(
-    paste('epigraHMM', mark, data, 'Output', paste0(bp, 'bp'), sep = '_')
-  ))$output, "viterbi")[, 1]
-gr.viterbi <-
-  rowRanges(get(paste(
-    'epigraHMM', mark, data, 'Output', paste0(bp, 'bp'), sep = '_'
-  )))[viterbi == 1]
-
 # Loading peaks: csaw
 csaw <-
   fread(list.files(
@@ -256,16 +246,16 @@ gr.rseg <-
 gr.rseg <-
   gr.rseg[seqnames(gr.rseg) %in% chromosome] # Selecting only chromosomes of interest
 
-# # Loading peaks: diffReps (it failed)
-# diffreps = fread(list.files(
-#   file.path('../../Public/diffReps', mark, data, paste0('Output', bp)),
-#   '.annotated',
-#   full.names = TRUE
-# ),
-# header = T)
-# gr.diffreps = with(diffreps, GenomicRanges::GRanges(Chrom, IRanges(Start, End)))
-# gr.diffreps$FDR = diffreps$padj #It already gives me adjusted p-values
-# gr.diffreps <- gr.diffreps[seqnames(gr.diffreps) %in% chromosome]
+# # Loading peaks: diffReps
+diffreps = fread(list.files(
+  file.path('../../Public/diffReps', mark, data, paste0('Output', bp)),
+  '.annotated',
+  full.names = TRUE
+),
+header = T)
+gr.diffreps = with(diffreps, GenomicRanges::GRanges(Chrom, IRanges(Start, End)))
+gr.diffreps$FDR = diffreps$padj #It already gives me adjusted p-values
[…]

### 2021-02-19 Supp/Figure_PR_LFC/Figure_PR_LFC.R

@@ -1,0 +1,12 @@
+load('Figure_PR_H3K36me3_Encode_twocells_500bp_10LFC.RData')
+load('Figure_PR_H3K36me3_Encode_twocells_500bp_15LFC.RData')
+load('Figure_PR_H3K36me3_Encode_twocells_500bp_25LFC.RData')
+load('Figure_PR_H3K36me3_Encode_twocells_500bp_30LFC.RData')
+
+fig <- ggarrange(fig_lfc10,
+                 fig_lfc15,
+                 fig_lfc25,
+                 fig_lfc30,
+                 nrow = 2,ncol = 2,legend = 'top',common.legend = TRUE)
+
+ggsave(fig,filename = 'Figure_PR_LFC.pdf',width = 9,height = 11.5)

### 2021-02-19 Supp/Figure_PR_Viterbi/Figure_PR_Viterbi.R

@@ -1,0 +1,9 @@
+load('Figure_PR_H3K36me3_Encode_twocells_250bp.RData')
+load('Figure_PR_H3K36me3_Encode_twocells_500bp.RData')
+load('Figure_PR_H3K36me3_Encode_twocells_750bp.RData')
+load('Figure_PR_H3K36me3_Encode_twocells_1000bp.RData')
+
+fig <- ggarrange(fig.ROC_250,fig.ROC_500,fig.ROC_750,fig.ROC_1000,
+                 nrow = 2,ncol = 2,legend = 'top',common.legend = TRUE)
+
+ggsave(fig,filename = 'Figure_PR_Viterbi.pdf',width = 9,height = 11.5)

### 2021-02-19 Supp/Figure_ROC_ReadCounts/Figure_ROC_ReadCounts_H3K27me3_1e+06.R

@@ -1,0 +1,196 @@
+library(data.table)
+library(ggplot2)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+
+mark = 'H3K27me3'
+size = 1e6
+
+### Simulation scnearios
+# Marker:   H3K27me3    H3K36me3
+# Groups:   2, 3, 4
+# Replicates: 1, 2, 4
+# Nwindow: 1e5, 5e5, 1e6
+# Nsim: 100
+
+scenario = expand.grid(Marker=c('H3K27me3','H3K36me3'),Groups=c(2,3,4),Replicates=c(1,2,4),Nwindow=c(1e5,5e5,1e6),Nsim=100,Pct=round(100*c(0.7,0.8,0.9,1)))
+scenario$Label = with(scenario,paste(Marker,paste0(Groups,'G'),paste0(Replicates,'R'),paste0(Nwindow,'W'),paste0(Pct,'SNR'),sep="_"))
+scenario <- scenario[order(scenario$Groups,scenario$Replicates,scenario$Nwindow,-scenario$Pct),]
+# print(scenario)
+
+### Loading FDRpp
+
+pct = c(70,80,90,100)
+
+fdrpp.lst = list()
+for(i in pct){
+  outRData <-
+    paste0('../../Sim/ReadCounts/Metrics_',
+           i,
+           'SNR/',
+           'Metrics_',
+           i,
+           'SNR_fdrpp.RData')
+  if (file.exists(outRData)) {
+    load(outRData)
+    fdrpp.lst[[paste0(i)]] = fdrpp.list
+  }
+}
+fdrpp.lst <- unlist(fdrpp.lst,recursive = F)
+
+fdrpp <- rbindlist(lapply(fdrpp.lst,FUN = function(x){
+  y = rbindlist(x,idcol = 'rn') #as.data.table(do.call(rbind,x),keep.rownames = T)
+  y[,Simulation := as.numeric(gsub("Sim","",rn))][,rn := NULL]
+  return(y)
+}),idcol = 'rn')
+
+label <- as.data.table(do.call(rbind,strsplit(fdrpp$rn,"_")))
+setnames(label,c('Marker','Groups','Replicates','Nwindow','Pct'))
+label[,Groups := as.numeric(gsub("G","",Groups))]
+label[,Replicates := as.numeric(gsub("R","",Replicates))]
+label[,Nwindow := as.numeric(gsub("W","",Nwindow))]
[…]

### 2021-02-19 Supp/Figure_ROC_ReadCounts/Figure_ROC_ReadCounts_H3K36me3_1e+06.R

@@ -1,0 +1,196 @@
+library(data.table)
+library(ggplot2)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+
+mark = 'H3K36me3'
+size = 1e6
+
+### Simulation scnearios
+# Marker:   H3K27me3    H3K36me3
+# Groups:   2, 3, 4
+# Replicates: 1, 2, 4
+# Nwindow: 1e5, 5e5, 1e6
+# Nsim: 100
+
+scenario = expand.grid(Marker=c('H3K27me3','H3K36me3'),Groups=c(2,3,4),Replicates=c(1,2,4),Nwindow=c(1e5,5e5,1e6),Nsim=100,Pct=round(100*c(0.7,0.8,0.9,1)))
+scenario$Label = with(scenario,paste(Marker,paste0(Groups,'G'),paste0(Replicates,'R'),paste0(Nwindow,'W'),paste0(Pct,'SNR'),sep="_"))
+scenario <- scenario[order(scenario$Groups,scenario$Replicates,scenario$Nwindow,-scenario$Pct),]
+# print(scenario)
+
+### Loading FDRpp
+
+pct = c(70,80,90,100)
+
+fdrpp.lst = list()
+for(i in pct){
+  outRData <-
+    paste0('../../Sim/ReadCounts/Metrics_',
+           i,
+           'SNR/',
+           'Metrics_',
+           i,
+           'SNR_fdrpp.RData')
+  if (file.exists(outRData)) {
+    load(outRData)
+    fdrpp.lst[[paste0(i)]] = fdrpp.list
+  }
+}
+fdrpp.lst <- unlist(fdrpp.lst,recursive = F)
+
+fdrpp <- rbindlist(lapply(fdrpp.lst,FUN = function(x){
+  y = rbindlist(x,idcol = 'rn') #as.data.table(do.call(rbind,x),keep.rownames = T)
+  y[,Simulation := as.numeric(gsub("Sim","",rn))][,rn := NULL]
+  return(y)
+}),idcol = 'rn')
+
+label <- as.data.table(do.call(rbind,strsplit(fdrpp$rn,"_")))
+setnames(label,c('Marker','Groups','Replicates','Nwindow','Pct'))
+label[,Groups := as.numeric(gsub("G","",Groups))]
+label[,Replicates := as.numeric(gsub("R","",Replicates))]
+label[,Nwindow := as.numeric(gsub("W","",Nwindow))]
[…]

### 2021-02-19 Supp/Figure_Ternary/Figure_Ternary_H3K27ac.R

@@ -1,0 +1,187 @@
+library(magrittr)
+library(data.table)
+library(tidyverse)
+library(tibble)
+library(RColorBrewer)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(vsn)
+library(DESeq2)
+library(ggtern)
+
+# General parameters
+bp = 250
+chromosome = paste0('chr',1:22)
+mark = 'H3K27ac'
+data = 'Encode_threecells'
+top = 1
+fdr = 0.05
+
+Modes <- function(x) {
+  ux <- unique(x)
+  tab <- tabulate(match(x, ux))
+  if(length(ux[tab == max(tab)])==1){
+    return(ux[tab == max(tab)])
+  } else{
+    out = ux[tab == max(tab)]
+    return(max(out))
+  }
+}
+
+agg = function(x,gr.peak,gr.genome,cells=c('Helas3.adj','Hepg2.adj','Huvec.adj')){
+  gr.summarized = as.data.frame(GenomicRanges::findOverlaps(gr.peak,gr.genome))
+  gr.summarized = as.data.table(cbind(gr.summarized,x[gr.summarized$subjectHits,cells,with=F]));colnames(gr.summarized)=c('Peak','Genome',cells)
+  gr.summarized[,Width :=  width(gr.peak[Peak])]
+  gr.summarized <- gr.summarized[,.(Summary_Helas3 = sum(Helas3.adj,na.rm = T),
+                                    Summary_Hepg2 = sum(Hepg2.adj,na.rm = T),
+                                    Summary_Huvec = sum(Huvec.adj,na.rm = T),
+                                    Summary_Width = unique(Width)),by=Peak]
+  return(gr.summarized)
+}
+
+pat.peaks = function(x,gr.peaks,gr.genome){
+  ovp = findOverlaps(gr.peaks,gr.genome)
+  out = rbindlist(lapply(unique(queryHits(ovp)), FUN = function(i){data.table(Peak = i,Mode = Modes(x[subjectHits(ovp)[queryHits(ovp)==i],]$Pattern))}))
+  Pattern = rep(NA,length(gr.peaks))
+  Pattern[out$Peak] = out$Mode
+  return(Pattern)
+}
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
[…]

### 2021-02-19 Supp/Figure_Ternary/Figure_Ternary_H3K4me3.R

@@ -1,0 +1,187 @@
+library(magrittr)
+library(data.table)
+library(tidyverse)
+library(tibble)
+library(RColorBrewer)
+library(GenomicRanges)
+library(SummarizedExperiment)
+library(vsn)
+library(DESeq2)
+library(ggtern)
+
+# General parameters
+bp = 250
+chromosome = paste0('chr',1:22)
+mark = 'H3K4me3'
+data = 'Encode_threecells'
+top = 1
+fdr = 0.05
+
+Modes <- function(x) {
+  ux <- unique(x)
+  tab <- tabulate(match(x, ux))
+  if(length(ux[tab == max(tab)])==1){
+    return(ux[tab == max(tab)])
+  } else{
+    out = ux[tab == max(tab)]
+    return(max(out))
+  }
+}
+
+agg = function(x,gr.peak,gr.genome,cells=c('Helas3.adj','Hepg2.adj','Huvec.adj')){
+  gr.summarized = as.data.frame(GenomicRanges::findOverlaps(gr.peak,gr.genome))
+  gr.summarized = as.data.table(cbind(gr.summarized,x[gr.summarized$subjectHits,cells,with=F]));colnames(gr.summarized)=c('Peak','Genome',cells)
+  gr.summarized[,Width :=  width(gr.peak[Peak])]
+  gr.summarized <- gr.summarized[,.(Summary_Helas3 = sum(Helas3.adj,na.rm = T),
+                                    Summary_Hepg2 = sum(Hepg2.adj,na.rm = T),
+                                    Summary_Huvec = sum(Huvec.adj,na.rm = T),
+                                    Summary_Width = unique(Width)),by=Peak]
+  return(gr.summarized)
+}
+
+pat.peaks = function(x,gr.peaks,gr.genome){
+  ovp = findOverlaps(gr.peaks,gr.genome)
+  out = rbindlist(lapply(unique(queryHits(ovp)), FUN = function(i){data.table(Peak = i,Mode = Modes(x[subjectHits(ovp)[queryHits(ovp)==i],]$Pattern))}))
+  Pattern = rep(NA,length(gr.peaks))
+  Pattern[out$Peak] = out$Mode
+  return(Pattern)
+}
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
[…]

### 2021-02-19 Supp/Table_Simulation_ReadCounts/Table_Simulation_ReadCounts_H3K27me3_1e+06.R

@@ -1,0 +1,294 @@
+# Table: simulation results
+
+library(data.table)
+library(ggplot2)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(RColorBrewer)
+library(kableExtra)
+
+mark = 'H3K27me3'
+size = 1e6
+
+### Simulation scnearios
+# Marker:   H3K27me3    H3K36me3
+# Groups:   2, 3, 4
+# Replicates: 1, 2, 4
+# Nwindow: 1e5, 5e5, 1e6
+# Nsim: 100
+
+scenario = expand.grid(
+  Marker = c('H3K27me3', 'H3K36me3'),
+  Groups = c(2, 3, 4),
+  Replicates = c(1, 2, 4),
+  Nwindow = c(1e5, 5e5, 1e6),
+  Nsim = 100,
+  Pct = round(100 * c(0.7, 0.8, 0.9, 1))
+)
+scenario$Label = with(scenario,
+                      paste(
+                        Marker,
+                        paste0(Groups, 'G'),
+                        paste0(Replicates, 'R'),
+                        paste0(Nwindow, 'W'),
+                        paste0(Pct, 'SNR'),
+                        sep = "_"
+                      ))
+scenario <-
+  scenario[order(scenario$Groups,
+                 scenario$Replicates,
+                 scenario$Nwindow,
+                 -scenario$Pct), ]
+# print(scenario)
+
+### True parameter functions
+
+genpar = function(Marker, Groups, pct) {
+  ## Initial probability (always begin with background) and Transition Matrix
+  pi = c(1, 0)
+  ## Number of states
+  N.states = 2 ^ Groups
+  if (Marker == 'H3K36me3') {
+    ## See 01_H3K36me3_Parameters.R
+    ## Transition Probability Matrix
+    d1 = 1 - 1 / (131.92967 + 1)
+    d2 = 1 - 1 / (43.75663 + 1)
+    gamma = matrix(0, N.states, N.states)
+    diag(gamma) = c(d1, rep(d2, N.states - 1))
+    gamma[1, 2:N.states] = (1 - d1) / (N.states - 1)
+    gamma[gamma == 0] = (1 - d2) / (N.states - 1)
+    
+    ## ChIP parameters
+    par.background = c(0.7560612, 0, 3.230474)
+    par.enrichment = c(2.4169519, 0, 2.891908)
[…]

### 2021-02-19 Supp/Table_Simulation_ReadCounts/Table_Simulation_ReadCounts_H3K36me3_1e+06.R

@@ -1,0 +1,294 @@
+# Table: simulation results
+
+library(data.table)
+library(ggplot2)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(RColorBrewer)
+library(kableExtra)
+
+mark = 'H3K36me3'
+size = 1e6
+
+### Simulation scnearios
+# Marker:   H3K27me3    H3K36me3
+# Groups:   2, 3, 4
+# Replicates: 1, 2, 4
+# Nwindow: 1e5, 5e5, 1e6
+# Nsim: 100
+
+scenario = expand.grid(
+  Marker = c('H3K27me3', 'H3K36me3'),
+  Groups = c(2, 3, 4),
+  Replicates = c(1, 2, 4),
+  Nwindow = c(1e5, 5e5, 1e6),
+  Nsim = 100,
+  Pct = round(100 * c(0.7, 0.8, 0.9, 1))
+)
+scenario$Label = with(scenario,
+                      paste(
+                        Marker,
+                        paste0(Groups, 'G'),
+                        paste0(Replicates, 'R'),
+                        paste0(Nwindow, 'W'),
+                        paste0(Pct, 'SNR'),
+                        sep = "_"
+                      ))
+scenario <-
+  scenario[order(scenario$Groups,
+                 scenario$Replicates,
+                 scenario$Nwindow,
+                 -scenario$Pct), ]
+# print(scenario)
+
+### True parameter functions
+
+genpar = function(Marker, Groups, pct) {
+  ## Initial probability (always begin with background) and Transition Matrix
+  pi = c(1, 0)
+  ## Number of states
+  N.states = 2 ^ Groups
+  if (Marker == 'H3K36me3') {
+    ## See 01_H3K36me3_Parameters.R
+    ## Transition Probability Matrix
+    d1 = 1 - 1 / (131.92967 + 1)
+    d2 = 1 - 1 / (43.75663 + 1)
+    gamma = matrix(0, N.states, N.states)
+    diag(gamma) = c(d1, rep(d2, N.states - 1))
+    gamma[1, 2:N.states] = (1 - d1) / (N.states - 1)
+    gamma[gamma == 0] = (1 - d2) / (N.states - 1)
+    
+    ## ChIP parameters
+    par.background = c(0.7560612, 0, 3.230474)
+    par.enrichment = c(2.4169519, 0, 2.891908)
[…]

### 2021-02-19 Latex/appendix/webappendix.Rnw
Updating tex

@@
 \end{table}
+
+\pagebreak
+\clearpage
 
@@
 
+\pagebreak
+\clearpage
+
 \subsection{GC-content bias in broad marks}
+
+\subsubsection{Simulation Study}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_Simulation_GC/Summary_Scatter_G113G233.png}}
+\end{center}
+\caption{Simulation study for GC-content bias correction with epigraHMM.
+In (A), example of a genomic region from the simulated data.
+In (B), simulated GC content effect on ChIP-seq counts.
+Scenario: low depth in condition 1, high depth in condition 2; positive GC effect on ChIP-seq counts in both conditions.
+\label{fig:sim_gc_scatter}}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_Simulation_GC/Summary_FDR_G113G233.png}}
+\end{center}
+\caption{Simulation study for GC-content bias correction with epigraHMM.
+Average observed true positive rate (y-axis) and observed false discovery rate (x-axis) in 100 simulated data sets for four different GC-content bias correction strategies: naive model without normalization for sequencing depth or GC-content bias, GC-content adjusted model, sequencing depth adjusted model, and GC-content and sequencing depth adjusted model.
+Optimal results are obtained only when the effect of GC-content is considered together with normalization for depth.
+The GC-content bias was accounted via model offsets in epigraHMM using the normalizing denominators from \cite{teng2017accounting}.
+\label{fig:sim_gc_fdr}}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_Simulation_GC/Summary_ROC_G113G233.png}}
+\end{center}
+\caption{Simulation study for GC-content bias correction with epigraHMM.
+Observed sensitivity (y-axis) and observed 1-specificity (x-axis) for each of the 100 simulated data sets four different GC-content bias correction strategies: naive model without normalization for sequencing depth or GC-content bias, GC-content adjusted model, sequencing depth adjusted model, and GC-content and sequencing depth adjusted model.
+Optimal results are obtained only when the effect of GC-content is considered together with normalization for depth.
+The GC-content bias was accounted via model offsets in epigraHMM using the normalizing denominators from \cite{teng2017accounting}.
+Dashed lines connect the observed metrics for each simulated dataset.
+\label{fig:sim_gc_roc}}
+\end{figure}
+
+\pagebreak
+\clearpage
 
@@
 
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_BIC/H3K36me3_3G_4R_1e+05W_100SNR_bic.png}}
+\end{center}
+\caption{Simulation study for model selection via BIC. Simulation scenario: H3K36me3 data, 3 conditions, 4 replicates per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR.
+\label{fig:bic_4rep}}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_BIC/H3K36me3_3G_2R_1e+05W_100SNR_bic.png}}
+\end{center}
+\caption{Simulation study for model selection via BIC. Simulation scenario: H3K36me3 data, 3 conditions, 2 replicates per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR.
+\label{fig:bic_2rep}}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_BIC/H3K36me3_3G_1R_1e+05W_100SNR_bic.png}}
+\end{center}
+\caption{Simulation study for model selection via BIC. Simulation scenario: H3K36me3 data, 3 conditions, 1 replicate per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR.
+\label{fig:bic_1rep}}
+\end{figure}
+
+\pagebreak
+\clearpage
+
 \section{Study on FDR Control and the Viterbi Algorithm}
@@
 \end{figure}
+
+\pagebreak
+\clearpage
 

### 2021-02-19 Latex/letter/letter.Rmd
Updating tex

---
title: "Response Letter for Manuscript ID BIOM2020344P - 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions'"
author: "Pedro L. Baldoni, Naim U. Rashid, Joseph G. Ibrahim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reviewer 1

**The authors faithfully referenced the relevant literature. However, it may be better to cite the recent review from a comprehensive comparison of tools for differential ChIP-seq analysis, and discuss the reason for selecting the candidate DPCs in the benchmark part.**

> We thank the reviewer for this constructive comment. We cite in our paper the benchmarking article 'A comprehensive comparison of tools for differential ChIP-seq analysis' (Briefins in Bioinformatics, 2016). However, we authors are not aware of a more recent comprehensive review of DPCs algorithms in the literature. As suggested we add the following sentence to our real data analysis section 'We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short data sets.' We believe that such a list covers the most used algorithms for DPC in thie field of bioinformatics.

**In the third challenge at "Third, the analysis of ChIP-seq data…", the author may also need to discuss the potential sequence GC content bias except for local total reads abundance bias, since GC bias may lead to false-positive peak call or differential peak call as mentioned in Accounting for GC-content bias reduces systematic errors and batch effects in ChIP-seq data.**

> We acknowledge the issue of GC-content bias in the referenced paragraph. We cite the work of Teng and Irizarry (2017) that proposed a correction method for GC-content bias in ChIP-seq analysis. Their method is applicable to any peak calling algorithm that measures the enrichment of reads in bins, such as our method. Later in the article, we point out to the reader that our method is able to account for GC-content bias in the same manner as suggested by Teng and Irizarry (2017) by using their count-normalizing denominators in our model as offsets. We describe this process in detail in Section B1.3 of the Web Appendix.

**The author may also need to consider citing the MACS2 GitHub page along with the Zhang et al.2008 since the first MACS version does not support broad peak calling.**

> This has been addressed. In the second paragraph of Section 5, we reference the current GitHub repository of MACS2 \url{https://github.com/macs3-project/MACS}.

**1. As mentioned before, GC bias is a nonlinear covariate that may lead to false-positive peak, mixNBHMM includes a GLM-based framework for adjusting local read abundance bias, could the mixNBHMM handle extra GC content in the bins, would correction of GC bias increase the specificity and sensitivity of mixNBHMM?**

> As mentioned above, our method is able to account go GC-bias via model offsets. These model offsets can be obtained by methods such as the one proposed by Teng and Irizarry (2017). We present a simulation study in Section B1.3 of the Web Appendix showing that epigraHMM leads to higher sensitivity/specificity when accounting for GC-bias even after read count normalization using the count-normalizing denominators from Teng and Irizarry (2017). However, for the broad marks that we analized in our manuscript, we did not observe a substantial effect of GC-content on read count distribution and did not observe any substantial improvement in accounting for GC-content in our model via offsets when utilizing the method from Teng and Irizarry (2017), a method that was developed only for narrow marks characterized by sharp peaks. In fact, the authors do not recommend the use of their method when analyze epigenomic marks with broad enrichment. We did not explore alternative methods for GC-bias correction for broad marks as we believe this task is beyond the scope of our work.

**2. What's the exact form of weighted NB regression used in the parameter estimation part? What is X and Y in the regression?**

> In the main text, we refer the reader to Section B5 of the Web Appendix. To estimate the model coefficients during the $t^{th}$ iteration of the EM-algorithm, one needs to find the optimal values $\beta_{1}^{(t)}$, $\beta_{3}^{(t)}$, $\lambda_{1}^{(t)}$, and $\lambda_{3}^{(t)}$ that maximize the quantity $Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) + Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) + Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right)$. This problem is analogous to estimating parameters in a weighted NB regression model, since $Q_{1}$, $Q_{2}$, and $Q_{3}$ can be seen as weighted log-likelihood functions from a NB generalized linear model with means and dispersions parametrized as described in the Section 3.1 of the main article. Exact forms of the $Q$-functions are presented in Equation B.4 of Web Appendix B5.

**3. What are the BIC measurement for Table 1 and corresponding web tables? Are these results showing the optimal component?**

> Results presented in Table 1 pretain to the full model with all possible combinatorial patterns. This is the 'optimal' model in the sense that the simulated data contains all possible combinatorial patterns of enrichment and background among the condition. The point of Figure 1 is to show that our efficient estimation method via fast RCEM leads to unbiased parameter estimates. Together with Figure 2, Table 1 raises the point that increasing the number of technical or biological replicates leads to better results in terms of sensitivity/specificity of peak calls as well as in the sense of asymptotic efficiency of parameter estimates. We leave the discussion of model selection via BIC in our model to Section B2 of the Web Appendix where we present a simulation study.

**4. In section 4.1.1, there is no clear evidence in Table 1 and corresponding web tables showing the trend that higher SNR has a lower bias as there are only two levels of SNR in one replicate cases, more levels of SNR may be needed for this, also what is the SNR level for the real data from ENCODE?**

> We thank the reviewer for point this out. We believe that comparing the relative bias of parameter estimates across different SNR levels is difficult because different SNRs are associated with different parameter values (some larger than others) in our simulation study. Although we did observe an improvement in performance in terms of model sensitivity/specificity of peak calls in higher SNR data sets, the main driving quantity for the reduction of parameter bias is the number of replicates included in our model. For this reason, we rephrase the mentioned sentence in our paper as "Depending on the number of conditions, the observed relative bias and the range of the reported percentiles tended to decrease as more replicates were included in the analyses. This effect was particularly significant in scenarios with four conditions with respect to parameters $\beta_{3}$ and $\lambda_{3}$. In scenarios with higher number of conditions, these results highlight the importance of experimental replicates to achieve precise parameter estimates.". The estimated SNR level from ENCODE was 3.21 (ratio of mean counts between peak and background regions) for H3K37me3. This is now reported in our manuscript. We present the estimate SNR value for other scenarios in our Web Appendix (Web Tables 2-3)

**5. In real data benchmarking,  some histone marks have a clear classification of broad or narrow peaks marks, others are not, for example, H3K27ac and H3K4me3 are the predictors for super-enhancer or super promoter, the genome-wide peaks are constituted with a mixture of broad and narrow peaks,  mixNBHMM could handle the independent narrow or broad peaks,  could mixNBHMM handle the heterogenous peaks situation?**

> We believe our method is able to handle such situations. Note, for instance, in Figure 4D of our manuscript that we report differential peak calls from H3K4me3 in which both narrow and broad differential peaks are deemed to be differential by epigraHMM at FDR level of 0.05 (see rightmost peaks). ChIPComp, for instance, missed the broad peak on the right while called as differential all narrow peaks on the left. In fact, note that the left most region of Figure 4D is comprised by a super enhancer region, in which several peaks are present in the cell line Hepg2. Our method was able to call this super enhancer region as differential while also covering the broad peak on the right of the panel.

**6. What is the exact difference between nominal FDR and observed FDR? Is nominal FDR the FDR estimated in the alpha or total FDR mentioned in the method part?**

> Nominal FDR is the total FDR as described in the last paragraph of the Methods section. The observed FDR is the actual proportion of false discoveries (i.e. the proportion of genomic windows incorrectly classified as differentially enriched out of the total number of genomic windows classified as differential). Once a nominal FDR level is specified, an $\alpha$ cutoff level can be determined and used to classify genomic windows into differential and non-differential. Then, the observed FDR can be straightforwardly calculated as the proportion of false discoveries. We make this distinction clear in the last paragraph of the 'Methods' section as well as in the first paragraph of the 'Simulation Results' section.

**7. For the benchmark measurement of real data, how about spearman correlation and LFC metrics for the differential broad peaks calling? Using differentially transcribed genes from LFC (>2) of H3K36me3 sounds like a cyclic logic for differential H3K36me3 peak call, authors may need to consider using the matched differentially expressed genes as the gold standard to evaluate the differential transcribed genes called from differential H3K36me3 activity. In addition, gene ontology at least should be provided for interpreting the biological significance of those differential peaks covered genes, are there any liver (HepG2) or Cervical (HelaS3) related terms predicted?**

> We believe that Spearman correlation and LFC are not good metrics to compare peak callers with broad marks for two reasons. First, methods that tend to call narrow peaks within brod regions of enrichment will exhibit an unrealistic advantage in performance in terms of LFC. This is so because these methods tend to call as narrow peaks any genomic window exhibiting excessively high read count enrichment (or, LFC in differential enrichment). For broad marks, interest lies not on these single punctate windows but instead in broad regions of enrichment. Second, and due to a similar reason, these methods will also exhibit unrealistic advantage in performance in terms of Spearman correlation because broad regions of enrichment are much noisier than punctate regions and, thefore, read count correlations calculated from broad peaks will be attenuated in comparison to correlations calculated from narrow and punctate windows.

> Regarding the use of 'matched differentially expressed genes as the gold standard to evaluate the differential transcribed genes called from differential H3K36me3 activity', we did not observe a sufficient correspondance between gene expression levels and H3K36me3 enrichment for the cancer cell lines analyzed in our paper. For instance, we have found several instances were genes were expressed but no enrichment of H3K36me3 was observed. Below, we present some of these problematic cases obtained from the UCSC Genome Browser. Note in Figure 1, for example, that both cell lines Helas3 and Hepg2 were not enrichment for H3K36me3 in the rightmost region of the plot while the associated gene body was clearly expressed. Note in Figure 2, for example, the lack of agreement between the transcription level of the reported gene and the enrichment for H3K36me3 in cell line Hepg2. Due to these reasons, we decided not to use gene expression levels when benchmarking H3K36me3 as suggested by the reviewer.

![First example of nonconcordant gene expression level and H3K36me3 ChIP-seq enrichment in Helas3](Fig_RNAseq_Screenshot.pdf)

![Second example of nonconcordant gene expression level and H3K36me3 ChIP-seq enrichment in Helas3](Fig_RNAseq_Screenshot2.pdf)

> Regarding the biological significance of our method, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3. Specifically, we show that all known marker genes of Helas3 had associated differential enrichment for the H3K36me3 mixture model component. The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone. These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.

**Authors may also need to consider improving the following regions of the paper to make it be more accessible to the readers (minor).**

**1. In the introduction part, "To quantify local genome activity" may be changed to "To quantify genome-wide epigenome activity".**

> I have changed "To quantify local genome activity" to "To quantify local epigenomic activity on a genome-wide scale".

**2. enhancer EZH2 is not a proper term for the description of the chromatin regulator since enhancer is a kind of cis-element, this should be corrected as Enhancer of zeste homolog 2 (EZH2).**

> This has been addressed

**3. For the layout of figures, web figure 1 could be Figure 1, and Figure 1 could be web figure 1, the workflow along with the method in the main text could make things clearer.**

> We thank the reviewer for this comment. We added Web Figure 1 to our main manuscript and it is now shown in Figure 1 of the paper.

**4. In section 3.1, the "N x 1 vector" and "NM x 1 vector" may be replaced with N vector and N x M matrix.**

> This has been addressed

**5. In section 4.2.1, how would we deal with the failure case from RSEG? Were they removed in the benchmark?**

> This has been addressed and made clear in text. For RSEG, the metrics were calculated based on the instances where it ran without issues (71 out of 100 simulations).

**6. In the supplementary documents (Web appendices), the figure or table labels can be prefixed with Web or Supplementary.**

> This has been addressed. The tables and figures of the supp. are already indexed with 'Web' in front of it (e.g. Web Figure 1).

**7. B22 section occurs twice in the web appendices.**

> This has been addressed

**8. On page 11 of main text, "t th step" of the EM algorithm can be "t th iteration" of the EM algorithm.**

> This has been addressed

**9. In section 4.1, genome length can be changed to window number.**

> This has been addressed

**10. How many threads are the software using for Figure 2C, 3E, 4E? Are they using similar memories?**

> All analyses in the paper were conducted using a single CPU per task without multithreading. Figures 4E and 3E now exhibit the peak memory comsumption by each method along with the total computing time. We observed that in broad marks (500bp) our method was among the most efficient ones, requiring only 5GB of memory and 30 min for a genome-wide analysis of 4 samples (2 conditions, 2 replicates per condition). In narrow marks (250bp), our method required nearly 10GB of memory for a computation time of about 1 hour. We believe that these are reasonable quantities and quite comparable to other  efficient algorithms. For example, the widely used method DiffBind required nearly as much memory even though it is a method that relies on peaks called a priori (MACS2 in our analysis) and had a much lower performance in broad histone modification marks.

**11. In 1st line of the 16th page, two technical replicates seem not right, usually, we computed the differential peak regions from biological replicates.**

> This has been addressed. The analyzed ENCODE data pertain to isogenic replicates (such as https://www.encodeproject.org/experiments/ENCSR000AOL/) because they are derived from cell lines of the same biosample/donor. These are neither technical nor biological replicates (https://www.encodeproject.org/data-standards/terms/), and we make this distinction in the referenced text.

**12. The main text appendix for Q function and forward and backward probability can be merged into the B5 in web appendices.**

> This has been addressed

**13. In Figure 5D. RNA-seq tracks may be added to see if it is a highly expressed gene set near the differential peak regions.**

> We have added the gene expression level of the covered gene body in Figure 5F of our manuscript.

**14. In B3, certain combinatorial "patters" should be "patterns".**

> This has been addressed

**15. In short differential peak calling, what might be the reason that mixNBHMM is worse than other methods?**

> We believe our method is quite comparable to other DPC in short differential peak calling. These results are shown in Figure 4 of our paper as well as in the Web Appendix (Web Figures 25-27). In terms of LFC and Spearman correlation, our method performed quite comparably to ChIPComp, a widely used method for narrow marks. Examples in Figure 4B, 4D, and 4F clearly shows that our model is correctly calling differential peaks similar to other methods.. We did see, however, a slight increase in memory consumption of 5GB when analyzing such marks due to the higher resolutions of 250bp. However, such an increase is still well within the available resourses of most high performance computing clusters and even in some personal laptops nowadays.

**16. BIC features as the model selection is a novel way for inference of chromatin segmentation, which is worth discussing in the discussion.**

> As mentioned in our paper, we added a discussion in Web Appendix regarding the model selection via BIC. Web Figures 8-10 include results from our simulation study exploring the presented strategy to choose the optimal number of combinatorial patterns via BIC. Due to the lack of space, we have added the following sentence to our Discussion "In situations where certain combinatorial patterns are rare, prunning of specific combinatorial patterns is possible via model selection via BIC (Web Appendix B2)".

**17. Biological replicate insight is really important, which is the leading factor in differential peak calling, this provides a guideline for future ChIP-seq experiment design about the necessity of generating biological replicate for differential epigenome study, which is worth mentioning in the discussion.**

> We thank the reviwer for the constructive comment. Due to the lack of space in our paper, we added the following brief sentence to our discussion "Our results highlight the importance of the inclusion of technical/biological replicates in the analysis of epigenomic data".

# Reviewer 2

**1. There are also other methods that can also perform combinatorial binding pattern detection such as jMOSAiCS (https://genomebiology.biomedcentral.com/articles/10.1186/gb-2013-14-4-r38).  dPCA (https://www.pnas.org/content/110/17/6789), SignalSpider (https://academic.oup.com/bioinformatics/article/31/1/17/2366199). The proposed method should be compared to these methods. In addition. HMM-based peak callers have been widely developed including histoneHMM (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4347972/);  HPeak (https://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-11-369); https://academic.oup.com/bioinformatics/article/24/20/2344/258202. chromHMM (http://compbio.mit.edu/ChromHMM/). These similar methods diminishes the novelty of mixNBHMM, which is also based on HMM and for the same peak-calling/genome segmentation purpose. Moreover, the difference in terms of methodology among all these HMM-based peak-calling methods and mixNBHMM are not clear.**

> TO DO.

**2. mixNBHMM does not consider background control sample in peak-calling, which may result highly false positive. While other methods such as ChIPComp consider both IP and control. I am curious how the simulations are set for these methods considering controls, and how the control samples are analyzed in the real ENCODE data analysis.**

> Input control samples were utilized by the methods that are designed and capable to do so. These methods include ChIPComp, THOR, diffReps, and DiffBind. We added the following sentence to the second paragraph of Section 5 of our paper: "Results presented in this section pertain to the analysis of two cell lines, namely Helas3 and Hepg2. Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR)". Similarly, input controls were also generated in our simulation study and accounted for in these algorithms. Throughout our analysis, we did not observe any significant improvement in performance by accounting for input control effect in differential peak calling. This effect was also reported by others (Lun \& Smyth, 2015).

**3. According to Figure 2A/Figure 3A, mixNBHMM has significantly underestimated FDR. While other methods such as RSEG has better FDR control. The results are contrary to the claim mixNBHMM outperforms other competing methods.**

> In Figure 2A, the observed FDR is shown for all methods and for all assessed nominal FDR levels, for all methods. The observed FDR shown in Figure 2A is quite close to the nominal FDR chosen (note vertical lines on the panel and corresponding nominal FDR points). In Figure 3A, no FDR was or is currently reported. Instead, we show 1-specificity on the x-axis. In Figure 3B, we show the FDR for all methods. As we can see, the observed and nominal FDR levels reported by our method were quite close to each other (0.05, 0.07, 0.08, 0.09, 0.10). In comparison, RSEG exhibited observed FDR levels of 0.23, 0.63, 0.69, 0.74,0.80 for nominal FDR levels of 0.01, 0.05, 0.10, 0.15, and 0.20. We thank the reviewer for this comment but we do believe our results are superior than the suggested FDR control of RSEG.

**4. In the real data analysis for the ENCODE data, there is no gold standard for DPC. However, nearest differential expressed gene among multiple conditions can be used to evaluate DPC because DPC and nearest DE genes could have the same change direction and magnitude.**

> We thank the reviewer for the constructive comment. As we replied to the first anonymous reviewe, we did not observe a sufficient agreement between gene expression levels for the analyzed cell lines in our paper (see Figures 1 and 2 above). In fact, not all the analyzed marks in this paper are associated with the expression level of the neared gene. For instance, H3K27me3 and EZH2 are two epigenomic marks known to exhibit read enrichment in intergenic regions without any nearby gene body. We believe that this analysis would be suboptimal to the one presented in this paper.

**5. The importance of mixNBHMM lies on its ability to detect different combinatorial pattern of enrichment across cell lines.  However, the biological meaning of combinatorial pattern is not clear, which limits the general application of mixNBHMM. A biological interpretation of the finding results are needed to demonstrate the potential biological application of the method, e.g. finding novel TF cofactors.**

> As we discuss above, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3. Specifically, we show that all known marker genes of Helas3 had associated differential enrichment for the H3K36me3 mixture model component. The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone. These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.

# Reviewer 3

**The real data analysis section is another simulation study. The authors assumed that they know the truth about the data and compared the result they get from mixNBHMM to the results they get from other competing pipelines. For instance they provided criteria ROC (Figure 3A), false positive rates, etc. If it is a real “real dataset” then nobody knows the truth and the authors must remove all comparisons to other methods and provide a real data analysis. If the authors know the truth, then they must move this section to simulation study and provide another real dataset and analyze it without comparing to other methods.**

> We appreciate the comment for our reviewer 3. Although Figure 3 of our paper does pertain to a simulation study, our paper also includes an extensive section on the analysis of real samples from the ENCODE project. We compare our method with competing algorithms in both simulated and real data scenarios. We utilized data driven metric to benchmark our algorithm with other and observed superior performance throughout our analysis. Section 5 of our paper presents results from the analysis of the ENCODE project.

### 2021-02-19 Latex/paper/paper.tex
Updating tex

@@
 with $y_{hij}\in\{0,1,2,\ldots\}$, such that $\log(\mu_{(1,hij)})=\beta_{1} + u_{hij}$, $\log(\phi_{1})=\lambda_{1}$, $\log(\mu_{(3,hij)})=\beta_{1} + \beta_{3} + u_{hij}$, and $\log(\phi_{3})=\lambda_{1} + \lambda_{3}$.
-The offset $u_{hij}$ adjusts for technical or biological artifacts (such as GC-content bias) and allows the non-linear normalization of the signal profile across genomic windows, conditions, and samples (Web Appendix B2).
+The offset $u_{hij}$ adjusts for technical or biological artifacts (such as the GC-content bias as presented in \cite{teng2017accounting}, Web Appendix B1) and allows the non-linear normalization of the signal profile across genomic windows, conditions, and samples (Web Appendix B1).
 When $u_{hij}=0$, $\beta_{1}$ and $\lambda_{1}$ represent the log-mean and log-dispersion, respectively, of read counts pertaining to consensus background state windows, whereas $\beta_{3}$ and $\lambda_{3}$ represent the difference in log-mean and log-dispersion of read counts from consensus enrichment state windows relative to consensus background state windows.
@@
 input controls; Web Appendix A3), for which a state-level testing of their effects on the read count distribution could be performed.
-In a simulation study and in real data analyses, however, we did not observe a significant improvement in performance in differential peak detection after accounting for the effect of input controls (Web Appendix B2), a fact that has also been observed by others \citep{lun2015csaw}.
+In a simulation study and in real data analyses, however, we did not observe a significant improvement in performance in differential peak detection after accounting for the effect of input controls (Web Appendix B), a fact that has also been observed by others \citep{lun2015csaw}.
 
@@
 \end{align}
-where $Q_{0}(\bs{\pi},\bs{\gamma}|\bs{\Psi}^{(t)})$, $Q_{1}(\bs{\psi}_{1}|\bs{\Psi}^{(t)})$, $Q_{2}(\bs{\delta},\bs{\psi}_{2}|\bs{\Psi}^{(t)})$, and $Q_{3}(\bs{\psi}_{3}|\bs{\Psi}^{(t)})$ are defined in Web Appendix B5.
+where $Q_{0}(\bs{\pi},\bs{\gamma}|\bs{\Psi}^{(t)})$, $Q_{1}(\bs{\psi}_{1}|\bs{\Psi}^{(t)})$, $Q_{2}(\bs{\delta},\bs{\psi}_{2}|\bs{\Psi}^{(t)})$, and $Q_{3}(\bs{\psi}_{3}|\bs{\Psi}^{(t)})$ are defined in Web Appendix B4.
 In the E-step of the EM algorithm, we compute the posterior probabilities from (\ref{eq:diffQfunc}).
-The quantities $Pr\left(Z_{j}=r|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)$ and $Pr\left(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)$, defined in Web Appendix B5, can be calculated through the Forward-Backward algorithm and $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})=f_{(2,l)}(\mb{y}_{..j}|\mb{x}_l;\bs{\psi}_{(2,l)}^{(t)})\delta_{l}^{(t)}/\sum_{k=1}^{L}f_{(2,k)}(\mb{y}_{..j}|\mb{x}_k;\bs{\psi}_{(2,k)}^{(t)})\delta_{k}^{(t)}$ for $l=1,\ldots,L$.
+The quantities $Pr\left(Z_{j}=r|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)$ and $Pr\left(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)$, defined in Web Appendix B4, can be calculated through the Forward-Backward algorithm and $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})=f_{(2,l)}(\mb{y}_{..j}|\mb{x}_l;\bs{\psi}_{(2,l)}^{(t)})\delta_{l}^{(t)}/\sum_{k=1}^{L}f_{(2,k)}(\mb{y}_{..j}|\mb{x}_k;\bs{\psi}_{(2,k)}^{(t)})\delta_{k}^{(t)}$ for $l=1,\ldots,L$.
 
@@
 First, mixture proportions can be estimated as $\hat{\delta}_{l}^{(t+1)}=\sum_{j=1}^{M}Pr(Z_{j}=2|\mb{y},\mb{x};\bs{\Psi}^{(t)})Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})/\sum_{j=1}^{M}Pr(Z_{j}=2|\mb{y},\mb{x};\bs{\Psi}^{(t)})$.
-Estimating $(\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$ from (\ref{eq:diffQfunc}) can be seen as obtaining parameter estimates from a series of weighted NB regression models with shared mean and dispersion parameters (Web Appendix B5).
+Estimating $(\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$ from (\ref{eq:diffQfunc}) can be seen as obtaining parameter estimates from a series of weighted NB regression models with shared mean and dispersion parameters (Web Appendix B4).
 We jointly estimate these quantities via the algorithm BFGS \citep{fletcher2013practical}.
@@
 If pruning differential combinatorial patterns of the differential mixture component is of interest, the optimal number of mixture components $L^{*}$, $L^{*}<L$, can be selected via the Bayesian Information Criterion (BIC) for HMMs.
-We observed that selecting the optimal number of mixture components based on BIC agrees with the pruning of rare differential combinatorial patterns that we would not biologically expect to observe in real data (Web Appendix B3).
+We observed that selecting the optimal number of mixture components based on BIC agrees with the pruning of rare differential combinatorial patterns that we would not biologically expect to observe in real data (Web Appendix B2).
 
@@
 Differential regions of enrichment are formed by merging adjacent windows that either meet a nominal FDR threshold level for the differential HMM state or belong to the same Viterbi's predicted state.
-A discussion about these two approaches under the proposed model is presented in the Web Appendix B4.
-Additional details of proposed EM algorithm and the implemented code are available in the Web Appendices B5 and A3, respectively.
+A discussion about these two approaches under the proposed model is presented in the Web Appendix B3.
+Additional details of proposed EM algorithm and the implemented code are available in the Web Appendices B4 and A4, respectively.
 
@@
 This effect was particularly significant in scenarios with four conditions with respect to parameters $\beta_{3}$ and $\lambda_{3}$.
-In general, scenarios with higher SNR showed lower relative bias and variability of the parameter estimates in comparison to scenarios with lower SNR, regardless of the number of conditions or replicates per condition.
 In scenarios with higher number of conditions, these results highlight the importance of experimental replicates to achieve precise parameter estimates.
@@
 
-We compared the genome-wide performance of the presented model with the current DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR.
+We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short data sets.
 For the methods that require a set of candidate regions to be specified \textit{a priori}, ChIPComp and DiffBind, peaks were called in advance using MACS2 (\cite{zhang2008model}; available at \url{https://github.com/macs3-project/MACS}) and used as input to the software for differential call.
@@
 Differential genomic windows were mostly representing either transcribed chromatin states or repressed chromatin states (Figure~\ref{fig:fig_multimarks}B).
-All differential combinatorial patterns expected to be rare had associated mixture proportion estimates less than 0.05 and were pruned using our model selection scheme via BIC presented in Web Appendix B3 (Figure~\ref{fig:fig_multimarks}C).
+All differential combinatorial patterns expected to be rare had associated mixture proportion estimates less than 0.05 and were pruned using our model selection scheme via BIC presented in Web Appendix B2 (Figure~\ref{fig:fig_multimarks}C).
 %.
@@
 Our method offers the benefit of simultaneously detecting differential peaks and classifying the combinatorial pattern of enrichment through mixture model posterior probabilities even in the context of genomic segmentation with highly diverse epigenomic marks (see Web Appendix D4).
-By using the BIC for model selection (Web Appendix B3), one can choose the number of biologically relevant mixture components to be included in the model, a task that may not be as straightforward in methods such as ChromHMM (see Supplementary Figure 4 in \citet{ernst2012chromhmm}).
+By using the BIC for model selection (Web Appendix B2), one can choose the number of biologically relevant mixture components to be included in the model, a task that may not be as straightforward in methods such as ChromHMM (see Supplementary Figure 4 in \citet{ernst2012chromhmm}).
 
@@
 Our implementation of the RCEM algorithm led to genome-wide analyses of ChIP-seq data under a computational time comparable to some of the fastest current methods and was at least 5 times faster than current HMM-based algorithms.
+Our results highlight the importance of the inclusion of technical/biological replicates in the analysis of epigenomic data.
 Lastly, our method allows the simultaneous detection and classification of differential combinatorial patterns of enrichment from its embedded mixture model and the associated posterior probabilities under any number of conditions.
-In situations where certain combinatorial patterns are rare, prunning of specific combinatorial patterns is possible via model selection via BIC.
+In situations where certain combinatorial patterns are rare, prunning of specific combinatorial patterns is possible via model selection via BIC (Web Appendix B2).
 Our software has been implemented into an R package and is available for download (Web Appendix A).

### 2021-02-19 Latex/appendix/webappendix.Rnw
Updating text

@@
 \subsection{Normalization for non-linear biases via model offsets}
+\label{s:loess}
 
@@
 Overall, we did not observe a significant improvement in performance by including input control in differential peak calling.
-Although several methods do offer the option of including controls in their analysis pipeline, we did not find that their inclusion was justifiable under our modelling assumptions.
+Although several methods do offer the option of including controls in their analysis pipeline, we did not find that their inclusion was justifiable under our modeling assumptions.
 Our findings are in agreement with \cite{lun2015csaw}.
@@
 
-\subsubsection{Simulation Study}
+We performed a simulation study to assess the benefits and ability of epigraHMM to adjust for GC-content bias.
+In this study, we borrowed the method from \cite{teng2017accounting} and used their normalizing denominators to adjust for GC-content bias in our simulations study.
+As \cite{teng2017accounting} suggest, we utilized their normalizing offsets into our own model to account for the bias results from the simulated GC content.
+Because the simulated data varied with respect to the sequencing depth, we further included in our model normalizing offsets to adjust for the excess of read enrichment using the loess normalization approach as described in Section \ref{s:loess}.
+
+Data was generated as described in Section 4.1 of our paper, with the addition that GC-content bias was introduced to the mean model.
+As suggested by \cite{teng2017accounting}, the simulated GC-content effect varied from condition to condition as well as the sequencing depth.
+In this simulation study, we assessed decreasing, neutral, and increasing effects of GC-content on the mean read count distribution.
+In Figure \ref{fig:sim_gc_scatter} we show an example of simulated data with increasing GC-content effect on ChIP-seq counts in both conditions.
+In this example, these conditions varied regarding the depth of sequencing reads.
 
@@
 
+We assessed the model performance while adjusting for sequencing depth only, GC-content effect only, and simultaneously adjusting for both GC-content and sequencing depth.
+We observed that the optimal results were achieve only after normalization for both GC and sequencing depth bias.
+Despite the improvement in performance after the non-linear normalization, we were able to achieve optimal results in terms of true positive rate and false discovery rate of peak calls only after accounting for GC-content bias.
+These results not only show that our model is able to account for such type of bias, here computed using the methodology from \cite{teng2017accounting}, but also highlight the importance of such an adjustment.
+These results are shown in Figure \ref{fig:sim_gc_fdr}.
+
 \begin{figure}[h]
@@
 \end{figure}
+
+To better visualize the results on a per data set basis, Figure \ref{fig:sim_gc_roc} shows the sensitivity and 1-specificity results from all 100 simulated datasets for the aforementioned example.
+Here, we see that for all data sets only after account for GC-content bias we were able to achieve optimal results.
 
@@
 \section{Bayesian Information Criterion (BIC) for Hidden Markov Models}
+\label{s:bicsec}
 
@@
 
+We assessed the performance of our model selection scheme via BIC with a simulation study.
+To this end, we re-generated data in the same fashion as described in Section 4 of our paper while removing certain combinatorial patterns.
+Specifically, to mimic the ENCODE data analyzed in Section 5 of our paper, we generated data for which 3 conditions were available but only 2 out of 6 possible differential combinatorial patterns were available.
+
+First, it is important to note that for an analysis of 3 conditions, there are $2^3-2 = 6$ possible combinatorial patterns of differential enrichment.
+In this context, for a mixture model with $L$ components, there are ${6\choose L}$ possible ways to model $L$ components.
+The choice of the number $L$ of components therefore needs to reflect the relevant biological patterns in the data.
+In this study, we assessed the algorithm presented in Section \ref{s:bicsec} by fitting all $\sum_{l=1}^{6}{6\choose l}$ possible model and computing their BIC for 100 simulated data sets.
+On average, the optimal chosen model should be the one with $L=2$ components \textit{and} with the components modeling those relevant ones.
+In this study, the two relevant patterns simulated were enrichment in condition 1 alone, and co-enrichment of conditions 2 and 3 (similar to Figure 6 of the main paper).
+As in Section 4 of our paper, we assessed the performance of our model under different SNR levels, different number of replicates per condition, and different genome sizes.
+
+We present our results in Figures \ref{fig:bic_4rep}, \ref{fig:bic_2rep}, and \ref{fig:bic_1rep}.
+We start our model selection by fitting the full model with 6 components (L=6, rightmost block of the triangle with the highlighted '1').
+We then proceed with our algorithm as described in this section and remove 1 combinatorial pattern at a time until the difference in BIC is large enough.
+As we can see from Figures \ref{fig:bic_4rep}, \ref{fig:bic_2rep}, and \ref{fig:bic_1rep}, this is achieved when L=2 and the combinatorial patterns being modeled is 1 (enrichment condition 1) and 2/3 (co-enrichment of condition 2 and 3).
+As we can see, our model selection scheme correctly selected the optimal model with the proper number of combinatorial patterns $L$ and the proper patterns themselves.
+
+In accordance with the results presented in Section 4 of our paper, we note that the number of replicates played a major role in the success of our model selections scheme.
+For example, by comparing Figure \ref{fig:bic_4rep} and Figure \ref{fig:bic_1rep}, we see that more technical/biological replicates lead to models that can be better distinguished in terms of BIC when one is searching for the optimal patterns and number of combinatorial patterns.
+In summary, the presented model selection scheme allows users to prune combinatorial patterns of interest when one is not interested in modeling certain rare biological patterns from the data.
+
 \begin{figure}[h]
@@
 \end{center}
-\caption{Simulation study for model selection via BIC. Simulation scenario: H3K36me3 data, 3 conditions, 4 replicates per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR.
+\caption{Simulation study for model selection via BIC.
+Simulation scenario: H3K36me3 data, 3 conditions, 4 replicates per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR.
+Dashed lines indicate order of model selection scheme beginning with the full model ('1') and ending with the reduced model ('6').
+The optimal model is the one with highlighted '2', since it had a significant lower BIC than any other model with a single differential component.
+Highlighted number in each block indicated the combinatorial pattern being modeled.
+For instance '1/2 2 3' indicates a model with three components and the following patterns: co-enrichment of condition 1 and 2, enrichment of condition 2 alone, and enrichment of condition 3 alone.
 \label{fig:bic_4rep}}
@@
 \end{center}
-\caption{Simulation study for model selection via BIC. Simulation scenario: H3K36me3 data, 3 conditions, 2 replicates per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR.
+\caption{Simulation study for model selection via BIC.
+Simulation scenario: H3K36me3 data, 3 conditions, 2 replicates per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR.
+Dashed lines indicate order of model selection scheme beginning with the full model ('1') and ending with the reduced model ('6').
+The optimal model is the one with highlighted '2', since it had a significant lower BIC than any other model with a single differential component.
+Highlighted number in each block indicated the combinatorial pattern being modeled.
+For instance '1/2 2 3' indicates a model with three components and the following patterns: co-enrichment of condition 1 and 2, enrichment of condition 2 alone, and enrichment of condition 3 alone.
 \label{fig:bic_2rep}}
@@
 \end{center}
-\caption{Simulation study for model selection via BIC. Simulation scenario: H3K36me3 data, 3 conditions, 1 replicate per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR.
+\caption{Simulation study for model selection via BIC.
+Simulation scenario: H3K36me3 data, 3 conditions, 1 replicate per condition, $10^5$ genomic windows, observed (estimated from ENCODE) SNR.
+Dashed lines indicate order of model selection scheme beginning with the full model ('1') and ending with the reduced model ('6').
+The optimal model is the one with highlighted '2', since it had a significant lower BIC than any other model with a single differential component.
+Highlighted number in each block indicated the combinatorial pattern being modeled.
+For instance '1/2 2 3' indicates a model with three components and the following patterns: co-enrichment of condition 1 and 2, enrichment of condition 2 alone, and enrichment of condition 3 alone.
 \label{fig:bic_1rep}}

### 2021-02-19 Latex/paper/paper.tex
Updating text

@@
 (B): estimated mixture probabilities and the associated differential combinatorial patterns.
-(C): density estimate from expression of genes intersecting differential peaks associated with the enrichment of H3K36me3 alone or the enrichment of H3K27me3 and EZH2 in consensus.
-(D): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively.
+(C): choice of best model with two mixture components via BIC.
+(D): density estimate from expression of genes intersecting differential peaks associated with the enrichment of H3K36me3 alone or the enrichment of H3K27me3 and EZH2 in consensus.
+(D): association with differential H3K36me3 enrichment based based on mixture model posterior proportions of known marker genes.
+(E): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively.
 This figure appears in color in the electronic version of this article.
@@
 Lastly, our method allows the simultaneous detection and classification of differential combinatorial patterns of enrichment from its embedded mixture model and the associated posterior probabilities under any number of conditions.
-In situations where certain combinatorial patterns are rare, prunning of specific combinatorial patterns is possible via model selection via BIC (Web Appendix B2).
+In situations where certain combinatorial patterns are rare, pruning of specific combinatorial patterns is possible via model selection via BIC (Web Appendix B2).
 Our software has been implemented into an R package and is available for download (Web Appendix A).

### 2021-02-24 Latex/letter/letter.Rmd
Update letter.Rmd
typo fixes

@@
 > We believe that Spearman correlation and LFC are not good metrics to compare peak callers with broad marks for two reasons.
-First, methods that tend to call narrow peaks within brod regions of enrichment will exhibit an unrealistic advantage in performance in terms of LFC.
+First, methods that tend to call narrow peaks within broad regions of enrichment will exhibit an unrealistic advantage in performance in terms of LFC.
 This is so because these methods tend to call as narrow peaks any genomic window exhibiting excessively high read count enrichment (or, LFC in differential enrichment).
@@
 Below, we present some of these problematic cases obtained from the UCSC Genome Browser.
-Note in Figure 1, for example, that both cell lines Helas3 and Hepg2 were not enrichment for H3K36me3 in the rightmost region of the plot while the associated gene body was clearly expressed.
+Note in Figure 1, for example, that both cell lines Helas3 and Hepg2 did not show enrichment for H3K36me3 in the rightmost region of the plot while the associated gene body was clearly expressed.
 Note in Figure 2, for example, the lack of agreement between the transcription level of the reported gene and the enrichment for H3K36me3 in cell line Hepg2.
@@
 
-> We believe our method is quite comparable to other DPC in short differential peak calling.
+> We believe our method is quite comparable to other DPC methods in short differential peak calling.
 These results are shown in Figure 4 of our paper as well as in the Web Appendix (Web Figures 25-27).
 In terms of LFC and Spearman correlation, our method performed quite comparably to ChIPComp, a widely used method for narrow marks.
-Examples in Figure 4B, 4D, and 4F clearly shows that our model is correctly calling differential peaks similar to other methods..
+Examples in Figure 4B, 4D, and 4F clearly shows that our model is correctly calling differential peaks similar to other methods.
 We did see, however, a slight increase in memory consumption of 5GB when analyzing such marks due to the higher resolutions of 250bp.
@@
 > We thank the reviewer for the constructive comment.
-As we replied to the first anonymous reviewe, we did not observe a sufficient agreement between gene expression levels for the analyzed cell lines in our paper (see Figures 1 and 2 above).
+As we replied to the first anonymous reviewer, we did not observe sufficient agreement between gene expression levels for the analyzed cell lines in our paper (see Figures 1 and 2 above).
 In fact, not all the analyzed marks in this paper are associated with the expression level of the neared gene.

### 2021-02-26 Latex/letter/letter.Rmd
txt updates

@@
 > We thank the reviewer for this constructive comment.
-We cite in our paper the benchmarking article 'A comprehensive comparison of tools for differential ChIP-seq analysis' (Briefins in Bioinformatics, 2016).
+We cite in our paper the benchmarking article 'A comprehensive comparison of tools for differential ChIP-seq analysis' (Briefings in Bioinformatics, 2016).
 However, we authors are not aware of a more recent comprehensive review of DPCs algorithms in the literature.
@@
 
-> TO DO.
+> We thank the anonymous reviewer for this constructive comment.
+We note that ChromHMM benchmarked against the our method is the most widely used tool to perform chromatin segmentation analyses, and it is present as a standard method for large consortium projects such as ENCODE, Roadmap Epigenomics, Blueprint, CEEHRC, Mouse ENCODE, and Drosophila modENCODE^[https://dx.doi.org/10.1038%2Fnprot.2017.124].
+In addition, we believe that the comparison of our method to the suggested algorithms would provide a secondary analysis for our paper due to the following reasons.
+
+> First, jMOSAiCS is a package for chromatin segmentation that was removed from the Bioconductor project during release 3.4 in October 2016^[http://bioconductor.org/about/removed-packages/].
+Second, methods have highlighted several limitations of dPCA^[https://dx.doi.org/10.1038%2Fncomms8973] in the analysis of chromatin segmentation, which includes the sensitivity to data scaling inherent to PCA, and the lack of covariate correction for batch and experiment-specific biases.
+Third, SignalSpider is an algorithm based on *Gaussian mixture models* for the clustering of enhancer and gene promoter regions across multiple ChIP-seq dat sets.
+We note that epigenomic marks associated with such regions are *narrow* in comparison to those analyzed in our paper, and their analysis via mixture models can be reasonably justified since read count enrichment is often present on narrow genomic positions.
+In contrast, our model is tailored for the analysis of marks for which read count enrichment expand through *broad* genomic domains.
+To this end, our model accounts for the local dependency of read enrichment from broad regions via *hidden Markov models* while still providing the combinatorial pattern detection via *negative binomial mixture models*.
+Hence, we believe that such a comparison would not be relevant for the readership of our article.
+
+> In comparison to other HMM-based differential peak calling algorithms, we believe that the methodology presented in our paper provides several novelties over current approaches.
+First, the hidden Markov model presented in our article has an embedded mixture model in one of its components to model the combinatorial patterns of any number of conditions.
+Methods such as THOR and RSEG (both benchmarked in our paper), as well as the suggested histoneHMM, are only capable of analyzing two conditions at a time.
+Our ability to annotate combinatorial patterns of enrichment via mixture model posterior proportions is a novelty in comparison to HMM-based methods in both two- and multi-sample analyses.
+Second, our method provides an additional degree of flexibility in terms of data normalization for non linear biases as well as GC-content bias via model offsets in comparison to the aforementioned methods.
+Third, the implementation of our HMM is novel in the sense that we utilize an efficient rejection-controlled EM algorithm that leads to a fast estimation procedure and makes our algorithm among the fastest differential peak callers present in the literature.
+The suggested algorithm HPeak is a single-sample peak caller, in contrast to our multi-sample multi-condition differential peak caller, for which no software is available for download as of February 25 2020^[See https://sph.umich.edu/csg/qin/HPeak and https://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-11-369].
 
@@
 As we replied to the first anonymous reviewer, we did not observe sufficient agreement between gene expression levels for the analyzed cell lines in our paper (see Figures 1 and 2 above).
-In fact, not all the analyzed marks in this paper are associated with the expression level of the neared gene.
+In Figure 2, we have an example of a H3K36me3 differentiallly enriched region and a not DE gene.
+In contrast, Figure 3 below presents an additional example of a DE gene and without H3K36me3 enrichment in both Helas3 and Hepg2 cell lines.
+Moreover, not all the analyzed marks in this paper are associated with the expression level of the nearest gene.
 For instance, H3K27me3 and EZH2 are two epigenomic marks known to exhibit read enrichment in intergenic regions without any nearby gene body.
-We believe that this analysis would be suboptimal to the one presented in this paper.
+In several instances, enrichment of the repressive mark H3K27me3 occurs on inter-genic regions with nearby expressed genes.
+Figure 4 below shows an example of such a region.
+Due to these reasons, we believe that the proposed analysis would be suboptimal to the one presented in this paper for the analyzed cell lines.
+
+![Third example of nonconcordant gene expression level and H3K36me3 ChIP-seq enrichment in Helas3.
+Here, a differentially expressed gene without differential erichment of H3K36me3](Fig_RNAseq_Screenshot3.pdf)
+
+![Examples of H3K27me3 (a repressive mark) enrichment on intergenic regions.
+Note, on the right, an example of H3K27me3 enrichment on an intergenic region next to an expressed gene in both Helas3 and Hepg2 cell lines.](Fig_RNAseq_Screenshot4.pdf)
 

### 2021-02-26 Latex/paper/paper.tex
txt updates

@@
 (D): density estimate from expression of genes intersecting differential peaks associated with the enrichment of H3K36me3 alone or the enrichment of H3K27me3 and EZH2 in consensus.
-(D): association with differential H3K36me3 enrichment based based on mixture model posterior proportions of known marker genes.
-(E): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively.
+(E): cancer marker genes and their associated mean mixture posterior proportion from differential H3K36me3 enrichment pattern of overlapping differential peaks.
+(F): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively.
 This figure appears in color in the electronic version of this article.
@@
 For genes overlapping differential regions associated with either transcribed or repressed chromatin states, we computed the distribution of transcripts per million (TPM) from matching RNA-seq experiments.
-Genes associated with transcribed chromatin states had a significantly higher distribution of TPM counts than genes associated with repressed chromatin states (Figure~\ref{fig:fig_multimarks}C).
-In fact, all known marker genes of the Helas3 cell line \citep{morin2008profiling} were associated with differential enrichment for H3K36me3, measured by the associated average mixture posterior proportions of the overlapping differential peaks (Figure~\ref{fig:fig_multimarks}D).
+Genes associated with transcribed chromatin states had a significantly higher distribution of TPM counts than genes associated with repressed chromatin states (Figure~\ref{fig:fig_multimarks}D).
+In fact, nearly all known Helas3 cycle-regulated genes (therefore expressed at some stage of the cell life cycle; \cite{whitfield2002identification}) were associated with differential enrichment of H3K36me3, here measured by the associated average mixture posterior proportions of the overlapping differential peaks (Figure~\ref{fig:fig_multimarks}E).
+The expression of such genes are known to correlate with the proliferative state of tumors and their study may help characterize their role in cancer \citep{whitfield2002identification}.
 We detected broad differential regions of enrichment and the classification of differential combinatorial patterns agreed with their biological roles as well as the expression levels of associated gene bodies.
-Figure~\ref{fig:fig_multimarks}E shows an example of a genomic region with differential enrichment for the analyzed marks.
+Figure~\ref{fig:fig_multimarks}F shows an example of a genomic region with differential enrichment for the analyzed marks.
 We compare our results to ChromHMM with 3 states, a method developed for chromatin segmentation.

### 2021-02-26 Main/Figure_MultiMarks/Figure_GeneSymbols.R
Changing figure

@@ -52,53 +52,30 @@ ENCODE.rnaseq.scaled <-
   ENCODE.rnaseq.scaled[overlapsAny(rowRanges(ENCODE.rnaseq.scaled),
                                    rowRanges(epigraHMM_object))]
 
-# Subsetting marker genes of Helas3 cell line (DOI: 10.2144/000112900)
-## GNAS (ENSG00000087460) does not match range from UCSC list
-## CDK4 (ENSG00000135446) does not match range from UCSC list
+# Subsetting marker genes of Helas3 cell line (https://doi.org/10.1091/mbc.02-02-0030)
+## CCNG2 does not match range from UCSC list
 
-helas3_genes <-
-  c(
-    'ANAPC7',
-    'ASPM',
-    'CDC2L2',
-    'CENPF',
-    'BRD3',
-    'CDK5RAP1',
-    'EIF2AK1',
-    'GPI',
-    'GSPT2',
-    'INPPL1',
-    'MARK3',
-    'NUP210',
-    'PER1',
-    'PKN3',
-    'RB1',
-    'SMC3',
-    'SMC4',
-    'STAG2',
-    'ULK1',
-    'YWHAE',
-    'CDC25C',
-    'CDKN3',
-    'CHEK2',
-    'EIF4A2',
-    'FANCD2',
-    'PCSK7',
-    'SDHC',
-    'SUFU',
-    'TAF15',
-    'YWHAZ',
-    #'CDK4',#'GNAS',
-    'PLK1',
-    'PSMD4',
-    'RAF1',
-    'RAP1D',
-    'BTRC',
-    'MLLT3'
-  )
+if (!file.exists('helas3_genes.txt')) {
+  helas3_genes <-
+    fread(
+      'http://genome-www.stanford.edu/Human-CellCycle/HeLa/data/table2_knowngenes.txt',
+      skip = 2,
+      header = FALSE
+    )
+  setnames(helas3_genes,
+           c('gene', 'accession', 'published', 'reference', 'phase'))
+  fwrite(helas3_genes, file = 'helas3_genes.txt')
+} else{
+  helas3_genes <- fread('helas3_genes.txt')
+}
+
+helas3_genes[gene == 'CDKN1A, p21', gene := 'CDKN1A']
+helas3_genes[gene == 'CDKN2D, p19', gene := 'CDKN2D']
+helas3_genes <- helas3_genes[-grep('Histone', gene), ]
[…]

### 2021-02-26 Main/Figure_MultiMarks/Figure_MultiMarks.R

@@ -15,7 +15,7 @@ fig.bot <- ggarrange(fig_genesymbols,fig_segmentation[['3']],nrow = 1,legend = '
 
 # Put them all together
 
-fig.final = ggarrange(fig.top,fig.mid,fig.bot,nrow=3,heights = c(0.3,0.3,0.4))
+fig.final = ggarrange(fig.top,fig.mid,fig.bot,nrow=3,heights = c(0.275,0.275,0.45))
 fig.final.label = ggarrange(fig.final,get_legend(fig_segmentation[['3']]),nrow=2,heights = c(0.95,0.05))
 
 ggsave(filename = 'Figure_MultiMarks.pdf',plot = fig.final.label,height = 11.5,width = 9,dpi = 'retina')

### 2021-02-26 Latex/paper/paper.tex
txt updates

@@
 (D): density estimate from expression of genes intersecting differential peaks associated with the enrichment of H3K36me3 alone or the enrichment of H3K27me3 and EZH2 in consensus.
-(E): cancer marker genes and their associated mean mixture posterior proportion from differential H3K36me3 enrichment pattern of overlapping differential peaks.
+(E): cycle-regulated genes and their associated mean mixture posterior proportion from differential H3K36me3 enrichment pattern of overlapping differential peaks.
 (F): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively.

### 2021-03-02 Latex/letter/letter.Rmd
Update letter.Rmd
typo fixes and added comments to be addressed

@@
 We cite in our paper the benchmarking article 'A comprehensive comparison of tools for differential ChIP-seq analysis' (Briefings in Bioinformatics, 2016).
-However, we authors are not aware of a more recent comprehensive review of DPCs algorithms in the literature.
-As suggested we add the following sentence to our real data analysis section 'We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short data sets.' We believe that such a list covers the most used algorithms for DPC in thie field of bioinformatics.
+However, we are not aware of a more recent comprehensive review of DPCs algorithms in the literature.
+As suggested, we add the following sentence to our real data analysis section 'We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short ChIP-seq enrichment profiles.' (PEDRO I UPDATED THE PRIOR LINE, PLEASE UPDATE IN THE MAIN TEXT) We believe that such a list covers the most used algorithms for DPC in thie field of bioinformatics.
 
@@
 We present a simulation study in Section B1.3 of the Web Appendix showing that epigraHMM leads to higher sensitivity/specificity when accounting for GC-bias even after read count normalization using the count-normalizing denominators from Teng and Irizarry (2017).
-However, for the broad marks that we analized in our manuscript, we did not observe a substantial effect of GC-content on read count distribution and did not observe any substantial improvement in accounting for GC-content in our model via offsets when utilizing the method from Teng and Irizarry (2017), a method that was developed only for narrow marks characterized by sharp peaks.
+However, for the broad marks that we analyzed in our manuscript, we did not observe a substantial effect of GC-content on read count distribution (IF WE EVALUATED AND SHOWED THIS IN THE PAPER, CITE THE SECTION OR FIGURE WHERE WE DID) and did not observe any substantial improvement in accounting for GC-content in our model via offsets when utilizing the method from Teng and Irizarry (2017), a method that was developed only for narrow marks characterized by sharp peaks.
 In fact, the authors do not recommend the use of their method when analyze epigenomic marks with broad enrichment.
-We did not explore alternative methods for GC-bias correction for broad marks as we believe this task is beyond the scope of our work.
+For these reasons, we did not explore alternative methods for GC-bias correction for broad marks.
+(NOTE: SEEMS LIKE YOU ARE SAYING THAT WE TRIED IT FOR BROAD MARKS AND I DID NOT MAKE A BIG DIFFERENCE, AND FOR SHORT MARKS WE CAN USE THE TENG AND IRRIZARY METHOD.
+DO WE WANT TO SAY WHY IT DOESNT WORK FOR BROAD MARKS HERE?
+SIGNAL IS TOO DIFFUSE SO LOWER CORRELATION WITH ENRICHMENT ACROSS WINDOWS COMPARED TO SHORT ENRICHMENT MARKS?
+I AM ASSUMING IT MADE A DIFFERENCE FOR SHORT MARKS?
+)
 
@@
 
-> In the main text, we refer the reader to Section B5 of the Web Appendix.
+> In the main text, we refer the reader to Section B5 of the Web Appendix (FOR THE FORM OF THE WEIGHTED NB REGRESSION?
+IF SO SAY IT TO BE CLEAR).
 To estimate the model coefficients during the $t^{th}$ iteration of the EM-algorithm, one needs to find the optimal values $\beta_{1}^{(t)}$, $\beta_{3}^{(t)}$, $\lambda_{1}^{(t)}$, and $\lambda_{3}^{(t)}$ that maximize the quantity $Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) + Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) + Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right)$.
@@
 Exact forms of the $Q$-functions are presented in Equation B.4 of Web Appendix B5.
+(I WOULD DIRECTLY EXPLAIN WHAT IS X AND Y HERE AS THE REVIEWER ASKED.
+I ALSO DO NOT SEE EXACTLY WHERE YOU EXPLAINED THE EXACT FORM OF THE WEIGHTED NB REGRESSION...IF THAT IS CLEAR FROM SECTION B5 IN THE APPENDIX THEN IGNORE THIS COMMENT).
 
@@
 
-> Results presented in Table 1 pretain to the full model with all possible combinatorial patterns.
-This is the 'optimal' model in the sense that the simulated data contains all possible combinatorial patterns of enrichment and background among the condition.
-The point of Figure 1 is to show that our efficient estimation method via fast RCEM leads to unbiased parameter estimates.
+> Results presented in Table 1 pertain to the full model that utilizes all possible combinatorial patterns in the differential state.
+This is the 'optimal' model in the sense that the simulated data contains all possible combinatorial patterns of enrichment and background by design in our simulation.
+The purpose of Figure 1 is to show that our efficient estimation method via fast RCEM leads to unbiased parameter estimates.
 Together with Figure 2, Table 1 raises the point that increasing the number of technical or biological replicates leads to better results in terms of sensitivity/specificity of peak calls as well as in the sense of asymptotic efficiency of parameter estimates.
-We leave the discussion of model selection via BIC in our model to Section B2 of the Web Appendix where we present a simulation study.
+We leave the discussion of model selection via BIC in our model to Section B2 of the Web Appendix where we present a simulation study that directly evaluates the performance of model selection in our method.
 
@@
 
-> We thank the reviewer for point this out.
+> We thank the reviewer for pointing this out.
 We believe that comparing the relative bias of parameter estimates across different SNR levels is difficult because different SNRs are associated with different parameter values (some larger than others) in our simulation study.
@@
 In scenarios with higher number of conditions, these results highlight the importance of experimental replicates to achieve precise parameter estimates.".
-The estimated SNR level from ENCODE was 3.21 (ratio of mean counts between peak and background regions) for H3K37me3.
+The estimated SNR level from ENCODE was 3.21 (ratio of mean counts between peak and background regions) for H3K27me3.
 This is now reported in our manuscript.
@@
 > We believe our method is able to handle such situations.
-Note, for instance, in Figure 4D of our manuscript that we report differential peak calls from H3K4me3 in which both narrow and broad differential peaks are deemed to be differential by epigraHMM at FDR level of 0.05 (see rightmost peaks).
+Note, for instance, in Figure 4D of our manuscript that we report differential peak calls from H3K4me3 in which both narrow and broad differential peaks are deemed to be differential by epigraHMM at an FDR level of 0.05 (see rightmost peaks).
 ChIPComp, for instance, missed the broad peak on the right while called as differential all narrow peaks on the left.
@@
 > We believe that Spearman correlation and LFC are not good metrics to compare peak callers with broad marks for two reasons.
-First, methods that tend to call narrow peaks within broad regions of enrichment will exhibit an unrealistic advantage in performance in terms of LFC.
-This is so because these methods tend to call as narrow peaks any genomic window exhibiting excessively high read count enrichment (or, LFC in differential enrichment).
+First, methods that tend to incorrectly call narrow peaks within broad regions of enrichment will exhibit inflated performance in terms of LFC, despite not capturing broad signal regions.
+This is because these methods tend to call as narrow peaks any genomic window exhibiting excessively high read count enrichment (or, LFC in differential enrichment) (IT WOULD BE GOOD TO POINT IN AN EXAMPLE IN A FIGURE FROM THE PAPER THAT SHOWS AN EXAMPLE).
 For broad marks, interest lies not on these single punctate windows but instead in broad regions of enrichment.
-Second, and due to a similar reason, these methods will also exhibit unrealistic advantage in performance in terms of Spearman correlation because broad regions of enrichment are much noisier than punctate regions and, thefore, read count correlations calculated from broad peaks will be attenuated in comparison to correlations calculated from narrow and punctate windows.
+Similarly, methods that accurately call broad regions of enrichment capture signal are much noisier than in punctate regions (SHOW EXAMPLE OF THIS) and, thefore, read count correlations calculated from broad peaks will be attenuated in comparison to correlations calculated from those called from narrow and punctate windows.
 
 > Regarding the use of 'matched differentially expressed genes as the gold standard to evaluate the differential transcribed genes called from differential H3K36me3 activity', we did not observe a sufficient correspondance between gene expression levels and H3K36me3 enrichment for the cancer cell lines analyzed in our paper.
-For instance, we have found several instances were genes were expressed but no enrichment of H3K36me3 was observed.
+For instance, we have found several instances where genes were expressed but no enrichment of H3K36me3 was observed.
 Below, we present some of these problematic cases obtained from the UCSC Genome Browser.
@@
 Due to these reasons, we decided not to use gene expression levels when benchmarking H3K36me3 as suggested by the reviewer.
+(I AM STILL NOT FOLLOWING THIS ARGUMENT.
+HOW ARE WE BENCHMARKING THEN?)
 
@@
 > Regarding the biological significance of our method, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3.
-Specifically, we show that all known marker genes of Helas3 had associated differential enrichment for the H3K36me3 mixture model component.
-The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone.
+Specifically, we show that all known marker genes of Helas3 were associated with differential enrichment for the H3K36me3 mixture model component (I DONT NKOW WHAT THIS MEANS - WINDOWS THAT OVERLAP THESE GENES WERE CLASSIFIED AS THE DIFFERENTIAL STATE?
+MAKE THIS CLEAR).
+The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone (THIS LINE IS HARD TO FOLLOW...DO YOU MEAN THE MEAN K36 ONLY COMBINATION POSTERIOR PROBABILITY OF WINDOWS COVERING THIS GENE?
+NOT CLEAR).
 These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.
@@
 Similarly, input controls were also generated in our simulation study and accounted for in these algorithms.
-Throughout our analysis, we did not observe any significant improvement in performance by accounting for input control effect in differential peak calling.
+Throughout our analysis, we did not observe any significant improvement in performance by accounting for input control effect in differential peak calling (IS THIS REFERRING TO ALL METHODS, OR JUST OURS?
+DO WE SHOW THIS FOR EITHER?
+IF SO WE SHOLD REFER TO THE FIGURE OR SECTION RELATED TO THIS POINT).
 This effect was also reported by others (Lun \& Smyth, 2015).
@@
 Due to these reasons, we believe that the proposed analysis would be suboptimal to the one presented in this paper for the analyzed cell lines.
+(SO THE ARGUMENT IS THAT IT ISNT A 1:1 RELATIONSHIP AND THATS WHY WE DIDNT USE DIFFERENTIAL EXPRESSION OF GE WITH DIFFERENTIAL EXPRESSION OF THE MARK?
+HOW WIDESPREAD IS THIS?
+DO A LOT OF OTHER PAPERS DO THIS?
+IF NOT, CAN WE CITE OTHER PAPERS ALSO, SAYING THAT THEY DO NOT USE THIS METRIC?)
 
@@
 Specifically, we show that all known marker genes of Helas3 had associated differential enrichment for the H3K36me3 mixture model component.
-The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone.
+The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone (SEE MY PRIOR COMMENT ABOUT MAKING THIS CLEAR...POSTERIOR PROBABILITY OF K36 ALONE, AVERAGED ACROSS WINDOWS SPANNING THIS GENE?).
 These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.

### 2021-03-03 Supp/Figure_GC_Counts/Figure_GC_Counts.R
Adding GC figure

@@ -1,0 +1,82 @@
+library(ggplot2)
+library(BSgenome.Hsapiens.UCSC.hg19)
+
+bp = 500
+ngroups = 2
+mark = 'EZH2'
+data = 'Encode_twocells'
+chromosome = paste0('chr',c(1:22,'X'))
+
+# Loading data
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3Ezh239875AlnRep1.markdup.q10.sorted.RData'
+  )
+)
+helas31 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_helas3/',
+    mark,
+    '/wgEncodeBroadHistoneHelas3Ezh239875AlnRep2.markdup.q10.sorted.RData'
+  )
+)
+helas32 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2Ezh239875AlnRep1.markdup.q10.sorted.RData'
+  )
+)
+hepg21 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+rm(counts)
+load(
+  paste0(
+    '../../Data/Encode_hepg2/',
+    mark,
+    '/wgEncodeBroadHistoneHepg2Ezh239875AlnRep2.markdup.q10.sorted.RData'
+  )
+)
+hepg22 = subset(counts[[paste0(bp)]], chr %in% chromosome)
+
+countsraw <- as.data.table(counts[[paste0(bp)]])
+countsraw <- countsraw[chr %in% chromosome, ]
+gr.countsraw <- countsraw[, GRanges(chr, IRanges(start, stop))]
+
+counts = subset(counts[[paste0(bp)]], chr %in% chromosome, select = c('chr', 'start', 'stop'))
+gr.counts = with(counts, GRanges(chr, IRanges(start = start, end = stop)))
+
+# Combining data
+ChIP = cbind(counts, Window = 1:nrow(counts))
+ChIP = cbind(ChIP, Helas3_1 = helas31$counts)
[…]

### 2021-03-03 Supp/Report_RNA_vs_ChIP/report.Rmd
Adding RNA-seq report

---
title: 'RNA- and ChIP-seq study: to which degree differential H3K36me3 associate with DGE?'
output:
  pdf_document:
    toc: true
    number_sections: yes
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Pedro L. Baldoni
urlcolor: blue
header-includes:
  - \usepackage{setspace}\doublespacing
editor_options: 
  chunk_output_type: inline
---

```{r include=FALSE}
knitr::opts_chunk$set(comment = NA)
knitr::opts_chunk$set(size = "small")
```

Here, I will assess to which degree differential enrichment for H3K36me3 associate with differential gene expression (DGE) levels. The main question I want to answer is: can we use differential H3K36me3 enrichment to infer DGE, and vice-versa?

To answer this question, I will use RNA-seq and ChIP-seq data from cell lines Helas3 and Hepg2 (two of the cell lines reported in our manuscripts). There are two isogenic replicates from each assay. RNA-seq data were previously quantified with Salmon. PCR duplicates and low-quality reads were removed from ChIP-seq data.

# RNA-seq data preprocessing

The analysis of RNA-seq data will follow the steps from the DESeq2 [vignette](http://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html).

## Preparing the data

```{r libraries,warning=FALSE,message=FALSE}
# Loading some libraries
library(EnsDb.Hsapiens.v75)
library(tximport)
library(pheatmap)
library(DESeq2)
library(vsn)
library(data.table)
library(ggplot2)
library(apeglm)
```

```{r salmon}
# Reading in Salmon's qf files

files <- list.files('../../Public/Salmon','^quant.sf$',recursive = T,full.names = T)

for(i in files){
  cellname = strsplit(i,'/')[[1]][length(strsplit(i,'/')[[1]])-2]
  repname = strsplit(i,'/')[[1]][length(strsplit(i,'/')[[1]])-1]
  repname = substr(repname,nchar(repname),nchar(repname))
  names(files)[which(files==i)] = paste0(cellname,'.Rep',repname)
}

files
```

```{r tximport}
# Creating annotation table & calling tximport
Tx <- transcripts(EnsDb.Hsapiens.v75,return.type='data.frame')
tx2gene = subset(Tx,select=c('tx_id','gene_id'))

txi.salmon <- tximport(files, type = "salmon", tx2gene = tx2gene)
head(txi.salmon$counts)
```

```{r deseq2_object}
# Creating DESeq2 object
sampleTable <-
  data.frame(condition = factor(unlist(lapply(
    strsplit(colnames(txi.salmon$counts), "\\."),
    FUN = function(x) {
      x[1]
    }
  ))))
rownames(sampleTable) <- colnames(txi.salmon$counts)
sampleTable

dds <-
  DESeqDataSetFromTximport(txi.salmon, sampleTable, design =  ~ condition)
rowRanges(dds) <- genes(EnsDb.Hsapiens.v75)[rownames(dds)]
seqlevelsStyle(dds) <- 'UCSC'
head(assay(dds))
```

## Pre-filtering

```{r filter_cells}
# Selecting only the relevant cell lines
dds <- dds[, colData(dds)$condition %in% c('Helas3','Hepg2')]
dds$condition <- droplevels(dds$condition)
dds$condition <- relevel(dds$condition, ref = "Helas3")
head(assay(dds))
```

```{r filter_counts}
# Filtering lowly expressed genes
keep.cts <- rowSums(counts(dds)) >= 25
dds <- dds[keep.cts,]
head(assay(dds))
```

```{r filter_protein}
# Keeping only protein coding genes
keep.coding <- rowRanges(dds)$gene_biotype == 'protein_coding'
dds <- dds[keep.coding,]
dim(dds)
```

```{r filter_chr}
# Selecting only relevant chromosomes (chrY is not present here)
keep.chr <- seqnames(dds) %in% paste0('chr',c(1:22,'X'))
dds <- dds[keep.chr,]
dim(dds)
```

## Visualization

```{r heatmap1}
# Sample-to-sample distances
vsd <- vst(dds, blind=FALSE)
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$condition, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(RColorBrewer::brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```

## DGE

Now, I will run DESeq2 to find DEG between the two cell lines. I expect to see lots of DGE genes since we are comparing two very distinct cell lines.

```{r deseq2}
# Calling DESeq2
dds <- DESeq(dds)
res <- results(dds)
res

# How many DGE are there? Let's set a p-value cutoff of 0.01
cutoff <- 0.01
sum(res$padj<cutoff)
```

Let's take a look at a heatmap of some of these genes.

```{r heatmap2}
# Normalizing cts
ntd <- normTransform(dds)

# Selecting some genes with largest absolute LFC differences for visualization
sig.res <- res[res$padj<cutoff,]
sig.res <- sig.res[order(sig.res$log2FoldChange),]
select <- c(rownames(head(sig.res,100)),rownames(tail(sig.res,100)))

# Set up annotation
df <- as.data.frame(colData(dds)[,c("condition")])
colnames(df) <- 'condition'
rownames(df) <- colnames(dds)

# Plotting
pheatmap(assay(ntd)[select,],cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)
```

# ChIP-seq data preprocessing

I will follow the same analysis steps for RNA-seq data here. H3K36me3 is expected to be enriched only on (expressed?) gene bodies, so it is reasonable to use gene bodies as an annotation reference. To this end, I will compute ChIP-seq read counts mapping onto the gene bodies as suggested in the DESeq2 vignette [here](http://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#why-un-normalized-counts)

## Preparing the data

```{r bam}
# Reading in ChIP-seq bam files

files <- list.files('../../Data/','^wgEncode*.*.bam$',recursive = T,full.names = T)
files <- files[grep('H3K36me3',files)]
files <- files[grep('helas3|hepg2',files)]

names(files) <- c(paste0('Helas3.Rep',1:2),paste0('Hepg2.Rep',1:2))

files
```

```{r bamsignals}
# Counting reads with bamsignals
chip.cts <- list()
for (i in seq_len(length(files))) {
  chip.cts[[i]] <- bamsignals::bamCount(files[i],rowRanges(dds))
}
chip.cts <- do.call(cbind,chip.cts)
colnames(chip.cts) <- names(files)
rownames(chip.cts) <- rownames(dds)

head(chip.cts)
```

```{r deseq2_object2}
# Creating DESeq2 object
sampleTable <-
  data.frame(condition = factor(unlist(lapply(
    strsplit(colnames(chip.cts), "\\."),
    FUN = function(x) {
      x[1]
    }
  ))))
rownames(sampleTable) <- colnames(chip.cts)
sampleTable

chip.dds <-
  DESeqDataSetFromMatrix(chip.cts, sampleTable, design =  ~ condition)
head(assay(chip.dds))
```

At this point, `chip.dds` already has all necessary information we need. Now, note that there are a few genes without any mapped ChIP-seq counts.

```{r nocts}
# Genes without H3K36me3 counts
nocts <- (rowSums(assay(chip.dds)) == 0)

# How many are there?
nrow(chip.dds[nocts,])
```

See below that the expression levels of these genes can be, sometimes, non negligible.

```{r heatmap_nocts}
# Plotting
pheatmap(assay(ntd[row.names(chip.dds[nocts,]),]),cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)
```

In any case, I will keep these genes in for now. When computing fold changes, they will not be considered, though.

## Visualization

```{r heatmap1.1}
# Sample-to-sample distances
vsd <- vst(chip.dds, blind=FALSE)
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$condition, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(RColorBrewer::brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```

At this point, we already start seeing some problems. Samples from Helas3 tend to be more similar between each other than with samples from Hepg2. However, note that one of the Hepg2 samples is as similar to its isogenic replicate as to replicates from Helas3. Maybe it is a poor quality ChIP-seq sample?

There is definitely a big difference in depth between cell lines, but I don't think it can justify alone the problematic case. The sample Hepg2.Rep2 also has very low depth but it is quite different from the Helas3 replicates from the distance heatmap above.

```{r lowdepth}
# Quick check of sequencing depth
colSums(assay(chip.dds))/1e6
```

## Differential enrichment

I am looking for a simple metric to compare DEG and differential enrichment for H3K36me3. I will use LFC to this end. If differential enrichment for H3K36me3 is necessary and sufficient to observe DGE, then the LFC of RNA-seq and ChIP-seq should be in agreement.

Now, I will run DESeq2 to get the LFC estimates from ChIP-seq data.

```{r deseq2.1}
# Calling DESeq2
chip.dds <- DESeq(chip.dds)
chip.res <- results(chip.dds)
chip.res

# How many differentially enriched genes are there? Using a p-value cutoff of 0.01
## I am removing the 56 genes where there was no enrichment signal for H3K36me3
sum(chip.res$padj<cutoff,na.rm = TRUE)
```

The analysis above is telling us that there are about 1500 genes with differential enrichment for H3K36me3. This number is about 5 times smaller than the number of DEG genes.

# Results

## Do log fold changes between ChIP-seq and RNA-seq agree?

I will compute shrunken LFC for better visualization and ranking of genes as detailed [here](http://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#log-fold-change-shrinkage-for-visualization-and-ranking).

```{r results1}
shk.chip.res <- lfcShrink(chip.dds,type="apeglm",coef = 'condition_Hepg2_vs_Helas3')
shk.res <- lfcShrink(dds,type="apeglm",coef = 'condition_Hepg2_vs_Helas3')

# First, make sure we are comparing the same genes
all.equal(rownames(shk.chip.res),rownames(shk.res))

dt.lfc <- data.table(Gene = rownames(shk.res),
                     LFC.RNAseq = shk.res$log2FoldChange,
                     LFC.ChIPseq = shk.chip.res$log2FoldChange,
                     AdjP.RNAseq = shk.res$padj,
                     AdjP.ChIPseq = shk.chip.res$padj)

ggplot(dt.lfc) +
  geom_point(aes(x = LFC.RNAseq,y = LFC.ChIPseq),alpha = 0.1) +
  theme_bw()

ggplot(dt.lfc) +
  geom_point(aes(x = LFC.RNAseq,y = LFC.ChIPseq,color = (AdjP.RNAseq<cutoff)),alpha = 0.25) +
  scale_color_brewer(palette = 'Set1',direction = -1)+
  theme_bw()

ggplot(dt.lfc) +
  geom_point(aes(x = LFC.RNAseq,y = LFC.ChIPseq,color = (AdjP.ChIPseq<cutoff)),alpha = 0.25) +
  scale_color_brewer(palette = 'Set1',direction = -1)+
  theme_bw()

```

We can see that there is a positive association between LFC of RNA- and ChIP-seq counts. However, several genes exhibit no association between RNA-seq and ChIP-seq counts.

```{r results2}
# Selecting problematic genes
dt.lfc.issues <- dt.lfc[((AdjP.RNAseq<cutoff) & (AdjP.ChIPseq>cutoff)) |
                          ((AdjP.RNAseq>cutoff) & (AdjP.ChIPseq<cutoff)),]

ggplot(dt.lfc.issues) +
  geom_point(aes(x = LFC.RNAseq,y = LFC.ChIPseq),alpha = 0.1) +
  theme_bw()

ggplot(dt.lfc.issues) +
  geom_point(aes(x = LFC.RNAseq,y = LFC.ChIPseq,color = (AdjP.ChIPseq<cutoff)),alpha = 0.25) +
  scale_color_brewer(palette = 'Set1',direction = -1)+
  theme_bw()
```

## UCSC Genome Browser view os problematic cases

Below, I present a few snap shots from the UCSC Genome Browser of problematic cases from the `dt.lfc.issues` table above.

### Example 1: Gene ENSG00000113389

Here we have an example of a DE gene without differential enrichment of H3K36me3. Note that, in Helas3, enrichment on the gene body is no different from enrichment on the intergenic (background) region. Therefore, observed differences in ChIP-seq are simply a sequencing depth artifact and would have to be accounted by the peak caller with proper normalization methods. If used as a benchmarking mark, differential peaks covering this genes would be incorrectly considered as a true positive calls.

![Example 1 (Gene ENSG00000113389)](ENSG00000113389.pdf)

\newpage

### Example 2 (Gene ENSG00000137285)

Another problematic example here. We have a gene that is expressed in both cell lines but the gene expression level is different.

There are two problems with the example. First, DEG level does not lead to differential ChIP-seq enrichment. The enrichment for H3K36me3 is similar in both cell lines. Second, it appears that the enrichment of H3K36me3 reads on the gene body is no different than intergenic (background) region.

If considered as a benchmarking mark, this gene would be considered as a true positive. One would expect differential peak callers to not call differential peaks on this region. If so, this gene would lead to an increased false negative rate in methods that correctly call this region as consensus background.

![Example 2 (Gene ENSG00000137285)](ENSG00000137285.pdf)

\newpage

### Example 3 (Gene ENSG00000139998)

This is another similar example of DEG not leading to differential H3K36me3.

However, there is one additional point to make here. If the gene below is DEG, and one assumes that there is a local ChIP-seq signal difference in between cell lines on the exonic regions, how do we justify the similar ChIP-seq coverage on the remaining regions on the gene body (right most part of the figure) between the two cell lines. Should differential peaks covering those regions be true positive calls (because they cover a DE gene)? Or should they be false positive calls (because there is no difference in ChIP-seq signal)? I believe the latter is a more reasonable choice in this example.

![Example 3 (Gene ENSG00000139998)](ENSG00000139998.pdf)

\newpage

### Example 4 (Gene ENSG00000144852)

Another example of DE gene not implying differential enrichment of H3K36me3.

![Example 4 (Gene ENSG00000144852)](ENSG00000144852.pdf)

\newpage

### Example 5 (Gene ENSG00000160202)

Another example of DE gene not implying differential enrichment of H3K36me3.

![Example 5 (Gene ENSG00000160202)](ENSG00000160202.pdf)

\newpage

### Example 6 (Gene ENSG00000178498)

Another example of DE gene not implying differential enrichment of H3K36me3.

![Example 6 (Gene ENSG00000178498)](ENSG00000178498.pdf)

\newpage

### Example 7 (Gene ENSG00000198610)

Another example of DE gene not implying differential enrichment of H3K36me3.

![Example 7 (Gene ENSG00000198610)](ENSG00000198610.pdf)

\newpage

# Summary

Based on the results above, there seems to be a positive association between DGE and differential enrichment for H3K36me3 when comparing Helas3 and Hepg2 cell lines. However, it is straightforward to find examples where DGE does not lead to differential ChIP-seq enrichment. When comparing methods regarding sensitivity/specificity, including these problematic genes in the set of gold-standard marks would lead to an invalid benchmarking study. Therefore, utilizing DE genes as a metric to compare methods regarding the genome-wide sensitivity/specificity of differential peak calls is not ideal. To pursue with this approach, one would need to manually curate the gene list while utilizing ChIP-seq data to determine the gold-standard set of true positive and true negative genes.

As detailed in our paper, we defined true and positive genes as those with differential enrichment for H3K36me3 ChIP-seq read counts. While this approach is not ideal either, it does not introduce problematic examples as those shown above. If one is concerned about the cutoff choice to define the gold-standard set of genes based on ChIP-seq counts, we show in our Supplementary Materials that our method, epigraHMM, outperforms all current differential peak callers in terms of sensitivity/specificity regardless the chosen LFC cutoff.

Lastly, it is important to note here that, as H3K36me3 is a broad epigenomic mark, the calculation of sensitivity/specificity of differential peak callers should not be based on 'hits on genes'. In other words, a differential peak hitting a DE gene does not imply that the differential peak is properly covering the entire differential region of ChIP-seq read enrichment. While the 'hits on genes' could be a valid approach to compute sensitivity/specificity from *short* marks (which is, by the way, the strategy used by many other papers; see csaw's paper from Lun & Smyth (2015) for an example), it is not ideal for *broad* marks for which the enrichment of counts can be quite complex and expand through large genomic domains. Therefore, the calculation of sensitivity/specificity for marks such as H3K36me3 should be based on either the average proportion of DE gene coverage by differential peaks or on the standard window-based calculation of sensitivity/specificity. In our paper, we utilize the latter strategy.

### 2021-03-03 Latex/paper/paper.tex
txt update

@@
 % Paragraph 1: Why does epigenetics matter? Nail down the problem 
-Epigenomics, the study of the human genome and its interactions with proteins and other cellular elements, has become of significant interest in recent years.
-Such interactions have been shown to regulate essential cellular functions such as gene expression and DNA packaging \citep{kim2018polycomb}, resulting in downstream phenotypic impact.
-Therefore, the interrogation of how these interactions may change across conditions, such as cell types or treatments, is of marked interest in biomedical research.
-Several landmark articles have identified specific genomic regions of changing (differential) epigenomic activity between conditions as drivers of cell differentiation \citep{creyghton2010histone}, cancer progression \citep{varambally2002polycomb}, and a number of human diseases \citep{portela2010epigenetic}.
-Within differential regions, the delineation of specific patterns of change across conditions is also of interest, for example classifying the gain-of- or loss-of-activity in genomic loci due to treatment \citep{clouaire2014cfp1}.
-The identification of specific combinations of processes acting locally may also be informative, such as for segmenting the genome into regulatory states \citep{kundaje2015integrative}.
+Epigenomics, the study of the genome and its interactions with proteins and cellular elements, has become of significant interest in recent years.
+Such interactions may regulate essential cellular functions, such as gene expression, resulting in downstream phenotypic impact \citep{kim2018polycomb}.
+Hence, the interrogation of how these interactions may change across conditions, such as cell types or treatments, is of marked interest in biomedical research.
+Several landmark articles have identified genomic regions of changing (differential) epigenomic activity between conditions as drivers of cell differentiation \citep{creyghton2010histone} and a number of human diseases \citep{portela2010epigenetic}.
+Within differential regions, delineating specific patterns of change across conditions is also of interest, for example classifying the gain-of- or loss-of-activity in genomic loci due to treatment \citep{clouaire2014cfp1}.
+%The identification of specific combinations of processes acting locally may also be informative, such as for segmenting the genome into regulatory states \citep{kundaje2015integrative}.
 
 % Paragraph 3: Introduce ChIP-seq
-To quantify local epigenomic activity on a genome-wide scale, a common high-throughput assay is chromatin immunoprecipitation followed by massively parallel sequencing (ChIP-seq).
-ChIP-seq experiments begin with cross-linking DNA and proteins within chromatin structures, which are then fragmented by sonication in a particular sample.
-DNA fragments bound to the protein of interest are isolated by chromatin immunoprecipitation, which are then sequenced via massively parallel high-throughput sequencing to generate short sequencing reads pertaining to the original fragments.
-Sequences are then mapped onto a reference genome through sequence alignment to determine their likely locations of origin.
-Genomic coordinates containing a high density of mapped reads, often referred to as enrichment regions (peaks), indicate likely locations of protein-DNA interaction sites, and all other regions are referred to as background regions.
-This local read density is often summarized by counting the number of reads mapped onto non-overlapping windows of fixed length tiling the genome (window read counts), forming the basis for downstream analyses.
+To quantify local epigenomic activity, a common high-throughput assay is chromatin immunoprecipitation followed by massively parallel sequencing (ChIP-seq).
+ChIP-seq begins with cross-linking DNA and proteins within chromatin structures, which are then fragmented by sonication.
+DNA fragments bound to the protein of interest are isolated by immunoprecipitation and sequenced via high-throughput sequencing to generate short reads pertaining to the original fragments.
+Sequences are then mapped onto a reference genome to determine their likely locations of origin.
+Genomic coordinates containing a high density of mapped reads (enrichment regions, peaks) indicate likely locations of protein-DNA interaction sites, while other regions are referred to as background regions.
+This read density is often summarized by counting reads mapped onto non-overlapping windows of fixed length (window read counts), forming the basis for downstream analyses.
 Across multiple conditions, regions exhibiting enrichment in at least one condition, but not across all conditions, indicate the presence of differential activity pertaining to the protein-DNA interaction of interest.
@@
 % Paragraph 4: Issues with the data/methods
-To date, many differential peak callers (DPCs) have been proposed \citep{song2011identifying,stark2011diffbind,shen2013diffreps,chen2015novel,lun2015csaw,allhoff2016differential}.
-However, several challenges affect their ability to accurately detect regions of differential activity from the wide range of ChIP-seq experiments (Section \ref{sec:background}).
-First, differential regions may be both short or broad in length, causing difficulty for methods optimized for a particular type of signal profile \citep{stark2011diffbind,chen2015novel}.
-Second, methods that pool experimental replicates together \citep{song2011identifying} often exhibit more false positive calls compared to methods that jointly model replicates from each condition \citep{steinhauser2016comprehensive}.
-Third, the analysis of ChIP-seq data is often subject to complex biases that may vary across the genome, as differences in local read enrichment may depend on quantities such as the total read abundance in a given region or GC content \citep{teng2017accounting}.
-DPCs that disregard such complex effects or that solely rely on sample-specific global scaling factors or control subtraction methods \citep{shen2013diffreps,chen2015novel,allhoff2016differential} may be prone to detecting spurious differences due to the lack of non-linear normalization methods \citep{lun2015csaw}.
-Reflecting these limitations, a recent comparison of DPCs demonstrated that current methods tend to detect either a large number of short peaks (low sensitivity) or exhibit a high number of false positive calls (low specificity) in ChIP-seq experiments with broad regions of enrichment \citep{steinhauser2016comprehensive}.
-Moreover, only few methods are able to simultaneously test for differential activity across three or more conditions \citep{stark2011diffbind,chen2015novel,lun2015csaw}, or can classify specific differential combinatorial patterns.
-Altogether, these limitations can impact the drawing of accurate insights from modern epigenomic studies.
+To date, many differential peak callers (DPCs) have been proposed.
+However, several challenges affect their ability to accurately detect regions of differential activity from the wide range of experiments (Section \ref{sec:background}).
+First, differential regions may be both short or broad, causing difficulty for methods optimized for a particular type of peak profile \citep{stark2011diffbind,chen2015novel}.
+Second, pooling experimental replicates often leads to lower specificity in comparison to joint modeling samples from each condition \citep{song2011identifying}.
+Third, ChIP-seq data is often subject to complex biases that vary across the genome, as differences in local read enrichment may depend on quantities such as local read abundance or GC content \citep{teng2017accounting}.
+DPCs that ignore such effects, or rely on global scaling factors or control subtraction methods \citep{shen2013diffreps,allhoff2016differential}, may be subject to spurious differences due to the lack of non-linear normalization methods \citep{lun2015csaw}.
+In ChIP-seq data with broad peaks, a recent comparison showed that methods often detect a large number of either short or false positive peaks \citep{steinhauser2016comprehensive}.
+Moreover, only few methods can detect or classify differential combinatorial patterns across any number of conditions \citep{stark2011diffbind,chen2015novel,lun2015csaw}.
 
 % Paragraph 4: What we propose? Broad peaks, mixture model, and multiple conditions
-Here, we propose an efficient and flexible statistical method to identify differential regions of enrichment from epigenomic experiments with diverse signal profiles and collected under common multi-replicate, multi-condition settings.
-Our method overcomes the limitations of current DPCs with three major features.
-First, it uses a hidden Markov model (HMM) to account for the diversity in differential enrichment profiles that may result from short and broad epigenomic ChIP-seq data sets.
-Second, it captures specific differential combinatorial patterns through a novel finite mixture model emission distribution within the HMM's differential state.
-Each mixture component pertains to a particular differential combinatorial pattern that is formed by the presence or absence of local enrichment across conditions, where a generalized linear model (GLM) is used to model the specific differential combinatorial pattern while accounting for sample- and window-specific normalization factors via offsets.
-Third, it enables the simultaneous detection and classification of epigenomic changes under three or more conditions, a novelty not yet available in any other DPC algorithm.
-The presented method, epigraHMM (Figure~\ref{fig:fig_motivation}), offers additional benefits over current HMM-based DPC algorithms \citep{song2011identifying,allhoff2016differential} that includes a flexible GLM-based framework with an embedded mixture model, which allows the modeling of covariates of interest, the inclusion of model offsets for non-linear normalization and GC-bias adjustment, and a fast and accurate parameter estimation scheme via rejection-controlled EM algorithm (RCEM).
+Here, we present epigraHMM (Figure~\ref{fig:fig_motivation}), an efficient and flexible statistical method to identify differential regions of enrichment from epigenomic experiments with diverse signal profiles.
+We overcome the limitations of current DPCs with three major features.
+First, it uses a hidden Markov model (HMM) to account for the diversity in differential enrichment from broad and short ChIP-seq data collected under multi-replicate, multi-condition settings.
+Second, it models specific combinatorial patterns of enrichment via a finite mixture model emission distribution within the HMM differential state.
+Each mixture component pertains to a particular differential combinatorial pattern originated by the presence or absence of local enrichment across conditions, where a generalized linear model (GLM) encodes specific differential combinatorial patterns while accounting for sample- and window-specific normalizing offsets.
+Third, it simultaneously detects and classifies epigenomic changes under three or more conditions, a novelty not yet available in any other DPC algorithm.
+epigraHMM offers additional benefits over current HMM algorithms \citep{song2011identifying,allhoff2016differential} with an embedded GLM mixture model that allows the modeling of covariates of interest, the inclusion of normalizing offsets for non-linear biases (such as GC-content bias), and a fast parameter estimation scheme via rejection-controlled EM algorithm (RCEM).
 
@@
 % What are histones and why are they important
-Histones are proteins that interact and condense  DNA in eukaryotic cells into structural units called nucleosomes.
-Multiple types of enzymatic modifications may be applied to histones, resulting in changes in local DNA packaging and chromatin accessibility mediated by nucleosomes \citep{bannister2011regulation}.
-In turn, cellular processes such as gene transcription, gene silencing, DNA repair, replication, and recombination are also affected.
+Histones are proteins that condense the DNA in eukaryotic cells into units called nucleosomes and, when enzymatically modified, mediate changes in local DNA packaging and chromatin accessibility.
+Hence, cellular processes such as gene transcription, gene silencing, and DNA repair are also affected.
 Proteins that interact with DNA and alter its functional properties are often referred to as epigenomic marks.
-For example, the trimethylation of histone H3 at lysines 36 and 27 (H3K36me3 and H3K27me3) are two types of histone modifications that tend to occur in genomic loci containing actively transcribed and repressed genes \citep{liu2016distinct}, respectively, and exhibit broad enrichment profiles.
-These marks have been investigated in cancer studies, where their absence is often observed in multiple cancer types \citep{wei2008loss}.
-As a result, H3K36me3 and H3K27me3 are considered to be key prognostic indicators in patients with breast, ovarian, and pancreatic cancer.
-The enhancer of zeste homolog 2 (EZH2), a major component of the polycomb complex PRC2 that catalyzes the methylation of H3K27me3 \citep{margueron2011polycomb}, is another example of a protein with experimental signal characterized by broad enrichment domains and co-occurs with the activity of H3K27me3.
+The histone modifications H3K36me3 and H3K27me3 are example of marks that associate with genomic loci containing transcribed and repressed genes \citep{liu2016distinct}, respectively, while exhibiting broad enrichment profiles.
+The enhancer of zeste homolog 2 (EZH2), a component of the complex PRC2 that catalyzes the methylation of H3K27me3 \citep{margueron2011polycomb}, is another example of a protein characterized by broad enrichment domains and co-occurs with the activity of H3K27me3.
+Conversely, short peaks from histone modifications H3K27ac and H3K4me3 are usually deposited on the promoter regions of transcribed genes and several studies have associated their role with gene transcription \citep{creyghton2010histone}.
+Similarly, the transcription factor CTCF is a protein that binds to short DNA motifs and is responsible for cellular processes that include the regulation of the chromatin 3D structure \citep{shukla2011ctcf}.
 
-Using ChIP-seq data pertaining to histone modifications H3K27me3 and H3K36me3, and to EZH2 from the ENCODE Consortium, we find that current DPCs have difficulty in accurately detecting broad regions of differential enrichment between several common cell lines (Figure~\ref{fig:fig_motivation}).
+Using ChIP-seq data pertaining to histone modifications H3K27me3 and H3K36me3, and to EZH2 from the ENCODE Consortium \citep{landt2012chip}, we find that current DPCs have difficulty in accurately detecting broad regions of differential enrichment between several common cell lines (Figure~\ref{fig:fig_motivation}).
 In line with previous findings \citep{steinhauser2016comprehensive}, we observe that even current DPCs designed for broad data \citep{song2011identifying,allhoff2016differential} tend to detect either overly fragmented differential peaks or call regions exhibiting no difference in experimental signal between conditions as differential (Sections \ref{sec:sim2} and \ref{sec:broad}).
-The low specificity and sensitivity of such methods may impair the biological interpretation of the resulting peak calls in downstream analyses.
 Methods that rely on candidate peaks may also exhibit a compromised performance due to the limitations of single-sample peak callers in broad data \citep{stark2011diffbind,chen2015novel}.
-In addition, most current DPCs restrict their application to the analysis of two experimental conditions.
-For methods that are tailored for the analysis of three or more conditions, the classification of specific differential combinatorial patterns across conditions (or across various epigenomic processes) is still an open problem.
-The classification of such patterns would allow researchers to, for example, quantify treatment responses on the epigenomic level \citep{clouaire2014cfp1}, or identify sets of processes working together to regulate local chromatin state.
-We find that the performance of such methods exhibit low sensitivity and specificity in calling differential regions in broad marks (Figure~\ref{fig:fig_motivation}).
+Moreover, most DPCs restrict their application to the analysis of two experimental conditions.
+For DPCs tailored for the analysis of three or more conditions, the classification of specific differential combinatorial patterns across conditions (or across various epigenomic processes) is still an open problem.
+The classification of such patterns would allow researchers to quantify treatment responses on the epigenomic level or identify sets of processes working together to regulate local chromatin state \citep{clouaire2014cfp1}.
+
 \begin{figure}
@@
 
-We assessed the performance of our model on ChIP-seq experiments characterized by broad peaks (H3K36me3, H3K27me3, and EZH2) and short peaks (H3K27ac, H3K4me3, and the transcription factor CTCF).
-In simulations (Section \ref{sec:simulation}) and in data sets from the ENCODE Consortium \citep{landt2012chip}, we show that our model addresses the issues of the current peak callers in broad data (Section \ref{sec:broad}), while being flexible for short peaks (Section \ref{sec:sharp}) and comparable to the fastest DPCs regarding the computation time.
+We assessed the performance of our model on ChIP-seq experiments characterized by broad peaks (H3K36me3, H3K27me3, and EZH2) and short peaks (H3K27ac, H3K4me3, and CTCF).
+In simulated and in real data from the ENCODE Consortium (Sections \ref{sec:simulation} and \ref{sec:realdata}), our model addresses the issues of the current peak callers in broad data, while being flexible for short peaks and comparable to the fastest DPCs regarding the computation time.
 We show that our method can also be utilized for genomic regulatory state segmentation when studying multiple types of epigenomic processes from a single condition or cell line (Section \ref{sec:predpostprob}).
-The Web Appendices A1 and A2 present the data accession codes and the data pre-processing steps, respectively.
-Code implementing the method and to replicate the presented results are available in the Web Appendices A3 and A4, respectively.
+Web Appendices A1-A4 present data accession codes, data pre-processing steps, code implementing the method, and scripts to replicate the presented results, respectively.
 
@@
 \label{subsec:hmmmixt}
-Let $Y_{hij}$ denote the random variable pertaining to the ChIP read count for genomic window $j$ from sample $i$ of condition $h$, where $j=1,\ldots,M$, $i=1,\ldots,n_{h}$, $h=1,\ldots,G$, and let $y_{hij}$ be the observed count.
+Let $Y_{hij}$ denote the random variable pertaining to the read count for genomic window $j$ from sample $i$ of condition $h$, where $j=1,\ldots,M$, $i=1,\ldots,n_{h}$, $h=1,\ldots,G$, and let $y_{hij}$ be the observed count.
 Here, $n_h$ is the number of samples in condition $h$ and $N = \sum_{h = 1}^G n_h$ is the total number of samples across the $G$ conditions.
-At the $j^{th}$ window, let $\mb{y}_{..j}=(y_{11j},\ldots,y_{Gn_{G}j})^\prime$ denote the $N \times 1$ vector of ChIP window read counts across all samples and conditions, and let $\mb{y}=(\mb{y}_{..1},\ldots,\mb{y}_{..M})$ denote the corresponding $N \times M$ matrix of window read counts spanning all windows, samples, and conditions.
-We assume that each window belongs to one of three possible hidden states: consensus background (state 1), differential (state 2), and consensus enrichment (state 3).
+At the $j^{th}$ window, let $\mb{y}_{..j}=(y_{11j},\ldots,y_{Gn_{G}j})^\prime$ denote the $N \times 1$ vector of window read counts across all samples and conditions, and let $\mb{y}=(\mb{y}_{..1},\ldots,\mb{y}_{..M})$ denote the corresponding $N \times M$ matrix of window read counts spanning all windows, samples, and conditions.
+We assume that each window belongs to one of three possible hidden states: consensus background, differential, and consensus enrichment.
 Windows exhibiting low (high) enrichment across all conditions will be modeled by an emission distribution pertaining to the consensus background (enrichment) state.
@@
 
-To model transitions between states, we assume a single latent discrete time stationary Markov chain $\mb{Z}=\{Z_{j}\}_{j=1}^{M}$, $Z_{j}\in\{1,2,3\}$, with state-to-state transition probabilities $\bs{\gamma}=(\gamma_{11},\gamma_{12},\ldots,\gamma_{33})^\prime$ and initial probabilities $\bs{\pi}=(\pi_{1},\pi_{2},\pi_{3})^\prime$, such that $\sum_{s=1}^{3}\gamma_{rs}=1$ and $\sum_{s=1}^{3}\pi_{s}=1$ for $r\in\{1,2,3\}$.
-To facilitate the notation, let $f_{r}(\mb{y}_{..j}| \bs{\psi}_{r})$ denote the emission distribution corresponding  to the $r^{th}$ hidden state, where $\bs{\Psi} = (\bs{\pi}^\prime,\bs{\gamma}^\prime,\bs{\delta}^\prime,\bs{\psi}^\prime)^\prime$ denotes the vector of all model parameters, $\bs{\psi}=(\bs{\psi}_{1}^\prime,\bs{\psi}_{2}^\prime,\bs{\psi}_{3}^\prime)^\prime$ denotes each state's set of emission distribution-specific parameters, and $\mathcal{Z}$ denotes the set of $3^M$ possible state paths of length $M$.
-Then, the likelihood function pertaining to the proposed HMM may be written as
+We assume a single latent discrete time stationary Markov chain $\mb{Z}=\{Z_{j}\}_{j=1}^{M}$, $Z_{j}\in\{1,2,3\}$, with state-to-state transition probabilities $\bs{\gamma}=(\gamma_{11},\gamma_{12},\ldots,\gamma_{33})^\prime$ and initial probabilities $\bs{\pi}=(\pi_{1},\pi_{2},\pi_{3})^\prime$, such that $\sum_{s=1}^{3}\gamma_{rs}=1$ and $\sum_{s=1}^{3}\pi_{s}=1$ for $r\in\{1,2,3\}$.
+Let $f_{r}(\mb{y}_{..j}| \bs{\psi}_{r})$ denote the emission distribution corresponding to the $r^{th}$ hidden state, $\bs{\Psi} = (\bs{\pi}^\prime,\bs{\gamma}^\prime,\bs{\delta}^\prime,\bs{\psi}^\prime)^\prime$ denote the vector of all model parameters, $\bs{\psi}=(\bs{\psi}_{1}^\prime,\bs{\psi}_{2}^\prime,\bs{\psi}_{3}^\prime)^\prime$ denote each state's set of emission distribution-specific parameters, and $\mathcal{Z}$ denote the set of $3^M$ possible state paths of length $M$.
+The likelihood function of the proposed HMM may be written as
+
 \begin{align}\label{eq:difflik}
@@
 \end{align}
+
 Here, $\mb{x}$ is a fixed $G \times L$ design matrix enumerating each of the $L$ possible differential combinatorial patterns in terms of the presence or absence of enrichment across each of the $G$ conditions, only in the emission distribution of the differential state.
-
 We assume that read counts pertaining to genomic windows from the consensus background ($r=1$) and consensus enrichment  ($r=3$) states follow a Negative Binomial (NB) distribution with state-specific parameters $\bs{\psi}_{r}=(\mu_{(r,hij)},\phi_{r})^\prime$, with mean $\mu_{(r,hij)}$ and variance $\mu_{(r,hij)}(1+\mu_{(r,hij)}/\phi_{r})$.
-Assuming independence of read counts across experiments and samples, conditional upon the HMM state, the emission distribution of the consensus background and consensus enrichment states, respectively, can be written as
+Conditional on the HMM state, we assume independence of read counts across samples and write the emission distribution of the consensus background and enrichment states as
 \begin{align}\label{eq:nondiffdensity}
@@
 \end{align}
-with $y_{hij}\in\{0,1,2,\ldots\}$, such that $\log(\mu_{(1,hij)})=\beta_{1} + u_{hij}$, $\log(\phi_{1})=\lambda_{1}$, $\log(\mu_{(3,hij)})=\beta_{1} + \beta_{3} + u_{hij}$, and $\log(\phi_{3})=\lambda_{1} + \lambda_{3}$.
-The offset $u_{hij}$ adjusts for technical or biological artifacts (such as the GC-content bias as presented in \cite{teng2017accounting}, Web Appendix B1) and allows the non-linear normalization of the signal profile across genomic windows, conditions, and samples (Web Appendix B1).
-When $u_{hij}=0$, $\beta_{1}$ and $\lambda_{1}$ represent the log-mean and log-dispersion, respectively, of read counts pertaining to consensus background state windows, whereas $\beta_{3}$ and $\lambda_{3}$ represent the difference in log-mean and log-dispersion of read counts from consensus enrichment state windows relative to consensus background state windows.
+such that $\log(\mu_{(1,hij)})=\beta_{1} + u_{hij}$, $\log(\phi_{1})=\lambda_{1}$, $\log(\mu_{(3,hij)})=\beta_{1} + \beta_{3} + u_{hij}$, and $\log(\phi_{3})=\lambda_{1} + \lambda_{3}$.
+The offset $u_{hij}$ may adjust for technical or biological artifacts such as the GC-content bias and allows the non-linear normalization of read counts across genomic windows, conditions, and replicates (Web Appendix B1; \cite{teng2017accounting}).
+%When $u_{hij}=0$, $\beta_{1}$ and $\lambda_{1}$ represent the log-mean and log-dispersion, respectively, of read counts pertaining to consensus background state windows, whereas $\beta_{3}$ and $\lambda_{3}$ represent the difference in log-mean and log-dispersion of read counts from consensus enrichment state windows relative to consensus background state windows.
 
-For windows belonging to the differential state ($r = 2$), we assume that the corresponding read counts are modeled by a $L$-component finite mixture model with mixture components that follow a Negative Binomial distribution, where each component corresponds to a particular differential combinatorial pattern.
-To define these patterns, let us consider the sets $S_{1},\ldots,S_{L}$ that delineate the subset of the $G$ conditions that are enriched in each of the $L$ differential combinatorial patterns.
-For instance, if $G=3$, the sets $S_{1} =\{1\}$, $S_{2} =\{2\}$, $S_{3} =\{3\}$, $S_{4} =\{1,2\}$, $S_{5} =\{1,3\}$, and $S_{6} =\{2,3\}$ define the six possible differential combinatorial patterns of enrichment and background across three conditions.
-That is, the set  $S_{1}$ denotes enrichment in only the first condition and background in all others, whereas the set $S_{6}$ denotes enrichment in conditions 2 and 3 and background in condition 1.
+In the differential state ($r = 2$), read counts are modeled by a $L$-component finite mixture model with mixture components that follow a Negative Binomial distribution, where each component corresponds to a particular differential combinatorial pattern.
+To define these patterns, consider sets $S_{1},\ldots,S_{L}$  delineating the subset of the $G$ conditions that are enriched in each of the $L$ differential combinatorial patterns.
+For instance, if $G=3$, the sets $S_{1} =\{1\}$, $S_{2} =\{2\}$, $S_{3} =\{3\}$, $S_{4} =\{1,2\}$, $S_{5} =\{1,3\}$, and $S_{6} =\{2,3\}$ define the six possible differential combinatorial patterns of enrichment and background across three conditions (\textit{e.g.}, $S_{6}$ denotes enrichment in conditions 2 and 3 and background in condition 1).
 The presence or absence of enrichment in each of the $L$ sets is encoded into each column of $\mb{x}=(\mb{x}_{1},\ldots,\mb{x}_{L})$, such that $\mb{x}_{l} = (x_{1l},\ldots,x_{Gl})'$, and $x_{hl} = I(h\in S_{l})$ for $l=1,\ldots,L$ and $h = 1,\ldots,G$.
 That is, $\mb{x}_{l}$ is the $G \times 1$ vector of binary indicator variables denoting which subset of conditions are enriched in pattern (mixture component) $l$.
-
-Let $\bs{\psi}_{2}$ denote the state-specific parameter vector pertaining to the differential state and let $\bs{\psi}_{(2,l)}$ denote the set of parameters pertaining to the $l^{th}$ mixture component.
-Assuming independence of read counts across conditions and samples, conditional upon the differential HMM state, the finite mixture model emission distribution can be written as
+Let $\bs{\psi}_{2}$ and $\bs{\psi}_{(2,l)}$ denote the parameter vectors pertaining to the differential state and to the $l^{th}$ mixture component, respectively.
+Conditional on the differential state, we assume independence of read counts across samples and write the finite mixture model emission distribution as
 \begin{align}\label{eq:diffdensity}
-f_{2}(\mb{y}_{..j}|\mb{x};\bs{\delta},\bs{\psi}_{2}) &= \sum_{l=1}^{L}\delta_{l}\left[\prod_{h=1}^{G}\prod_{i=1}^{n_{h}}\frac{\Gamma\left(y_{hij}+\phi_{(2,l,h)}\right)}{y_{hij}!\Gamma\left(\phi_{(2,l,h)}\right)}\left(\frac{\phi_{(2,l,h)}}{\mu_{(2,l,hij)}+\phi_{(2,l,h)}}\right)^{\phi_{(2,l,h)}}\times\right.\nonumber\\
-&\left.\times\left(\frac{\mu_{(2,l,hij)}}{\mu_{(2,l,hij)}+\phi_{(2,l,h)}}\right)^{y_{hij}}\right],\quad y_{hij}\in\{0,1,2,\ldots\}, % &= \sum_{l=1}^{L}\delta_{l}f_{(2,l)}\left(\mb{y}_{..j}|\mb{x}_l;\bs{\psi}_{(2,l)}\right),\nonumber\\
+\resizebox{.925\hsize}{!}{$f_{2}(\mb{y}_{..j}|\mb{x};\bs{\delta},\bs{\psi}_{2}) = \sum_{l=1}^{L}\delta_{l}\prod_{h=1}^{G}\prod_{i=1}^{n_{h}}\frac{\Gamma\left(y_{hij}+\phi_{(2,lh)}\right)}{y_{hij}!\Gamma\left(\phi_{(2,lh)}\right)}\left(\frac{\phi_{(2,lh)}}{\mu_{(2,lhij)}+\phi_{(2,lh)}}\right)^{\phi_{(2,lh)}}\left(\frac{\mu_{(2,lhij)}}{\mu_{(2,lhij)}+\phi_{(2,lh)}}\right)^{y_{hij}},$}
 \end{align}
-where $\mu_{(2,l,hij)}$ and $\phi_{(2,l,h)}$ are the mean and dispersion, respectively, pertaining to read counts originating from window $j$ and sample $i$ in condition $h$ from the mixture component $l$.
-We assume that $\log(\mu_{(2,l,hij)}) = \beta_{1} + \beta_{3}x_{hl} + u_{hij}$ and $\log(\phi_{(2,l,h)}) = \lambda_{1} + \lambda_{3}x_{hl}$.
-That is, in the mixture component $l$, we utilize the same consensus background (consensus enriched) log-mean and log-dispersion from (\ref{eq:nondiffdensity}) in all conditions that are specified by $\mb{x}_{l}$ to be background (enriched) in the $l^{th}$ differential combinatorial pattern.
-There are several advantages to such a parametrization for the differential emission distribution.
-For example, it ensures that windows exhibiting differential enrichment across conditions share means and dispersions that are common between the consensus background and consensus enrichment states, a reasonable assumption that significantly increases computational efficiency.
-Utilizing a mixture model as the differential state emission distribution avoids the computational burden that would come from assuming separate hidden states for each of the $L$ differential combinatorial patterns, particularly as $G$ increases.
+where $\mu_{(2,lhij)}$ and $\phi_{(2,lh)}$ are the mean and dispersion, respectively, pertaining to read counts originating from window $j$ and sample $i$ in condition $h$ from the mixture component $l$.
+We assume that $\log(\mu_{(2,lhij)}) = \beta_{1} + \beta_{3}x_{hl} + u_{hij}$ and $\log(\phi_{(2,lh)}) = \lambda_{1} + \lambda_{3}x_{hl}$.
+That is, in the mixture component $l$, we utilize the same consensus background (consensus enriched) parametrization from (\ref{eq:nondiffdensity}) in all conditions that $\mb{x}_{l}$ specifies to be background (enriched) in the $l^{th}$ differential combinatorial pattern.
+Such a parametrization ensures that windows exhibiting differential enrichment across conditions share common means and dispersions between the consensus background enrichment states, an assumption that increases computational efficiency.
+Utilizing a mixture model as the differential state emission distribution avoids the computational burden that would come from assuming separate hidden states for each of the $L$ differential combinatorial patterns.
 We evaluate the strength of these assumptions through multiple simulations and a real data benchmarking analysis in Sections \ref{sec:simulation} and \ref{sec:realdata}.
-
-Two novel features result from our proposed approach that are relevant to the context of differential enrichment detection from ChIP-seq experiments.
-By using a modified version of the Expectation-Maximization (EM) algorithm to estimate the model parameters, we are able not only to detect differential enrichment regions across multiple conditions,  but we can also classify various differential combinatorial patterns of enrichment within broad and short differential enrichment domains.
-With state-specific parameters, the current implementation of the method allows the direct modeling of continuous covariates (e.g.
-input controls; Web Appendix A3), for which a state-level testing of their effects on the read count distribution could be performed.
-In a simulation study and in real data analyses, however, we did not observe a significant improvement in performance in differential peak detection after accounting for the effect of input controls (Web Appendix B), a fact that has also been observed by others \citep{lun2015csaw}.
 
@@
 \label{subsec:hmmest}
-To simplify the parameter estimation in (\ref{eq:obslik}), we introduce another set of latent variables $\mb{W}=(\mb{W}_{1}^\prime,\ldots,\mb{W}_{M}^\prime)^\prime$, such that $\mb{W}_{j}=(W_{j1},\ldots,W_{jL})^\prime$ for $j=1,\ldots,M$.
+Here, consider a set of latent variables $\mb{W}=(\mb{W}_{1}^\prime,\ldots,\mb{W}_{M}^\prime)^\prime$, such that $\mb{W}_{j}=(W_{j1},\ldots,W_{jL})^\prime$ for $j=1,\ldots,M$.
 We assume that $\mb{W}$ is a sequence of independent random vectors such that $\mb{W}_{j}|(Z_{j}=2)\sim\text{Multinomial}(1,\bs{\delta})$ and $\mb{W}_{j}|(Z_{j}=r)=0$ with probability 1 if $r=\{1,3\}$.
-Under this setup, one may define the data generating mechanism when $Z_{j}=2$ (differential state) and $W_{jl}=1$ ($l^{th}$ differential combinatorial pattern) such that read counts pertaining to genomic window $j$ are sampled from $f_{(2,l)}$ given $\bs{\psi}_{(2,l)}$ and $\mb{x}_{l}$.
-Let $\mathcal{W}$ denote the set of $L^M$ possible combinations of latent vectors $\mb{W}$.
-Hence, the likelihood function of the observed data (\ref{eq:difflik}) can be rewritten as
+The data generating mechanism can be then interpreted as read counts pertaining to window $j$ being sampled from $f_{(2,l)}$, given $\bs{\psi}_{(2,l)}$ and $\mb{x}_{l}$, when $Z_{j}=2$ (differential state) and $W_{jl}=1$ ($l^{th}$ differential combinatorial pattern).
+Denoting $\mathcal{W}$ as the set of $L^M$ possible combinations of latent vectors $\mb{W}$, the likelihood function of the observed data (\ref{eq:difflik}) can be rewritten as
+
 \begin{align}\label{eq:obslik}
-f(\mathbf{y}|\mathbf{x};\bs{\Psi}) &= \sum_{\mb{Z}\in\mathcal{Z}}\sum_{\mb{W}\in\mathcal{W}}\left\{\left[\prod_{r=1}^{3}\pi_{r}^{I(Z_{1}=r)}\prod_{j=2}^{M}\prod_{r=1}^{3}\prod_{s=1}^{3}\gamma_{rs}^{I(Z_{j-1}=r,Z_{j}=s)}\right.\right]\times\left[\prod_{j=1}^{M}\left(\prod_{l=1}^{L}\delta_{l}^{W_{jl}}\right)^{I(Z_{j}=2)}\right]\times\nonumber\\
-&\times \left.\left[\prod_{j=1}^{M}f_{1}\left(\mb{y}_{..j}|\bs{\psi}_{1}\right)^{I(Z_{j}=1)}\left(\prod_{l=1}^{L}f_{(2,l)}(\mb{y}_{..j}|\mb{x}_l;\bs{\psi}_{(2,l)})^{W_{jl}}\right)^{I(Z_{j}=2)}f_{3}\left(\mb{y}_{..j}|\bs{\psi}_{3}\right)^{I(Z_{j}=3)}\right]\right\}, %&= \sum_{\mb{Z}\in\mathcal{Z}}\left\{\left[\sum_{\mb{W}\in\mathcal{W}}f(\mb{y}|\mb{w},\mb{z},\mb{x};\bs{\psi})f(\mb{w}|\mb{z},\mb{x};\bs{\delta})\right]f(\mb{z}|\mb{x};\bs{\pi},\bs{\gamma})\right\},\\
+f(\mathbf{y}|\mathbf{x};\bs{\Psi}) &= \sum_{\mb{Z}\in\mathcal{Z}}\sum_{\mb{W}\in\mathcal{W}}\left\{\left[\prod_{r=1}^{3}\pi_{r}^{I(Z_{1}=r)}\prod_{j=2}^{M}\prod_{r=1}^{3}\prod_{s=1}^{3}\gamma_{rs}^{I(Z_{j-1}=r,Z_{j}=s)}\right.\right]\prod_{j=1}^{M}\prod_{l=1}^{L}\delta_{l}^{W_{jl}I(Z_{j}=2)}\times\\
+&\times \left.\left[\prod_{j=1}^{M}f_{1}\left(\mb{y}_{..j}|\bs{\psi}_{1}\right)^{I(Z_{j}=1)}\left(\prod_{l=1}^{L}f_{(2,l)}(\mb{y}_{..j}|\mb{x}_l;\bs{\psi}_{(2,l)})^{W_{jl}}\right)^{I(Z_{j}=2)}f_{3}\left(\mb{y}_{..j}|\bs{\psi}_{3}\right)^{I(Z_{j}=3)}\right]\right\},\nonumber %&= \sum_{\mb{Z}\in\mathcal{Z}}\left\{\left[\sum_{\mb{W}\in\mathcal{W}}f(\mb{y}|\mb{w},\mb{z},\mb{x};\bs{\psi})f(\mb{w}|\mb{z},\mb{x};\bs{\delta})\right]f(\mb{z}|\mb{x};\bs{\pi},\bs{\gamma})\right\},\\
 \end{align}
@@
 In the E-step of the EM algorithm, we compute the posterior probabilities from (\ref{eq:diffQfunc}).
-The quantities $Pr\left(Z_{j}=r|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)$ and $Pr\left(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)$, defined in Web Appendix B4, can be calculated through the Forward-Backward algorithm and $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})=f_{(2,l)}(\mb{y}_{..j}|\mb{x}_l;\bs{\psi}_{(2,l)}^{(t)})\delta_{l}^{(t)}/\sum_{k=1}^{L}f_{(2,k)}(\mb{y}_{..j}|\mb{x}_k;\bs{\psi}_{(2,k)}^{(t)})\delta_{k}^{(t)}$ for $l=1,\ldots,L$.
+The quantities $Pr\left(Z_{j}=r|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)$ and $Pr\left(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)$ can be calculated through the Forward-Backward algorithm (Web Appendix B4) and $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})=f_{(2,l)}(\mb{y}_{..j}|\mb{x}_l;\bs{\psi}_{(2,l)}^{(t)})\delta_{l}^{(t)}/\sum_{k=1}^{L}f_{(2,k)}(\mb{y}_{..j}|\mb{x}_k;\bs{\psi}_{(2,k)}^{(t)})\delta_{k}^{(t)}$ for $l=1,\ldots,L$.
 
 The $Q$ function is maximized with respect to the parameters $\bs{\Psi}=(\bs{\pi}^\prime,\bs{\gamma}^\prime,\bs{\delta}^\prime,\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$ during the M-step of the algorithm.
-Estimates for the initial and transition probabilities can be directly calculated as $\hat{\pi}^{(t+1)}_{r}=Pr\left(Z_{1}=r|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)$ and $\hat{\gamma}^{(t+1)}_{rs}=\sum_{j=2}^{M}Pr(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t)})/\sum_{j=2}^{M}Pr(Z_{j-1}=r|\mb{y},\mb{x};\bs{\Psi}^{(t)})$, respectively, restricted to $\sum_{r=1}^{3}\hat{\pi}^{(t+1)}_{r}=1$ and $\sum_{s=1}^{3}\hat{\gamma}^{(t+1)}_{rs} = 1$, for $r\in\{1,2,3\}$.
-We perform conditional maximizations to compute estimates of the remaining model parameters $(\bs{\delta}^\prime,\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$.
+Estimates of the initial and transition probabilities can be directly calculated as $\hat{\pi}^{(t+1)}_{r}=Pr\left(Z_{1}=r|\mb{y},\mb{x};\bs{\Psi}^{(t)}\right)$ and $\hat{\gamma}^{(t+1)}_{rs}=\sum_{j=2}^{M}Pr(Z_{j-1}=r,Z_{j}=s|\mb{y},\mb{x};\bs{\Psi}^{(t)})/\sum_{j=2}^{M}Pr(Z_{j-1}=r|\mb{y},\mb{x};\bs{\Psi}^{(t)})$, respectively, restricted to $\sum_{r=1}^{3}\hat{\pi}^{(t+1)}_{r}=1$ and $\sum_{s=1}^{3}\hat{\gamma}^{(t+1)}_{rs} = 1$, for $r\in\{1,2,3\}$.
+We perform conditional maximizations to estimate the remaining parameters $(\bs{\delta}^\prime,\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$.
 First, mixture proportions can be estimated as $\hat{\delta}_{l}^{(t+1)}=\sum_{j=1}^{M}Pr(Z_{j}=2|\mb{y},\mb{x};\bs{\Psi}^{(t)})Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})/\sum_{j=1}^{M}Pr(Z_{j}=2|\mb{y},\mb{x};\bs{\Psi}^{(t)})$.
-Estimating $(\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$ from (\ref{eq:diffQfunc}) can be seen as obtaining parameter estimates from a series of weighted NB regression models with shared mean and dispersion parameters (Web Appendix B4).
-We jointly estimate these quantities via the algorithm BFGS \citep{fletcher2013practical}.
+Then, estimates of $(\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$ are obtained in a similar fashion to the parameter estimation of a weighted NB regression model (Web Appendix B4).
 
-The estimation scheme is robust to situations where certain differential combinatorial patterns of enrichment are rare (Figure \ref{fig:fig_multimarks}).
-This unique characteristic results from the fact that ChIP-seq experiments often provide enough data (usually $M>10^7$ non-overlapping windows of 250 bp fixed size for the human reference genome) to estimate the parameters $(\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$, which are shared across all $L$ mixture components and HMM states.
-If pruning differential combinatorial patterns of the differential mixture component is of interest, the optimal number of mixture components $L^{*}$, $L^{*}<L$, can be selected via the Bayesian Information Criterion (BIC) for HMMs.
-We observed that selecting the optimal number of mixture components based on BIC agrees with the pruning of rare differential combinatorial patterns that we would not biologically expect to observe in real data (Web Appendix B2).
+The EM algorithm iterates until the maximum absolute relative change in the parameter estimates three iterations apart is less than $10^{-3}$ for three consecutive iterations.
+Here, we use a RCEM algorithm with threshold 0.05, which substantially reduces the dimensionality of the data during the M-step by randomly assigning a zero posterior probability to genomic windows unlikely to belong to each of the HMM states.
+Moreover, our estimation scheme allows genomic windows with equal distribution of counts across replicates and conditions to have their posterior probability aggregated during the M-step of the algorithm, which leads to a fast gradient-based optimization.
+Upon convergence, HMM posterior probabilities can be used to segment the genome into consensus background, differential, or consensus enrichment windows.
+Approaches that control the total false discovery rate (FDR) via posterior probabilities \citep{efron2001empirical} or that estimate the most likely sequence of hidden states can be used.
+Let $\hat{\rho}_{j2} = Pr\left(Z_{j}=2|\mb{y},\mb{x};\hat{\bs{\Psi}}\right)$ denote the estimated posterior probability that the $j^{th}$ genomic window belongs to the differential HMM state, $j=1,\ldots,M$.
+A posterior probability cutoff $\alpha$ is then chosen by controlling the total FDR $\sum_{j=1}^{M}(1-\hat{\rho}_{j2})I(\hat{\rho}_{j2}\geq 1-\alpha)/\sum_{j=1}^{M}I(\hat{\rho}_{j2}\geq 1-\alpha)$, where $I(\cdot)$ is an indicator function.
+Differential peaks are formed by merging adjacent windows that either meet a nominal FDR level for the differential HMM state or belong to the same predicted state (Web Appendix B3).
+Under this set up, we are able not only to detect differential enrichment regions across multiple conditions, but we can also classify various differential combinatorial patterns of enrichment within broad and short domains with mixture model posterior proportions.
 
-To obtain the parameter estimates $\hat{\bs{\Psi}}$, the EM algorithm iterates until the maximum absolute relative change in the parameter estimates three iterations apart is less than $10^{-3}$ for three consecutive iterations.
-To reduce the computation time, we make use of a RCEM algorithm with threshold 0.05.
-Briefly, the RCEM algorithm substantially reduces the dimensionality of the data during the M-step by randomly assigning a zero posterior probability to genomic windows unlikely to belong to each of the HMM states.
-The current estimation set up allows genomic windows exhibiting equal distribution of read counts to have their posterior probability aggregated during the M-step of the algorithm.
-Often, the distribution of read counts along the genome is highly concentrated on a particular set of values, such as 0, 1, and 2 for instance.
-Genomic windows exhibiting a particular pattern of counts across samples and conditions can have their posterior probability aggregated during the M-step, which further reduces the dimensionality of the objective function during the numerical optimization and leads to a fast gradient-based optimization.
-%Genomic windows with assigned zero posterior probability due to the RCEM approach do not contribute to the final set of aggregated posterior probabilities.
-
-Once the algorithm reaches convergence, the final set of HMM posterior probabilities can be used to segment the genome into consensus background, differential, or consensus enrichment windows.
-Approaches that control the total false discovery rate (FDR) via posterior probabilities \citep{efron2001empirical} or that estimate the most likely sequence of hidden states \citep{viterbi1967error} can be used for such purposes.
-Let $\hat{\rho}_{j2} = Pr\left(Z_{j}=2|\mb{y},\mb{x};\hat{\bs{\Psi}}\right)$ denote the estimated posterior probability that the $j^{th}$ genomic window belongs to the differential HMM state, $j=1,\ldots,M$.
-For a cutoff of posterior probability $\alpha$, the nominal FDR is $\sum_{j=1}^{M}(1-\hat{\rho}_{j2})I(\hat{\rho}_{j2}\geq 1-\alpha)/\sum_{j=1}^{M}I(\hat{\rho}_{j2}\geq 1-\alpha)$, where $I(\cdot)$ is an indicator function.
-The posterior probability cutoff is then chosen by controlling the total FDR.
-Differential regions of enrichment are formed by merging adjacent windows that either meet a nominal FDR threshold level for the differential HMM state or belong to the same Viterbi's predicted state.
-A discussion about these two approaches under the proposed model is presented in the Web Appendix B3.
-Additional details of proposed EM algorithm and the implemented code are available in the Web Appendices B4 and A4, respectively.
+The estimation scheme is robust to situations where certain differential combinatorial patterns of enrichment are rare (Figure \ref{fig:fig_multimarks}), which results from the fact that ChIP-seq experiments often provide enough data to estimate the parameters $(\beta_{1},\beta_{3},\lambda_{1},\lambda_{3})^\prime$ shared across all $L$ mixture components and HMM states.
+If pruning differential combinatorial patterns of the mixture model is of interest, the optimal number of mixture components $L^{*}$, $L^{*}<L$, can be selected via the Bayesian Information Criterion (BIC) for HMMs.
+We observed that selecting the optimal $L^{*}$ and combinatorial patterns via BIC agrees with pruning of rare differential combinatorial patterns that one would not expect to observe biologically (Web Appendix B2).
+In addition, the current implementation of epigraHMM allows the adjustment for external covariates, such as GC-content and input controls, via model offsets (Web Appendix A3).
+In real data analyses, however, we did not observe a significant influence of input controls on the sensitivity and specificity of peak calls, a fact that has also been noted by others \citep{lun2015csaw}.
 
@@
 \label{sec:simulation}
-We evaluate the presented model in two independent simulation studies of broad epigenomic marks.
-In the first study (Section~\ref{sec:sim1}), we simulated read count-based data to assess the precision of the parameter estimation scheme, the performance of differential peak detection, and the accuracy of the classification of specific differential combinatorial patterns of enrichment within differential peaks.
-In the second simulation study (Section~\ref{sec:sim2}), we utilize the simulation pipeline presented in \cite{lun2015csaw} to generate synthetic ChIP-seq reads from \textit{in silico} experiments with broad differential peaks.
-The aim of the second simulation study was to compare our model with other DPCs in a more realistic scenario with broad peaks, while also avoiding the assumption of a parametric model for the data.
+
+We evaluated epigraHMM in two simulation studies.
+First, we simulated read count-based data to assess the precision of the estimation scheme, the performance of differential peak detection, and the accuracy of the classification of differential combinatorial patterns (Section~\ref{sec:sim1}).
+Next, we utilized the simulation pipeline from \cite{lun2015csaw} to simulate ChIP-seq reads from experiments with broad differential peaks (Section~\ref{sec:sim2}).
+The aim of the second simulation study was to compare epigraHMM with other DPCs in a more realistic scenario with broad peaks, while avoiding the choice of a parametric model for the data.
 
@@
 \label{sec:sim1}
-Read counts were simulated under different scenarios that varied regarding the type of histone modification mark (H3K36me3 and H3K27me3), number of genomic windows ($M$, $10^5$, $5\times 10^5$, and $10^6$ windows), number of conditions ($G$, 2, 3, and 4), and number of replicates per condition ($n$, 1, 2, and 4).
-We further assessed our model under different Signal-to-Noise Ratio (SNR) levels.
-We define the SNR as the ratio between the means of consensus enrichment and consensus background emission distributions.
-Mean and dispersion parameters used in this simulation study were estimated from ENCODE data and are presented in the Web Appendix C1 for all the scenarios.
-Different SNR levels were defined by decreasing the ratio of the means in decrements of 10\% while maintaining the mean-variance relationship.
-Read counts were assumed to follow a NB distribution and were simulated using a first-order Markov chain with $2^G$ states, representing every combination of background and enrichment across $G$ conditions.
-We aimed to assess whether our model was able to assign all $2^G-2$ simulated differential states to the differential HMM state, while maintaining a precise parameter estimation scheme and accurate classification of differential combinatorial patterns.
+Read counts were simulated under scenarios that varied regarding the type of histone modification mark (H3K36me3 and H3K27me3), number of windows ($M$, $10^5$, $5\times 10^5$, and $10^6$ windows), number of conditions ($G$, 2, 3, and 4), and number of replicates per condition ($n$, 1, 2, and 4).
+We further assessed epigraHMM under different Signal-to-Noise Ratio (SNR) levels, here defined as the ratio between the means of consensus enrichment and background emission distributions, by decreasing the mean ratio while maintaining the mean-variance relationship.
+Parameters used in this study were estimated from ENCODE data (Web Appendix C1).
+Simulated counts followed a NB distribution and were generated using a first-order Markov chain with $2^G$ states, representing every combination of background and enrichment across $G$ conditions.
+We assessed whether epigraHMM was able to assign all $2^G-2$ differential states to the differential HMM state, while precisely estimating model parameters and accurately classifying differential combinatorial patterns.
 
-\subsubsection{Simulation Results}
-Table~\ref{Table:table_simulation} shows the true values and the average relative bias of parameter estimates (and the $2.5^{th}$, $97.5^{th}$ percentiles) from a hundred simulated data sets relative to the scenario of H3K27me3 with $10^6$ genomic windows (see Web Appendix C1 for additional details).
-Results are shown for the SNR level estimated from the ENCODE project (mean count ratio between peak and enrichment regions of 3.21), different number of conditions, and different number of replicates per condition.
-Overall, no significant differences regarding the relative bias of parameter estimates were observed across simulations under different number of genomic windows.
-Depending on the number of conditions, the observed relative bias and the range of the reported percentiles tended to decrease as more replicates were included in the analyses.
-This effect was particularly significant in scenarios with four conditions with respect to parameters $\beta_{3}$ and $\lambda_{3}$.
-In scenarios with higher number of conditions, these results highlight the importance of experimental replicates to achieve precise parameter estimates.
-The proposed estimation approach via EM algorithm led to precise parameter estimates and was robust to a data generating mechanism that was different than the one assumed by the proposed model.
+Table~\ref{Table:table_simulation} presents results relative to the H3K27me3 simulation scenario with $10^6$ genomic windows and ENCODE-estimated SNR (mean ratio of 3.21; see Web Appendix C1 for additional details).
+Depending on the number of conditions, the relative bias and the range of the reported percentiles tended to decrease as more replicates were included in the analyses.
+This effect was more pronounced when estimating $\beta_{3}$ and $\lambda_{3}$ in scenarios with four conditions, which highlights the importance of experimental replicates to achieve precise parameter estimates.
+Overall, the proposed estimation scheme led to precise parameter estimates and was robust to a data generating mechanism that was different than the one assumed by the proposed model.
+No significant differences regarding the relative bias of parameter estimates were observed across simulations under different number of genomic windows.
+
 \begin{center}
@@
 \label{sec:sim2}
-We performed a second simulation study aiming to compare the proposed model with the current DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR.
-We used the simulation pipeline presented by \cite{lun2015csaw} where data were generated in a more general scheme without a particular read count model assumption.
-Here, sequencing reads from broad ChIP-seq experiments were generated for two conditions and two replicates per condition.
-For the differential peaks callers ChIPComp and DiffBind that require sets of candidate regions, we followed the analyses presented by \cite{lun2015csaw} and called peaks in advance using HOMER.
-Peaks were then used as input in the respective software for differential call.
-A hundred simulated data sets were generated and peaks were called by all the methods under multiple nominal FDR thresholds.
-For our method and RSEG, window-based posterior probabilities were used to control the total FDR as described in Section \ref{subsec:hmmest}.% We refer to the method presented in this article as epigraHMM.
-We attempted to utilize MACS in this simulation study but the peak caller failed to output peaks for the simulated data.
 
-\subsubsection{Simulation Results}
-Figure~\ref{fig:fig_simulation_csaw} shows the main results of our second simulation study.
-Out of 100 simulated data sets, RSEG either failed to analyze the data due to internal errors or called the entire genome as differential in 26 and in 3 instances, respectively.
-Problematic cases from RSEG were not considered in this simulation study.
-Similar issues have been previously reported in other studies \citep{starmer2016detecting}.
-We observed that our method showed the highest observed sensitivity among all DPCs, regardless of the nominal FDR thresholding level, while maintaining a moderate observed FDR (Figure~\ref{fig:fig_simulation_csaw}A).
-Here, the observed FDR is the proportion of genomic windows incorrectly called as differential out of the total number of windows called as differential.
-Methods such as diffReps, RSEG, and THOR showed higher observed FDR levels than the nominal threshold due to the excessive number of differential peaks called outside true differential regions (shaded area in Figure~\ref{fig:fig_simulation_csaw}D).
-While diffReps and THOR called an excessive number of short and discontiguous peaks, RSEG called regions that were usually wider than the observed differential enrichment regions.
-These results are further illustrated in Figure~\ref{fig:fig_simulation_csaw}B, where we present the average ratio of the number of called and simulated peaks (y-axis) and the average number of called peaks intersecting true differential regions (x-axis).
-Regarding the computation time, the HMM-based algorithms RSEG and THOR appeared to be the most computationally intensive and required longer amounts of time to analyze the data.
-In Figure~\ref{fig:fig_simulation_csaw}C, we present the box plots of computing time (in minutes) across a hundred simulated data sets for all benchmarked methods.
-While still being an HMM-based algorithm, our method was among the fastest tools for differential peak detection due to the implemented strategies to improve the computation time of the EM algorithm (Section \ref{subsec:hmmest}).
-Figure~\ref{fig:fig_simulation_csaw}D shows an example of a genomic region with simulated data and called peaks from various methods using nominal FDR threshold 0.05.
-As shown, our method was able to consistently cover most of true differential regions with broad peaks while exhibiting a limited number of false discoveries (see Web Appendix C2 for additional results).
+Here, we compared epigraHMM with the DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR on data simulated with the pipeline from \cite{lun2015csaw} without a particular model assumption.
+Sequencing reads from ChIP-seq experiments were generated for two conditions and two replicates per condition.
+For the two-step DPCs ChIPComp and DiffBind, we followed \cite{lun2015csaw} and called peaks in advance using HOMER \citep{heinz2010simple}, which were then used as input in the respective software for differential call (MACS2 failed to output peaks on the simulated data).
+A hundred simulated data sets were generated and peaks were called under multiple nominal FDR thresholds.
+For epigraHMM and RSEG, window-based posterior probabilities were used to control the total FDR (Section \ref{subsec:hmmest}).
+
+Figure~\ref{fig:fig_simulation_csaw} shows the results of our second simulation study.
+epigraHMM showed the highest observed sensitivity among all DPCs while maintaining an observed FDR close to the nominal value (Figure~\ref{fig:fig_simulation_csaw}A).
+Here, the observed FDR is the proportion of windows incorrectly called as differential out of the total number of windows called as differential.
+Due to the excessive number of false positive differential peak calls, diffReps, RSEG, and THOR showed higher observed FDR levels than the nominal FDR values (Figure~\ref{fig:fig_simulation_csaw}D).
+Overall, diffReps and THOR called an excessive number of short and discontiguous peaks while RSEG called regions that were usually wider than the simulated differential peaks (Figure~\ref{fig:fig_simulation_csaw}B).
+Regarding the computation time, epigraHMM was among the fastest DPCs due to the efficient RCEM implementation (Figure~\ref{fig:fig_simulation_csaw}C).
+Other HMM-based algorithms, RSEG and THOR, appeared to be the most computationally intensive and required longer amounts of time to analyze the data.
+In general, epigraHMM was able to consistently cover most of true differential regions with broad peaks while exhibiting a limited number of false discoveries (Figure~\ref{fig:fig_simulation_csaw}D; Web Appendix C2).
+Results shown for RSEG do not include 27 cases in which the algorithm failed to analyze the data due to internal errors or called the entire genome as differential.
+Similar issues of RSEG have been reported by others \citep{starmer2016detecting}.
+
 \begin{figure}
@@
 \label{sec:realdata}
-We applied our method to ChIP-seq data from the ENCODE Consortium (Section \ref{sec:background}) to detect differential regions of enrichment of several epigenomic marks across distinct cell lines.
-First, we analyzed broad data from the histone modifications H3K36me3 and H3K27me3 as well as data from EZH2 (Section \ref{sec:broad}).
-Secondly, we assessed the performance of the presented model on ChIP-seq experiments from the transcription factor CTCF and the histone modifications H3K27ac and H3K4me3 characterized by short peaks (Section \ref{sec:sharp}).
-For H3K27ac and H3K4me3, enrichment peaks are usually deposited on the promoter regions of actively transcribed genes and several studies have associated their role with gene transcription \citep{creyghton2010histone,lauberth2013h3k4me3}.
-The transcription factor CTCF is a protein that binds to short DNA motifs and is responsible for several cellular processes that include the regulation of the chromatin 3D structure and mRNA splicing \citep{shukla2011ctcf}.
-Two isogenic replicates for each epigenomic mark were used in the analysis.
-Using RNA-seq data, we assessed the practical significance of our results by associating the detection and classification of differential combinatorial patterns from called peaks of H3K36me3, H3K27me3, and EZH2 with the direction of gene expression (Section \ref{sec:predpostprob}; see Web Appendix D1 for the RNA-seq data processing).
 
-We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short data sets.
-For the methods that require a set of candidate regions to be specified \textit{a priori}, ChIPComp and DiffBind, peaks were called in advance using MACS2 (\cite{zhang2008model}; available at \url{https://github.com/macs3-project/MACS}) and used as input to the software for differential call.
-We benchmarked methods regarding the coverage of differentially transcribed gene bodies, the number and average size of differential peak calls, $\log_{2}$ fold change (LFC) of read counts, Spearman correlation of $\log_{2}$-transformed read counts between cell lines, and computation time.
-Metrics for sensitivity and specificity were defined on the window level and based on the coverage of differentially transcribed gene bodies by called peaks (Web Appendix D1).
-For broad marks, read counts were computed using non-overlapping windows of 500bp.
-For the remaining short marks, we computed read counts using non-overlapping windows of 250bp.
-Results presented in this section pertain to the analysis of two cell lines, namely Helas3 and Hepg2.
+We applied epigraHMM on ChIP-seq data from the ENCODE Consortium to detect differential peaks of several epigenomic marks across multiple cell lines.
+We analyzed data characterized by broad peaks (H3K36me3, H3K27me3, and EZH2; Section \ref{sec:broad}) and short peaks (CTCF, H3K27ac, and H3K4me3; Section \ref{sec:sharp}).
+Two isogenic replicates of each cell line were used in the analysis.
+Using RNA-seq data, we assessed the practical significance of our results by associating the detection and classification of differential combinatorial patterns from called peaks with gene expression (Section \ref{sec:predpostprob}; Web Appendix D1).
+
+We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithms for both broad and short ChIP-seq enrichment profiles.
+For two-step DPCs, ChIPComp and DiffBind, we used peak calls from MACS2 as candidate peaks \citep{zhang2008model}.
+Methods were compared in terms of the coverage of differentially enriched genes, the number and average size of differential peak calls, the $\log_{2}$ fold change (LFC) and Spearman correlation of $\log_{2}$ reads counts mapped onto differential peak calls, the computation time, and maximum memory utilized.
+Metrics for sensitivity and specificity were defined on the window level and based on the coverage of differentially enriched genes by called peaks (Web Appendix D1).
+Read counts were computed using non-overlapping windows of 500bp and 250bp for broad and short marks, respectively (see Web Appendix D2 for a discussion on window size).
+Results presented here pertain to the analysis of cell lines Helas3 and Hepg2 (see Web Appendix D3 for additional results).
 Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR).
-We did not observe a significant improvement in performance by accounting for input control effect or GC-content bias with epigraHMM and did not adjust our analyses for these effects.
-A discussion about the choice of the window size is presented in the Web Appendix D2.
-Results from the analysis of more than two cell lines are presented in the Web Appendix D3.
-Data accessing code, data pre-processing steps, method-specific parameters, and code to replicate the presented results are detailed in the Web Appendix A.
+We did not observe a significant improvement in performance by accounting for input control effect or GC-content bias with epigraHMM and did not adjust our analyses for these effects (Web Appendix B1).
 
@@
 \label{sec:broad}
-Methods were benchmarked regarding the coverage of differentially transcribed gene bodies.
-The histone modification H3K36me3 is known to be associated with gene transcription and enriched regions of this mark are usually deposited on transcribed gene bodies.
-Hence, the location of differential peaks of H3K36me3 is expected to agree with the location of differentially expressed genes.
-Following the analysis presented by \cite{steinhauser2016comprehensive}, we defined a set of protein coding genes exhibiting $|\text{LFC}|>2$ of ChIP-seq read counts between the two analyzed cell lines as true differentially transcribed genes.
-Results using different threshold levels are presented in the Web Appendix D1 and agree with those presented here.
-Protein-coding genes with total read count across cell lines under the $25^{th}$ percentile were excluded from the analysis.
-Normalization by the median log-ratios of each replicate over the geometric mean was performed to avoid spurious differences due to sequencing depth.%The analysis of H3K27me3 is challenging as its enrichment is often associated with both intergenic and silenced genes.
-Methods that utilize a gold-standard for benchmarking purposes for this mark rely on real-time PCR results on specific loci \citep{zang2009clustering}.
-Here, we compare methods for this mark with respect to the characteristics of called peaks.
+First, we compared methods regarding the coverage of differentially enriched genes by differential peak calls for H3K36me3.
+Following the analysis of \cite{ji2013differential}, we defined a set of protein coding genes exhibiting $|\text{LFC}|>2$ of ChIP-seq counts between Helas3 and Hepg2 cell lines as true differential binding sites (see Web Appendix D1 for results using different thresholding values).
+Median ratio normalization of counts was performed on protein-coding genes after excluding those with total counts under the $25^{th}$ percentile of the distribution.
+Figure~\ref{fig:fig_broad}A shows receiver operating characteristic (ROC) curves for various methods and different nominal FDR levels for H3K36me3 differential peak calls.
+Observed true (false) positive rates were computed on the window-level as the proportion of windows called as differential out of all windows associated (not associated) with differentially enriched genes (Web Appendix D1).
+Overall, epigraHMM outperformed other DPCs by covering most of the differentially enriched genes while calling a limited number of false positive peaks.
 
-In Figure~\ref{fig:fig_broad}A, we show receiver operating characteristic (ROC) curves for various methods and different nominal levels of FDR threshold for differential peaks of H3K36me3.
-Similar to the analysis of broad histone modification marks presented by \citet{xing2012genome}, we computed the observed true positive rates (false positive rates) on the window-level as the proportion of windows called as differential out of the total number of windows associated (not associated) with differentially transcribed genes (Web Appendix D1).
-Our method had the best overall performance among all DPCs as its differential peaks were able to cover most of differentially transcribed gene bodies while still maintaining a low number of false positives.
-Methods that tended to call short peaks, such as ChIPComp and DiffBind, were the ones with the lowest sensitivity among all methods.
-ChIPComp and DiffBind have been previously shown to be dependent on the set of candidate peaks and to perform best in scenarios with short peaks (Figure~\ref{fig:fig_broad}B; \citet{steinhauser2016comprehensive,lun2015csaw}).
-In Figure~\ref{fig:fig_broad}C, we show the observed sensitivity (y-axis) and the average differential peak size (kbp; x-axis) for various methods under different nominal FDR levels (the observed FDR is annotated next to each data point).
-Our model and RSEG, two HMM-based methods, tended to call broader differential peaks and exhibited better sensitivity than other methods.
-Yet, differential peaks called by RSEG often did not correspond to differential regions of enrichment (Figure~\ref{fig:fig_broad}D), a behavior that has been noted by others \citep{starmer2016detecting} and also seen in simulated data (Figure~\ref{fig:fig_simulation_csaw}).
-Our HMM-based method with a non-linear normalization scheme via model offsets allowed us to maintain a low observed FDR and a higher sensitivity than other DPCs.
-In Figure~\ref{fig:fig_broad}F, we show examples of differential peak calls for EZH2.
-Our method was among the most efficient algorithms due to our computational scheme, taking approximately 30 minutes to analyze genome-wide data (Figure~\ref{fig:fig_broad}E).
+Here, DiffBind, which has been shown to be dependent on the set of candidate peaks \citep{lun2015csaw}, called the shortest peaks overall (Figures \ref{fig:fig_broad}B and \ref{fig:fig_broad}C) and ChIPComp, which has been shown to perform best in scenarios with short marks \citep{steinhauser2016comprehensive}, appeared to have limited performance in broad marks such as H3K27me3 and EZH2 (Figures \ref{fig:fig_broad}D and \ref{fig:fig_broad}F).
+epigraHMM and RSEG, two HMM-based methods, called broader differential peaks and exhibited better sensitivity than other methods.
+Yet, differential peaks from RSEG often did not correspond to observed differential enrichment, a fact that explains its high observed FDR and agrees with our simulations (Figures \ref{fig:fig_broad}C and \ref{fig:fig_simulation_csaw}) and others \citep{starmer2016detecting}.
+Our HMM-based approach with a non-linear normalization via model offsets allowed us to maintain a low observed FDR and a higher sensitivity than other DPCs.
+Overall, epigraHMM was among the most efficient algorithms due to our computational scheme, taking approximately 30 minutes to analyze genome-wide data (Figure~\ref{fig:fig_broad}E).
+
 \begin{figure}
@@
 \label{sec:sharp}
-We further evaluated the performance of the proposed method on data sets characterized by short peaks, namely the histone modifications H3K4me3 and H3K27ac and the transcription factor CTCF.
-The goal of our analysis was to assess whether our method was robust to different types of data and still able to call short differential regions of enrichment.
-In these scenarios, differential peaks are usually observed in isolated genomic regions and exhibit a high SNR.
-It has been shown that certain HMM-based approaches, including RSEG, have low accuracy under short histone modification marks and TF data \citep{hocking2016optimizing}.
-As we show, the model proposed in this article performs comparably to the evaluated DPCs known to perform best in short data (ChIPComp, DiffBind; \cite{steinhauser2016comprehensive}).
 
-We calculated the LFC and the Spearman correlation between cell lines Helas3 and Hepg2 based on ChIP-seq read counts mapped onto differential peaks called by each method.
-Read counts were previously normalized by the median log-ratios of each replicate over the geometric mean to avoid spurious differences due to sequencing depth.
-As these marks are characterized by short peaks, ideal methods would show high absolute LFC and negative correlation between read counts mapped on differential peaks.
-Figure~\ref{fig:fig_sharp} shows the main results from our analysis using short data sets.
-In Figures~\ref{fig:fig_sharp} A and C we show the median LFC and the Spearman correlation of ChIP-seq counts for differential CTCF and H3K4me3 peak calls (sorted by the absolute LFC), respectively, under a nominal FDR control of 0.05.
-We present separate curves regarding the signal of observed enrichment to better characterize the direction of change.
-The results show that the HMM-based methods RSEG and THOR were among those with the lowest absolute LFC among all methods, which confirms their sub optimal performance in the scenario of short peaks \citep{hocking2016optimizing}.
-In addition, we observed that ChIPComp had the best performance overall as it was able to call differential peaks with the highest absolute LFC and the lowest correlation between read counts of the two analyzed cell lines.
-Our model was able to properly call truly short differential peaks (Figures~\ref{fig:fig_sharp} B, D, and F) and was comparable to the non-HMM based methods regarding the computation time (Figure~\ref{fig:fig_sharp}E), jointly calling differential peaks in less than 1.5 hour.
+We further evaluated the performance epigraHMM on ChIP-seq samples characterized by short peaks (CTCF, H3K4me3, and H3K27ac).
+The goal of our analysis was to assess whether epigraHMM was flexible to different types of data and still able to call short differential regions of enrichment.
+Here, differential peaks are usually observed in isolated genomic regions and exhibit a high SNR.
+It has been shown that certain HMM-based algorithms, including RSEG, have low accuracy with such short marks \citep{hocking2016optimizing}.
+
+We calculated the LFC and the Spearman correlation between cell lines Helas3 and Hepg2 of ChIP-seq counts mapped onto differential peaks called by each method.
+Median ratio normalization of counts was performed prior to computing such metrics.
+Here, ideal methods would show high absolute LFC and negative correlation between read counts mapped onto their differential peaks.
+We observed that other HMM-based algorithms, RSEG and THOR, were among those with the lowest absolute LFC among all methods, which confirms their sub optimal performance in the scenario of short peaks (Figures \ref{fig:fig_sharp}A and \ref{fig:fig_sharp}C).
+ChIPComp, DiffBind, and csaw, were among those with the best performance overall, as their differential peak calls had the highest absolute LFCs and the lowest Spearman correlation of counts between cell lines.
+Overall, epigraHMM performed comparably to these methods that are known to perform best in short epigenomic marks while calling differential peaks in less than 1.5 hour (Figures \ref{fig:fig_sharp}B and \ref{fig:fig_sharp}D-\ref{fig:fig_sharp}F; \cite{steinhauser2016comprehensive}).
+
 \begin{figure}
@@
 \label{sec:predpostprob}
-Lastly, we analyzed data from the cell line Helas3 to segment its genome regarding the activity of marks H3K36me3, H3K27me3, and EZH2.
-We considered each mark as a separate experimental condition ($G=3$) and sought to jointly classify local chromatin states in Helas3 based upon the presence or absence of enrichment from each mark.
-It is known that EZH2 catalyzes the methylation of H3K27me3, a repressive mark, and H3K36me3 is associated with transcribed genes (Section \ref{sec:background}).
-Hence, we expected regions of enrichment in consensus for these marks to be rare and differential regions to be mostly represented by either transcribed chromatin states (enrichment for H3K36me3 alone) or repressed chromatin states (enrichment co-occurrence for H3K27me3 and EZH2).
-The analyses presented in this section highlight the applicability of our method in the context of genomic segmentation \citep{ernst2012chromhmm}, a distinct problem not tackled by current DPCs.
 
-First, we segmented the genome using the Viterbi sequence of most likely HMM states to understand the distribution of genomic regions associated with consensus background, differential, and consensus enrichment states (Figure~\ref{fig:fig_multimarks}).
-While the majority of the genomic regions exhibited no enrichment for any of the analyzed marks, regions of consensus enrichment were rare and represented only 2\% of the analyzed genome (Figure~\ref{fig:fig_multimarks}A), as expected.
-Consensus background and differential regions mostly covered intergenic (66\%) and protein-coding genic regions (61\%), respectively.
-Differential genomic windows were mostly representing either transcribed chromatin states or repressed chromatin states (Figure~\ref{fig:fig_multimarks}B).
-All differential combinatorial patterns expected to be rare had associated mixture proportion estimates less than 0.05 and were pruned using our model selection scheme via BIC presented in Web Appendix B2 (Figure~\ref{fig:fig_multimarks}C).
-%.
-The Viterbi algorithm allows the simultaneous segmentation of the genome regarding the three HMM states without the necessity of choosing state-specific FDR control thresholds
+Lastly, we analyzed data from the cell line Helas3 to segment its genome regarding the activity of H3K36me3, H3K27me3, and EZH2.
+We considered each mark as a separate experimental condition ($G=3$) and sought to jointly classify local chromatin states based upon the presence or absence of enrichment from each mark.
+Assuming that EZH2 catalyzes the methylation of H3K27me3, a repressive mark, and H3K36me3 associates with transcribed genes (Section \ref{sec:background}), we expected consensus enrichment regions to be rare and differential regions to be mostly represented by either transcribed chromatin states (enrichment of H3K36me3 alone) or repressed chromatin states (co-enrichment for H3K27me3 and EZH2).
+The analyses presented here highlight the applicability of our method in the context of genomic segmentation \citep{ernst2012chromhmm}, a distinct problem not tackled by current DPCs.
+
+First, we observed that while the majority of genomic regions exhibited no enrichment for any of the analyzed marks, regions of consensus enrichment were rare and covered only 2\% of the genome (Figure~\ref{fig:fig_multimarks}A), as expected.
+Consensus background and differential regions mostly covered intergenic and protein-coding genic regions, respectively.
+In fact, differential genomic windows were mostly representing either transcribed chromatin states or repressed chromatin states (Figure~\ref{fig:fig_multimarks}B).
+All differential combinatorial patterns expected to be rare had estimated mixture model proportions less than 0.05, which were then pruned using a model selection scheme via BIC (Figure~\ref{fig:fig_multimarks}C; Web Appendix B2).
+
 \begin{figure}
@@
 
-These results suggest that protein-coding genes overlapping differential regions should be either silenced (e.g.
-genes associated with repressed chromatin states) or highly expressed (e.g.
-genes associated with transcribed chromatin states).
+To assess the biological significance of our results, we associated the distribution of transcripts per million (TPM, from matching RNA-seq experiments) of protein-coding genes with combinatorial patterns of overlapping differential peaks.
 To assign combinatorial patterns to differential peaks, we chose the combination pertaining to the most frequent mixture component across windows by using the maximum estimated mixture model posterior probability, $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})$, $j=1,\ldots,M$.
-For genes overlapping differential regions associated with either transcribed or repressed chromatin states, we computed the distribution of transcripts per million (TPM) from matching RNA-seq experiments.
 Genes associated with transcribed chromatin states had a significantly higher distribution of TPM counts than genes associated with repressed chromatin states (Figure~\ref{fig:fig_multimarks}D).
-In fact, nearly all known Helas3 cycle-regulated genes (therefore expressed at some stage of the cell life cycle; \cite{whitfield2002identification}) were associated with differential enrichment of H3K36me3, here measured by the associated average mixture posterior proportions of the overlapping differential peaks (Figure~\ref{fig:fig_multimarks}E).
-The expression of such genes are known to correlate with the proliferative state of tumors and their study may help characterize their role in cancer \citep{whitfield2002identification}.
-We detected broad differential regions of enrichment and the classification of differential combinatorial patterns agreed with their biological roles as well as the expression levels of associated gene bodies.
-Figure~\ref{fig:fig_multimarks}F shows an example of a genomic region with differential enrichment for the analyzed marks.
-We compare our results to ChromHMM with 3 states, a method developed for chromatin segmentation.
-Our method offers the benefit of simultaneously detecting differential peaks and classifying the combinatorial pattern of enrichment through mixture model posterior probabilities even in the context of genomic segmentation with highly diverse epigenomic marks (see Web Appendix D4).
-By using the BIC for model selection (Web Appendix B2), one can choose the number of biologically relevant mixture components to be included in the model, a task that may not be as straightforward in methods such as ChromHMM (see Supplementary Figure 4 in \citet{ernst2012chromhmm}).
+In fact, nearly all known Helas3 cycle-regulated genes (therefore genes expressed at some stage of the cell life cycle; \cite{whitfield2002identification}) were associated with high average mixture posterior proportion of differential enrichment for H3K36me3 alone (Figure~\ref{fig:fig_multimarks}E).
+
+The expression of such genes are known to correlate with proliferative states of tumors and their study may help characterize their role in cancer \citep{whitfield2002identification}.
+Broad differential peaks called by epigraHMM, and their assigned differential combinatorial pattern, often agreed with the biological roles as well as the expression levels of associated gene bodies (Figure~\ref{fig:fig_multimarks}F; results from ChromHMM with 3 states for comparison).
+epigraHMM offers the benefit of simultaneously detecting differential peaks and classifying differential combinatorial patterns of enrichment even in the context of genomic segmentation of highly diverse epigenomic marks (Web Appendix D4).
+By using the BIC for model selection, one can choose the number of biologically relevant mixture components to be included in the model, a task that may not be as straightforward in methods such as ChromHMM (Web Appendix B2).
 
@@
 First, it uses an HMM-based approach that accounts for the local dependency of ChIP-seq counts and is able to precisely detect broad and short differential regions of enrichment.
-Second, it utilizes a GLM-based framework with model offsets that account for potential non-linear biases in the data as well as a constrained parametrization across HMM states and mixture components.
-Our implementation of the RCEM algorithm led to genome-wide analyses of ChIP-seq data under a computational time comparable to some of the fastest current methods and was at least 5 times faster than current HMM-based algorithms.
+Second, it utilizes a GLM-based framework with model offsets that accounts for non-linear biases and other potential nuisance factors such as GC-content bias or input control samples.
+Our efficient implementation of the RCEM algorithm led to genome-wide analyses of ChIP-seq data under a computational time comparable to some of the fastest current algorithms.
 Our results highlight the importance of the inclusion of technical/biological replicates in the analysis of epigenomic data.
 Lastly, our method allows the simultaneous detection and classification of differential combinatorial patterns of enrichment from its embedded mixture model and the associated posterior probabilities under any number of conditions.
-In situations where certain combinatorial patterns are rare, pruning of specific combinatorial patterns is possible via model selection via BIC (Web Appendix B2).
+In situations where certain differential combinatorial patterns are rare, our model selection scheme via BIC provides means of pruning specific combinatorial patterns of interest (Web Appendix B2).
 Our software has been implemented into an R package and is available for download (Web Appendix A).

### 2021-03-03 Latex/letter/letter.Rmd
txt update

--- 2021-02-26 Latex/letter/letter.Rmd
@@
 We cite in our paper the benchmarking article 'A comprehensive comparison of tools for differential ChIP-seq analysis' (Briefings in Bioinformatics, 2016).
-However, we authors are not aware of a more recent comprehensive review of DPCs algorithms in the literature.
-As suggested we add the following sentence to our real data analysis section 'We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short data sets.' We believe that such a list covers the most used algorithms for DPC in thie field of bioinformatics.
+However, we are not aware of a more recent comprehensive review of DPCs algorithms in the literature.
+As suggested, we add the following sentence to our real data analysis section 'We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short ChIP-seq enrichment profiles.'.
+We believe that such a list covers the most used algorithms for DPC in the field of bioinformatics.
 
@@
 We present a simulation study in Section B1.3 of the Web Appendix showing that epigraHMM leads to higher sensitivity/specificity when accounting for GC-bias even after read count normalization using the count-normalizing denominators from Teng and Irizarry (2017).
-However, for the broad marks that we analized in our manuscript, we did not observe a substantial effect of GC-content on read count distribution and did not observe any substantial improvement in accounting for GC-content in our model via offsets when utilizing the method from Teng and Irizarry (2017), a method that was developed only for narrow marks characterized by sharp peaks.
+However, for the broad marks that we analyzed in our manuscript, we did not observe a substantial effect of GC-content on read count distribution and did not observe any substantial improvement in accounting for GC-content in our model via offsets when utilizing the method from Teng and Irizarry (2017), a method that was developed only for narrow marks characterized by sharp peaks.
 In fact, the authors do not recommend the use of their method when analyze epigenomic marks with broad enrichment.
-We did not explore alternative methods for GC-bias correction for broad marks as we believe this task is beyond the scope of our work.
+A possible explanation for such a lack of effect of GC-content on the analyzed broad ChIP-seq data is that, in such scenarios, the enrichment of counts has a much lower signal-to-noise ratio with respect to background regions in comparison to short epigenomic marks.
+In addition, the high resolution of conts computed on non-overlapping windows of 500bp makes the effect of GC-content on enrichment regions indistinguishable from the GC-content effect on background regions.
+We shown in Web Figure 7 results from our GC-content analysis utizilin the epigenomic mark EZH2.
+Due to these reasons, we did not explore alternative methods for GC-bias correction for broad marks.
 
@@
 
-> In the main text, we refer the reader to Section B5 of the Web Appendix.
-To estimate the model coefficients during the $t^{th}$ iteration of the EM-algorithm, one needs to find the optimal values $\beta_{1}^{(t)}$, $\beta_{3}^{(t)}$, $\lambda_{1}^{(t)}$, and $\lambda_{3}^{(t)}$ that maximize the quantity $Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) + Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) + Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right)$.
-This problem is analogous to estimating parameters in a weighted NB regression model, since $Q_{1}$, $Q_{2}$, and $Q_{3}$ can be seen as weighted log-likelihood functions from a NB generalized linear model with means and dispersions parametrized as described in the Section 3.1 of the main article.
-Exact forms of the $Q$-functions are presented in Equation B.4 of Web Appendix B5.
+> During the $t^{th}$ iteration of the EM-algorithm, one needs to obtain $\beta_{1}^{(t)}, \beta_{3}^{(t)}, \lambda_{1}^{(t)}$, and $\lambda_{3}^{(t)}$ that maximize $Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) + Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) + Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right)$ during the maximization step, such that
+
+$$Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) = \sum_{j=1}^{M}Pr\left(Z_{j}=1|\mathbf{y},\mathbf{x};\boldsymbol{\Psi}^{(t)}\right)\log f_{1}\left(\mathbf{y}_{..j}|\boldsymbol{\psi}_{1}\right), $$
+
+$$ Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) = \sum_{j=1}^{M}Pr\left(Z_{j}=2|\mathbf{y},\mathbf{x};\boldsymbol{\Psi}^{(t)}\right)\sum_{l=1}^{L}Pr(W_{jl}=1|Z_{j}=2,\mathbf{y}_{..j},\mathbf{x};\boldsymbol{\Psi}^{(t)})\delta_{l}\log f_{(2,l)}\left(\mathbf{y}_{..j}|\mathbf{x}_l;\boldsymbol{\psi}_{(2,l)}\right),\text{ and,}$$
+$$ Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right) = \sum_{j=1}^{M}Pr\left(Z_{j}=3|\mathbf{y},\mathbf{x};\boldsymbol{\Psi}^{(t)}\right)\log f_{3}\left(\mathbf{y}_{..j}|\boldsymbol{\psi}_{3}\right),$$
+
+> where the notation is precisely defined in Section 3 of the main text.
+Note that $Q_{1}$, $Q_{2}$, and $Q_{3}$ above can all be seen as weighted NB regression models where the weights are formed by a combination of HMM posterior probabilties and/or mixture model posterior proportions/probabilities.
+Given the chosen constrained parametrization across all HMM states, we use numerical gradient-based optimization to obtain the parameter estimates.
+In the equations above, $\mathbf{y}_{..j}$ is the sequence of read counts observed across all conditions and replicates on genomic window $j$.
+On the other hand, the design matrix $x$ is a a matrix of zeros and ones of dimention $G\times L$ that encodes all differential combinatorial patterns of enrichment $L$ across $G$ conditions (see Figure 1 of the main paper for an example).
+Further details of the implemented RCEM algorithm are detailed in Web Appendix B4.
 
@@
 
-> Results presented in Table 1 pretain to the full model with all possible combinatorial patterns.
-This is the 'optimal' model in the sense that the simulated data contains all possible combinatorial patterns of enrichment and background among the condition.
-The point of Figure 1 is to show that our efficient estimation method via fast RCEM leads to unbiased parameter estimates.
-Together with Figure 2, Table 1 raises the point that increasing the number of technical or biological replicates leads to better results in terms of sensitivity/specificity of peak calls as well as in the sense of asymptotic efficiency of parameter estimates.
-We leave the discussion of model selection via BIC in our model to Section B2 of the Web Appendix where we present a simulation study.
+> Results presented in Table 1 pertain to the full model that utilizes all possible combinatorial patterns in the differential state.
+This is the 'optimal' model in the sense that the simulated data contains all possible combinatorial patterns of enrichment and background by design in our simulation.
+The purpose of Table 1 is to show that our efficient estimation method via fast RCEM leads to unbiased parameter estimates.
+Together with Web Figures 14 and 15, Table 1 raises the point that increasing the number of technical or biological replicates leads to better results in terms of sensitivity/specificity of peak calls as well as in the sense of asymptotic efficiency of parameter estimates.
+We leave the discussion of model selection via BIC in our model to Section B2 of the Web Appendix where we present a simulation study that directly evaluates the performance of model selection in our method.
 
@@
 > We believe that Spearman correlation and LFC are not good metrics to compare peak callers with broad marks for two reasons.
-First, methods that tend to call narrow peaks within broad regions of enrichment will exhibit an unrealistic advantage in performance in terms of LFC.
-This is so because these methods tend to call as narrow peaks any genomic window exhibiting excessively high read count enrichment (or, LFC in differential enrichment).
-For broad marks, interest lies not on these single punctate windows but instead in broad regions of enrichment.
-Second, and due to a similar reason, these methods will also exhibit unrealistic advantage in performance in terms of Spearman correlation because broad regions of enrichment are much noisier than punctate regions and, thefore, read count correlations calculated from broad peaks will be attenuated in comparison to correlations calculated from narrow and punctate windows.
+First, methods that tend to incorrectly call narrow peaks within broad regions of enrichment will exhibit inflated performance in terms of LFC, despite not capturing broad signal regions.
+This is because these methods tend to call as narrow peaks any genomic window exhibiting excessively high read count enrichment (or, LFC in differential enrichment).
+See example of these  cases in Figure 3 of the paper (narrow H3K36me3 peak calls from diffReps on the rightmost enrichment region of panel B, narrow H3K37me3 peak calls from ChIPComp on the highlighted area of panel D, and punctate EZH2 calls from DiffBind in on the highlighted area of panel F).
+For broad marks, interest lies not on these single punctate windows but, instead, in broad regions of enrichment.
+Similarly, for methods that call broad regions of enrichment, which are much noisier than punctate regions, read count correlations calculated from their differential peaks will be inflated in comparison to correlations calculated from those of narrow and punctate windows.
+Note in Figure 4C, for example, that RSEG exhibits the highest average Spearman correlation among all methods while incorrectly calling broad peaks on short epigenomic marks.
+Conversely, algortihms that call punctate peaks within broad regions (Figure 3) still exhibit low/negative Spearman correlation, a metric that does not reflect the poor performance of these methods in broad peaks.
+Lastly, we note that our method, epigraHMM, provides accurate differential peak calls in both broad and short marks.
 
 > Regarding the use of 'matched differentially expressed genes as the gold standard to evaluate the differential transcribed genes called from differential H3K36me3 activity', we did not observe a sufficient correspondance between gene expression levels and H3K36me3 enrichment for the cancer cell lines analyzed in our paper.
-For instance, we have found several instances were genes were expressed but no enrichment of H3K36me3 was observed.
+For instance, we have found several instances where genes were expressed but no enrichment of H3K36me3 was observed.
+Also, note that the strategy of utilizing ChIP-seq data to define sets of silver-standard marks for benchmarking purposes has been already used by others (Ji et.
+al, 2013; PNAS).
 Below, we present some of these problematic cases obtained from the UCSC Genome Browser.
@@
 Due to these reasons, we decided not to use gene expression levels when benchmarking H3K36me3 as suggested by the reviewer.
+
+> NAIM, PLEASE SEE MY REPORT IN THE REPO UNDER ./Supp/Report_RNA_vs_ChIP/ FOR A COMPREHENSIVE DISCUSSION ABOUT HOW USING RNA-SEQ TO DEFINE DE GENES AND THEN BENCHMARKING DIFFERENTIAL H3K36ME3 CHIP-SEQ PEAKS CAN BE PROBLEMATIC.
+WE DO NOT BENCHMARK DIFFERENTIAL PEAK CALLING METHODS IN THE PAPER USING RNA-SEQ.
+AS I MENTION IN THE PAPER:  'Following the analysis of \cite{ji2013differential}, we defined a set of protein coding genes exhibiting $|\text{LFC}|>2$ of ChIP-seq counts between Helas3 and Hepg2 cell lines as true differential binding sites.
+Results using different threshold levels are presented in the Web Appendix D1 and agree with those presented here.'.
+I CITE THE PNAS PAPER FROM HONGKAI JI (2013), WHERE THEY DEFINED GOLD-STANDARD BENCHMARKING MARKS BASED ON LOG FOLD CHANGES OF CHIP-SEQ COUNTS, NOT ON RNA-SEQ COUNTS (SEE PAGE 6971 OF THEIR PAPER).
+THE ONLY PART OF THE PAPER WHERE WE USE RNA-SEQ IS IN CHROMATIN SEGMENTATION (FIGURE 5).
 
@@
 > Regarding the biological significance of our method, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3.
-Specifically, we show that all known marker genes of Helas3 had associated differential enrichment for the H3K36me3 mixture model component.
+Specifically, we show that all known marker genes of Helas3 were associated with differential enrichment for the H3K36me3 mixture model component.
+That is, genes that are expected to be actively transcribed in Helas3 (because they are marker genes) were associated with differential peaks and, in addition, were associated with the differential combinatorial pattern of enrichment for H3K36me3 alone.
+Note that consensus enrichment is a rare state and not expected to be observed in this analysis.
 The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone.
+That is, for each marker gene and the overlapping differential peaks, we calculate the average window-based mixture proportion from the differential combinatorial pattern associated with enrichment for H3K36me3 alone.
+Note that our selected model had two mixture components in the differential HMM state (either enrichment of H3K36me3 alone, or enrichment of H3K27me3 and EZH2 together).
 These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.
@@
 > We believe our method is quite comparable to other DPC methods in short differential peak calling.
-These results are shown in Figure 4 of our paper as well as in the Web Appendix (Web Figures 25-27).
+These results are shown in Figure 4 of our paper as well as in the Web Appendix (Web Figures 26-28).
 In terms of LFC and Spearman correlation, our method performed quite comparably to ChIPComp, a widely used method for narrow marks.
@@
 Similarly, input controls were also generated in our simulation study and accounted for in these algorithms.
-Throughout our analysis, we did not observe any significant improvement in performance by accounting for input control effect in differential peak calling.
+Throughout our analysis, we did not observe any significant improvement in performance in epigraHMM by accounting for input control effect in differential peak calling (See Web Figure 2 for these results).
 This effect was also reported by others (Lun \& Smyth, 2015).
+We did not compare in our paper the performance of other methods with and without accounting for input controls.
+As stated above, analyses reported in our paper were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR).
+For all other methods (epigraHMM, RSEG, and csaw), input controls were not considered and their results are reported as such.
 
@@
 Due to these reasons, we believe that the proposed analysis would be suboptimal to the one presented in this paper for the analyzed cell lines.
+
+> NAIM, PLEASE SEE MY COMMENT ABOVE ABOUT THIS GENE EXPRESSION ANALYSIS.
+I CITE THE PNAS PAPER FROM HONGKAI JI (2013), WHERE THEY DEFINED GOLD-STANDARD BENCHMARKING MARKS BASED ON LOG FOLD CHANGES OF CHIP-SEQ COUNTS, NOT ON RNA-SEQ COUNTS (SEE PAGE 6971 OF THEIR PAPER).
+PLEASE, SEE MY REPORT UNDER ./Supp/Report_RNA_vs_ChIP/ AND LET ME KNOW IF ANY OF THOSE RESULTS SHOULD BE INCLUDED HERE TO SUPPORT OUR RESPONSES.
 

### 2021-03-03 Latex/appendix/webappendix.Rnw
updating appendix

@@
 \label{fig:sim_gc_roc}}
+\end{figure}
+
+In Figure \ref{fig:encode_gc_counts} below, we present smoothed scatter plots of log-transformed ChIP-seq counts (y-axis) and the respective GC content (x-axis) for the ENCODE EZH2 data that we analyzed.
+While using a 500bp window size to compute read counts, we did not observe enough evidence to justify account for such an effect in our analyses.
+Note that, when using such read count resolution, the effect of GC content can be hardly seen.
+In addition, there is no evident separation of differential GC-content effect for peak and background regions, as reported in \cite{teng2017accounting}.
+
+\begin{figure}[h]
+\begin{center}
+\centerline{\includegraphics[scale = 0.75]{../../Supp/Figure_GC_Counts/Figure_GC_Counts.pdf}}
+\end{center}
+\caption{Smoothed scatter plots of log-transformed EZH2 ENCODE ChIP-seq counts and GC content.
+\label{fig:encode_gc_counts}}
 \end{figure}

### 2021-03-04 Latex/appendix/webappendix.Rnw
Fix reference

@@
 Additionally, the observed FDR was calculated as the proportion of genomic windows incorrectly called as differential out of the total number of called differential windows.
-In Figures~\ref{fig:sim1.fig1}-\ref{fig:sim1.fig3}, panel A shows the average observed true positive rate (y-axis) and the observed FDR (x-axis) for different nominal FDR levels across a hundred simulated data.
+In Figures~\ref{fig:sim1.fig1}-\ref{fig:sim1.fig2}, panel A shows the average observed true positive rate (y-axis) and the observed FDR (x-axis) for different nominal FDR levels across a hundred simulated data.
 Results are shown for different levels of SNR, number of conditions, and number of replicates per condition.

### 2021-03-04 Latex/appendix/webappendix.Rnw
Fix reference

@@
 Then, we compared the window-based classification with the true window-based simulated states from the Markov Chain (states $2,\ldots,G-1$).
-In Figures~\ref{fig:sim1.fig1}-\ref{fig:sim1.fig3}, panel B shows the confusion matrices of classified (x-axis) and simulated (y-axis) differential windows for a scenario with three conditions.
+In Figures~\ref{fig:sim1.fig1}-\ref{fig:sim1.fig2}, panel B shows the confusion matrices of classified (x-axis) and simulated (y-axis) differential windows for a scenario with three conditions.
 Differential combinatorial patterns of enrichment are represented by the sequences of letters 'E' (enrichment) and 'B' (background), such that each letter corresponds to the status of a given condition.

### 2021-03-05 Latex/appendix/webappendix.Rnw
txt update

@@
 \begin{table}[h!]
-\footnotesize
+\tiny
 \centering
 \caption{GEO sample accession codes of the analyzed data from the ENCODE Consortium}
-\begin{tabular}{lccccccc}
+\begin{tabular}{lcccccccc}
 \hline
-Cell Line & H3K27me3 & H3K36me3 & EZH2 & H3K4me3 & H3K27ac & CTCF & RNA-seq\\
+Cell Line & Control & H3K27me3 & H3K36me3 & EZH2 & H3K4me3 & H3K27ac & CTCF & RNA-seq\\
 \hline
-H1hesc & GSM733748 & GSM733725 & GSM1003524 & GSM733657 & GSM733718 & GSM733672 & GSM758566\\
-HelaS3 & GSM733696 & GSM733711 & GSM1003520 & GSM733682 & GSM733684 & GSM733785 & GSM765402\\
-Hepg2 & GSM733754 & GSM733685 & GSM1003487 & GSM733737 & GSM733743 & GSM733645 & GSM758575\\
-Huvec & GSM733688 & GSM733757 & GSM1003518 & GSM733673 & GSM733691 & GSM733716 & GSM758563\\
+H1hesc & GSM733770 & GSM733748 & GSM733725 & GSM1003524 & GSM733657 & GSM733718 & GSM733672 & GSM758566\\
+HelaS3 & GSM733659 & GSM733696 & GSM733711 & GSM1003520 & GSM733682 & GSM733684 & GSM733785 & GSM765402\\
+Hepg2 & GSM733732 & GSM733754 & GSM733685 & GSM1003487 & GSM733737 & GSM733743 & GSM733645 & GSM758575\\
+Huvec & GSM733715 & GSM733688 & GSM733757 & GSM1003518 & GSM733673 & GSM733691 & GSM733716 & GSM758563\\
 \hline
@@
 such that $options = \{\text{--broad --broad-cutoff 0.1}\}$ if H3K27me3, H3K36me3, or EZH2, or $options = \{\text{-q 0.01}\}$ otherwise.
+
+All ENCODE data analysis presented in the main text, and in this Web Appendix, were adjusted for input control samples in methods that are designed to do so.
+Specifically, differential peaks from ChIPComp, DiffBind, diffReps, and THOR were obtained on ChIP-seq samples while passing input control samples in their software call as an optional input argument argument.
+In contrast, input control samples were not considered in epigraHMM, csaw, and RSEG analyses.
 

### 2021-03-05 Latex/letter/letter.Rmd
txt update

@@
 output: pdf_document
+urlcolor: blue
 ---
@@
 ```
+
+Dear reviewers,
+
+We thank you for the constructive comments regarding our paper "Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions".
+
+Our revised paper has 20 pages in total, as requested, including appendices and references, and excluding the title page and abstract, figures and tables.
+We emphasize that the our presented method, previously called mixNBHMM, is now called epigraHMM.
+We refer to our method as epigraHMM throughout this revised version of our paper.
+
+Detailed point-by-point responses to all comments are provided below.
 
@@
 However, we are not aware of a more recent comprehensive review of DPCs algorithms in the literature.
-As suggested, we add the following sentence to our real data analysis section 'We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short ChIP-seq enrichment profiles.'.
+As suggested, we add the following sentence to our real data analysis section 'Here, we compared epigraHMM with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithms for both broad and short ChIP-seq enrichment profiles (...)'.
 We believe that such a list covers the most used algorithms for DPC in the field of bioinformatics.
@@
 We cite the work of Teng and Irizarry (2017) that proposed a correction method for GC-content bias in ChIP-seq analysis.
-Their method is applicable to any peak calling algorithm that measures the enrichment of reads in bins, such as our method.
+Their method is applicable to any peak calling algorithm that measures the enrichment of reads in bins, such as our method epigraHMM.
 Later in the article, we point out to the reader that our method is able to account for GC-content bias in the same manner as suggested by Teng and Irizarry (2017) by using their count-normalizing denominators in our model as offsets.
@@
 
-> This has been addressed. In the second paragraph of Section 5, we reference the current GitHub repository of MACS2 \url{https://github.com/macs3-project/MACS}.
+> This has been addressed.
+As requested by the Editorial Office, we reference the current GitHub repository of MACS2 \url{https://github.com/macs3-project/MACS} in the Supporting Information section of our revised paper.
 
@@
 We present a simulation study in Section B1.3 of the Web Appendix showing that epigraHMM leads to higher sensitivity/specificity when accounting for GC-bias even after read count normalization using the count-normalizing denominators from Teng and Irizarry (2017).
-However, for the broad marks that we analyzed in our manuscript, we did not observe a substantial effect of GC-content on read count distribution and did not observe any substantial improvement in accounting for GC-content in our model via offsets when utilizing the method from Teng and Irizarry (2017), a method that was developed only for narrow marks characterized by sharp peaks.
-In fact, the authors do not recommend the use of their method when analyze epigenomic marks with broad enrichment.
+However, for the broad marks that we analyzed in our manuscript, we did not observe a substantial effect of GC-content on read count distribution.
+Also, we did not observe any substantial improvement in accounting for GC-content in our model via offsets when utilizing the method from Teng and Irizarry (2017), which was developed only for narrow marks characterized by sharp peaks.
+In fact, the authors do not recommend the use of their method when analyzing epigenomic marks with broad enrichment.
 A possible explanation for such a lack of effect of GC-content on the analyzed broad ChIP-seq data is that, in such scenarios, the enrichment of counts has a much lower signal-to-noise ratio with respect to background regions in comparison to short epigenomic marks.
-In addition, the high resolution of conts computed on non-overlapping windows of 500bp makes the effect of GC-content on enrichment regions indistinguishable from the GC-content effect on background regions.
-We shown in Web Figure 7 results from our GC-content analysis utizilin the epigenomic mark EZH2.
+In addition, the high resolution of conts computed on non-overlapping windows of 500bp may lead to a non-differential effect of GC-content on enrichment regions in comparison to the GC-content effect on background regions.
+We shown in Web Figure 7 results from our GC-content analysis utilizing the epigenomic mark EZH2.
 Due to these reasons, we did not explore alternative methods for GC-bias correction for broad marks.
@@
 
-> During the $t^{th}$ iteration of the EM-algorithm, one needs to obtain $\beta_{1}^{(t)}, \beta_{3}^{(t)}, \lambda_{1}^{(t)}$, and $\lambda_{3}^{(t)}$ that maximize $Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) + Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) + Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right)$ during the maximization step, such that
+> During the $t^{th}$ iteration of the EM-algorithm, one needs to obtain $\beta_{1}^{(t)}, \beta_{3}^{(t)}, \lambda_{1}^{(t)}$, and $\lambda_{3}^{(t)}$ that maximizes $Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) + Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) + Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right)$ during the maximization step, where
 
@@
 
-$$ Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) = \sum_{j=1}^{M}Pr\left(Z_{j}=2|\mathbf{y},\mathbf{x};\boldsymbol{\Psi}^{(t)}\right)\sum_{l=1}^{L}Pr(W_{jl}=1|Z_{j}=2,\mathbf{y}_{..j},\mathbf{x};\boldsymbol{\Psi}^{(t)})\delta_{l}\log f_{(2,l)}\left(\mathbf{y}_{..j}|\mathbf{x}_l;\boldsymbol{\psi}_{(2,l)}\right),\text{ and,}$$
+$$ Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) = \sum_{j=1}^{M}Pr\left(Z_{j}=2|\mathbf{y},\mathbf{x};\boldsymbol{\Psi}^{(t)}\right)\sum_{l=1}^{L}Pr(W_{jl}=1|Z_{j}=2,\mathbf{y}_{..j},\mathbf{x};\boldsymbol{\Psi}^{(t)})\log \delta_{l}f_{(2,l)}\left(\mathbf{y}_{..j}|\mathbf{x}_l;\boldsymbol{\psi}_{(2,l)}\right),\text{ and,}$$
 $$ Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right) = \sum_{j=1}^{M}Pr\left(Z_{j}=3|\mathbf{y},\mathbf{x};\boldsymbol{\Psi}^{(t)}\right)\log f_{3}\left(\mathbf{y}_{..j}|\boldsymbol{\psi}_{3}\right),$$
@@
 > where the notation is precisely defined in Section 3 of the main text.
-Note that $Q_{1}$, $Q_{2}$, and $Q_{3}$ above can all be seen as weighted NB regression models where the weights are formed by a combination of HMM posterior probabilties and/or mixture model posterior proportions/probabilities.
+Note that $Q_{1}$, $Q_{2}$, and $Q_{3}$ above can all be seen as weighted NB regression models where the weights are formed by HMM marginal posterior probabilties ($Q_{1}$ and $Q_{3}$), and a combination of HMM marginal posterior probabilities, mixture model mixture posterior probabilities, and mixture model proportions ($Q_{2}$).
 Given the chosen constrained parametrization across all HMM states, we use numerical gradient-based optimization to obtain the parameter estimates.
@@
 Although we did observe an improvement in performance in terms of model sensitivity/specificity of peak calls in higher SNR data sets, the main driving quantity for the reduction of parameter bias is the number of replicates included in our model.
-For this reason, we rephrase the mentioned sentence in our paper as "Depending on the number of conditions, the observed relative bias and the range of the reported percentiles tended to decrease as more replicates were included in the analyses.
-This effect was particularly significant in scenarios with four conditions with respect to parameters $\beta_{3}$ and $\lambda_{3}$.
-In scenarios with higher number of conditions, these results highlight the importance of experimental replicates to achieve precise parameter estimates.".
+For this reason, we rephrase the mentioned sentence in our paper as "Depending on the number of conditions, the relative bias and the range of the reported percentiles tended to decrease as more replicates were included in the analyses.
+This effect was more pronounced when estimating $\beta_{3}$ and $\lambda_{3}$ in scenarios with four conditions, which highlights the importance of experimental replicates to achieve precise parameter estimates.".
 The estimated SNR level from ENCODE was 3.21 (ratio of mean counts between peak and background regions) for H3K27me3.
@@
 
+> We thank the reviewer for the constructive comment.
+We prepared a report to support our response to this comment, which can be assessed at https://github.com/plbaldoni/epigraHMMPaper/blob/main/Supp/Report_RNA_vs_ChIP/report.pdf
+
 > We believe that Spearman correlation and LFC are not good metrics to compare peak callers with broad marks for two reasons.
@@
 
-> Regarding the use of 'matched differentially expressed genes as the gold standard to evaluate the differential transcribed genes called from differential H3K36me3 activity', we did not observe a sufficient correspondance between gene expression levels and H3K36me3 enrichment for the cancer cell lines analyzed in our paper.
+> Regarding the use of 'matched differentially expressed genes as the gold standard to evaluate the differential transcribed genes called from differential H3K36me3 activity', we did not observe a sufficient correspondance between gene expression levels and H3K36me3 enrichment for the cancer cell lines analyzed in our paper (https://github.com/plbaldoni/epigraHMMPaper/blob/main/Supp/Report_RNA_vs_ChIP/report.pdf).
 For instance, we have found several instances where genes were expressed but no enrichment of H3K36me3 was observed.
-Also, note that the strategy of utilizing ChIP-seq data to define sets of silver-standard marks for benchmarking purposes has been already used by others (Ji et.
+Also, note that the strategy of utilizing ChIP-seq data to define sets of silver-standard marks for benchmarking purposes has been already used by others (see Figure 2E in Ji et.
 al, 2013; PNAS).
@@
 Due to these reasons, we decided not to use gene expression levels when benchmarking H3K36me3 as suggested by the reviewer.
-
-> NAIM, PLEASE SEE MY REPORT IN THE REPO UNDER ./Supp/Report_RNA_vs_ChIP/ FOR A COMPREHENSIVE DISCUSSION ABOUT HOW USING RNA-SEQ TO DEFINE DE GENES AND THEN BENCHMARKING DIFFERENTIAL H3K36ME3 CHIP-SEQ PEAKS CAN BE PROBLEMATIC.
-WE DO NOT BENCHMARK DIFFERENTIAL PEAK CALLING METHODS IN THE PAPER USING RNA-SEQ.
-AS I MENTION IN THE PAPER:  'Following the analysis of \cite{ji2013differential}, we defined a set of protein coding genes exhibiting $|\text{LFC}|>2$ of ChIP-seq counts between Helas3 and Hepg2 cell lines as true differential binding sites.
-Results using different threshold levels are presented in the Web Appendix D1 and agree with those presented here.'.
-I CITE THE PNAS PAPER FROM HONGKAI JI (2013), WHERE THEY DEFINED GOLD-STANDARD BENCHMARKING MARKS BASED ON LOG FOLD CHANGES OF CHIP-SEQ COUNTS, NOT ON RNA-SEQ COUNTS (SEE PAGE 6971 OF THEIR PAPER).
-THE ONLY PART OF THE PAPER WHERE WE USE RNA-SEQ IS IN CHROMATIN SEGMENTATION (FIGURE 5).
+The linked report provides several other problematic cases with the peak call tracks and the entire RNA-seq data analysis.
 
@@
 > Regarding the biological significance of our method, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3.
-Specifically, we show that all known marker genes of Helas3 were associated with differential enrichment for the H3K36me3 mixture model component.
+Specifically, we show that all known marker genes of Helas3 were associated with the differential combinatorial pattern associated with enrichment for H3K36me3 alone in epigraHMM's mixture model.
 That is, genes that are expected to be actively transcribed in Helas3 (because they are marker genes) were associated with differential peaks and, in addition, were associated with the differential combinatorial pattern of enrichment for H3K36me3 alone.
 Note that consensus enrichment is a rare state and not expected to be observed in this analysis.
-The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone.
-That is, for each marker gene and the overlapping differential peaks, we calculate the average window-based mixture proportion from the differential combinatorial pattern associated with enrichment for H3K36me3 alone.
+The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior probability associated with the mixture component that model the enrichment of H3K36me3 alone.
+That is, for each marker gene and the overlapping differential peaks, we calculate the average window-based posterior probability from the differential combinatorial pattern associated with enrichment for H3K36me3 alone.
 Note that our selected model had two mixture components in the differential HMM state (either enrichment of H3K36me3 alone, or enrichment of H3K27me3 and EZH2 together).
@@
 
-> I have changed "To quantify local genome activity" to "To quantify local epigenomic activity on a genome-wide scale".
+> We have changed "To quantify local genome activity" to "To quantify local epigenomic activity on a genome-wide scale".
 
@@
 
-> This has been addressed
+> This has been addressed in our paper and we utilize the proper terminology in Section 2.
 
@@
 
-> This has been addressed
+> This has been addressed in our revised paper.
 
@@
 
-> This has been addressed and made clear in text. For RSEG, the metrics were calculated based on the instances where it ran without issues (71 out of 100 simulations).
+> This has been addressed and made clear in text.
+For RSEG, the metrics were calculated based on the instances where it ran without issues.
+In other words, we did not consider problematic cases from RSEG when computing metrics of sensitivity, FDR, computing time, etc.
+in our simulation study.
 
@@
 
-> This has been addressed. The tables and figures of the supp. are already indexed with 'Web' in front of it (e.g. Web Figure 1).
+> This has been addressed. The tables and figures of the Supplementary Information file are indexed with 'Web' in front of it (e.g. Web Figure 1).
 
@@
 We observed that in broad marks (500bp) our method was among the most efficient ones, requiring only 5GB of memory and 30 min for a genome-wide analysis of 4 samples (2 conditions, 2 replicates per condition).
-In narrow marks (250bp), our method required nearly 10GB of memory for a computation time of about 1 hour.
+In narrow marks (250bp), our method required nearly 10GB of memory for a computation time of about 1.5 hour.
 We believe that these are reasonable quantities and quite comparable to other efficient algorithms.
@@
 Web Figures 8-10 include results from our simulation study exploring the presented strategy to choose the optimal number of combinatorial patterns via BIC.
-Due to the lack of space, we have added the following sentence to our Discussion "In situations where certain combinatorial patterns are rare, prunning of specific combinatorial patterns is possible via model selection via BIC (Web Appendix B2)".
+Due to the lack of space, we have added the following sentence to our Discussion "When certain differential combinatorial patterns are rare, our model selection scheme via BIC provides means of pruning specific patterns of interest (Web Appendix B2).".
 
@@
 These methods include ChIPComp, THOR, diffReps, and DiffBind.
-We added the following sentence to the second paragraph of Section 5 of our paper: "Results presented in this section pertain to the analysis of two cell lines, namely Helas3 and Hepg2.
-Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR)".
-Similarly, input controls were also generated in our simulation study and accounted for in these algorithms.
+We added the following sentence to the second paragraph of Section 5 of our paper: "Results presented here pertain to the analysis of cell lines Helas3 and Hepg2 (see Web Appendix D3 for additional results).
+Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR).".
+Similarly, input controls were also generated in our simulation study (Sequencing Read Simulation Section) and accounted for in these algorithms.
 Throughout our analysis, we did not observe any significant improvement in performance in epigraHMM by accounting for input control effect in differential peak calling (See Web Figure 2 for these results).
-This effect was also reported by others (Lun \& Smyth, 2015).
+This effect was also reported by others (Lun \& Smyth, 2015), in which the authors note that 'These controls are mostly irrelevant when testing for DB between ChIP samples' (see http://bioconductor.org/books/3.13/csawBook/chap-filter.html#with-negative-controls).
 We did not compare in our paper the performance of other methods with and without accounting for input controls.
@@
 
-> In Figure 2A, the observed FDR is shown for all methods and for all assessed nominal FDR levels, for all methods.
-The observed FDR shown in Figure 2A is quite close to the nominal FDR chosen (note vertical lines on the panel and corresponding nominal FDR points).
+> In Figure 2A, the observed FDR is shown for all methods and for all assessed nominal FDR levels.
+The observed FDR shown in Figure 2A is, on average, very close to the nominal FDR chosen (note vertical lines on the panel and corresponding nominal FDR points).
+Note that, overall, epigraHMM is the method that has the highest sensitivity with observed FDR closest to the nominal FDR.
 In Figure 3A, no FDR was or is currently reported.
 Instead, we show 1-specificity on the x-axis.
-In Figure 3B, we show the FDR for all methods.
+Instead, the observed FDR is reported in Figure 3B for all methods.
 As we can see, the observed and nominal FDR levels reported by our method were quite close to each other (0.05, 0.07, 0.08, 0.09, 0.10).
@@
 > We thank the reviewer for the constructive comment.
-As we replied to the first anonymous reviewer, we did not observe sufficient agreement between gene expression levels for the analyzed cell lines in our paper (see Figures 1 and 2 above).
+We prepared a report to support our response to this comment, which can be assessed at https://github.com/plbaldoni/epigraHMMPaper/blob/main/Supp/Report_RNA_vs_ChIP/report.pdf
+
+> As we replied to the first anonymous reviewer, we did not observe sufficient agreement between gene expression levels for the analyzed cell lines in our paper (see Figures 1 and 2 above).
 In Figure 2, we have an example of a H3K36me3 differentiallly enriched region and a not DE gene.
@@
 Figure 4 below shows an example of such a region.
-Due to these reasons, we believe that the proposed analysis would be suboptimal to the one presented in this paper for the analyzed cell lines.
-
-> NAIM, PLEASE SEE MY COMMENT ABOVE ABOUT THIS GENE EXPRESSION ANALYSIS.
-I CITE THE PNAS PAPER FROM HONGKAI JI (2013), WHERE THEY DEFINED GOLD-STANDARD BENCHMARKING MARKS BASED ON LOG FOLD CHANGES OF CHIP-SEQ COUNTS, NOT ON RNA-SEQ COUNTS (SEE PAGE 6971 OF THEIR PAPER).
-PLEASE, SEE MY REPORT UNDER ./Supp/Report_RNA_vs_ChIP/ AND LET ME KNOW IF ANY OF THOSE RESULTS SHOULD BE INCLUDED HERE TO SUPPORT OUR RESPONSES.
+Due to these reasons, we believe that the proposed analysis would be less optimal for benchmarking purposes to the one presented in this paper for the analyzed cell lines.
+The linked report provides several other problematic cases with the peak call tracks and the entire RNA-seq data analysis.
 
@@
 
-> As we discuss above, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3.
-Specifically, we show that all known marker genes of Helas3 had associated differential enrichment for the H3K36me3 mixture model component.
-The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone.
+> As we discuss in our response to comment 7 of the first reviewer, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3.
+Specifically, we show that all known marker genes of Helas3 had associated differential combinatorial pattern associated with differential enrichment for H3K36me3 alone in our embedded mixture model.
+The presented differential enrichment level shown in Figure 5E is the average mixture model posterior probability associated with the mixture component that model the enrichment of H3K36me3 alone.
 These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.
+We have added the following paragraph to our paper "In fact, nearly all known Helas3 cycle-regulated
+genes (therefore genes expressed at some stage of the cell life cycle; Whitfield et al. (2002)) were associated with high average mixture posterior probabilities of differential enrichment
+for H3K36me3 alone (Figure 5E).
+The expression of such genes are known to correlate with proliferative states of tumors and their study may help characterize their role in cancer (Whitfield et al., 2002)".
 

### 2021-03-05 Latex/paper/paper.tex
txt update

@@
 Genomic coordinates containing a high density of mapped reads (enrichment regions, peaks) indicate likely locations of protein-DNA interaction sites, while other regions are referred to as background regions.
-This read density is often summarized by counting reads mapped onto non-overlapping windows of fixed length (window read counts), forming the basis for downstream analyses.
+The read density is often summarized by counting reads mapped onto non-overlapping windows of fixed length (window read counts), forming the basis for downstream analyses.
 Across multiple conditions, regions exhibiting enrichment in at least one condition, but not across all conditions, indicate the presence of differential activity pertaining to the protein-DNA interaction of interest.
@@
 Third, it simultaneously detects and classifies epigenomic changes under three or more conditions, a novelty not yet available in any other DPC algorithm.
-epigraHMM offers additional benefits over current HMM algorithms \citep{song2011identifying,allhoff2016differential} with an embedded GLM mixture model that allows the modeling of covariates of interest, the inclusion of normalizing offsets for non-linear biases (such as GC-content bias), and a fast parameter estimation scheme via rejection-controlled EM algorithm (RCEM).
+epigraHMM offers benefits over current HMM algorithms \citep{song2011identifying,allhoff2016differential} with an embedded GLM mixture model that allows the modeling of covariates of interest, the inclusion of normalizing offsets for non-linear biases (such as GC-content bias), and a fast estimation scheme via rejection-controlled EM algorithm (RCEM; \cite{ma2006data}).
 
@@
 Using ChIP-seq data pertaining to histone modifications H3K27me3 and H3K36me3, and to EZH2 from the ENCODE Consortium \citep{landt2012chip}, we find that current DPCs have difficulty in accurately detecting broad regions of differential enrichment between several common cell lines (Figure~\ref{fig:fig_motivation}).
-In line with previous findings \citep{steinhauser2016comprehensive}, we observe that even current DPCs designed for broad data \citep{song2011identifying,allhoff2016differential} tend to detect either overly fragmented differential peaks or call regions exhibiting no difference in experimental signal between conditions as differential (Sections \ref{sec:sim2} and \ref{sec:broad}).
+In line with previous findings \citep{steinhauser2016comprehensive}, we observe that even current DPCs designed for broad data \citep{song2011identifying,allhoff2016differential} tend to either call overly fragmented differential peaks or call regions exhibiting no difference in experimental signal between conditions as differential (Sections \ref{sec:sim2} and \ref{sec:broad}).
 Methods that rely on candidate peaks may also exhibit a compromised performance due to the limitations of single-sample peak callers in broad data \citep{stark2011diffbind,chen2015novel}.
@@
 such that $\log(\mu_{(1,hij)})=\beta_{1} + u_{hij}$, $\log(\phi_{1})=\lambda_{1}$, $\log(\mu_{(3,hij)})=\beta_{1} + \beta_{3} + u_{hij}$, and $\log(\phi_{3})=\lambda_{1} + \lambda_{3}$.
-The offset $u_{hij}$ may adjust for technical or biological artifacts such as the GC-content bias and allows the non-linear normalization of read counts across genomic windows, conditions, and replicates (Web Appendix B1; \cite{teng2017accounting}).
+The offset $u_{hij}$ may adjust for technical or biological artifacts such as the GC-content bias and allows the non-linear normalization of read counts across genomic windows, replicates, and conditions (Web Appendix B1; \cite{teng2017accounting}).
 %When $u_{hij}=0$, $\beta_{1}$ and $\lambda_{1}$ represent the log-mean and log-dispersion, respectively, of read counts pertaining to consensus background state windows, whereas $\beta_{3}$ and $\lambda_{3}$ represent the difference in log-mean and log-dispersion of read counts from consensus enrichment state windows relative to consensus background state windows.
 
-In the differential state ($r = 2$), read counts are modeled by a $L$-component finite mixture model with mixture components that follow a Negative Binomial distribution, where each component corresponds to a particular differential combinatorial pattern.
+In the differential state ($r = 2$), read counts are modeled by a $L$-component finite mixture model with mixture components that follow a NB distribution, where each component corresponds to a particular differential combinatorial pattern.
 To define these patterns, consider sets $S_{1},\ldots,S_{L}$  delineating the subset of the $G$ conditions that are enriched in each of the $L$ differential combinatorial patterns.
@@
 Differential peaks are formed by merging adjacent windows that either meet a nominal FDR level for the differential HMM state or belong to the same predicted state (Web Appendix B3).
-Under this set up, we are able not only to detect differential enrichment regions across multiple conditions, but we can also classify various differential combinatorial patterns of enrichment within broad and short domains with mixture model posterior proportions.
+Under this set up, we are able not only to detect differential enrichment regions across multiple conditions, but we can also classify various differential combinatorial patterns of enrichment within broad and short domains with mixture model posterior probabilities.
 
@@
 We observed that selecting the optimal $L^{*}$ and combinatorial patterns via BIC agrees with pruning of rare differential combinatorial patterns that one would not expect to observe biologically (Web Appendix B2).
-In addition, the current implementation of epigraHMM allows the adjustment for external covariates, such as GC-content and input controls, via model offsets (Web Appendix A3).
-In real data analyses, however, we did not observe a significant influence of input controls on the sensitivity and specificity of peak calls, a fact that has also been noted by others \citep{lun2015csaw}.
+In addition, the current implementation of epigraHMM allows for the adjustment for external covariates, such as GC-content and input controls, via model offsets (Web Appendix A3).
+In real data analyses, however, we did not observe a significant influence of input controls on the sensitivity and specificity of differential peak calls, a fact that has also been noted by others \citep{lun2015csaw}.
 
@@
 We further assessed epigraHMM under different Signal-to-Noise Ratio (SNR) levels, here defined as the ratio between the means of consensus enrichment and background emission distributions, by decreasing the mean ratio while maintaining the mean-variance relationship.
-Parameters used in this study were estimated from ENCODE data (Web Appendix C1).
+Model parameters were estimated from ENCODE data (Web Appendix C1).
 Simulated counts followed a NB distribution and were generated using a first-order Markov chain with $2^G$ states, representing every combination of background and enrichment across $G$ conditions.
@@
 \label{sec:sim2}
-
-Here, we compared epigraHMM with the DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR on data simulated with the pipeline from \cite{lun2015csaw} without a particular model assumption.
+Here, we compared epigraHMM with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithms for both broad and short ChIP-seq enrichment profiles, on data simulated with the pipeline from \cite{lun2015csaw} without a particular model assumption.
 Sequencing reads from ChIP-seq experiments were generated for two conditions and two replicates per condition.
-For the two-step DPCs ChIPComp and DiffBind, we followed \cite{lun2015csaw} and called peaks in advance using HOMER \citep{heinz2010simple}, which were then used as input in the respective software for differential call (MACS2 failed to output peaks on the simulated data).
+For the two-step DPCs ChIPComp and DiffBind, we followed \cite{lun2015csaw} and called peaks in advance using HOMER \citep{heinz2010simple}, which were then used as input in the respective software for differential call.
 A hundred simulated data sets were generated and peaks were called under multiple nominal FDR thresholds.
@@
 
-Figure~\ref{fig:fig_simulation_csaw} shows the results of our second simulation study.
-epigraHMM showed the highest observed sensitivity among all DPCs while maintaining an observed FDR close to the nominal value (Figure~\ref{fig:fig_simulation_csaw}A).
+Overall, epigraHMM showed the highest observed sensitivity among all DPCs while maintaining an observed FDR close to the nominal value across a wide range of nominal FDR levels (Figure~\ref{fig:fig_simulation_csaw}A).
 Here, the observed FDR is the proportion of windows incorrectly called as differential out of the total number of windows called as differential.
 Due to the excessive number of false positive differential peak calls, diffReps, RSEG, and THOR showed higher observed FDR levels than the nominal FDR values (Figure~\ref{fig:fig_simulation_csaw}D).
-Overall, diffReps and THOR called an excessive number of short and discontiguous peaks while RSEG called regions that were usually wider than the simulated differential peaks (Figure~\ref{fig:fig_simulation_csaw}B).
+Overall, diffReps and THOR called an excessive number of short and discontiguous peaks while RSEG called regions that were usually wider than the simulated differential peaks and did not correspond to simulated differential enrichment (Figure~\ref{fig:fig_simulation_csaw}B).
 Regarding the computation time, epigraHMM was among the fastest DPCs due to the efficient RCEM implementation (Figure~\ref{fig:fig_simulation_csaw}C).
@@
 
-We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithms for both broad and short ChIP-seq enrichment profiles.
+We compared the genome-wide performance of epigraHMM with ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR.
 For two-step DPCs, ChIPComp and DiffBind, we used peak calls from MACS2 as candidate peaks \citep{zhang2008model}.
@@
 \label{sec:broad}
-First, we compared methods regarding the coverage of differentially enriched genes by differential peak calls for H3K36me3.
-Following the analysis of \cite{ji2013differential}, we defined a set of protein coding genes exhibiting $|\text{LFC}|>2$ of ChIP-seq counts between Helas3 and Hepg2 cell lines as true differential binding sites (see Web Appendix D1 for results using different thresholding values).
+We compared methods regarding the coverage of differentially enriched genes by differential peak calls for H3K36me3, an epigenomic mark associated with transcribed genes.
+Following \cite{ji2013differential}, in which the authors define true differential binding sites from LFCs of ChIP-seq counts to assess model sensitivity (Figure 2E in \cite{ji2013differential}), we defined a set of protein coding genes exhibiting $|\text{LFC}|>2$ of ChIP-seq counts between Helas3 and Hepg2 cell lines as true differential binding sites (see Web Appendix D1 for results using different thresholding values).
 Median ratio normalization of counts was performed on protein-coding genes after excluding those with total counts under the $25^{th}$ percentile of the distribution.
@@
 
-Here, DiffBind, which has been shown to be dependent on the set of candidate peaks \citep{lun2015csaw}, called the shortest peaks overall (Figures \ref{fig:fig_broad}B and \ref{fig:fig_broad}C) and ChIPComp, which has been shown to perform best in scenarios with short marks \citep{steinhauser2016comprehensive}, appeared to have limited performance in broad marks such as H3K27me3 and EZH2 (Figures \ref{fig:fig_broad}D and \ref{fig:fig_broad}F).
+DiffBind, which has been shown to be dependent on the set of candidate peaks \citep{lun2015csaw}, called the shortest peaks overall (Figures \ref{fig:fig_broad}B and \ref{fig:fig_broad}C) and ChIPComp, which has been shown to perform best in scenarios with short marks \citep{steinhauser2016comprehensive}, appeared to have limited performance in broad marks such as H3K27me3 and EZH2 (Figures \ref{fig:fig_broad}D and \ref{fig:fig_broad}F).
 epigraHMM and RSEG, two HMM-based methods, called broader differential peaks and exhibited better sensitivity than other methods.
@@
 \caption{Analysis of broad ENCODE data.
-(A): ROC curves of H3K36me3 differential peak calls.
+(A): ROC curves of DPCs covering differentially enriched genes ($|\text{LFC}|>2$ of ChIP-seq counts) for H3K36me3.
 (C): sensitivity (y-axis) and average H3K36me3 differential peak size (kbp; x-axis) of various methods under different nominal FDR thresholds (observed FDR annotated next to data points).
@@
 
-Lastly, we analyzed data from the cell line Helas3 to segment its genome regarding the activity of H3K36me3, H3K27me3, and EZH2.
-We considered each mark as a separate experimental condition ($G=3$) and sought to jointly classify local chromatin states based upon the presence or absence of enrichment from each mark.
+We analyzed data from the cell line Helas3 to segment its genome regarding the joint activity of H3K36me3, H3K27me3, and EZH2.
+We considered each mark as a separate experimental condition ($G=3$) and jointly classified local chromatin states based upon the presence or absence of enrichment from each mark.
 Assuming that EZH2 catalyzes the methylation of H3K27me3, a repressive mark, and H3K36me3 associates with transcribed genes (Section \ref{sec:background}), we expected consensus enrichment regions to be rare and differential regions to be mostly represented by either transcribed chromatin states (enrichment of H3K36me3 alone) or repressed chromatin states (co-enrichment for H3K27me3 and EZH2).
@@
 (D): density estimate from expression of genes intersecting differential peaks associated with the enrichment of H3K36me3 alone or the enrichment of H3K27me3 and EZH2 in consensus.
-(E): cycle-regulated genes and their associated mean mixture posterior proportion from differential H3K36me3 enrichment pattern of overlapping differential peaks.
+(E): cycle-regulated genes and their associated average mixture posterior probabilities associated with differential H3K36me3 enrichment pattern of overlapping differential peaks.
 (F): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively.
@@
 To assess the biological significance of our results, we associated the distribution of transcripts per million (TPM, from matching RNA-seq experiments) of protein-coding genes with combinatorial patterns of overlapping differential peaks.
-To assign combinatorial patterns to differential peaks, we chose the combination pertaining to the most frequent mixture component across windows by using the maximum estimated mixture model posterior probability, $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})$, $j=1,\ldots,M$.
+To assign combinatorial patterns to differential peaks, we chose the combination pertaining to the most prevalent mixture component across windows by using the maximum estimated mixture model posterior probability, $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})$, $j=1,\ldots,M$.
 Genes associated with transcribed chromatin states had a significantly higher distribution of TPM counts than genes associated with repressed chromatin states (Figure~\ref{fig:fig_multimarks}D).
-In fact, nearly all known Helas3 cycle-regulated genes (therefore genes expressed at some stage of the cell life cycle; \cite{whitfield2002identification}) were associated with high average mixture posterior proportion of differential enrichment for H3K36me3 alone (Figure~\ref{fig:fig_multimarks}E).
+In fact, nearly all known Helas3 cycle-regulated genes (therefore genes expressed at some stage of the cell life cycle; \cite{whitfield2002identification}) were associated with high average mixture posterior probabilities of differential enrichment for H3K36me3 alone (Figure~\ref{fig:fig_multimarks}E).
 
 The expression of such genes are known to correlate with proliferative states of tumors and their study may help characterize their role in cancer \citep{whitfield2002identification}.
-Broad differential peaks called by epigraHMM, and their assigned differential combinatorial pattern, often agreed with the biological roles as well as the expression levels of associated gene bodies (Figure~\ref{fig:fig_multimarks}F; results from ChromHMM with 3 states for comparison).
+Differential peaks called by epigraHMM, and their assigned differential combinatorial pattern, often agreed with the biological roles as well as the expression levels of associated genes (Figure~\ref{fig:fig_multimarks}F; results from ChromHMM with 3 states for comparison).
 epigraHMM offers the benefit of simultaneously detecting differential peaks and classifying differential combinatorial patterns of enrichment even in the context of genomic segmentation of highly diverse epigenomic marks (Web Appendix D4).
@@
 First, it uses an HMM-based approach that accounts for the local dependency of ChIP-seq counts and is able to precisely detect broad and short differential regions of enrichment.
-Second, it utilizes a GLM-based framework with model offsets that accounts for non-linear biases and other potential nuisance factors such as GC-content bias or input control samples.
+Second, it utilizes a GLM-based framework with model offsets that accounts for non-linear biases and other potential factors such as GC-content bias or input control samples.
 Our efficient implementation of the RCEM algorithm led to genome-wide analyses of ChIP-seq data under a computational time comparable to some of the fastest current algorithms.
@@
 Lastly, our method allows the simultaneous detection and classification of differential combinatorial patterns of enrichment from its embedded mixture model and the associated posterior probabilities under any number of conditions.
-In situations where certain differential combinatorial patterns are rare, our model selection scheme via BIC provides means of pruning specific combinatorial patterns of interest (Web Appendix B2).
-Our software has been implemented into an R package and is available for download (Web Appendix A).
+When certain differential combinatorial patterns are rare, our model selection scheme via BIC provides means of pruning specific patterns of interest (Web Appendix B2).
 % Data resulting from these assays are similar in nature but aim to identify regions that alternatively are not bound to proteins.
@@
 \section*{Supporting Information}
-Web Appendices A-D, referenced in Sections 1-6, and a link to the implemented software
-are available with this article at the Biometrics website on Wiley Online Library.
+Web Appendices A-D, referenced in Sections 1-6, and a link to the implemented software are available with this paper at the Biometrics website on Wiley Online Library.
+MACS2 peak caller is available at https://github.com/macs3-project/MACS.
+epigraHMM has been implemented as an R package and it is available for download at https://github.com/plbaldoni/epigraHMM.
 

### 2021-03-05 Main/Figure_MultiMarks/Figure_GeneSymbols.R
Fix typo in label

@@ -101,6 +101,6 @@ fig_genesymbols <-
   theme_light() +
   coord_flip() +
   theme(panel.grid.major.y = element_blank()) +
-  labs(x = 'Known Cycle-Regulated Helas3 Genes', y = 'Differential H3K36me3 Enrichment (Mixture Prop.)')
+  labs(x = 'Known Cycle-Regulated Helas3 Genes', y = 'Differential H3K36me3 Enrichment (Mixture Prob.)')
 
 save(fig_genesymbols, file = './Figure_GeneSymbols.RData')

### 2021-03-05 Supp/Report_RNA_vs_ChIP/report.Rmd
Adding peak tracks

@@
 
-## UCSC Genome Browser view os problematic cases
+## UCSC Genome Browser view of problematic cases
 
@@
 
-### Example 1: Gene ENSG00000113389
+### Example 1 (Gene ENSG00000137285)
 
-Here we have an example of a DE gene without differential enrichment of H3K36me3.
-Note that, in Helas3, enrichment on the gene body is no different from enrichment on the intergenic (background) region.
-Therefore, observed differences in ChIP-seq are simply a sequencing depth artifact and would have to be accounted by the peak caller with proper normalization methods.
-If used as a benchmarking mark, differential peaks covering this genes would be incorrectly considered as a true positive calls.
-
-![Example 1 (Gene ENSG00000113389)](ENSG00000113389.pdf)
-
-\newpage
-
-### Example 2 (Gene ENSG00000137285)
-
-Another problematic example here. We have a gene that is expressed in both cell lines but the gene expression level is different.
+Here we have an example of a gene that is expressed in both Helas3 and Hepg2 but the gene expression level is different between cell lines.
 
@@
 Second, it appears that the enrichment of H3K36me3 reads on the gene body is no different than intergenic (background) region.
+In other words, RNA-seq abundance has no association with ChIP-seq H3K36me3 enrichment of read counts.
 
-If considered as a benchmarking mark, this gene would be considered as a true positive.
+If considered as a benchmarking mark, this gene would be considered as a true positive (DE gene).
 One would expect differential peak callers to not call differential peaks on this region.
-If so, this gene would lead to an increased false negative rate in methods that correctly call this region as consensus background.
+If so, this gene would lead to an increased false negative rate in methods that correctly call this region as consensus background (all methods but RSEG).
+As we can see, RSEG would have an artificial gain in sensitivity by calling this entire region as differential when, in fact, there is no difference in H3K36me3 enrichment.
 
-![Example 2 (Gene ENSG00000137285)](ENSG00000137285.pdf)
+![Example 1 (Gene ENSG00000137285)](ENSG00000137285.pdf)
 
@@
 
-### Example 3 (Gene ENSG00000139998)
+### Example 2 (Gene ENSG00000139998)
 
@@
 
-However, there is one additional point to make here.
+However, there is one additional point to be made here.
 If the gene below is DEG, and one assumes that there is a local ChIP-seq signal difference in between cell lines on the exonic regions, how do we justify the similar ChIP-seq coverage on the remaining regions on the gene body (right most part of the figure) between the two cell lines.
@@
 
-![Example 3 (Gene ENSG00000139998)](ENSG00000139998.pdf)
+![Example 2 (Gene ENSG00000139998)](ENSG00000139998.pdf)
 
@@
 
-### Example 4 (Gene ENSG00000144852)
+### Example 3 (Gene ENSG00000144852)
 
@@
 
-![Example 4 (Gene ENSG00000144852)](ENSG00000144852.pdf)
+![Example 3 (Gene ENSG00000144852)](ENSG00000144852.pdf)
 
@@
 
-### Example 5 (Gene ENSG00000160202)
+### Example 4 (Gene ENSG00000160202)
 
@@
 
-![Example 5 (Gene ENSG00000160202)](ENSG00000160202.pdf)
+![Example 4 (Gene ENSG00000160202)](ENSG00000160202.pdf)
 
@@
 
-### Example 6 (Gene ENSG00000178498)
+### Example 5 (Gene ENSG00000178498)
 
@@
 
-![Example 6 (Gene ENSG00000178498)](ENSG00000178498.pdf)
+![Example 5 (Gene ENSG00000178498)](ENSG00000178498.pdf)
 
@@
 
-### Example 7 (Gene ENSG00000198610)
+### Example 6 (Gene ENSG00000198610)
 
@@
 
-![Example 7 (Gene ENSG00000198610)](ENSG00000198610.pdf)
+![Example 6 (Gene ENSG00000198610)](ENSG00000198610.pdf)
 

### 2021-03-05 Supp/Report_RNA_vs_ChIP/Figure_PeakCalls_H3K36me3_Encode_twocells_500bp.R

@@ -1,0 +1,129 @@
+library(dplyr)
+library(ggplot2)
+library(GenomicRanges)
+library(reshape2)
+library(tidyr)
+library(magrittr)
+library(plyr)
+library(ggpubr)
+library(data.table)
+library(RColorBrewer)
+library(Polychrome)
+library(SummarizedExperiment)
+
+bp = 500
+mark = 'H3K36me3'
+data = 'Encode_twocells'
+fdr = 0.05
+chromosome = paste0('chr',1:22)
+
+# Loading epigraHMM
+load(list.files(
+  file.path('../../Public/epigraHMM', mark, data, 'Output'),
+  paste0('Output_', bp, 'bp.RData'),
+  full.names = TRUE
+))
+
+gr.epigrahmm <- epigraHMM::callPeaks(object = epigraHMM_H3K36me3_Encode_twocells_Output_500bp,
+                                     hdf5 = metadata(epigraHMM_H3K36me3_Encode_twocells_Output_500bp)$output,
+                                     method = 0.05)
+
+# Loading csaw
+csaw <-
+  fread(list.files(
+    file.path('../../Public/csaw', mark, data, paste0('Output', bp)),
+    '.tsv.gz',
+    full.names = TRUE
+  ))
+gr.csaw <-
+  with(csaw, GRanges(seqnames, IRanges(start = start, end = end)))
+gr.csaw$FDR <- csaw$FDR
+gr.csaw <- gr.csaw[seqnames(gr.csaw) %in% chromosome]
+gr.csaw <- gr.csaw[gr.csaw$FDR < fdr]
+
+# Loading RSEG
+load(list.files(
+  file.path('../../Public/RSEG', mark, data, paste0('Output', bp)),
+  'Output*.*.RData$',
+  full.names = TRUE
+))
+rseg[, postprob :=  (V4 + V5)]
+gr.rseg <- with(rseg, GRanges(V1, IRanges(start = V2, end = V3)))
+gr.rseg$FDR = rseg$postprob
+gr.rseg <-
+  gr.rseg[gr.rseg$FDR >= 0 &
+            gr.rseg$FDR <= 1] # There are a few > 1 cases, which I am excluding
+gr.rseg <-
+  gr.rseg[epigraHMM:::fdrControl(gr.rseg$FDR, fdr = fdr)] # Filtering significant windows
+gr.rseg <-
[…]

### 2021-04-02 Latex/appendix/webappendix.Rnw
Update text

@@
 $^*$Department of Biostatistics, University of North Carolina at Chapel Hill, Chapel Hill, North Carolina, USA\\%
-\today
+%\today
 }
@@
 \date{%
-\today
+%\today
 }
@@
 such that $options = \{\text{--broad --broad-cutoff 0.1}\}$ if H3K27me3, H3K36me3, or EZH2, or $options = \{\text{-q 0.01}\}$ otherwise.
+
+In the chromatin segmentation analysis, ChromHMM was run with the commands:
+\begin{itemize}
+\item ChromHMM-BinarizeBed: \textit{java -jar $\sim$/ChromHMM/ChromHMM.jar BinarizeBam -b 500 -c 'controldir' ~/ChromHMM/CHROMSIZES/hg19.txt 'inputbamdir' 'input.txt' 'outputbinarydir'} and
+\item ChromHMM-LearnModel: \textit{unset DISPLAY \&\& java -jar $\sim$/ChromHMM/ChromHMM.jar LearnModel -p 0 -b 500 'outputbinarydir' 'outputdir' 'numstates' hg19},
+\end{itemize}
+
+such that 'controldir' is the directory with control experiments, 'inputbamdir' is the directory with input bam files, 'input.txt' is an experimental table, 'outputbinarydir' is the output directory with binarized files, 'outputdir' is the model output, and 'numstates' is the number of states in the model.
 

### 2021-04-02 Latex/letter/210305_letter/letter.Rmd
Update text

--- 2021-03-05 Latex/letter/letter.Rmd
@@
 
-We thank you for the constructive comments regarding our paper "Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions".
+We appreciate the constructive comments regarding our paper "Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions".
 
 Our revised paper has 20 pages in total, as requested, including appendices and references, and excluding the title page and abstract, figures and tables.
-We emphasize that the our presented method, previously called mixNBHMM, is now called epigraHMM.
+We emphasize that our presented method, previously called mixNBHMM, is now called epigraHMM.
 We refer to our method as epigraHMM throughout this revised version of our paper.

### 2021-04-02 Latex/letter/210402_letter/letter.Rmd
Update text

---
title: "Response Letter for Manuscript ID BIOM2020344P.R1 - 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions'"
author: "Pedro L. Baldoni, Naim U. Rashid, Joseph G. Ibrahim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dear reviewers,

We appreciate the constructive comments on our paper "Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions" and the positive review. Detailed point-by-point responses to the outstanding comments/questions are provided below.

# Reviewer 1

1.	In Figure 4B, D, F, there is missing color legend for the gray which represents the differential peak regions.

> We have added the statement 'Shaded areas highlight differential peak regions' in the caption of Figures 3 and 4.

2.	In the new Figure 5E, the cycle is better to be changed to cell cycle, also for the “cell life cycle” and “cycle” in the main text.

> We have replaced 'cycle' to 'cell cycle' in the referenced parts of our paper (y-axis of Figure 5E, caption of Figure 5, and in the second to last paragraph os Section 5.3)

3.	Though the chromHMM running command is provided in https://github.com/plbaldoni/epigraHMMPaper, there is lacking description of the chromHMM running commands in the web appendices A.

> We have added the utilized ChromHMM commands in the referenced Web Appendix A.

> For a fair comparison between our method epigraHMM and ChromHMM, Figure 6F of our paper now reports results from ChromHMM with 4 chromatin states. Previously, we were reporting ChromHMM with only 3 states. As we detail in our paper, one should expect to observe 4 chromatin states in the joint analysis of H3K36me3, H3K37me3 and EZH2. That is:  (1) consensus background across all three marks, (2) consensus enrichment across all three marks, (3) enrichment of H3K36me3 alone, and (4) co-enrichment of H3K27me3 and EZH2. epigraHMM models states 3 and 4 into a single HMM state through its mixture model. Conversely, ChromHMM models each of the four states separately. The overall interpretation of Figure 6F remains unchanged.

4.	https://github.com/plbaldoni/epigraHMM might include software installation and tutorials in the readme, and author can consider upload onto Bioconductor for better distribution of the package. The GitHub link https://github.com/plbaldoni/epigraHMMPaper can be added in the Supporting Information in the main text as well.

> Thank you for the suggestion. Our R package is currently in the submisison process of the Bioconductor project (https://github.com/Bioconductor/Contributions/issues/1993). We hope to have our package included in the next Bioconductor for better distribution of our work. As suggested, we added a brief README file in our GitHub repository with installation instructions and a link to our vignette document that present the functionalities of our package. We have added the following sentence to the Supporting Information section of our paper "All of the scripts used in the analyses presented in this paper can be downloaded at https://github.com/plbaldoni/epigraHMMPaper."

5.	The DESeq link in the https://github.com/plbaldoni/epigraHMMPaper/blob/main/Supp/Report_RNA_vs_ChIP/report.pdf can be corrected to http://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html.

> The DESeq2 links utilized in our report now point to the suggested location if the file is open on a pdf viewer such as Adobe Reader or Preview (MacOS only).

### 2021-04-02 Latex/paper/paper.tex
Update text

@@
 At the bottom, general overview of the presented method for simultaneous detection and classification of broad and short differential regions of enrichment.
-This figure appears in color in the electronic version of this article.
+This figure appears in color in the electronic version of this article, and any mention of color refers to that version.
 \label{fig:fig_motivation}}
@@
 Shaded areas indicate true differential peaks.
-This figure appears in color in the electronic version of this article.
+This figure appears in color in the electronic version of this article, and any mention of color refers to that version.
 \label{fig:fig_simulation_csaw}}
@@
 Posterior probabilities of the HMM differential state are shown at the bottom of each panel.
+Shaded areas highlight differential peak regions.
 (E): computing time and peak memory usage of genome-wide analysis from various methods.
-This figure appears in color in the electronic version of this article.
+This figure appears in color in the electronic version of this article, and any mention of color refers to that version.
 \label{fig:fig_broad}}
@@
 Posterior probabilities of the HMM differential state are shown at the bottom of each panel.
+Shaded areas highlight differential peak regions.
 (E): computing time and peak memory usage of genome-wide analysis from various methods.
 Results are shown under a nominal FDR control of 0.05.
-This figure appears in color in the electronic version of this article.
+This figure appears in color in the electronic version of this article, and any mention of color refers to that version.
 \label{fig:fig_sharp}}
@@
 (D): density estimate from expression of genes intersecting differential peaks associated with the enrichment of H3K36me3 alone or the enrichment of H3K27me3 and EZH2 in consensus.
-(E): cycle-regulated genes and their associated average mixture posterior probabilities associated with differential H3K36me3 enrichment pattern of overlapping differential peaks.
+(E): cell cycle-regulated genes and their associated average mixture posterior probabilities associated with differential H3K36me3 enrichment pattern of overlapping differential peaks.
 (F): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively.
-This figure appears in color in the electronic version of this article.
+This figure appears in color in the electronic version of this article, and any mention of color refers to that version.
 \label{fig:fig_multimarks}}
@@
 Genes associated with transcribed chromatin states had a significantly higher distribution of TPM counts than genes associated with repressed chromatin states (Figure~\ref{fig:fig_multimarks}D).
-In fact, nearly all known Helas3 cycle-regulated genes (therefore genes expressed at some stage of the cell life cycle; \cite{whitfield2002identification}) were associated with high average mixture posterior probabilities of differential enrichment for H3K36me3 alone (Figure~\ref{fig:fig_multimarks}E).
+In fact, nearly all known Helas3 cell cycle-regulated genes (therefore genes expressed at some stage of the cell cycle; \cite{whitfield2002identification}) were associated with high average mixture posterior probabilities of differential enrichment for H3K36me3 alone (Figure~\ref{fig:fig_multimarks}E).
 
 The expression of such genes are known to correlate with proliferative states of tumors and their study may help characterize their role in cancer \citep{whitfield2002identification}.
-Differential peaks called by epigraHMM, and their assigned differential combinatorial pattern, often agreed with the biological roles as well as the expression levels of associated genes (Figure~\ref{fig:fig_multimarks}F; results from ChromHMM with 3 states for comparison).
+Differential peaks called by epigraHMM, and their assigned differential combinatorial pattern, often agreed with the biological roles as well as the expression levels of associated genes (Figure~\ref{fig:fig_multimarks}F; results from ChromHMM with 4 states for comparison).
 epigraHMM offers the benefit of simultaneously detecting differential peaks and classifying differential combinatorial patterns of enrichment even in the context of genomic segmentation of highly diverse epigenomic marks (Web Appendix D4).
@@
 \backmatter
-% \section*{Acknowledgements}
-% The authors thank Professor A. Sen for some helpful suggestions,
-% Dr C. R. Rangarajan for a critical reading of the original version of the
-% paper, and an anonymous referee for very useful comments that improved
-% the presentation of the paper.
+
+\section*{Acknowledgements}
+The authors wish to thank the editor, associate editor and three referees for helpful comments and suggestions, which have led to an improvement of this article.
+This research was partially supported by NIH grants GM70335, P01CA142538, P30CA016086, and P50CA058223 and by the Brazilian funding agency CAPES (13195/13‐1).
+
+\section*{Data Avalability Statement}
+
+The data that support the findings of this paper are openly available in the ENCODE Project portal \citep{sloan2016encode} at https://www.encodeproject.org.
 
@@
 epigraHMM has been implemented as an R package and it is available for download at https://github.com/plbaldoni/epigraHMM.
+All of the scripts used in the analyses presented in
+this paper can be downloaded at https://github.com/plbaldoni/epigraHMMPaper.
 

### 2021-04-02 Main/Figure_MultiMarks/Figure_GeneSymbols.R
Updating figure

@@ -101,6 +101,6 @@ fig_genesymbols <-
   theme_light() +
   coord_flip() +
   theme(panel.grid.major.y = element_blank()) +
-  labs(x = 'Known Cycle-Regulated Helas3 Genes', y = 'Differential H3K36me3 Enrichment (Mixture Prob.)')
+  labs(x = 'Known Cell Cycle-Regulated Helas3 Genes', y = 'Differential H3K36me3 Enrichment (Mixture Prob.)')
 
 save(fig_genesymbols, file = './Figure_GeneSymbols.RData')

### 2021-04-02 Main/Figure_MultiMarks/Figure_MultiMarks.R

@@ -11,11 +11,11 @@ load('Figure_Segmentation.RData')
 # Figure right
 fig.top <- ggarrange(fig_viterbi,fig_parameters,nrow = 1,labels = list('A','B'))
 fig.mid <- ggarrange(fig_bic,fig_expression,nrow = 1,legend = 'none',labels = list('C','D'))
-fig.bot <- ggarrange(fig_genesymbols,fig_segmentation[['3']],nrow = 1,legend = 'none',labels = list('E','F'))
+fig.bot <- ggarrange(fig_genesymbols,fig_segmentation[['4']],nrow = 1,legend = 'none',labels = list('E','F'))
 
 # Put them all together
 
 fig.final = ggarrange(fig.top,fig.mid,fig.bot,nrow=3,heights = c(0.275,0.275,0.45))
-fig.final.label = ggarrange(fig.final,get_legend(fig_segmentation[['3']]),nrow=2,heights = c(0.95,0.05))
+fig.final.label = ggarrange(fig.final,get_legend(fig_segmentation[['4']]),nrow=2,heights = c(0.95,0.05))
 
 ggsave(filename = 'Figure_MultiMarks.pdf',plot = fig.final.label,height = 11.5,width = 9,dpi = 'retina')

### 2021-04-02 Main/Figure_MultiMarks/Figure_Segmentation.R

@@ -224,7 +224,7 @@ fig_segmentation <- lapply(3:8,function(x){
     geom_text(data = dt.anno2,aes(x = x, y = y, label = Label),size = size.text) +
     geom_text_repel(data=refseq.out,aes(x = Lbl.x,y = Counts,label = gene_name),direction = 'y',size = size.text,nudge_y = -10,segment.color = 'grey') +
     theme_bw() +
-    scale_fill_manual(values = cbPalette[c(7,1,4)],name = 'ChromHMM\nSegmentation')+
+    scale_fill_manual(values = cbPalette[c(7,1,4,8)],name = 'ChromHMM\nSegmentation')+
     scale_color_viridis_c(option = 'B',name = 'Gene Expression\nlog2(TPM+1)')+
     labs(x = paste0('Genomic Window (',chromosome,')'),y = 'Normalized ChIP-seq counts') +
     scale_x_continuous(labels = scales::comma,limits = c(161580000,max(ChIP[ChIP$Window%in%idx,]$start))) +

### 2021-04-10 Latex/paper/paper.tex
Update text
Update ENCODE acronym

@@
 
-The data that support the findings of this paper are openly available in the ENCODE Project portal \citep{sloan2016encode} at https://www.encodeproject.org.
+The data that support the findings of this paper are openly available in the Encyclopedia of DNA Elements Project portal at https://www.encodeproject.org \citep{sloan2016encode}.
 
@@
 MACS2 peak caller is available at https://github.com/macs3-project/MACS.
+
 epigraHMM has been implemented as an R package and it is available for download at https://github.com/plbaldoni/epigraHMM.
-All of the scripts used in the analyses presented in
-this paper can be downloaded at https://github.com/plbaldoni/epigraHMMPaper.
+All of the scripts used in the analyses presented in this paper can be downloaded at https://github.com/plbaldoni/epigraHMMPaper.
 
